
\appendix

\chapter{Notations\label{chap:Appendix-Notations}}

\global\long\def\gunderline#1{\mathunderline{greenunder}{#1}}%
\global\long\def\bef{\forwardcompose}%
\global\long\def\bbnum#1{\custombb{#1}}%
\global\long\def\pplus{{\displaystyle }{+\negmedspace+}}%
Certain notations and terms were chosen in this book differently from
what the functional programming community currently uses. The proposed
notation is well adapted to reasoning about types and code, and especially
for designing data types and proving the laws of various typeclasses.

\section{Summary of notations for types and code}
\begin{description}
\item [{$F^{A}$}] type constructor $F$ with type argument $A$. In Scala,
\lstinline!F[A]!
\item [{$x^{:A}$}] value $x$ has type $A$; in Scala, \lstinline!x:A!
\item [{$\bbnum 1,\,1$}] the unit type and its value; in Scala, \lstinline!Unit!
and \lstinline!()!
\item [{$\bbnum 0$}] the void type. In Scala, \lstinline!Nothing!
\item [{$A+B$}] a disjunctive type (co-product). In Scala, this type is
\lstinline!Either[A, B]! 
\item [{$x^{:A}+\bbnum 0^{:B}$}] a value of a disjunctive type $A+B$.
In Scala, \lstinline!Left(x)!
\item [{$A\times B$}] a product (tuple) type. In Scala, this type is \lstinline!(A,B)!
\item [{$a^{:A}\times b^{:B}$}] value of a tuple type $A\times B$. In
Scala, \lstinline!(a, b)!
\item [{$A\rightarrow B$}] the function type, mapping from $A$ to $B$
\item [{$x^{:A}\rightarrow f$}] a nameless function (as a value). In Scala,
\lstinline!{ x:A => f }!
\item [{$\text{id}$}] the identity function; in Scala, \lstinline!identity[A]!
\item [{$\triangleq$}] \textsf{``}is defined to be\textsf{''} or \textsf{``}is equal by definition\textsf{''}
\item [{$\overset{!}{=}$}] \textsf{``}must be equal according to what we know\textsf{''}
\item [{$\overset{?}{=}$}] \textsf{``}we ask \textemdash{} is it equal? \textemdash{}
because we still need to prove that\textsf{''}
\item [{$\square$}] \textsf{``}this proof or this derivation is finished\textsf{''}
\item [{$\cong$}] for types, a natural isomorphism between types; for
values, \textsf{``}equivalent\textsf{''} values according to an already established
isomorphism
\item [{$A^{:F^{B}}$}] special type annotation, used for defining unfunctors
(GADTs)
\item [{$\wedge$}] logical conjunction; $\alpha\wedge\beta$ means \textsf{``}both
$\alpha$ and $\beta$ are true\textsf{''}
\item [{$\vee$}] logical disjunction; $\alpha\vee\beta$ means \textsf{``}either
$\alpha$ or $\beta$ or both are true\textsf{''}
\item [{$\Rightarrow$}] logical implication; $\alpha\Rightarrow\beta$
means \textsf{``}if $\alpha$ is true then $\beta$ is true\textsf{''}
\item [{$\text{fmap}_{F}$}] the standard method \lstinline!fmap! of a
functor $F$. In Scala, \lstinline!Functor[F].fmap!
\item [{$\text{flm}_{F},\text{ftn}_{F},\text{pu}_{F}$}] the standard methods
\lstinline!flatMap!, \lstinline!flatten!, and \lstinline!pure!
of a monad $F$
\item [{$F^{\bullet}$}] the type constructor $F$ understood as a type-level
function. In Scala, \lstinline!F[_]! 
\item [{$F^{\bullet}\leadsto G^{\bullet}$}] or $F\leadsto G$ a natural
transformation between functors $F$ and $G$. In Scala, \lstinline!F ~> G!
\item [{$\forall A.\,P^{A}$}] a universally quantified type expression.
In Scala 3, \lstinline![A] => P[A]!
\item [{$\exists A.\,P^{A}$}] an existentially quantified type expression.
In Scala, \lstinline!{ type A; val x: P[A] }! 
\item [{$\bef$}] the forward composition of functions: $f\bef g$ is $x\rightarrow g(f(x))$.
In Scala, \lstinline!f andThen g!
\item [{$\circ$}] the backward composition of functions: $f\circ g$ is
$x\rightarrow f(g(x))$. In Scala, \lstinline!f compose g!
\item [{$\circ$}] the backward composition of type constructors: $F\circ G$
is $F^{G^{\bullet}}$. In Scala, \lstinline!F[G[A]]! 
\item [{$\triangleright$}] use a value as the argument of a function:
$x\triangleright f$ is $f(x)$. In Scala, \lstinline!x.pipe(f)!
\item [{$f^{\uparrow G}$}] a function $f$ lifted to a functor $G$; same
as $\text{fmap}_{G}(f)$
\item [{$f^{\uparrow G\uparrow H}$}] a function lifted first to $G$ and
then to $H$. In Scala, \lstinline!h.map(_.map(f))! 
\item [{$f^{\downarrow H}$}] a function $f$ lifted to a contrafunctor
$H$ 
\item [{$\diamond_{M}$}] the Kleisli product operation for the monad $M$
\item [{$L\varangle M$}] or equivalently $T_{L}^{M}$ \textemdash{} the
monad $L$\textsf{'}s transformer applied to a monad $M$
\item [{$\oplus$}] the binary operation of a monoid. In Scala, \lstinline!x |+| y!
\item [{$\Delta$}] the \textsf{``}diagonal\textsf{''} function of type $\forall A.\,A\rightarrow A\times A$
\item [{$\pi_{1},\pi_{2},...$}] the projections from a tuple to its first,
second, ..., parts
\item [{$\boxtimes$}] pair product of functions: $(f\boxtimes g)(a\times b)\triangleq f(a)\times g(b)$
\item [{$\left[a,b,c\right]$}] an ordered sequence of values. In Scala,
\lstinline!Seq(a, b, c)!
\item [{$\begin{array}{||cc|}
x\rightarrow x & \bbnum 0\\
\bbnum 0 & a\rightarrow a\times a
\end{array}$}] ~a function that works with disjunctive types (a \textsf{``}\index{disjunctive functions}disjunctive
function\textsf{''})
\end{description}

\section{Detailed explanations}

$F^{A}$ means a type constructor $F$ with a type parameter $A$.
In Scala, this is \lstinline!F[A]!. Type constructors with multiple
type parameters are denoted by $F^{A,B,C}$. Nested type parameters
such as Scala\textsf{'}s \lstinline!F[G[A]]! are denoted by $F^{G^{A}}$,
meaning $F^{(G^{A})}$.

$x^{:A}$ means a value $x$ that has type $A$; this is a \textbf{\index{type annotation}type
annotation}. In Scala, a type annotation is \lstinline!x:A!. The
colon symbol, $:$, in the superscript shows that $A$ is not a type
argument (as it would be in a type constructor, $F^{A}$). The notation
$x:A$ can be used as well, but $x^{:A}$ is easier to read when $x$
is inside a larger code expression. 

$\bbnum 1$ means the unit type\index{unit type}, and $1$ means
the value of the unit type. In Scala, the unit type is \lstinline!Unit!,
and its value is \lstinline!()!. Example of using the unit type is
$\bbnum 1+A$, which corresponds to \lstinline!Option[A]! in Scala.

$\bbnum 0$ means the void\index{void type} type (the type with no
values). In Scala, this is the type \lstinline!Nothing!. Example
of using the void type is to denote the empty part of a disjunction.
For example, in the disjunction $\bbnum 1+A$ the non-empty part is
$\bbnum 0+A$, which in Scala corresponds to \lstinline!Some[A]!.
The empty part $\bbnum 1+\bbnum 0$ corresponds to \lstinline!None!.
Similarly, $A+\bbnum 0$ denotes the left part of the type $A+B$
(in Scala, \lstinline!Left[A]!), while $\bbnum 0+B$ denotes its
right part (in Scala, \lstinline!Right[A]!). Values of disjunctive
types are denoted similarly. For instance, $x^{:A}+\bbnum 0^{:B}$
denotes a value of the left part of the type $A+B$; in Scala, this
value is written as \lstinline!Left[A,B](x)!.

$A+B$ means the disjunctive type made from types $A$ and $B$ (or,
a disjunction of $A$ and $B$). In Scala, this is the type \texttt{}\lstinline!Either[A, B]!.

$x^{:A}+\bbnum 0^{:B}$ denotes a value of a disjunctive type $A+B$,
where $x$ is the value of type $A$, which is the chosen case, and
$\bbnum 0$ stands for other possible cases. For example, $x^{:A}+\bbnum 0^{B}$
is \lstinline!Left[A,B](x)! in Scala. Type annotations $^{:A}$ and
$^{:B}$ may be omitted if the types are unambiguous from the context.

$A\times B$ means the product type made from types $A$ and $B$.
In Scala, this is the tuple type \lstinline!(A,B)!.

$a^{:A}\times b^{:B}$ means a value of a tuple type $A\times B$;
in Scala, this is the tuple value \lstinline!(a, b)!. Type annotations
$^{:A}$ and $^{:B}$ may be omitted if the types are unambiguous
from the context.

$A\rightarrow B$ means the type of functions $A$ to $B$. In Scala,
this is the type \lstinline!A => B!. The function type\textsf{'}s arrow binds
weaker than $+$, which binds weaker than $\times$. So, $A+B\rightarrow C\times D$
means $(A+B)\rightarrow(C\times D)$.

$x^{:A}\rightarrow y$ means a nameless function with argument $x$
of type $A$ and function body $y$. (Usually, the body $y$ will
be an expression that uses $x$. In Scala, this is \lstinline!{ x: A => y }!.
Type annotation $^{:A}$ may be omitted if the type is unambiguous
from the context.

$\text{id}$ means the identity function. The type of its argument
should be either specified as $\text{id}^{A}$ or $\text{id}^{:A\rightarrow A}$,
or else should be unambiguous from the context. In Scala,  \lstinline!identity[A]!
corresponds to $\text{id}^{A}$.

$\triangleq$ means \textsf{``}equal by definition\textsf{''}. A definition of a function
$f$ is written as $f\triangleq(x^{:\text{Int}}\rightarrow x+10)$;
in Scala, this is \lstinline!val f = { x: Int => x + 10 }!. A definition
of a type constructor $F$ is written as $F^{A}\triangleq\bbnum 1+A$;
in Scala, this is \lstinline!type F[A] = Option[A]!.

$\cong$ for types means an equivalence (an isomorphism) of types.
For example, $A+A\times B\cong A\times\left(\bbnum 1+B\right)$. The
same symbol $\cong$ for \emph{values} means \textsf{``}equivalent\textsf{''} according
to an equivalence relation that needs to be established in the text.
For example, if we have established an equivalence that allows nested
tuples to be reordered whenever needed, we can write $\left(a\times b\right)\times c\cong a\times\left(b\times c\right)$,
meaning that these values are mapped to each other by the established
isomorphism functions. 

$A^{:F^{B}}$ in type expressions means that the type constructor
$F^{\bullet}$ assigns the type $F^{B}$ to the type expression $A$.
This notation is used for defining unfunctors (GADTs). For example,
the Scala code:

\begin{lstlisting}
sealed trait F[A]
case class F1() extends F[Int]
case class F2[A](a: A) extends F[(A, String)]
\end{lstlisting}
defines an unfunctor\index{unfunctor} denoted by $F^{A}\triangleq\bbnum 1^{:F^{\text{Int}}}+A^{:F^{A\times\text{String}}}$.

$\wedge$ (conjunction), $\vee$ (disjunction), and $\Rightarrow$
(implication) are used in formulas of Boolean as well as constructive
logic in Chapter~\ref{chap:5-Curry-Howard}, e.g., $\alpha\wedge\beta$,
where Greek letters stand for logical propositions.

$\text{fmap}_{F}$ means the standard method $\text{fmap}$ of the
\lstinline!Functor! typeclass, implemented for the functor $F$.
In Scala, this may be written as \texttt{}\lstinline!Functor[F].fmap!.
Since each functor $F$ has its own specific implementation of $\text{fmap}_{F}$,
the subscript \textsf{``}$F$\textsf{''} is not a type parameter of $\text{fmap}_{F}$.
The method $\text{fmap}_{F}$ actually has \emph{two} type parameters,
which can be written out as $\text{fmap}_{F}^{A,B}$. Then the type
signature of $\text{fmap}$ is written in full as $\text{fmap}_{F}^{A,B}:\left(A\rightarrow B\right)\rightarrow F^{A}\rightarrow F^{B}$.
For clarity, we may sometimes write out the type parameters $A,B$
in the expression $\text{fmap}_{F}^{A,B}$, but in most cases these
type parameters $A$, $B$ can be omitted without loss of clarity.

$\text{pu}_{F}$ denotes a monad $F$\textsf{'}s method \lstinline!pure!.
This function has type signature $A\rightarrow F^{A}$ and has a type
parameter $A$. In the code notation, the type parameter may be either
omitted or denoted as $\text{pu}_{F}^{A}$. If we are using \lstinline!pure!
with a complicated type, e.g., $\bbnum 1+P^{B}$, as of the type parameter
$A$, we may write the type parameter for clarity as $\text{pu}_{F}^{\bbnum 1+P^{B}}$.
The type signature of that function then becomes: 
\[
\text{pu}_{F}^{1+P^{B}}:\bbnum 1+P^{B}\rightarrow F^{\bbnum 1+P^{B}}\quad.
\]
But in most cases we will not need to write out the type parameters.

$\text{flm}_{F}$ denotes the curried version of a monad $F$\textsf{'}s method
\lstinline!flatMap!. The type signature of $\text{flm}_{F}$ is $\text{flm}_{F}:(A\rightarrow F^{B})\rightarrow F^{A}\rightarrow F^{B}$.
Note that Scala\textsf{'}s standard \lstinline!flatMap! type signature is
not curried. The curried method $\text{flm}_{F}$ is easier to use
in calculations involving the monad laws.

$\text{ftn}_{F}$ denotes a monad $F$\textsf{'}s method \lstinline!flatten!
with the type signature $\text{ftn}_{F}:F^{F^{A}}\rightarrow F^{A}$.

$F^{\bullet}$ means the type constructor $F$ understood as a type-level
function, \textemdash{} that is, with a type parameter unspecified.
In Scala, this is \lstinline!F[_]!. The bullet symbol, $\bullet$,
is used as a placeholder for the missing type parameter. When no type
parameter is needed, $F$ means the same as $F^{\bullet}$. (For example,
\textsf{``}a functor $F$\textsf{''} and \textsf{``}a functor $F^{\bullet}$\textsf{''} mean the same
thing.) However, it is useful for clarity to be able to indicate the
place where the type parameter would appear. For instance, functor
composition is denoted as $F^{G^{\bullet}}$; in Scala 2, this is
\texttt{}\lstinline!Lambda[X => F[G[X]]]! when using the \textsf{``}kind
projector\textsf{''}\index{kind@\textsf{``}kind projector\textsf{''} plugin} plugin.\footnote{\texttt{\href{https://github.com/typelevel/kind-projector}{https://github.com/typelevel/kind-projector}}}
When the type parameter $B$ of a bifunctor $P^{A,B}$ is fixed to
$Z$, we get a functor (with respect to $A$) denoted by $P^{\bullet,Z}$.
Another example: $T_{L}^{M,\bullet}$ denotes a monad transformer
for the base monad $L$ and the foreign monad $M$. The foreign monad
$M$ is a type parameter in $T_{L}^{M,\bullet}$. The symbol $\bullet$
stands for the transformer\textsf{'}s second type parameter. (The base monad
$L$ is not a type parameter in $T_{L}^{M,\bullet}$ because the construction
of the monad transformer depends on the internal details of $L$.)

$F^{\bullet}\leadsto G^{\bullet}$ or $F\leadsto G$ means a natural
transformation between two functors $F$ and $G$. In some Scala libraries,
this is denoted by \lstinline!F ~> G!.

$\forall A.\,P^{A}$ is a universally quantified type expression,
in which $A$ is a bound type parameter.

$\exists A.\,P^{A}$ is an existentially quantified type expression,
in which $A$ is a bound type parameter.

$\bef$ means the forward composition\index{forward composition}
of functions: $f\bef g$ (reads \textsf{``}$f$ before $g$\textsf{''}) is the function
defined as $x\rightarrow g(f(x))$.

$\circ$ means the backward composition\index{backward composition}
of functions: $f\circ g$ (reads \textsf{``}$f$ after $g$\textsf{''}) is the function
defined as $x\rightarrow f(g(x))$.

$\circ$ with type constructors means their (backward) composition,
for example $F\circ G$ denotes the type constructor $F^{G^{\bullet}}$.
In Scala, this is \lstinline!F[G[A]]!. 

$x\triangleright f$ (the \textbf{pipe notation})\index{pipe notation}\index{\$@$\triangleright$-notation!see \textsf{``}pipe notation\textsf{''}}
is a different syntax for $f(x)$. The value $x$ is passed as the
argument to the function $f$. In Scala, the expression $x\triangleright f$
is written as \lstinline!x.pipe(f)! or, if \lstinline!f! is a method,
\lstinline!x.f!. This syntax is used with many standard methods such
as \lstinline!size! or \lstinline!toSeq!. Because the argument $x$
is to the left of the function $f$ in this notation, forward compositions
of functions such as $x\triangleright f\triangleright g$ are naturally
grouped to the left as it is done in Scala code, for example \lstinline!x.toSeq.sorted!.
The operation $\triangleright$ (pronounced \textsf{``}pipe\textsf{''}) groups weaker
than the forward composition ($\bef$), and so we have $x\triangleright f\bef g=x\triangleright f\triangleright g$
in this notation. Reasoning about code in the pipe notation uses the
identities:
\begin{align*}
x\triangleright f=f(x)\quad,\quad\quad & \left(x\triangleright f\right)\triangleright g=x\triangleright f\triangleright g\quad,\\
x\triangleright f\bef g=x\triangleright\left(f\bef g\right)\quad,\quad\quad & x\triangleright f\triangleright g=x\triangleright f\bef g\quad.
\end{align*}
The pipe symbol groups stronger than the function arrow, so $x\rightarrow y\triangleright f$
is the same as $x\rightarrow(y\triangleright f)$. Here are some examples
of reasoning with functions in the pipe notation:
\begin{align*}
 & \left(a\rightarrow a\triangleright f\right)=\left(a\rightarrow f(a)\right)=f\quad,\\
 & f\triangleright\left(y\rightarrow a\triangleright y\right)=a\triangleright f=f(a)\quad,\\
 & f(y(x))=x\triangleright y\triangleright f\neq x\triangleright\left(y\triangleright f\right)=f(y)(x)\quad.
\end{align*}
The correspondence between the forward composition and the backward
composition:
\begin{align*}
 & f\bef g=g\circ f\quad,\\
 & x\triangleright(f\bef g)=x\triangleright f\bef g=x\triangleright f\triangleright g=g(f(x))=(g\circ f)(x)\quad.
\end{align*}

$f^{\uparrow G}$ means a function $f$ lifted to a functor $G$.
For a function $f^{:A\rightarrow B}$, the application of $f^{\uparrow G}$
to a value $g^{:G^{A}}$ is written as $f^{\uparrow G}(g)$ or as
$g\triangleright f^{\uparrow G}$. In Scala, this is \lstinline!g.map(f)!.
Nested lifting (i.e., lifting to the functor composition $H\circ G$)
can be written as $f^{\uparrow G\uparrow H}$, which means $\left(f^{\uparrow G}\right)^{\uparrow H}$,
and produces a function of type $H^{G^{A}}\rightarrow H^{G^{B}}$.
Applying a nested lifting to a value $h$ of type $H^{G^{A}}$ is
written as $h\triangleright f^{\uparrow G\uparrow H}$. In Scala,
this is \lstinline!h.map(_.map(f))!. The functor composition law
is written as:
\[
p^{\uparrow G}\bef q^{\uparrow G}=\left(p\bef q\right)^{\uparrow G}\quad.
\]
The notation $x\triangleright p^{\uparrow G}\triangleright q^{\uparrow G}$
is intended to be similar to the Scala code \lstinline!x.map(p).map(q)!.

$f^{\downarrow H}$ means a function $f$ lifted to a contrafunctor
$H$. For a function $f^{:A\rightarrow B}$, the application of $f^{\downarrow H}$
to a value $h:H^{B}$ is written as $h\triangleright f^{\downarrow H}$
and yields a value of type $H^{A}$. In Scala, this may be written
as \lstinline!h.contramap(f)!. Nested lifting is denoted as, e.g.,
$f^{\downarrow H\uparrow G}\triangleq(f^{\downarrow H})^{\uparrow G}$.

$\diamond_{M}$ means the Kleisli product operation for a given monad
$M$. This is a binary operation working on two Kleisli functions
of types $A\rightarrow M^{B}$ and $B\rightarrow M^{C}$ and yields
a new function of type $A\rightarrow M^{C}$.

$L\varangle M$ denotes the monad $L$\textsf{'}s transformer applied to a
foreign monad $M$. We define $(K\varangle L)\varangle M\triangleq K\varangle(L\varangle M)$,
which makes the monad transformer application into an associative
operation.

$\oplus$ means the binary operation of a monoid, e.g., $x\oplus y$.
For this expression to make sense, a specific monoid type should be
defined . In Scala libraries, $x\oplus y$ is often denoted as \lstinline!x |+| y!.

$\Delta$ means the standard \textsf{``}diagonal\textsf{''} function of type $\forall A.\,A\rightarrow A\times A$,
i.e., $\Delta\triangleq a^{:A}\rightarrow a\times a$. In Scala:
\begin{lstlisting}
def delta[A](a: A): (A, A) = (a, a)
\end{lstlisting}

$\pi_{1},\pi_{2},...$ denote the functions extracting the first,
second, ..., parts in a tuple. In Scala, $\pi_{1}$ is \lstinline!(_._1)!.

$\boxtimes$ means the pair product\index{pair product of functions}
of functions, defined by $(f\boxtimes g)(a\times b)=f(a)\times g(b)$.
In Scala, the pair product can be implemented as a higher-order function:
\begin{lstlisting}
def pair_product[A,B,P,Q](f: A => P, g: B => Q): ((A, B)) => (P, Q) = {
  case (a, b) => (f(a), g(b))
}
\end{lstlisting}
The operations $\Delta$, $\pi_{i}$ (where $i=1,2,...$), and $\boxtimes$
allow us to express any function operating on tuples. Useful properties
for reasoning about code of such functions: 
\begin{align*}
{\color{greenunder}\text{identity law}:}\quad & \Delta\bef\pi_{i}=\text{id}\quad,\\
{\color{greenunder}\text{naturality law}:}\quad & f\bef\Delta=\Delta\bef(f\boxtimes f)\quad,\\
{\color{greenunder}\text{left and right projection laws}:}\quad & (f\boxtimes g)\bef\pi_{1}=\pi_{1}\bef f\quad,\quad\quad(f\boxtimes g)\bef\pi_{2}=\pi_{2}\bef g\quad,\\
{\color{greenunder}\text{composition law}:}\quad & (f\boxtimes g)\bef(p\boxtimes q)=(f\bef p)\boxtimes(g\bef q)\quad,
\end{align*}
as well as the functor lifting laws for $\Delta$ and $\pi_{i}$:
\begin{align*}
 & f^{\uparrow F}\bef\Delta=\Delta\bef f^{\uparrow(F\times F)}=\Delta\bef(f^{\uparrow F}\boxtimes f^{\uparrow F})\quad,\\
 & (f^{\uparrow F}\boxtimes f^{\uparrow G})\bef\pi_{1}=f^{\uparrow(F\times G)}\bef\pi_{1}=\pi_{1}\bef f^{\uparrow F}\quad.
\end{align*}

$\left[a,b,c\right]$ means an ordered sequence of values, such as
a list or an array. In Scala, this can be \lstinline!List(a, b, c)!,
\lstinline!Vector(a, b, c)!, \lstinline!Array(a, b, c)!, or another
collection type.

$f^{:Z+A\rightarrow Z+A\times A}\triangleq\,\begin{array}{||cc|}
z\rightarrow z & \bbnum 0\\
\bbnum 0 & a\rightarrow a\times a
\end{array}\,\,$ is the \textbf{matrix notation}\index{matrix notation} for a function
whose input and/or output type is a disjunctive type (\index{disjunctive functions}a
\textbf{disjunctive function}). In Scala, the function $f$ is written
as:
\begin{lstlisting}
def f[Z, A]: Either[Z, A] => Either[Z, (A, A)] = {
  case Left(z)   => Left(z)       // Identity function on Z.
  case Right(a)  => Right((a, a)) // Delta on A.
}
\end{lstlisting}
The rows of the matrix indicate the different \lstinline!case!s in
the function\textsf{'}s code, corresponding to the different parts of the input
disjunctive type. If the input type is not disjunctive, there will
be only one row. The columns of the matrix indicate the parts of the
output disjunctive type. If the output type is not disjunctive, there
will be only one column.

A matrix may show all parts of the disjunctive types in separate \textsf{``}type
row\textsf{''} and \textsf{``}type column\textsf{''}:
\begin{equation}
f^{:Z+A\rightarrow Z+A\times A}\triangleq\,\begin{array}{|c||cc|}
 & Z & A\times A\\
\hline Z & \text{id} & \bbnum 0\\
A & \bbnum 0 & a\rightarrow a\times a
\end{array}\quad.
\end{equation}
This notation clearly indicates the input and the output types of
the function and is useful for reasoning about the code. The vertical
double line separates the \emph{input} types from the function code
(output types have a single line). In the code above, the \textsf{``}type
column\textsf{''} shows the parts of the input disjunctive type $Z+A$. The
\textsf{``}type row\textsf{''} shows the parts of the output disjunctive type $Z+A\times A$.

The matrix notation is adapted to \emph{forward} function composition
($f\bef g$). Assume that $A$ is a monoid type, and consider the
composition of the function $f$ shown above and the function $g$
defined as:
\begin{lstlisting}
def g[Z, A: Monoid]: Either[Z, (A, A)] => A = {
  case Left(_)          => Monoid[A].empty
  case Right((a1, a2))  => a1 |+| a2
}
\end{lstlisting}
In the matrix notation, the function $g$ is written (with and without
types) as:
\[
g\triangleq\,\begin{array}{|c||c|}
 & A\\
\hline Z & \_\rightarrow e^{:A}\\
A\times A & a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}
\end{array}\quad,\quad\quad g\triangleq\,\begin{array}{||c|}
\_\rightarrow e^{:A}\\
a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}
\end{array}\quad.
\]
The forward composition $f\bef g$ is computed by forward-composing
the matrix elements using the rules of the ordinary matrix multiplication,
omitting any terms containing $\bbnum 0$:
\begin{align*}
f\bef g & =\,\begin{array}{||cc|}
\text{id} & \bbnum 0\\
\bbnum 0 & a\rightarrow a\times a
\end{array}\,\bef\,\begin{array}{||c|}
\_\rightarrow e^{:A}\\
a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}
\end{array}\\
 & =\,\,\begin{array}{||c|}
\text{id}\bef(\_\rightarrow e^{:A})\\
\left(a\rightarrow a\times a\right)\bef\left(a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}\right)
\end{array}\,=\,\begin{array}{||c|}
\_\rightarrow e^{:A}\\
a\rightarrow a\oplus a
\end{array}\quad.
\end{align*}
Applying a function to a disjunctive value such as $x^{:Z+A}$ is
computed by writing $x$ as a row vector:
\[
x=z^{:Z}+\bbnum 0^{:A}=\,\begin{array}{|cc|}
z^{:Z} & \bbnum 0\end{array}\quad.
\]
Then the computation $x\triangleright f\bef g$ again follows the
rules of matrix multiplication:
\[
x\triangleright f\bef g=\,\begin{array}{|cc|}
z^{:Z} & \bbnum 0\end{array}\,\triangleright\,\begin{array}{||c|}
\_\rightarrow e^{:A}\\
a\rightarrow a\oplus a
\end{array}\,=z\triangleright(\_\rightarrow e)=e\quad.
\]
Since the standard rules of matrix multiplication are associative,
the properties of the $\triangleright$-notation such as $x\triangleright(f\bef g)=(x\triangleright f)\triangleright g$
are guaranteed to hold with matrices.

To use the matrix notation with \emph{backward} compositions ($f\circ g$),
all code matrices need to be transposed. (A standard identity of matrix
calculus is that the transposition reverses the order of composition,
$\left(AB\right)^{T}=B^{T}A^{T}$.) The input types will then appear
in the top row and the output types in the left column; the double
line is above the matrix since that is where the function inputs come
from. The above calculations are then rewritten as:
\begin{align*}
g\circ f & =\,\begin{array}{|c|cc|}
 & Z & A\times A\\
\hline\hline A & \_\rightarrow e^{:A} & a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}
\end{array}\,\circ\,\begin{array}{|c|cc|}
 & Z & A\\
\hline\hline Z & \text{id} & \bbnum 0\\
A\times A & \bbnum 0 & a\rightarrow a\times a
\end{array}\\
 & =\,\,\begin{array}{|cc|}
\hline\hline \text{id}\bef(\_\rightarrow e^{:A}) & \left(a\rightarrow a\times a\right)\bef\left(a_{1}\times a_{2}\rightarrow a_{1}\oplus a_{2}\right)\end{array}\,=\,\begin{array}{|cc|}
\hline\hline \_\rightarrow e^{:A} & a\rightarrow a\oplus a\end{array}\quad.\\
(g\circ f)(x) & =\,\begin{array}{|cc|}
\hline\hline \_\rightarrow e^{:A} & a\rightarrow a\oplus a\end{array}\,\,\begin{array}{|c|}
z^{:Z}\\
\bbnum 0
\end{array}\,=(\_\rightarrow e^{:A})(z)=e\quad.
\end{align*}
The \emph{forward} composition seems to be easier to read and to reason
about in the matrix notation.

\chapter{Glossary of terms\label{chap:Appendix-Glossary-of-terms}}
\begin{description}
\item [{Code~notation}] \index{code notation}A mathematical notation
developed in this book for deriving properties of code in functional
programs. Variables have optional type annotations, such as $x^{:A}$
or $f^{:A\rightarrow B}$. Nameless functions are denoted by$x^{:A}\rightarrow f$,
products by $a\times b$, and values of a disjunctive type $A+B$
are written as $x^{:A}+\bbnum 0^{:B}$ or $\bbnum 0^{:A}+y^{:B}$.
Functions working with disjunctive types are denoted by matrices.
Lifting of functions to functors, such as $\text{fmap}_{L}(f)$, is
denoted by $f^{\uparrow L}$; function compositions are denoted by
$f\bef g$ (forward composition) and $f\circ g$ (backward composition);
and function applications by $f(x)$ or equivalently $x\triangleright f$.
See Appendix~\ref{chap:Appendix-Notations} for details.
\item [{\index{contrafunctor}Contrafunctor}] A type constructor having
the properties of a contravariant functor\index{contrafunctor} with
respect to a type parameter. Instead of \textsf{``}contravariant functor\textsf{''},
this book uses the shorter name \textsf{``}contrafunctor\textsf{''}.
\item [{Disjunctive~type}] \index{disjunctive type}A type representing
one of several distinct possibilities. In Scala, this is usually implemented
as a sealed trait extended by several case classes. The standard Scala
disjunction types are \lstinline!Option[A]! and \lstinline!Either[A, B]!.
Also known as \index{sum type!see \textsf{``}disjunctive type\textsf{''}}\textbf{sum
}type, \textbf{tagged union}\index{tagged union type!see \textsf{``}disjunctive type\textsf{''}}
type, \textbf{co-product}\index{co-product type!see \textsf{``}disjunctive type\textsf{''}}
type, and variant type (in Object Pascal and in OCaml). The shortest
name is \textsf{``}sum type,\textsf{''} but the English word \textsf{``}disjunctive\textsf{''} is
less ambiguous to the ear than \textsf{``}sum\textsf{''}.
\item [{Exponential-polynomial~type}] \index{exponential-polynomial type}A
type constructor built using products, disjunctions (sums or co-products),
and function types (\textsf{``}exponentials\textsf{''}), as well as type parameters
and fixed types. For example,  \lstinline!type F[A] = Either[(A,A), Int=>A]!
is an exponential-polynomial type constructor. Such type constructors
are always profunctors and can also be functors or contrafunctors.
\item [{\index{functor block}Functor~block}] A short syntax for composing
several \lstinline!map!, \lstinline!flatMap!, and \lstinline!filter!
operations applied to a functor-typed value. The type constructor
corresponding to that value must be a functor and is fixed throughout
the entire functor block. For example, the Scala code
\begin{lstlisting}
for { x <- List(1,2,3); y <- List(10, x); if y > 2 }
  yield 2 * y
\end{lstlisting}
is equivalent to the code
\begin{lstlisting}
List(1, 2, 3).flatMap(x => List(10, x))
  .filter(y => y > 1).map(y => 2 * y)
\end{lstlisting}
and computes the value \lstinline!List(20, 20, 20, 6)!. This is a
functor block that \textsf{``}raises\textsf{''} computations to the \lstinline!List!
functor. Similar syntax exists in a number of languages and is called
a \textbf{\textsf{``}for-comprehension\textsf{''}}\index{for-comprehensions (Python)@\texttt{for}-comprehensions (Python)}
or a \textsf{``}list comprehension\textsf{''} in Python, \textbf{\textsf{``}do-notation\textsf{''}}\index{do-notation (Haskell)@\texttt{do}-notation (Haskell)}
in Haskell, and \textbf{\textsf{``}computation expressions\textsf{''}}\index{computation expressions (F#)@computation expressions (F\#)}
in F\#. I use the name \textsf{``}functor block\textsf{''} in this book because it
is shorter and more descriptive. (The type constructor used in a functor
block needs to be at least a functor but does not have to be a monad.)
\item [{Kleisli~function}] \index{Kleisli!functions} A function with
type signature $A\rightarrow M^{B}$ (in Scala, \lstinline!A => M[B]!)
for some fixed monad $M$. Also called a Kleisli morphism\index{Kleisli!morphisms}
(a morphism in the Kleisli category corresponding to the monad $M$).
The monadic method $\text{pure}_{M}:A\rightarrow M^{A}$ has the type
signature of a Kleisli function. The Kleisli composition operation,
$\diamond_{M}$, is a binary operation that combines two Kleisli functions
(of types $A\rightarrow M^{B}$ and $B\rightarrow M^{C}$) into a
new Kleisli function (of type $A\rightarrow M^{C}$).
\item [{\index{method}Method}] This word is used in two ways: 1) A method$_{1}$
is a Scala function defined as a member of a typeclass. For example,
\lstinline!flatMap! is a method defined in the \lstinline!Monad!
typeclass. 2) A method$_{2}$ is a Scala function defined as a member
of a data type declared as a Java-compatible \lstinline!class! or
\lstinline!trait!. Trait methods$_{2}$ are necessary in Scala when
implementing functions whose arguments have type parameters (because
Scala function values defined via \lstinline!val! cannot have type
parameters). So, many typeclasses such as \lstinline!Functor! or
\lstinline!Monad!, whose methods$_{1}$ require type parameters,
will use Scala \lstinline!traits! with methods$_{2}$ for their implementation.
The same applies to type constructions with quantified types, such
as the Church encoding. 
\item [{Nameless~function}] \index{nameless function}An expression of
function type, representing a function. For example, \lstinline!(x: Int) => x * 2!.
Also known as function expression, function literal, anonymous function,\index{anonymous function!see \textsf{``}nameless functions\textsf{''}}
closure, \index{lambda-function!see \textsf{``}nameless function\textsf{''}}lambda-function,
lambda-expression, or simply a \textsf{``}lambda\textsf{''}.
\item [{Partial~type-to-value~function~(PTVF)}] A function with a type
parameter but defined only for a certain subset of types.\index{partial type-to-value function}
In Scala, PTVFs are implemented via a typeclass constraint:
\begin{lstlisting}
def double[T: Semigroup](t: T): T = implicitly[Semigroup[T]].combine(t, t)
\end{lstlisting}
This PTVF is defined only for types \lstinline!T! for which a \lstinline!Semigroup!
typeclass instance is available.
\item [{Polynomial~functor}] \index{polynomial functor}A type constructor
built using disjunctions (sums), products (tuples), type parameters
and fixed types. For example, in Scala, \lstinline!type F[A] = Either[(Int, A), A]!
is a polynomial functor with respect to the type parameter \lstinline!A!,
while \lstinline!Int! is a fixed type (not a type parameter). Polynomial
functors are also known as \textbf{algebraic data types}\index{algebraic data type}.
\item [{Product~type}] \index{product type}A type representing several
values given at once. In Scala, product types are the tuple types,
for example \lstinline!(Int, String)!, and case classes. Also known
as \index{tuples}\textbf{tuple} type, \textbf{struct} (in C and C++),
and \textbf{record}.
\item [{\index{profunctor}Profunctor}] A type constructor whose type parameter
occurs in both covariant and contravariant positions and satisfying
the appropriate laws; see Section~\ref{subsec:f-Profunctors}.
\item [{Type~notation}] \index{type notation}A mathematical notation
for type expressions developed in this book for easier reasoning about
types in functional programs. Disjunctive types are denoted by $+$,
product types by $\times$, and function types by $\rightarrow$.
The unit type is denoted by $\bbnum 1$, and the void type by $\bbnum 0$.
The function arrow $\rightarrow$ groups weaker than $+$, which in
turn groups weaker than $\times$. This means
\[
Z+A\rightarrow Z+A\times A\quad\text{is the same as}\quad\left(Z+A\right)\rightarrow\left(Z+\left(A\times A\right)\right)\quad.
\]
 Type parameters are denoted by superscripts. As an example, the Scala
definition\texttt{}
\begin{lstlisting}
type F[A] = Either[(A, A => Option[Int]), String => List[A]]
\end{lstlisting}
is written in the type notation as 
\[
F^{A}\triangleq A\times\left(A\rightarrow\bbnum 1+\text{Int}\right)+(\text{String}\rightarrow\text{List}^{A})\quad.
\]
\item [{\index{unfunctor}Unfunctor}] A type constructor that cannot possibly
be a functor, nor a contrafunctor, nor a profunctor. An example is
a type constructor with explicitly indexed type parameters, such as
$F^{A}\triangleq\left(A\times A\right)^{:F^{\text{Int}}}+\left(\text{Int}\times A\right)^{:F^{\bbnum 1}}$.
The Scala code for this type constructor is
\begin{lstlisting}
sealed trait F[A]
final case class F1[A](x: A, y: A)   extends F[Int]
final case class F2[A](s: Int, t: A) extends F[Unit]
\end{lstlisting}
This can be seen as a \index{GADT!see \textsf{``}unfunctor\textsf{''}}\textbf{GADT}
(generalized algebraic data type) that uses specific values of type
parameters instead of the type parameter $A$ in at least some of
its case classes.
\end{description}

\section{How the term \textquotedblleft algebra\textquotedblright{} is used
in functional programming}

This book avoids using the terms \textsf{``}algebra\index{algebra}\textsf{''} or
\textsf{``}algebraic\index{algebraic}\textsf{''} because these terms are too ambiguous.
As we will see, the FP community uses the word \textsf{``}algebra\textsf{''} in at
least \emph{three} incompatible ways.

\paragraph{Definition 0.}

In mathematics, an \textquotedblleft algebra\textquotedblright{} is
a vector space with a multiplication operation ($*$) satisfying certain
properties, such as distributivity: $z*(x+y)=z*x+z*y$. For example,
the set of all $10\times10$ matrices with real coefficients is a
$100$-dimensional vector space that satisfies the required properties.
Such matrices form an \textquotedblleft algebra\textquotedblright{}
in the mathematical sense. (This meaning of the word \textsf{``}algebra\textsf{''}
is \emph{not} used in FP.)

\paragraph{Definition 1.}

For a given functor $F$, an \textsf{``}$F$-algebra\textsf{''}\index{$F$-algebra}
is a type $T$ and a function with type signature $F^{T}\rightarrow T$.
(This definition comes from category theory. This book does not use
$F$-algebras because knowing their category-theoretic properties
does not help in practical programming.) The name \textsf{``}$F$-algebra\textsf{''}
emphasizes the dependence on a chosen functor $F$. There is no direct
connection between this \textsf{``}algebra\textsf{''} and Definition~0, except when
the functor $F$ is defined by $F^{A}\triangleq A\times A$: a function
of type $T\times T\rightarrow T$ may then be interpreted as a \textsf{``}multiplication\textsf{''}
operation for values of type $T$. However, even in that case $T$
will not necessarily satisfy the laws of a vector space.

A recursive type $T$ defined by $F^{T}\cong T$ has an isomorphism
$F^{T}\rightarrow T$ and so is an $F$-algebra. Recursive polynomial
types are known as \textsf{``}algebraic\textsf{''} in this sense:\footnote{For some proofs, see \texttt{\href{https://homepages.inf.ed.ac.uk/wadler/papers/free-rectypes/free-rectypes.txt}{https://homepages.inf.ed.ac.uk/wadler/papers/free-rectypes/free-rectypes.txt}} }
they are $F$-algebras with a polynomial recursion scheme $F^{\bullet}$
and an added requirement of isomorphism for the map $F^{T}\rightarrow T$.

The \textsf{``}algebra\textsf{''} defined in this sense is also relevant to the Church
encoding of a free monad (also known as the \textsf{``}tagless final\index{tagless final}\textsf{''}
pattern), which is the type $\forall E^{\bullet}.\,(S^{E^{\bullet}}\leadsto E^{\bullet})\rightarrow E^{A}$.
This type uses a higher-order type $S^{E^{\bullet}}$ parameterized
by a \emph{type constructor} parameter $E^{\bullet}$. In this context,
one calls an \textsf{``}algebra\textsf{''} a value of type $S^{E^{\bullet}}\leadsto E^{\bullet}$,
which is an $S$-algebra in the category of type constructors. However,
knowing that $S^{E^{\bullet}}\leadsto E^{\bullet}$ is an $S$-algebra
in the category of type constructors does not provide any help or
additional insights for practical work with the Church encoding of
free monads.

The higher-order type constructor $S$ is used to parameterize the
effects described by a Church-encoded free monad, so this book calls
it the \textsf{``}effect constructor\textsf{''}.

\paragraph{Definition 2.}

Polynomial data types (for example, \lstinline!type F[A] = Option[(A, A)]!)
are often called \textquotedblleft algebraic data types\textquotedblright .
Here, the word \textsf{``}algebraic\textsf{''} suggests \textsf{``}school-level algebra\textsf{''}
dealing with polynomials: those data types are built from \textsf{``}sums\textsf{''}
and \textsf{``}products\textsf{''} of types. The type \lstinline!F[A]! may admit
some binary or unary operations, but this does not turn it into an
algebra in the sense of Definitions~0 or 1. Instead of calling types
\textsf{``}algebraic\textsf{''}, this book uses more precise terms \textquotedblleft polynomial
types\textquotedblright{} and \textquotedblleft exponential-polynomial
types\textquotedblright . Data types containing functions, such as
\lstinline!Option[Int => (A, A)]!, are exponential-polynomial (but
not polynomial).

\paragraph{Definition 3.}

One talks about the \textquotedblleft algebra\textquotedblright{}
of properties of functions such as \lstinline!map! or \lstinline!flatMap!,
meaning that these functions satisfy certain \textsf{``}algebraic\textsf{''} laws
(e.g., the composition, naturality, or associativity laws). But these
laws do not make functions \lstinline!map! or \lstinline!flatMap!
into an algebra in the sense of Definition~0 or in the sense of Definition~1.
There is also no relation to the algebraic data types of Definition~2.
So, this is a different usage of the word \textsf{``}algebra\textsf{''}. However,
there is no general \textsf{``}algebra of laws\textsf{''} that we can use; every derivation
proceeds in a different way, specific to the laws being proved. In
mathematics, \textsf{``}algebraic\textsf{''} equations are distinguished from differential
or integral equations. In that sense, the laws used in FP are \emph{always}
algebraic: they are just equations with compositions and applications
of functions. So, there is no advantage in calling laws \textsf{``}algebraic\textsf{''}
within the context of FP. This book calls them \textsf{``}equations\textsf{''} or
just \textsf{``}laws\textsf{''}.

We find that the current usage of the word \textsf{``}algebra\textsf{''} in FP is
both inconsistent and unhelpful to software practitioners. In this
book, the word \textsf{``}algebra\textsf{''} always means a branch of mathematics,
as in \textsf{``}high-school algebra\textsf{''}. Instead of \textsf{``}algebra\textsf{''} as in Definitions~1
to~3, this book talks about \textsf{``}polynomial types\textsf{''} or \textsf{``}polynomial
functors\textsf{''} or \textsf{``}exponential-polynomial functors\textsf{''} etc.; \textsf{``}equations\textsf{''}
or \textsf{``}laws\textsf{''}; and an \textsf{``}effect constructor\textsf{''} $S$. 

The next section presents the LJT algorithm in detail.

\chapter{Inferring code from types. The LJT algorithm\label{app:The-LJT-algorithm}}

Section~\ref{subsec:Example:-Proving-a-ch-proposition} showed that
the rules in Table~\ref{tab:Proof-rules-of-constructive-and-boolean}
do not provide an algorithm for finding a proof for a given sequent.
To illustrate the problem on another example, let us try proving the
sequent:
\[
A,B\vee C\vdash(A\wedge B)\vee C\quad.
\]
This sequent is provable because we can write the corresponding Scala
code:
\begin{lstlisting}
def f[A, B, C](a: A): Either[B, C] => Either[(A, B), C] = {
  case Left(b)    => Left((a, b))
  case Right(c)   => Right(c)
}
\end{lstlisting}
How can we obtain a proof of this sequent according to the rules in
Table~\ref{tab:Proof-rules-of-constructive-and-boolean}? We find
that we could potentially apply the rules \textsf{``}create \lstinline!Left!\textsf{''},
\textsf{``}create \lstinline!Right!\textsf{''}, \textsf{``}use \lstinline!Either!\textsf{''}, and
\textsf{``}use function\textsf{''}. However, we get stuck at the next step. To apply
\textsf{``}create \lstinline!Left!\textsf{''}, we first need to prove the sequent
$A,B\vee C\vdash A\wedge B$. But this sequent cannot be proved: we
do not necessarily have values of both types $A$ and $B$ if we are
only given values of type $A$ and of type \lstinline!Either[B, C]!.
To apply \textsf{``}create \lstinline!Right!\textsf{''}, we need to prove the sequent
$A,B\vee C\vdash C$. Again, we find that this sequent cannot be proved.
The next choice is the rule \textsf{``}use \lstinline!Either!\textsf{''} that matches
any goal of the sequent as the proposition $\gamma$. But we are then
required to choose two new propositions ($\alpha$ and $\beta$) such
that we can prove $A,B\vee C\vdash\alpha\vee\beta$ as well as $A,B\vee C,\alpha\vdash(A\wedge B)\vee C$
and $A,B\vee C,\beta\vdash(A\wedge B)\vee C$. It is not clear how
we should choose $\alpha$ and $\beta$ in order to make progress
in the proof. The remaining rule, \textsf{``}use function\textsf{''}, similarly requires
us to choose a new proposition $\alpha$ such that we can prove $A,B\vee C\vdash\alpha$
and $A,B\vee C\vdash\alpha\Rightarrow((A\wedge B)\vee C)$. Again,
the algorithm gives us no guidance about making a correct choice of
$\alpha$.

We see that the rules in Table~\ref{tab:Proof-rules-for-constructive-logic}
are not helpful for proof search. The main difficulty is that certain
rules (\textsf{``}use function\textsf{''} and \textsf{``}use Either\textsf{''}) require us to choose
new arbitrary propositions and prove more complicated sequents. For
instance, the \textsf{``}use function\textsf{''} rule gives a proof of $\Gamma\vdash\beta$
if we first choose some other proposition $\alpha$ and prove $\Gamma\vdash\alpha$
and $\Gamma\vdash\alpha\Rightarrow\beta$. The rule does not say how
to choose the proposition $\alpha$ correctly. We need to guess the
correct $\alpha$ in each case. Even if we choose $\alpha$ in some
way, we will then have a more complicated sequent to prove. We are
not guaranteed that we are getting closer to finding the proof of
the initial sequent. 

It is far from obvious how to overcome that difficulty. Mathematicians
have studied the constructive logic for more than 60 years, trying
to replace the rules in Table~\ref{tab:Proof-rules-of-constructive-and-boolean}
by a different but equivalent set of derivation rules that require
no guessing when looking for a proof. The first partial success was
G.~Gentzen\textsf{'}s algorithm\footnote{See \texttt{\href{https://en.wikipedia.org/wiki/Sequent_calculus\#Overview}{https://en.wikipedia.org/wiki/Sequent\_calculus\#Overview}}}
that\index{Gerhard Gentzen} he called \textsf{``}LJ\textsf{''}. The LJ algorithm
still had a significant problem: one of its derivation rules could
be applied infinitely many times. So, the LJ algorithm was not guaranteed
to terminate without some heuristics for stopping a possible infinite
loop. A terminating version of the LJ algorithm, called \index{LJT algorithm}LJT,
was formulated in 1992.\footnote{The history of this research is outlined in \texttt{\href{https://research-repository.st-andrews.ac.uk/handle/10023/8824}{https://research-repository.st-andrews.ac.uk/handle/10023/8824}}.
An often cited paper by R.~Dyckhoff\index{Roy Dyckhoff} is \texttt{\href{https://philpapers.org/rec/DYCCSC}{https://philpapers.org/rec/DYCCSC}}.} 

We will first present the LJ algorithm. Although that algorithm does
not guarantee termination, it is simpler to apply by hand. Then we
will show how to pass from the LJ algorithm to LJT.

\section{The LJ algorithm}

Figure~\ref{fig:Rules-of-the-LJ-algorithm} shows the axioms and
the derivation rules of the LJ algorithm. Each rule says that the
bottom sequent will be proved if proofs are given for sequent(s) at
the top. For each possible sub-expression (conjunction $X\wedge Y$,
disjunction $X\vee Y$, and implication $X\Rightarrow Y$) there is
one rule where that sub-expression is a premise and one rule where
that sub-expression is the goal. These sub-expressions are marked
blue in Figure~\ref{fig:Rules-of-the-LJ-algorithm} to help us look
for a proof. To find out which rules apply, we match some part of
the sequent to be proved against a blue sub-expression in the rules.

\begin{figure}
\begin{centering}
\fbox{\begin{minipage}[t]{0.7\linewidth}%
\begin{align*}
\frac{}{\Gamma,X\vdash{\color{blue}X}}~(\text{Id})\qquad & \qquad\frac{}{\Gamma\vdash{\color{blue}\top}}~(\text{True})\\
\frac{\Gamma,A\Rightarrow B\vdash A\quad\quad\Gamma,B\vdash C}{\Gamma,{\color{blue}A\Rightarrow B}\vdash C}~(\text{Left}\Rightarrow)\qquad & \qquad\frac{\Gamma,A\vdash B}{\Gamma\vdash{\color{blue}A\Rightarrow B}}~(\text{Right}\Rightarrow)\\
\frac{\Gamma,A_{i}\vdash C}{\Gamma,{\color{blue}A_{1}\wedge A_{2}}\vdash C}~(\text{Left}\wedge_{i})\qquad & \qquad\frac{\Gamma\vdash A\quad\quad\Gamma\vdash B}{\Gamma\vdash{\color{blue}A\wedge B}}~(\text{Right}\wedge)\\
\frac{\Gamma,A\vdash C\quad\quad\Gamma,B\vdash C}{\Gamma,{\color{blue}A\vee B}\vdash C}~(\text{Left}\vee)\qquad & \qquad\frac{\Gamma\vdash A_{i}}{\Gamma\vdash{\color{blue}A_{1}\vee A_{2}}}~(\text{Right}\vee_{i})
\end{align*}
%
\end{minipage}}
\par\end{centering}
\caption{Axioms and derivation rules of the LJ algorithm. The rules \textquotedblleft ($\text{Left}\wedge_{i}$)\textquotedblright{}
and \textquotedblleft ($\text{Right}\vee_{i}$)\textquotedblright{}
have two versions each, with $i=1$ or $i=2$. \label{fig:Rules-of-the-LJ-algorithm}}
\end{figure}

G.~Gentzen proved that the rules in Figure~\ref{fig:Rules-of-the-LJ-algorithm}
are equivalent to the rules in Table~\ref{tab:Proof-rules-for-constructive-logic}.
The proof is beyond the scope of this book. We only remark that this
equivalence is far from obvious. To prove it, one needs to demonstrate
that all sequents derived through the first set of rules is also derivable
through the second set, and vice versa.

To illustrate the LJ algorithm, we prove the sequent~(\ref{eq:ch-example-sequent-2}).
Denote that sequent by $S_{0}$:
\[
S_{0}\triangleq\emptyset\vdash\left(\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta\right)\Rightarrow\beta\quad.
\]
 Since the goal of $S_{0}$ contains an implication, the rule \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}
applies and gives the sequent $S_{1}$:
\[
S_{1}\triangleq\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta\vdash\beta\quad.
\]
Now the premise contains an implication, so we apply the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}
and get two new sequents:
\[
S_{2}\triangleq\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta\vdash\alpha\Rightarrow\alpha\quad,\quad\quad S_{3}\triangleq\beta\vdash\beta\quad.
\]
Sequent $S_{3}$ follows from the \textsf{``}(Id)\textsf{''} axiom, so it remains
to prove $S_{2}$. Since $S_{2}$ contains an implication both as
a premise and as the goal, we may apply either the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}
or the rule \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}. We choose to apply
\textsf{``}($\text{Left}\Rightarrow$)\textsf{''} and get two new sequents:
\[
S_{4}\triangleq\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta\vdash\alpha\Rightarrow\alpha\quad,\quad\quad S_{5}:\beta\vdash\alpha\Rightarrow\alpha\quad.
\]
Notice that $S_{4}=S_{2}$. This means our proof search is getting
into an infinite loop trying to prove the same sequent $S_{2}$ over
and over again. We can prove $S_{5}$ but this will not help break
the loop.

Once we recognize the problem, we backtrack to the point where we
chose to apply \textsf{``}($\text{Left}\Rightarrow$)\textsf{''} to $S_{2}$. That
was a bad choice, so let us instead apply \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}
to $S_{2}$. This yields a new sequent $S_{6}$:
\[
S_{6}\triangleq\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta,\alpha\vdash\alpha\quad.
\]
This sequent follows from the \textsf{``}(Id)\textsf{''}axiom. There are no more sequents
to prove, so the proof of $S_{0}$ is finished. It can be drawn as
a tree like this:
\[
\xymatrix{\xyScaleY{1.5pc}\xyScaleX{4pc} &  &  & (\text{Id})\\
\ar[r]\sp(0.35){S_{0}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.5){S_{1}} & (\text{Left}\Rightarrow)\ar[r]\sp(0.5){S_{2}}\ar[ru]\sp(0.6){S_{3}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.65){S_{6}} & (\text{Id})
}
\]
The nodes of the tree are axioms or derivation rules, and the edges
are intermediate sequents required by the rules. Each rule node may
branch into one or more sequents because some rules require more than
one new sequent to be proved. The leaves of the tree are axioms that
do not require a proof of any further sequent. 

\subsection{Extracting code from proofs}

In the Curry-Howard correspondence, a sequent (such as $A,B,...,C\vdash X$)
represents the task of writing a fully parametric code expression
of type $X$ that uses some given values of types $A$, $B$, ...,
$C$. The sequent is true (i.e., provable) if that code expression
can be found. So, the code serves as an \textsf{``}evidence of proof\textsf{''} for
the sequent.

\begin{figure}
\begin{centering}
\noindent\fbox{\begin{minipage}[t]{1\columnwidth - 2\fboxsep - 2\fboxrule}%
\begin{align*}
\frac{}{\Gamma,A\vdash{\color{blue}A}}~(\text{Id})\quad & \quad\text{Proof}\,(\Gamma,A\vdash A)_{\text{given }p^{:\Gamma},x^{:A}}=x\\
\frac{}{\Gamma\vdash{\color{blue}\top}}~(\text{True})\quad & \quad\text{Proof}\,(\Gamma\vdash\top)_{\text{given }p^{:\Gamma}}=1\\
\frac{\Gamma,A\Rightarrow B\vdash A\quad\quad\Gamma,B\vdash C}{\Gamma,{\color{blue}A\Rightarrow B}\vdash C}~(\text{Left}\Rightarrow)\quad & \quad\text{Proof}\,(\Gamma,A\Rightarrow B\vdash C)_{\text{given }p^{:\Gamma},q^{:A\rightarrow B}}=\text{Proof}\,(\Gamma,B\vdash C)_{\text{given }p,b^{:B}}\\
 & \quad\quad\text{where}\quad b^{:B}\triangleq q\big(\text{Proof}\,(\Gamma,A\Rightarrow B\vdash A)_{\text{given }p,q}\big)\\
\frac{\Gamma,A\vdash B}{\Gamma\vdash{\color{blue}A\Rightarrow B}}~(\text{Right}\Rightarrow)\quad & \quad\text{Proof}\,(\Gamma\vdash A\Rightarrow B)_{\text{given }p^{:\Gamma}}=x^{:A}\rightarrow\text{Proof}\,(\Gamma,A\vdash B)_{\text{given }p^{:\Gamma},x^{:A}}\\
\frac{\Gamma,A\vdash C}{\Gamma,{\color{blue}A\wedge B}\vdash C}~(\text{Left}\wedge_{1})\quad & \quad\text{Proof}\,(\Gamma,A\wedge B\vdash C)_{\text{given }p^{:\Gamma},(a^{:A}\times b^{:B})}\\
 & \quad\quad=\text{Proof}\,(\Gamma,A\vdash C)_{\text{given }p^{:\Gamma},a^{:A}}\\
\frac{\Gamma,B\vdash C}{\Gamma,{\color{blue}A\wedge B}\vdash C}~(\text{Left}\wedge_{2})\quad & \quad\text{Proof}\,(\Gamma,A\wedge B\vdash C)_{\text{given }p^{:\Gamma},(a^{:A}\times b^{:B})}\\
 & \quad\quad=\text{Proof}\,(\Gamma,B\vdash C)_{\text{given }p^{:\Gamma},b^{:B}}\\
\frac{\Gamma\vdash A\quad\quad\Gamma\vdash B}{\Gamma\vdash{\color{blue}A\wedge B}}~(\text{Right}\wedge)\quad & \quad\text{Proof}\,(\Gamma\vdash A\wedge B)_{\text{given }p^{:\Gamma}}\\
 & \quad\quad=\text{Proof}\,(\Gamma\vdash A)_{\text{given }p^{:\Gamma}}\times\text{Proof}\,(\Gamma\vdash B)_{\text{given }p^{:\Gamma}}\\
\frac{\Gamma,A\vdash C\quad\quad\Gamma,B\vdash C}{\Gamma,{\color{blue}A\vee B}\vdash C}~(\text{Left}\vee)\quad & \quad\text{Proof}\,(\Gamma,A\vee B\vdash C)_{\text{given }p^{:\Gamma},q^{:A+B}}\\
 & \quad\quad=q\triangleright\begin{array}{|c||c|}
 & C\\
\hline A & x^{:A}\rightarrow\text{Proof}\,(\Gamma,A\vdash C)_{\text{given }p,x}\\
B & y^{:B}\rightarrow\text{Proof}\,(\Gamma,B\vdash C)_{\text{given }p,y}
\end{array}\\
\frac{\Gamma\vdash A}{\Gamma\vdash{\color{blue}A\vee B}}~(\text{Right}\vee_{1})\quad & \quad\text{Proof}\,(\Gamma\vdash A\vee B)_{\text{given }p^{:\Gamma}}=\text{Proof}\,(\Gamma\vdash A)+\bbnum 0^{:B}\\
\frac{\Gamma\vdash B}{\Gamma\vdash{\color{blue}A\vee B}}~(\text{Right}\vee_{2})\quad & \quad\text{Proof}\,(\Gamma\vdash A\vee B)_{\text{given }p^{:\Gamma}}=\bbnum 0^{:A}+\text{Proof}\,(\Gamma\vdash B)
\end{align*}
%
\end{minipage}}
\par\end{centering}
\caption{\label{fig:proof-transformers-for-LJ-rules}Proof transformers for
the rules of the LJ algorithm.}
\end{figure}

In the previous section, we have found a proof of the sequent $S_{0}$,
which represents the task of writing a fully parametric function with
type signature $(\left(A\rightarrow A\right)\rightarrow B)\rightarrow B$).
Let us now see how we can extract the code of that function from the
proof of the sequent $S_{0}$.

We start from the leaves of the proof tree and move step by step towards
the initial sequent. At each step, we shorten the proof tree by replacing
some sequent by its corresponding evidence-of-proof code. Eventually
we will replace the initial sequent by its corresponding code. Let
us see how this procedure works for the proof tree of the sequent
$S_{0}$ shown in the previous section.

Since the leaves occur after axiom edges, let us write the code corresponding
to the LJ axioms:
\begin{align*}
 & \frac{}{\Gamma,X\vdash X}~(\text{Id})\quad,\quad\quad\text{Proof}\,(\Gamma,X\vdash X)_{\text{given }p^{:\Gamma},x^{:X}}=x\quad;\\
 & \frac{}{\Gamma\vdash\top}~(\text{True})\quad,\quad\quad\text{Proof}\,(\Gamma\vdash\top)_{\text{given }p^{:\Gamma}}=1\quad.
\end{align*}
Here we denote explicitly the values (such as $p$ and $x$) given
as premises to the sequent. The notation $p^{:\Gamma}$ means all
values given in the set of premises $\Gamma$. Below we will assume
that the propositions $\alpha$ and $\beta$ correspond to types $A$
and $B$; that is, $\alpha\triangleq{\cal CH}(A)$ and $\beta\triangleq{\cal CH}(B)$.

The leaves in the proof tree for $S_{0}$ are the \textsf{``}($\text{Id}$)\textsf{''}
axioms used to prove the sequents $S_{3}$ and $S_{6}$. Let us write
the code that serves as the \textsf{``}evidence of proof\textsf{''} for these sequents.
For brevity, we denote $\gamma\triangleq\left(\alpha\Rightarrow\alpha\right)\Rightarrow\beta$
and $C\triangleq\left(A\rightarrow A\right)\rightarrow B$, so that
$\gamma={\cal CH}(C)$. Then we can write:
\begin{align*}
 & S_{3}\triangleq\beta\vdash\beta\quad,\quad\quad\text{Proof}\,(S_{3})_{\text{given }y^{:B}}=y\quad,\\
 & S_{6}\triangleq\gamma,\alpha\vdash\alpha\quad,\quad\quad\text{Proof}\,(S_{6})_{\text{given }q^{:C},x^{:A}}=x\quad.
\end{align*}
Note that the proof of $S_{6}$ does not use the first given value
$p^{:C}$ (corresponding to the premise $\gamma$).

We can shorten the proof tree by replacing the sequents $S_{3}$ and
$S_{6}$ by their \textsf{``}evidence of proof\textsf{''}:
\[
\xymatrix{\xyScaleY{1.5pc}\xyScaleX{4pc} &  &  & \square\\
\ar[r]\sp(0.35){S_{0}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.5){S_{1}} & (\text{Left}\Rightarrow)\ar[r]\sp(0.5){S_{2}}\ar[ru]\sp(0.6){(y)_{\text{given }y^{:B}}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.65){(x)_{\text{given }q^{:C},x^{:A}}} & \square
}
\]

The next step is to consider the proof of $S_{2}$, which is found
by applying the rule \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}. This rule
promises to give a proof of $S_{2}$ if we have a proof of $S_{6}$.
In order to extract code from that rule, we can write a function that
transforms a proof of $S_{6}$ into a proof of $S_{2}$. We call this
function the \textbf{proof transformer}\index{Curry-Howard correspondence!proof transformer}\index{proof transformer}
corresponding to the rule \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}. That
rule and its transformer are defined as:
\[
\frac{\Gamma,A\vdash B}{\Gamma\vdash A\Rightarrow B}~(\text{Right}\Rightarrow)\quad,\quad\quad\text{Proof}\,(\Gamma\vdash A\Rightarrow B)_{\text{given }p^{:\Gamma}}=x^{:A}\rightarrow\text{Proof}\,(\Gamma,A\vdash B)_{\text{given }p^{:\Gamma},x^{:A}}\quad.
\]
Applying the proof transformer to the known proof of $S_{6}$, we
obtain a proof of $S_{2}$:
\[
\text{Proof}\,(S_{2})_{\text{given }q^{:C}}=x^{:A}\rightarrow\text{Proof}\,(S_{6})_{\text{given }q^{:C},x^{:A}}=(x^{:A}\rightarrow x)_{\text{given }q^{:C}}\quad.
\]
The proof tree can be now shortened to:
\[
\xymatrix{\xyScaleY{1.5pc}\xyScaleX{3.5pc} &  &  & \square\\
\ar[r]\sp(0.35){S_{0}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.5){S_{1}} & (\text{Left}\Rightarrow)\ar[rr]\sp(0.62){(x^{:A}\rightarrow x)_{\text{given }q^{:C}}}\ar[ru]\sp(0.6){(y)_{\text{given }y^{:B}}} &  & \square
}
\]

The next step is to get the proof of $S_{1}$obtained by applying
the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}. That rule requires two
previous sequents, so its transformer is a function of two previously
obtained proofs:
\begin{align*}
 & \frac{\Gamma,A\Rightarrow B\vdash A\quad\quad\Gamma,B\vdash C}{\Gamma,A\Rightarrow B\vdash C}~(\text{Left}\Rightarrow)\quad,\\
 & \text{Proof}\,(\Gamma,A\Rightarrow B\vdash C)_{\text{given }p^{:\Gamma},q^{:A\rightarrow B}}=\text{Proof}\,(\Gamma,B\vdash C)_{\text{given }p,b^{:B}}\\
 & \text{where}\quad b^{:B}\triangleq q\big(\text{Proof}\,(\Gamma,A\Rightarrow B\vdash A)_{\text{given }p,q}\big)\quad.
\end{align*}
Using the proof tree shown above, we obtain the proof of $S_{1}$
by applying this proof transformer to the proofs of $S_{2}$ and $S_{3}$:
\begin{align*}
 & \text{Proof}\,(S_{1})_{\text{given }q^{:C}}=\text{Proof}\,(S_{3})_{\text{given }b}\text{ where }b^{:B}\triangleq q(\text{Proof}\,(S_{2})_{\text{given }q})\\
 & =b\text{ where }b^{:B}\triangleq q(x^{:A}\rightarrow x)=q(x^{:A}\rightarrow x)\quad.
\end{align*}
Substituting this proof, we shorten the proof tree to:
\[
\xymatrix{\xyScaleY{1.5pc}\xyScaleX{3.5pc}\ar[r]\sp(0.35){S_{0}} & (\text{Right}\Rightarrow)\ar[r]\sp(0.65){q(x^{:A}\rightarrow x)} & \square}
\]

It remains to obtain the proof of $S_{0}$ by applying the proof transformer
of the rule \textsf{``}($\text{Right}\Rightarrow$)\textsf{''}:
\begin{align*}
 & \text{Proof}\,(S_{0})=\text{Proof}\,(\emptyset\vdash(\alpha\Rightarrow\alpha)\Rightarrow\beta)\Rightarrow\beta)\\
 & =q^{:(A\rightarrow A)\rightarrow B}\rightarrow\text{Proof}\,(S_{1})_{\text{given }q}=q^{:(A\rightarrow A)\rightarrow B}\rightarrow q(x^{:A}\rightarrow x)\quad.
\end{align*}
The proof tree is now shortened to a single code expression, $q^{:(A\rightarrow A)\rightarrow B}\rightarrow q(x^{:A}\rightarrow x)$,
that serves as an evidence of proof for $S_{0}$ because this code
has type $\left(\left(A\rightarrow A\right)\rightarrow B\right)\rightarrow B$.
In this way, we have derived the code of a fully parametric function
from its type signature.

Figure~\ref{fig:proof-transformers-for-LJ-rules} shows the proof
transformers for all the rules of the LJ algorithm. Apart from the
special rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}, all other rules have
proof transformers using just one of the code constructions (\textsf{``}create
function\textsf{''}, \textsf{``}create tuple\textsf{''}, \textsf{``}use tuple\textsf{''}, etc.) allowed within
fully parametric code.

\section{The LJT algorithm}

As we have seen, the LJ algorithm can enter a loop if we apply the
rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''} and obtain a sequent we already
had at a previous step. That rule requires us to prove two new sequents:
\[
\frac{\Gamma,A\Rightarrow B\vdash A\quad\quad\Gamma,B\vdash C}{\Gamma,A\Rightarrow B\vdash C}~(\text{Left}\Rightarrow)\quad.
\]
A sign of trouble is that the first of these sequents ($\Gamma,A\Rightarrow B\vdash A$)
does not have a simpler form than the initial sequent ($\Gamma,A\Rightarrow B\vdash C$).
So, it is not clear that we are getting closer to completing the proof.
If $A=C$, the new sequent will simply repeat the initial sequent,
immediately creating a loop.

In some cases, a repeated sequent will occur after more than one step.
It is not easy to formulate the condition for stopping the loop or
to demonstrate that the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''} may
be avoided.

The LJT algorithm solves this problem by removing the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}
from the LJ algorithm. Instead, \emph{four} new rules are introduced.
Each of these rules contains a different pattern instead of $A$ in
the premise $A\Rightarrow C$:
\begin{align*}
\text{(}A\text{ is atomic)\,}\frac{\Gamma,A,B\vdash D}{\Gamma,A,{\color{blue}A\Rightarrow B}\vdash D}~(\text{Left}\Rightarrow_{A})\qquad & \qquad\frac{\Gamma,A\Rightarrow B\Rightarrow C\vdash D}{\Gamma,{\color{blue}(A\wedge B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\wedge})\\
\frac{\Gamma,B\Rightarrow C\vdash A\Rightarrow B\quad\quad\Gamma,C\vdash D}{\Gamma,{\color{blue}(A\Rightarrow B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\Rightarrow})\qquad & \qquad\frac{\Gamma,A\Rightarrow C,B\Rightarrow C\vdash D}{\Gamma,{\color{blue}(A\vee B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\vee})
\end{align*}
The rule \textsf{``}$\text{Left}\Rightarrow_{A}$\textsf{''} applies only if the implication
starts with an \textsf{``}atomic\textsf{''} type expression, i.e., a single type parameter
or a unit type. In all other cases, the implication must start with
a conjunction, a disjunction, or an implication, which means that
one of the three remaining rules will apply.

The LJT algorithm retains all the rules in Figure~\ref{fig:proof-transformers-for-LJ-rules}
except the rule \textsf{``}($\text{Left}\Rightarrow$)\textsf{''}, which is replaced
by the four new rules. It is far from obvious that the new rules are
equivalent to the old ones. It took mathematicians several decades
to come up with the LJT rules and to prove their validity. We will
rely on that result and will not attempt to prove it.

The proof transformers for the new rules are shown in Figure~\ref{fig:proof-transformers-for-LJT-rules}.
Figures~\ref{fig:proof-transformers-for-LJ-rules}\textendash \ref{fig:proof-transformers-for-LJ-rules}
define the set of proof transformers sufficient for using the LJT
algorithm in practice. The \lstinline!curryhoward! library\texttt{}\footnote{See \texttt{\href{https://github.com/Chymyst/curryhoward}{https://github.com/Chymyst/curryhoward}}}
implements these proof transformers.

The most complicated of the new rules is the rule \textsf{``}($\text{Left}\Rightarrow_{\Rightarrow}$)\textsf{''}.
Applying that rule\textsf{'}s transformer results in evidence-of-proof code
that is longer than the code obtained via LJ\textsf{'}s rule transformers.
The resulting code will need to be simplified symbolically. 

As an example of this, let us prove the sequent $S_{0}=\emptyset\vdash((\alpha\Rightarrow\alpha)\Rightarrow\beta)\Rightarrow\beta$
from the previous section. At each step, only one LJT rule applies
to each sequent. The initial part of the proof tree looks like this:
\[
\xymatrix{\xyScaleY{1.5pc}\xyScaleX{6pc} &  &  & ~\\
\ar[r]\sp(0.4){\emptyset\vdash((\alpha\Rightarrow\alpha)\Rightarrow\beta)\Rightarrow\beta} & (\text{Right}\Rightarrow)\ar[r]\sp(0.5){(\alpha\Rightarrow\alpha)\Rightarrow\beta\vdash\beta} & (\text{Left}\Rightarrow_{\Rightarrow})\ar[ru]\sp(0.65){\beta\vdash\beta}\ar[r]\sp(0.65){\alpha\Rightarrow\beta\vdash\alpha\Rightarrow\alpha} & ~
}
\]
The proofs for the sequents $\beta\vdash\beta$ and $\alpha\Rightarrow\beta\vdash\alpha\Rightarrow\alpha$
are the same as before:
\[
\text{Proof}\,(\beta\vdash\beta)_{\text{given }y^{:B}}=y\quad,\quad\quad\text{Proof}\,(\alpha\Rightarrow\beta\vdash\alpha\Rightarrow\alpha)_{\text{given }r^{:\alpha\Rightarrow\beta}}=x^{:A}\rightarrow x\quad.
\]
Substituting these proofs into the proof transformer of the rule \textsf{``}($\text{Left}\Rightarrow_{\Rightarrow}$)\textsf{''}
produces this code:
\begin{align*}
 & \text{Proof}\,((\alpha\Rightarrow\alpha)\Rightarrow\beta\vdash\beta)_{\text{given }q^{:(A\rightarrow A)\rightarrow B}}=q\big(\text{Proof}\,(\alpha\Rightarrow\beta\vdash\alpha\Rightarrow\alpha)_{\text{given }r^{:A\rightarrow B}}\big)\\
 & \quad\text{where }r^{:A\rightarrow B}=a^{:A}\rightarrow q(\_^{:A}\rightarrow a)\\
 & =q(x^{:A}\rightarrow x)\quad.
\end{align*}
The proof of $\alpha\Rightarrow\beta\vdash\alpha\Rightarrow\alpha$
does not actually use the intermediate value $r^{:A\rightarrow B}$
provided by the proof transformer. So, a symbolic simplification will
simply omit that code. The \lstinline!curryhoward! library always
performs symbolic simplification after applying the LJT algorithm. 

\begin{figure}
\begin{centering}
\noindent\fbox{\begin{minipage}[t]{1\columnwidth - 2\fboxsep - 2\fboxrule}%
\begin{align*}
\frac{\Gamma,A,B\vdash D}{\Gamma,A,{\color{blue}A\Rightarrow B}\vdash D}~(\text{Left}\Rightarrow_{A})\quad & \quad\text{Proof}\,(\Gamma,A,A\Rightarrow B\vdash D)_{\text{given }p^{:\Gamma},x^{:A},q^{:A\rightarrow B}}\\
 & \quad\quad=\text{Proof}\,(\Gamma,A,B\vdash D)_{\text{given }p,x,q(x)}\\
\frac{\Gamma,A\Rightarrow B\Rightarrow C\vdash D}{\Gamma,{\color{blue}(A\wedge B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\wedge})\quad & \quad\text{Proof}\,(\Gamma,(A\wedge B)\Rightarrow C\vdash D)_{\text{given }p^{:\Gamma},q^{:A\times B\rightarrow C}}\\
 & \quad\quad=\text{Proof}\,(\Gamma,A\Rightarrow B\Rightarrow C\vdash D)_{\text{given }p,(a^{:A}\rightarrow b^{:B}\rightarrow q(a\times b))}\\
\frac{\Gamma,A\Rightarrow C,B\Rightarrow C\vdash D}{\Gamma,{\color{blue}(A\vee B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\vee})\quad & \quad\text{Proof}\,(\Gamma,(A\vee B)\Rightarrow C\vdash D)_{\text{given }p^{:\Gamma},q^{:A+B\rightarrow C}}\\
 & \quad\quad=\text{Proof}\,(\Gamma,A\Rightarrow C,B\Rightarrow C\vdash D)_{\text{given }p,r,s}\\
 & \quad\quad\text{where}~r\triangleq a^{:A}\rightarrow q(a+\bbnum 0)\text{ and }s\triangleq b^{:B}\rightarrow q(\bbnum 0+b)\\
\frac{\Gamma,B\Rightarrow C\vdash A\Rightarrow B\quad\quad\Gamma,C\vdash D}{\Gamma,{\color{blue}(A\Rightarrow B)\Rightarrow C}\vdash D}~(\text{Left}\Rightarrow_{\Rightarrow})\quad & \quad\text{Proof}\,(\Gamma,(A\Rightarrow B)\Rightarrow C\vdash D)_{\text{given }p^{:\Gamma},q^{:\left(A\rightarrow B\right)\rightarrow C}}\\
 & \quad\quad=\text{Proof}\,(\Gamma,C\vdash D)_{\text{given }p,c}\\
 & \quad\quad\text{ where}~c^{:C}\triangleq q\big(\text{Proof}\,(\Gamma,B\Rightarrow C\vdash A\Rightarrow B)_{\text{given }p,r}\big)\\
 & \quad\quad\text{ and }r^{:B\rightarrow C}\triangleq b^{:B}\rightarrow q(\_^{:A}\rightarrow b)
\end{align*}
%
\end{minipage}}
\par\end{centering}
\caption{\label{fig:proof-transformers-for-LJT-rules}Proof transformers for
the four new rules of the \index{LJT algorithm|textit}LJT algorithm.}
\end{figure}

The reason the LJT algorithm terminates is that each rule replaces
a given sequent by one or more sequents with simpler premises or goals.\footnote{The paper \texttt{\href{http://citeseer.ist.psu.edu/viewdoc/summary?doi=10.1.1.35.2618}{http://citeseer.ist.psu.edu/viewdoc/summary?doi=10.1.1.35.2618}}
shows that the LJT algorithm terminates by giving gives an explicit
decreasing measure on the proof tree.} This guarantees that the proof search will terminate either with
a complete proof or with a sequent to which no further rules can be
applied. An example of such a sequent is $\alpha\vdash\beta$ where
$\alpha$ and $\beta$ are different, unrelated propositions. In that
situation, the LJT algorithm concludes that the initial sequent cannot
be proved.

To \emph{prove} that there is no proof, one needs to use methods
of proof theory that are beyond the scope of this book. A good introduction
to the required technique is the book \textsf{``}\emph{Proof and Disproof
in Formal Logic}\textsf{''} by R.~Bornat\index{Richard Bornat} (see footnote~\ref{fn:Bornat-proof-book}
on page~\pageref{fn:Bornat-proof-book}).

\chapter{Parametricity theorem and naturality laws\label{app:Proofs-of-naturality-parametricity}}

Functional programming (FP) focuses on a small set of language features
\textemdash{} the six type constructions and the nine code constructions\index{nine code constructions},
introduced in Sections~\ref{subsec:Type-notation-and-standard-type-constructions}
and~\ref{subsec:The-rules-of-proof}. These constructions, summarized
again in Tables~\ref{tab:Mathematical-notation-for-basic-code-constructions}
and \ref{tab:six-pure-type-constructions}\textendash \ref{tab:nine-pure-code-constructions},
create \textbf{fully parametric} programs\index{fully parametric!code}
and are sufficient to implement all design patterns of FP. At the
same time, restricting programs to be fully parametric will enable
several ways of mathematical reasoning about code. Examples of such
reasoning are treating programs as mathematical values (referential
transparency\index{referential transparency}); deriving the most
general type from code (type inference\index{type inference}); and
deriving code from type (code inference\index{code inference}).

Additionally, all fully parametric programs will automatically satisfy
certain laws derived from the so-called \textsf{``}parametricity theorem\textsf{''}.
The most often used laws of this kind are naturality laws for natural
transformations, i.e., for functions of type $\forall A.\,F^{A}\rightarrow G^{A}$
where both $F^{\bullet}$ and $G^{\bullet}$ are functors (or both
contrafunctors). Not having to verify naturality laws by hand saves
a lot of time.

Other laws that follow automatically from parametricity are composition
laws for functors, commutativity laws for bifunctors, and dinaturality
laws for profunctors. The goal of this Appendix is to derive those
laws as well as to prove the parametricity theorem. 

\begin{wraptable}{l}{0.545\columnwidth}%
\begin{centering}
\vspace{-0.2\baselineskip}
\begin{tabular}{|c|c|c|}
\hline 
\textbf{\small{}Type construction} & \textbf{\small{}Scala example} & \textbf{\small{}Type notation}\tabularnewline
\hline 
\hline 
{\small{}unit or \textsf{``}named unit\textsf{''}} & {\small{}}\lstinline!Unit!{\small{} or }\lstinline!None!{\small{} } & {\small{}$\bbnum 1$}\tabularnewline
\hline 
{\small{}type parameter} & {\small{}}\lstinline!A!{\small{} as in }\lstinline!F[A]!{\small{} } & {\small{}$A$ as in $F^{A}$}\tabularnewline
\hline 
{\small{}product type} & {\small{}}\lstinline!(A, B)! & {\small{}$A\times B$}\tabularnewline
\hline 
{\small{}co-product type} & {\small{}}\lstinline!Either[A, B]! & {\small{}$A+B$}\tabularnewline
\hline 
{\small{}function type} & {\small{}}\lstinline!A => B! & {\small{}$A\rightarrow B$}\tabularnewline
\hline 
{\small{}recursive type} & {\small{}}\lstinline!Fix[F[_]]! & {\small{}$\text{Fix}^{F^{\bullet}}$}\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\caption{\index{fully parametric!type constructions}The six type constructions
that may be used in fully parametric programs.\label{tab:six-pure-type-constructions}}
\vspace{-0.2\baselineskip}
\end{wraptable}%

\begin{table}
\begin{centering}
\begin{tabular}{|c|c|c|}
\hline 
\textbf{\small{}Code construction} & \textbf{\small{}Scala example} & \textbf{\small{}Code notation}\tabularnewline
\hline 
\hline 
{\small{}1. Use unit value} & {\small{}}\lstinline!()! & {\small{}$1$}\tabularnewline
\hline 
{\small{}2. Use given argument} & {\small{}}\lstinline!x! & {\small{}$x$}\tabularnewline
\hline 
{\small{}3. Create function} & {\small{}}\lstinline!x => expression! & {\small{}$x\rightarrow expression$}\tabularnewline
\hline 
{\small{}4. Use function} & {\small{}}\lstinline!f(x)! & $f(x)$ ~or~ $x\triangleright f$\tabularnewline
\hline 
{\small{}5. Create tuple} & {\small{}}\lstinline!(a, b)! & {\small{}$a\times b$}\tabularnewline
\hline 
{\small{}6. Use tuple} & {\small{}}\lstinline!{ case (a, b) => ... }!{\small{} or }\lstinline!p._1!{\small{}
or }\lstinline!p._2!{\small{} } & {\small{}$a\times b\rightarrow...$ ~or~ $p\triangleright\pi_{1}$
~or~ $p\triangleright\pi_{2}$}\tabularnewline
\hline 
{\small{}7. Create disjunctive value} & {\small{}}\lstinline!Left[A, B](a)! & {\small{}}%
\begin{minipage}[c]{0.19\columnwidth}%
{\small{}\vspace{0.2\baselineskip}
$a^{:A}+\bbnum 0^{:B}$ ~or~~ $\begin{array}{|cc|}
a & \bbnum 0\end{array}$\vspace{0.2\baselineskip}
}%
\end{minipage}\tabularnewline
\hline 
{\small{}8. Use disjunctive value} & {\small{}}%
\begin{minipage}[c]{0.33\columnwidth}%
{\small{}}\lstinline!p match { case Left(a)   => f(a)            case Right(b)  => g(b)          }!%
\end{minipage} & {\small{}}%
\begin{minipage}[c]{0.23\columnwidth}%
{\small{}\vspace{0.2\baselineskip}
$p\,\triangleright\,\begin{array}{|c||c|}
 & C\\
\hline A & a\rightarrow f(a)\\
B & b\rightarrow g(b)
\end{array}$\vspace{0.2\baselineskip}
}%
\end{minipage}\tabularnewline
\hline 
{\small{}9. Recursive call} & {\small{}}\lstinline!def f(x) = { ... f(y) ... }! & {\small{}$f(x)\triangleq...~\overline{f}(y)~...$}\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\caption{\index{fully parametric!code constructions}The nine code constructions
of fully parametric programs.\label{tab:nine-pure-code-constructions}}
\end{table}

The practical uses of parametricity were shown in \index{Philip Wadler}P.~Wadler\textsf{'}s
paper\footnote{See \texttt{\href{https://homepages.inf.ed.ac.uk/wadler/papers/free/free.ps}{https://homepages.inf.ed.ac.uk/wadler/papers/free/free.ps}}
and some further explanations in the blog posts \texttt{\href{https://reasonablypolymorphic.com/blog/theorems-for-free/}{https://reasonablypolymorphic.com/blog/theorems-for-free/}}
and \texttt{\href{https://bartoszmilewski.com/2014/09/22/}{https://bartoszmilewski.com/2014/09/22/}}\\
The title of the paper (\textsf{``}Theorems for free\textsf{''}\index{theorems for free})
is misleading. As we will see shortly, the theorems are obtained only
at the cost of restricting the code to the nine FP code constructions,
which would exclude most of the code in most real-world applications
today.} that was based on results derived by J.~Reynolds in 1983.\index{John Reynolds}\footnote{This book does not use those results. See the paper \texttt{\href{https://people.mpi-sws.org/~dreyer/tor/papers/reynolds.pdf}{https://people.mpi-sws.org/$\sim$dreyer/tor/papers/reynolds.pdf}}}
The Reynolds-Wadler approach to parametricity begins by using many-to-many
relations between types instead of functions. One first needs to learn
how to work with such relations.\footnote{Beginner-level tutorials on relations and relational parametricity
are rare. One is in a paper by \index{Ronald Backhouse}R.~Backhouse,
see \texttt{\href{https://www.researchgate.net/publication/262348393_On_a_Relation_on_Functions}{https://www.researchgate.net/publication/262348393\_On\_a\_Relation\_on\_Functions}}} The parametricity theorem yields a certain property of relations.
Then those relations must be somehow replaced by functions because
we are usually interested in laws having the form of equations for
functions. 

Certain parametricity properties can be proved without using relations,
but relations are proofs are technically simpler. However, a practicing
programmer who only uses the \emph{formulations} of the resulting
laws will not need to learn those proofs. Adapting the parametricity
theorem to the needs of FP practitioners, this Appendix will begin
by summarizing the naturality laws and other properties of fully parametric
code. Then Section~\ref{sec:Parametricity-theorem-for-relations}
will explain the relational approach to proving the parametricity
properties. %
\begin{comment}
Section~\ref{sec:Commutativity-laws-for-type-constructors} proves
that fully parametric type constructors obey commutativity laws. Section~\ref{sec:Naturality-laws-for-fully-parametric-functions}
shows, without using relations, that fully parametric functions satisfy
dinaturality laws (a generalization of naturality laws to arbitrary
type signatures). An important consequence, proved in Section~\ref{sec:Uniqueness-of-functor-and-contrafunctor},
is that the code implementing the functor and contrafunctor typeclasses
is unique. Finally, Section~\ref{sec:Parametricity-theorem-for-relations}
explains the relational approach to parametricity and proves the relational
parametricity theorem. 

Additional literature:

{[}1{]}: Girard, J.-Y.; Scedrov, A. \& Scott, P. J. Normal Forms and
Cut-Free Proofs as Natural Transformations. Logic From Computer Science,
Mathematical Science Research Institute Publications 21, Springer-Verlag,
1992, 217-241. http://citeseer.ist.psu.edu/viewdoc/summary?doi=10.1.1.41.811

{[}2{]}: Bainbridge, E. S.; Freyd, P. J.; Scedrov, A. \& Scott, P.
J. Functorial polymorphism. Theoretical computer science, Elsevier,
1990, 70, 35-64. https://core.ac.uk/display/82270459

{[}3{]}: De Lataillade, J. Dinatural Terms in System F. Logic in Computer
Science, 24th Annual IEEE Symposium, 267-276, 2009. https://www.irif.fr/\textasciitilde delatail/dinat.pdf

{[}4{]}: Pistone, P. On completeness and parametricity in the realizability
semantics of System F. https://arxiv.org/abs/1802.05143

{[}5{]}: https://libres.uncg.edu/ir/asu/f/Johann\_Patricia\_2014\_A\_Relationally\_Parametric\_Model\_Of\_Dependent\_Type\_Theory..pdf

See discussion here: https://cstheory.stackexchange.com/questions/42256/is-case-analysis-on-normal-forms-of-lambda-terms-sufficient-to-prove-parametrici
\end{comment}

Parametricity properties only apply to \emph{fully parametric} programs.\index{fully parametric!program}
This restriction prohibits, for instance, any use of mutable values,
\lstinline!null! values, multithreading, exceptions, run-time JVM
reflection, or any use of code defined in external libraries that
are not known to be fully parametric. Code that uses those features
of Scala is not covered by the parametricity theorem and may fail
naturality laws. An example of such code is the following function:
\begin{lstlisting}
def headOptBad[A]: List[A] => Option[A] = {
  case Nil                   => None
  case (head: Int) :: tail   => Some((head + 100).asInstanceOf[A])
  case head :: tail          => Some(head)
}
\end{lstlisting}
This code has no side effects, is purely functional and referentially\index{referential transparency}
transparent. But it is clear that \lstinline!headOptBad[A]! does
not work in the same way for all type parameters \lstinline!A!. When
\lstinline!A! is set to \lstinline!Int!, the behavior is different
than for all other types. This leads to a failure of the naturality
law:
\begin{lstlisting}
scala> headOptBad[String](  List(1, 2, 3).map(x => s"value = $x")  )
res0: Option[String] = Some(value = 1)

scala> headOptBad[Int]( List(1, 2, 3) ).map(x => s"value = $x")
res1: Option[String] = Some(value = 101)
\end{lstlisting}
So, full parametricity is a stronger restriction on code than referential
transparency.

Here is an example of fully parametric code that uses all nine constructions:
\begin{lstlisting}
def fmap[A, B](f: A => B): List[(A, A)] => List[(B, B)] = {      // 3
   case Nil            => Nil
//   8   1                1,7 
   case head :: tail   => (f (head._1), f (head._2)) :: fmap(f)(tail)
//   8       6             2 4     6  5 2 4     6    7   9
}  // This code uses each of the nine fully parametric constructions.
\end{lstlisting}
For instance, the expression \lstinline!head :: tail! is a tuple
pattern that extracts the two parts of a tuple (construction 6). The
recursive call \lstinline!fmap(f)(tail)! corresponds to construction
9.

\section{Practical use of parametricity properties}

\subsection{Naturality and dinaturality laws\label{sec:Naturality-laws-for-fully-parametric-functions} }

Parametricity properties apply to fully parametric expressions (most
often, functions) with at least one type parameter. Examples of such
functions are the \lstinline!map!, \lstinline!filter!, and \lstinline!fold!
methods for the \lstinline!Option! functor (denoted by $\text{Opt}^{A}\triangleq\bbnum 1+A$),
with the following type signatures:
\begin{align*}
 & \text{fmap}_{\text{Opt}}^{A,B}:\left(A\rightarrow B\right)\rightarrow\text{Opt}^{A}\rightarrow\text{Opt}^{B}\quad,\\
 & \text{filt}_{\text{Opt}}^{A}:\left(A\rightarrow\bbnum 2\right)\rightarrow\text{Opt}^{A}\rightarrow\text{Opt}^{A}\quad,\\
 & \text{fold}_{\text{Opt}}^{A,B}:B\times\left(A\times B\rightarrow B\right)\rightarrow\text{Opt}^{A}\rightarrow B\quad.
\end{align*}
These methods satisfy appropriate naturality laws \textemdash{} one
law per type parameter. 

When a function\textsf{'}s type signature is that of a \emph{natural transformation}\index{natural transformation}
between functors or between contrafunctors, the naturality laws have
the form derived in Section~\ref{subsec:Naturality-laws-and-natural-transformations}.
If $p:\forall A.\,G^{A}\rightarrow H^{A}$ where $G^{\bullet}$ and
$H^{\bullet}$ are functors then $p$ obeys the naturality law:
\begin{equation}
\text{for all }A,B,f^{:A\rightarrow B}\quad:\quad p^{A}\bef f^{\uparrow H}=f^{\uparrow G}\bef p^{B}\quad.\label{eq:naturality-law-for-functors}
\end{equation}
If $p:\forall A.\,G^{A}\rightarrow H^{A}$ where $G^{\bullet}$ and
$H^{\bullet}$ are contrafunctors then $p$ obeys the naturality law:
\begin{equation}
\text{for all }A,B,f^{:A\rightarrow B}\quad:\quad f^{\downarrow G}\bef p^{A}=p^{B}\bef f^{\downarrow H}\quad.\label{eq:naturality-law-for-contrafunctors}
\end{equation}

For example, if we fix the type parameter $A$ in the \lstinline!fmap!
method, we obtain a type signature of the form $F^{B}\rightarrow G^{B}$
where $F^{\bullet}$ and $G^{\bullet}$ are functors:
\[
\text{fmap}_{\text{Opt}}^{A,B}:F^{B}\rightarrow G^{B}\quad,\quad\quad F^{B}\triangleq A\rightarrow B\quad,\quad\quad G^{B}\triangleq\text{Opt}^{A}\rightarrow\text{Opt}^{B}\quad.
\]
The corresponding naturality law is:
\[
\text{fmap}_{\text{Opt}}\bef f^{\uparrow G}=f^{\uparrow F}\bef\text{fmap}_{\text{Opt}}\quad.
\]

Fixing the type parameter $B$ in $\text{fold}_{\text{Opt}}^{A,B}$,
we find a type signature of the form:
\[
\text{fold}_{\text{Opt}}^{A,B}:H^{A}\rightarrow K^{A}\quad,\quad\quad H^{A}\triangleq B\times\left(A\times B\rightarrow B\right)\quad,\quad\quad K^{A}\triangleq\text{Opt}^{A}\rightarrow B\quad,
\]
where $K$ and $H$ are contrafunctors. The corresponding naturality
law is:
\[
f^{\downarrow H}\bef\text{fold}_{\text{Opt}}=\text{fold}_{\text{Opt}}\bef f^{\downarrow K}\quad.
\]

However, fixing the type parameter $A$ in $\text{fold}_{\text{Opt}}^{A,B}$
gives the type signature that is \emph{not} of the form $P^{B}\rightarrow Q^{B}$
with any functors or contrafunctors $P$, $Q$ because $B$ is found
in too many places. In general, it is not obvious how to write naturality
laws for transformations with such type signatures. 

To make progress, note that each occurrence of a type parameter in
a type signature must be either in a covariant or in a contravariant
position. So, any fully parametric type signature may be written in
the form $\forall A.\:P^{A,A}$ using some \emph{profunctor} $P^{X,Y}$
(contravariant in $X$ and covariant in $Y$). For instance, we can
write in this way the type signature of \lstinline!fold! with the
parameter $A$ fixed:
\[
\text{fold}_{\text{Opt}}^{A,B}:P^{B,B}\quad,\quad\quad\text{where}\quad P^{X,Y}\triangleq X\times(A\times Y\rightarrow X)\rightarrow\text{Opt}^{A}\rightarrow Y\quad.
\]
The profunctor $R^{X,Y}$ is complicated but the form of the type
signature is conceptually simpler.

Given a fully parametric expression $p$ with the type signature $p:\forall A.\,P^{A,A}$,
where $P^{X,Y}$ is some profunctor, one can write a the \textbf{wedge
law}\index{wedge law!of profunctors} of $p$:

\begin{wrapfigure}[9]{l}{0.22\columnwidth}%
\vspace{-1.5\baselineskip}
\[
\xymatrix{\xyScaleY{2.0pc}\xyScaleX{2.6pc} & P^{B,B}\ar[d]\sp(0.37){f^{\downarrow P^{\bullet,B}}}\\
P^{A,A}\ar[r]\sp(0.55){f^{\uparrow P^{A,\bullet}}} & P^{A,B}
}
\]

\vspace{-0.4\baselineskip}
\end{wrapfigure}%

~\vspace{-0.8\baselineskip}

\begin{equation}
\text{for all }A,B,f^{:A\rightarrow B}\quad:\quad p^{A}\triangleright f^{\uparrow P^{A,\bullet}}=p^{B}\triangleright f^{\downarrow P^{\bullet,B}}\quad.\label{eq:wedge-law-for-profunctors}
\end{equation}
\vspace{-0.8\baselineskip}

\noindent The diagram is read by starting with a single value $p:\forall A.\:P^{A,A}$.
We apply $p$ to types $A$ and $B$ and obtain two values, $p^{A}$
and $p^{B}$, of types $P^{A,A}$ and $P^{B,B}$. The values $p^{A}$
and $p^{B}$ must then satisfy Eq.~(\ref{eq:wedge-law-for-profunctors}). 

We will prove later in this Appendix that the wedge law~(\ref{eq:wedge-law-for-profunctors})
is a consequence of the parametricity theorem. In turn, the naturality
laws~(\ref{eq:naturality-law-for-functors}) and~(\ref{eq:naturality-law-for-contrafunctors})
are consequences of the wedge law, as will be shown in Statement~\ref{subsec:Statement-naturality-laws-from-wedge-law}
below.

The wedge law is not easy to use in practice because the profunctors
$P$ are often complicated (as the example of \lstinline!fold! shows).
Since one is most often interested in naturality laws for \emph{functions},
it helps to specialize f the wedge law to the case when $P^{A,A}=G^{A,A}\rightarrow H^{A,A}$
with some profunctors $G^{X,Y}$ and $H^{X,Y}$. The result (also
derived in Statement~\ref{subsec:Statement-naturality-laws-from-wedge-law})
is called the \textbf{dinaturality law}\index{dinaturality law}:
\begin{equation}
\text{for all }A,B,f^{:A\rightarrow B}\quad:\quad f^{\downarrow G^{\bullet,A}}\bef p^{A}\bef f^{\uparrow H^{A,\bullet}}=f^{\uparrow G^{B,\bullet}}\bef p^{B}\bef f^{\downarrow H^{\bullet,B}}\quad.\label{eq:dinaturality-law-for-profunctors}
\end{equation}

\begin{wrapfigure}{l}{0.38\columnwidth}%
\vspace{-2.55\baselineskip}
\[
\xymatrix{\xyScaleY{1.8pc}\xyScaleX{2.0pc} & G^{A,A}\ar[r]\sb(0.5){p^{A}} & H^{A,A}\ar[rd]\sb(0.4){f^{\uparrow H^{A,\bullet}}}\\
G^{B,A}\negthickspace\negthickspace\negthickspace\ar[rd]\sp(0.55){\;\;f^{\uparrow G^{B,\bullet}}}\ar[ru]\sb(0.6){f^{\downarrow G^{\bullet,A}}} &  &  & \negthickspace\negthickspace\negthickspace H^{A,B}\\
 & G^{B,B}\ar[r]\sp(0.5){p^{B}} & H^{B,B}\ar[ru]\sp(0.45){f^{\downarrow H^{B,\bullet}}}
}
\]

\vspace{-1.7\baselineskip}
\end{wrapfigure}%

The diagram at left illustrates the dinaturality law as an equation
between functions of type $G^{B,A}\rightarrow H^{A,B}$. To build
up intuition for that law, notice that Eq.~(\ref{eq:dinaturality-law-for-profunctors})
combines the laws~(\ref{eq:naturality-law-for-functors})\textendash (\ref{eq:naturality-law-for-contrafunctors})
in the way required for all types to match. On the other hand, the
laws~(\ref{eq:naturality-law-for-functors}) and~(\ref{eq:naturality-law-for-contrafunctors})
will follow from Eq.~(\ref{eq:dinaturality-law-for-profunctors})
when $G^{A,A}$ and $H^{A,A}$ are both functors or both contrafunctors
with respect to $A$.

Functions $p:\forall A.\,G^{A,A}\rightarrow H^{A,A}$ satisfying Eq.~(\ref{eq:dinaturality-law-for-profunctors})
are called \textbf{dinatural transformations}\index{dinatural transformation}.
The property of dinaturality is weaker than that of naturality.\emph{
}A \emph{natural} transformation between profunctors $G$ and $H$
would have a different type signature, namely $\forall(A,B).\,G^{A,B}\rightarrow H^{A,B}$,
where the type parameters $A$, $B$ are independent. In other words,
this requires us to have a function $t:G^{A,B}\rightarrow H^{A,B}$
defined for arbitrary (not necessarily equal) type parameters $A$,
$B$. If we are only given a transformation with the type signature
$p:G^{A,A}\rightarrow H^{A,A}$, as a rule, the code of $p$ cannot
be extended to some $t$ that works with arbitrary type parameters
$A$, $B$ and has type signature $t:G^{A,B}\rightarrow H^{A,B}$.

With the formulas~(\ref{eq:naturality-law-for-functors})\textendash (\ref{eq:dinaturality-law-for-profunctors}),
we can write naturality laws more quickly, starting from any given
type signature $\forall A.\:P^{A,A}$. The following examples illustrate
how we may derive a naturality law by specializing the general law~(\ref{eq:dinaturality-law-for-profunctors})
to certain profunctors $G$ and $H$.

\subsubsection{Example \label{subsec:Example-derive-naturality-of-filter-from-dinaturality}\ref{subsec:Example-derive-naturality-of-filter-from-dinaturality}
(naturality law of \lstinline!filter!)\index{solved examples}}

To derive the naturality law of \lstinline!filter!, express \lstinline!filter!\textsf{'}s
type signature through profunctors $G$ and $H$ as:
\[
\text{filt}_{F}^{A}:G^{A,A}\rightarrow H^{A,A}\quad,\quad\quad G^{X,Y}\triangleq(X\rightarrow\bbnum 2)\quad,\quad\quad H^{X,Y}\triangleq F^{X}\rightarrow F^{Y}\quad,
\]
and then write the law~(\ref{eq:dinaturality-law-for-profunctors}):
\begin{equation}
f^{\downarrow G^{\bullet,A}}\bef\text{filt}_{F}^{A}\bef f^{\uparrow H^{A,\bullet}}\overset{?}{=}f^{\uparrow G^{B,\bullet}}\bef\text{filt}_{F}^{B}\bef f^{\downarrow H^{\bullet,B}}\quad.\label{eq:filter-law-via-dinatural-transformation-derivation1}
\end{equation}
It remains to substitute the code for the liftings using the specific
types of $H$ and $G$:
\begin{align*}
(f^{:A\rightarrow B})^{\downarrow G^{\bullet,A}}=p^{:B\rightarrow\bbnum 2}\rightarrow f\bef p\quad, & \quad\quad f^{\uparrow G^{B,\bullet}}=\text{id}\quad,\\
(f^{:A\rightarrow B})^{\downarrow H^{\bullet,B}}=q^{:F^{B}\rightarrow F^{B}}\rightarrow f^{\uparrow F}\bef q\quad, & \quad\quad f^{\uparrow H^{A,\bullet}}=q^{:F^{A}\rightarrow F^{A}}\rightarrow q\bef f^{\uparrow F}\quad.
\end{align*}
Then we rewrite Eq.~(\ref{eq:filter-law-via-dinatural-transformation-derivation1})
as:
\[
(p\rightarrow f\bef p)\bef\text{filt}_{F}\bef(q\rightarrow q\bef f^{\uparrow F})\overset{?}{=}\text{id}\bef\text{filt}_{F}\bef(q\rightarrow f^{\uparrow F}\bef q)\quad.
\]
To simplify the form of the naturality law, apply both sides to an
arbitrary $p^{:P^{B,A}}=p^{:B\rightarrow\bbnum 2}$:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p\triangleright(p\rightarrow f\bef p)\bef\text{filt}_{F}\bef(q\rightarrow q\bef f^{\uparrow F})\\
{\color{greenunder}\triangleright\text{-notation}:}\quad & \quad=\gunderline{p\triangleright(p}\rightarrow f\bef p)\triangleright\text{filt}_{F}\triangleright(q\rightarrow q\bef f^{\uparrow F})\\
{\color{greenunder}\text{apply functions}:}\quad & \quad=\gunderline{(f\bef p)\triangleright\text{filt}_{F}}\triangleright(q\rightarrow q\bef f^{\uparrow F})=\gunderline{\text{filt}_{F}(f\bef p)\triangleright(q}\rightarrow q\bef f^{\uparrow F})=\text{filt}_{F}(f\bef p)\bef f^{\uparrow F}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & p\triangleright\gunderline{\text{id}\bef}\text{filt}_{F}\bef(q\rightarrow f^{\uparrow F}\bef q)=p\triangleright\text{filt}_{F}\triangleright(q\rightarrow f^{\uparrow F}\bef q)\\
 & \quad=\gunderline{\text{filt}_{F}(p)\triangleright(q}\rightarrow f^{\uparrow F}\bef q)=f^{\uparrow F}\bef\text{filt}_{F}(p)\quad.
\end{align*}
We obtained the naturality law~(\ref{eq:naturality-law-of-filter})
of \lstinline!filter!:
\[
\text{filt}_{F}(f\bef p)\bef f^{\uparrow F}=f^{\uparrow F}\bef\text{filt}_{F}(p)\quad.
\]


\subsubsection{Example \label{subsec:Example-derive-naturality-of-fold-from-dinaturality}\ref{subsec:Example-derive-naturality-of-fold-from-dinaturality}
(naturality law of \lstinline!fold!)}

To derive the dinaturality law of \lstinline!fold! with respect to
the type parameter $B$, we begin by writing the type signature of
\lstinline!fold! as a transformation of type $G^{B,B}\rightarrow H^{B,B}$
between profunctors $G$, $H$:
\[
\text{fold}_{F}^{A,B}:G^{B,B}\rightarrow H^{B,B}\quad,\quad\quad\text{where}\quad G^{X,Y}\triangleq Y\times(A\times X\rightarrow Y)\quad\text{ and }H^{X,Y}\triangleq F^{A}\rightarrow Y\quad.
\]
Since the type parameter $A$ is fixed, let us write the dinaturality
law~(\ref{eq:dinaturality-law-for-profunctors}) with an arbitrary
function $f^{:B\rightarrow C}$:
\begin{equation}
(f^{:B\rightarrow C})^{\downarrow G^{\bullet,B}}\bef\text{fold}_{F}^{B}\bef f^{\uparrow H^{B,\bullet}}\overset{!}{=}f^{\uparrow G^{C,\bullet}}\bef\text{fold}_{F}^{C}\bef f^{\downarrow H^{\bullet,C}}\quad.\label{eq:fold-naturality-from-profunctor-derivation2}
\end{equation}
The lifting code required for the profunctors $G^{X,Y}\triangleq Y\times\left(A\times X\rightarrow Y\right)$
and $H^{X,Y}\triangleq F^{A}\rightarrow Y$ is
\begin{align*}
(f^{:B\rightarrow C})^{\downarrow G^{\bullet,B}}=\text{id}^{B}\boxtimes(h^{:A\times C\rightarrow B}\rightarrow a^{:A}\times b^{:B}\rightarrow h(a\times f(b)))\quad, & \quad\quad f^{\uparrow G^{C,\bullet}}=f\boxtimes(h^{:A\times C\rightarrow B}\rightarrow h\bef f)\quad,\\
(f^{:B\rightarrow C})^{\downarrow H^{\bullet,C}}=\text{id}\quad, & \quad\quad f^{\uparrow H^{B,\bullet}}=q^{:F^{A}\rightarrow B}\rightarrow q\bef f\quad.
\end{align*}
Substituting this code into the law~(\ref{eq:fold-naturality-from-profunctor-derivation2})
and applying to an arbitrary $p^{:G^{C,B}}=z^{:B}\times h^{:A\times C\rightarrow B}$,
we get:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (z\times h)\triangleright\gunderline{(f^{:B\rightarrow C})^{\downarrow G^{\bullet,B}}}\bef\text{fold}_{F}\bef\gunderline{f^{\uparrow H^{B,\bullet}}}\\
{\color{greenunder}\text{definitions of liftings}:}\quad & \quad=(z\times h)\triangleright(\text{id}\boxtimes(h\rightarrow a\times b\rightarrow h(a\times f(b))))\gunderline{\bef}\text{fold}_{F}\gunderline{\bef}(q\rightarrow q\bef f)\\
{\color{greenunder}\triangleright\text{-notation}:}\quad & \quad=(z\times h)\triangleright(\text{id}\boxtimes(h\rightarrow a\times b\rightarrow h(a\times f(b))))\triangleright\text{fold}_{F}\triangleright(q\rightarrow q\bef f)\\
{\color{greenunder}\text{apply functions}:}\quad & \quad=\text{fold}_{F}(z\times(a\times b\rightarrow h(a\times f(b))))\bef f\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & (z\times h)\triangleright\gunderline{f^{\uparrow G^{C,\bullet}}}\bef\text{fold}_{F}\bef\gunderline{f^{\downarrow H^{\bullet,C}}}=(z\times h)\triangleright(f\boxtimes(h^{:A\times C\rightarrow B}\rightarrow h\bef f))\bef\text{fold}_{F}\bef\text{id}\\
{\color{greenunder}\text{apply functions}:}\quad & \quad=\text{fold}_{F}(f(z)\times(h\bef f))\quad.
\end{align*}
We obtain the following law of \lstinline!fold!:
\[
\text{fold}_{F}(f(z)\times(h\bef f))=\text{fold}_{F}(z\times(a\times b\rightarrow h(a\times f(b))))\bef f\quad.
\]


\subsubsection{Statement \label{subsec:Statement-naturality-laws-from-wedge-law}\ref{subsec:Statement-naturality-laws-from-wedge-law}}

The naturality laws~(\ref{eq:naturality-law-for-functors})\textendash (\ref{eq:naturality-law-for-contrafunctors})
and the dinaturality law~(\ref{eq:dinaturality-law-for-profunctors})
are special cases of the wedge law~(\ref{eq:wedge-law-for-profunctors}). 

\subparagraph{Proof}

We assume that the wedge law holds for any values $p$ of type $\forall A.\,P^{A,A}$.

To prove Eq.~(\ref{eq:naturality-law-for-functors}), we define $P^{X,Y}\triangleq G^{X}\rightarrow H^{Y}$.
The two liftings of a function $f^{:A\rightarrow B}$ to $P$ are
expressed as:
\[
f^{\uparrow P^{A,\bullet}}=p^{:G^{A}\rightarrow H^{A}}\rightarrow p\bef f^{\uparrow H}\quad,\quad\quad f^{\downarrow P^{B,\bullet}}=p^{:G^{B}\rightarrow H^{B}}\rightarrow f^{\uparrow G}\bef p\quad.
\]
So, the wedge law~(\ref{eq:wedge-law-for-profunctors}) gives:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p^{A}\triangleright f^{\uparrow P^{A,\bullet}}=p^{A}\bef f^{\uparrow H}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & p^{B}\triangleright f^{\downarrow P^{B,\bullet}}=f^{\uparrow G}\bef p^{B}\quad.
\end{align*}
We obtain $p^{A}\bef f^{\uparrow H}=f^{\uparrow G}\bef p^{B}$, which
is the same as Eq.~(\ref{eq:naturality-law-for-functors}).

To prove Eq.~(\ref{eq:naturality-law-for-contrafunctors}), we define
$P^{X,Y}\triangleq G^{Y}\rightarrow H^{X}$. The two liftings of a
function $f^{:A\rightarrow B}$ to $P$ are expressed as:
\[
f^{\uparrow P^{A,\bullet}}=p^{:G^{A}\rightarrow H^{A}}\rightarrow f^{\downarrow G}\bef p\quad,\quad\quad f^{\downarrow P^{B,\bullet}}=p^{:G^{B}\rightarrow H^{B}}\rightarrow p\bef f^{\downarrow H}\quad.
\]
So, the wedge law~(\ref{eq:wedge-law-for-profunctors}) gives:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p^{A}\triangleright f^{\uparrow P^{A,\bullet}}=f^{\downarrow G}\bef p^{A}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & p^{B}\triangleright f^{\downarrow P^{B,\bullet}}=p^{B}\bef f^{\downarrow H}\quad.
\end{align*}
We obtain $f^{\downarrow G}\bef p^{A}=p^{B}\bef f^{\downarrow H}$,
which is the same as Eq.~(\ref{eq:naturality-law-for-contrafunctors}).

To prove Eq.~(\ref{eq:dinaturality-law-for-profunctors}), we define
$P^{X,Y}\triangleq G^{Y,X}\rightarrow H^{X,Y}$. Note that we need
to swap $X$ and $Y$ in $G^{Y,X}$ in order to conform to the required
variance of $P^{X,Y}$ (contravariant in $X$ and covariant in $Y$). 

The two liftings of a function $f^{:A\rightarrow B}$ to $P$ are
expressed as:
\[
f^{\uparrow P^{A,\bullet}}=p^{:G^{A,A}\rightarrow H^{A,A}}\rightarrow f^{\downarrow G^{\bullet,A}}\bef p\bef f^{\uparrow H^{A,\bullet}}\quad,\quad\quad f^{\downarrow P^{B,\bullet}}=p^{:G^{B,B}\rightarrow H^{B,B}}\rightarrow f^{\uparrow G^{B,\bullet}}\bef p\bef f^{\downarrow H^{\bullet,B}}\quad.
\]
So, the wedge law~(\ref{eq:wedge-law-for-profunctors}) gives:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p^{A}\triangleright f^{\uparrow P^{A,\bullet}}=f^{\downarrow G^{\bullet,A}}\bef p^{A}\bef f^{\uparrow H^{A,\bullet}}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & p^{B}\triangleright f^{\downarrow P^{B,\bullet}}=f^{\uparrow G^{B,\bullet}}\bef p^{B}\bef f^{\downarrow H^{\bullet,B}}\quad.
\end{align*}
We obtain $f^{\downarrow G^{\bullet,A}}\bef p^{A}\bef f^{\uparrow H^{A,\bullet}}=f^{\uparrow G^{B,\bullet}}\bef p^{B}\bef f^{\downarrow H^{\bullet,B}}$,
which is the same as Eq.~(\ref{eq:dinaturality-law-for-profunctors}).
$\square$

\subsection{Uniqueness of functor and contrafunctor liftings\label{sec:Uniqueness-of-functor-and-contrafunctor} }

All the laws shown here need to use function liftings such as $f^{\uparrow G}$
or $f^{\uparrow P^{A,\bullet}}$. How are these liftings defined?
Sections~\ref{subsec:f-Functor-constructions} and~\ref{subsec:f-Contrafunctor-constructions}
derived lawful and fully parametric implementations of the \lstinline!fmap!
and \lstinline!cmap! methods for all functors and contrafunctors
as long as they are build up from the six type constructions. The
naturality laws obtained from the parametricity theorem will use precisely
those \textsf{``}standard\textsf{''} implementations of \lstinline!fmap! and \lstinline!cmap!
because the proof of the parametricity theorem significantly depends
on the code of those implementations.

The type structure of a given fully parametric type constructor $P^{A}$
dictates a unique implementation of a lifting $f^{\uparrow P}$ or
$f^{\downarrow P}$. Let us summarize these implementations for the
six type constructions:

\paragraph{Constant type}

If $F^{A}\triangleq Z$ where $Z$ is a fixed type then $f^{\uparrow F}=\text{id}^{:Z\rightarrow Z}$.

\paragraph{Type parameter}

If $F^{A}\triangleq A$ then $f^{\uparrow F}=f$. If $F^{A}\triangleq G^{H^{A}}$
then $f^{\uparrow F}=(f^{\uparrow H})^{\uparrow G}$ if both $G$
and $H$ are functors and $f^{\uparrow F}=(f^{\downarrow H})^{\downarrow G}$
if both $G$ and $H$ are contrafunctors.

\paragraph{Products}

If $F^{A}\triangleq G^{A}\times H^{A}$ then $f^{\uparrow F}=f^{\uparrow G}\boxtimes f^{\uparrow H}$.

\paragraph{Co-products}

If $F^{A}\triangleq G^{A}+H^{A}$ then $f^{\uparrow F}=f^{\uparrow G}\boxplus f^{\uparrow H}$.

\paragraph{Function types}

If $F^{A}\triangleq G^{A}\rightarrow H^{A}$ then $f^{\uparrow F}=p^{:G^{A}\rightarrow H^{A}}\rightarrow f^{\downarrow G}\bef p\bef f^{\uparrow H}$.

\paragraph{Recursive type}

If $F^{A}\triangleq S^{A,F^{A}}$ then $f^{\uparrow F}=f^{\uparrow S^{\bullet,F^{A}}}\bef\big(\overline{f^{\uparrow F}}\big)^{\uparrow S^{B,\bullet}}$.
Here $\overline{f^{\uparrow F}}$ is a recursive call to $f^{\uparrow F}$.

In addition, we will need a seventh type construction: the universally
quantified\index{types!universally quantified} type.

\paragraph{Quantified type}

If $F^{A}\triangleq\forall X.\,P^{X,A}$ where $P^{X,A}$ is covariant
in $X$ then: 
\[
f^{\uparrow F}=\forall Y.\,p^{:\forall X.\,P^{X,A}}\rightarrow p^{Y}\triangleright f^{\uparrow P^{Y,\bullet}}\quad.
\]

It turns out that there no other lawful implementations of \lstinline!fmap!
for these type constructions:

\subsubsection{Statement \label{subsec:Statement-functor-is-unique}\ref{subsec:Statement-functor-is-unique}}

There is only one implementation of a given functor $F$\textsf{'}s \lstinline!fmap!
method that satisfies the identity, composition, and naturality laws.
In particular, fully parametric lawful implementations of \lstinline!fmap!
are unique.

\subparagraph{Proof}

Section~\ref{subsec:f-Functor-constructions} derived lawful implementations
of the \lstinline!fmap! method for all functors $F$ build up from
the six type constructions. Throughout this book, this standard lifting
code is denoted by $\text{fmap}_{F}(f)$ or by $f^{\uparrow F}$.
Now suppose that there exists \emph{another} lawful implementation
of \lstinline!fmap! for $F$, denoted by $\text{fmap}_{F}^{\prime}(f)$:
\[
\text{fmap}_{F}^{\prime}:\left(A\rightarrow B\right)\rightarrow F^{A}\rightarrow F^{B}\quad,\quad\quad\text{fmap}_{F}^{\prime}(f^{:A\rightarrow B})=\text{???}^{:F^{A}\rightarrow F^{B}}\quad.
\]
We will now show that $\text{fmap}_{F}^{\prime}=\text{fmap}_{F}$.
Let us fix the type parameter $A$ and apply the naturality law to
$\text{fmap}_{F}^{\prime}$ with respect to $B$. The resulting law
involves an arbitrary $g^{:B\rightarrow C}$:
\[
\text{fmap}_{F}^{\prime}(f^{:A\rightarrow B}\bef g^{:B\rightarrow C})\overset{!}{=}\text{fmap}_{F}^{\prime}(f)\bef g^{\uparrow F}\quad.
\]
In this naturality law, the lifting $g^{\uparrow F}$ must use the
\textsf{``}standard\textsf{''} lifting code $g^{\uparrow F}\triangleq\text{fmap}_{F}(g)$.
By assumption, $\text{fmap}_{F}^{\prime}$ obeys the composition law,
so we may write:
\[
\text{fmap}_{F}^{\prime}(f\bef g)=\text{fmap}_{F}^{\prime}(f)\bef\text{fmap}_{F}^{\prime}(g)\overset{!}{=}\text{fmap}_{F}^{\prime}(f)\bef g^{\uparrow F}\quad.
\]
Since $f^{:A\rightarrow B}$ is arbitrary, we can choose $A=B$ and
$f=\text{id}^{:B\rightarrow B}$ to obtain:
\[
\text{fmap}_{F}^{\prime}(\text{id})\bef\text{fmap}_{F}^{\prime}(g)\overset{!}{=}\text{fmap}_{F}^{\prime}(\text{id})\bef g^{\uparrow F}\quad.
\]
The identity law for $\text{fmap}_{F}^{\prime}$ gives $\text{fmap}_{F}^{\prime}(\text{id})=\text{id}$,
so we can simplify the last equation to:
\[
\text{fmap}_{F}^{\prime}(g)\overset{!}{=}g^{\uparrow F}=\text{fmap}_{F}(g)\quad.
\]
This must hold for arbitrary $g^{:B\rightarrow C}$, which proves
that $\text{fmap}_{F}^{\prime}=\text{fmap}_{F}$.

\subsubsection{Statement \label{subsec:Statement-contrafunctor-is-unique}\ref{subsec:Statement-contrafunctor-is-unique}}

Any contrafunctor $H$ has a unique implementation of a lawful \lstinline!cmap!
method. 

\subparagraph{Proof}

We use similar arguments as in the proof of Statement~\ref{subsec:Statement-functor-is-unique}.
For any lawful alternative implementation $\text{cmap}_{H}^{\prime}$,
the naturality law is:
\[
\text{cmap}_{H}^{\prime}(f^{:A\rightarrow B}\bef g^{:B\rightarrow C})\overset{!}{=}(g^{:B\rightarrow C})^{\downarrow H}\bef\text{cmap}_{H}^{\prime}(f)\quad.
\]
By assumption, the identity and composition law hold for $\text{cmap}_{H}^{\prime}$.
Setting $f=\text{id}^{:B\rightarrow B}$, we get:
\[
\text{cmap}_{H}^{\prime}(\text{id}\bef g)=\text{cmap}_{H}^{\prime}(g)\overset{!}{=}g^{\downarrow H}\bef\text{cmap}_{H}^{\prime}(\text{id})=g^{\downarrow H}\quad.
\]
This must hold for arbitrary $g^{:B\rightarrow C}$, which shows that
$\text{cmap}_{H}^{\prime}(g)=g^{\downarrow H}=\text{cmap}_{H}(g)$
as required. $\square$

In this section, we proved that lawful implementations of \lstinline!fmap!
and \lstinline!cmap! are unique assuming that they satisfy the naturality
laws that use the standard lifting code (this is a consequence of
full parametricity). However, this does not prove that any implementation
of \lstinline!fmap! and \lstinline!cmap! satisfying the functor
laws must be fully parametric (Problem~\ref{subsec:Problem-unique-functor-liftings}).

\section{Commutativity laws for bifunctors and profunctors\label{sec:Commutativity-laws-for-type-constructors}}

A special property that holds as a consequence of parametricity is
the bifunctor commutativity law~(\ref{eq:f-fmap-fmap-bifunctor-commutativity})
introduced in Section~\ref{subsec:Bifunctors}. If a type constructor
$P^{A,B}$ is a functor with respect to $A$ and $B$ separately then
the liftings with respect to the two type parameters will commute:

\begin{wrapfigure}{l}{0.28\columnwidth}%
\vspace{-1.9\baselineskip}
\[
\xymatrix{\xyScaleY{2.5pc}\xyScaleX{3.0pc}P^{A,B}\ar[r]\sp(0.5){f^{\uparrow P^{\bullet,B}}}\ar[d]\sp(0.45){g^{\uparrow P^{A,\bullet}}} & P^{C,B}\ar[d]\sb(0.45){g^{\uparrow P^{C,\bullet}}}\\
P^{A,D}\ar[r]\sp(0.5){f^{\uparrow P^{\bullet,D}}} & P^{C,D}
}
\]

\vspace{-1\baselineskip}
\end{wrapfigure}%

~\vspace{-0.5\baselineskip}
\[
\text{for all }A,B,C,D,f^{:A\rightarrow C},g^{:B\rightarrow D}\quad:\quad f^{\uparrow P^{\bullet,B}}\bef g^{\uparrow P^{C,\bullet}}=g^{\uparrow P^{A,\bullet}}\bef f^{\uparrow P^{\bullet,D}}\quad.
\]

Similar properties hold for type constructors with any number of type
parameters that are either covariant or contravariant with respect
to each of these type parameters.

We will now prove that the commutativity law holds for any fully parametric
bifunctors and profunctors. The proof proceeds by induction on the
type structure of a given type constructor.\footnote{In the blog post \texttt{\href{https://byorgey.wordpress.com/2018/03/30/}{https://byorgey.wordpress.com/2018/03/30/}},
Brent Yorgey\index{Brent Yorgey} gave a proof of the commutativity
law for bifunctors based on the Reynolds-Wadler relational parametricity
theorem. } It turns out that the proof also requires the commutativity laws
for type constructors with \emph{contravariant} type parameters. All
those proofs are completely analogous, so we will first prove the
commutativity law for profunctors, which are type constructors with
one contravariant and one covariant type parameter. The profunctor
commutativity law will be used below in Statement~\ref{subsec:Statement-wedge-law-from-parametricity}.

\subsection{Proof of the profunctor commutativity law\label{subsec:Proof-of-the-profunctor-commutativity-law}}

Given an arbitrary type constructor $P^{A,B}$ contravariant with
respect to $A$ and covariant with respect to $B$, we formulate the
profunctor \textbf{commutativity law}\index{commutativity law!of profunctors}\index{profunctor!commutativity law}
by analogy with the bifunctor law~(\ref{eq:f-fmap-fmap-bifunctor-commutativity}):
\begin{align}
{\color{greenunder}\text{commutativity law of }P:}\quad & \text{cmap}_{P^{\bullet,B}}(f^{:A\rightarrow C})\bef\text{fmap}_{P^{A,\bullet}}(g^{:B\rightarrow D})=\text{fmap}_{P^{C,\bullet}}(g)\bef\text{cmap}_{P^{\bullet,D}}(f)\quad,\nonumber \\
{\color{greenunder}\text{in a shorter notation}:}\quad & f^{\downarrow P^{\bullet,B}}\bef g^{\uparrow P^{A,\bullet}}=g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}\quad.\label{eq:profunctor-commutativity-law}
\end{align}

\begin{wrapfigure}{l}{0.38\columnwidth}%
\vspace{-1.7\baselineskip}
\[
\xymatrix{\xyScaleY{2.5pc}\xyScaleX{6.0pc}P^{C,B}\ar[r]\sp(0.55){\text{cmap}_{P^{\bullet,B}}(f^{:A\rightarrow C})~~~}\ar[d]\sb(0.45){\text{fmap}_{P^{C,\bullet}}(g)} & P^{A,B}\ar[d]\sb(0.45){\text{fmap}_{P^{A,\bullet}}(g^{:B\rightarrow D})}\\
P^{C,D}\ar[r]\sb(0.45){~~~~\text{cmap}_{P^{\bullet,D}}(f)} & P^{A,D}
}
\]

\vspace{-1.7\baselineskip}
\end{wrapfigure}%

We will now prove that Eq.~(\ref{eq:profunctor-commutativity-law})
holds for any \textbf{fully parametric profunctor}\index{fully parametric!profunctor}
$P^{A,B}$, i.e., a profunctor defined via a combination of the six
type constructions from Table~\ref{tab:six-pure-type-constructions}.
We will assume that all contrafunctor liftings such as $f^{\downarrow P^{\bullet,B}}$
and all functor liftings such as $g^{\uparrow P^{A,\bullet}}$ are
implemented as shown in Chapter~\ref{chap:Functors,-contrafunctors,-and}
for functors and contrafunctors corresponding to each of the type
constructions. (Section~\ref{sec:Uniqueness-of-functor-and-contrafunctor}
will show that the code from Chapter~\ref{chap:Functors,-contrafunctors,-and}
is the only possible lawful and fully parametric implementation of
the liftings.)

The proof goes by induction on the type structure of $P^{A,B}$. The
base case contains the first two constructions (the unit type and
the type parameter), which gives two possibilities: $P^{A,B}\triangleq\bbnum 1$
(constant profunctor) and $P^{A,B}\triangleq B$ (identity profunctor).
The other possibility of using a type parameter, $P^{A,B}\triangleq A$,
does not give a profunctor since $P^{A,B}$ must be contravariant
in $A$.

The commutativity law holds for $P^{A,B}\triangleq\bbnum 1$ because
all liftings are identity functions: $\text{cmap}_{P^{\bullet,B}}(f)=\text{id}$
and $\text{fmap}_{P^{A,\bullet}}(g)=\text{id}$. For the same reason,
the law will also hold for the constant profunctor $P^{A,B}\triangleq Z$
where $Z$ is a fixed type (or a type parameter other than $A$ or
$B$).

For the profunctor $P^{A,B}\triangleq B$, the law holds because $\text{cmap}_{P^{\bullet,B}}(f)=\text{id}$.
So, the base case is proved.

The inductive step looks at the outer construction in the type expression
of $P^{A,B}$, which must be one of product, co-product, function,
or recursion. These constructions create larger type expressions out
of smaller ones; for example, $P^{A,B}\triangleq\left(A\rightarrow B\right)\times B$
uses the product construction with sub-expressions $A\rightarrow B$
and $B$, which can be viewed as profunctors $R^{A,B}\triangleq A\rightarrow B$
and $S^{A,B}\triangleq B$. The inductive assumption is that the commutativity
law is already proved for all sub-expression types (such as $R^{A,B}$
or $S^{A,B}$). We will then need to prove the law for the entire
type expression $P^{A,B}$.

\paragraph{Product type}

If the outer level of $P^{A,B}$ is a product, we must have $P^{A,B}=R^{A,B}\times S^{A,B}$
where $R^{A,B}$ and $S^{A,B}$ are some profunctors. The code for
liftings follows from Statement~\ref{subsec:functor-Statement-functor-product}
and Exercise~\ref{subsec:functor-Exercise-functor-laws}:
\[
f^{\downarrow P^{\bullet,B}}\triangleq f^{\downarrow R^{\bullet,B}}\boxtimes f^{\downarrow S^{\bullet,B}}\quad,\quad\quad g^{\uparrow P^{A,\bullet}}\triangleq g^{\uparrow R^{A,\bullet}}\boxtimes g^{\uparrow S^{A,\bullet}}\quad.
\]
By inductive assumption, $R$ and $S$ already satisfy the commutativity
law~(\ref{eq:profunctor-commutativity-law}). So, we compute:
\begin{align*}
{\color{greenunder}\text{expect to equal }g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}:}\quad & \gunderline{f^{\downarrow P^{\bullet,B}}}\bef\gunderline{g^{\uparrow P^{A,\bullet}}}=\big(f^{\downarrow R^{\bullet,B}}\boxtimes f^{\downarrow S^{\bullet,B}}\big)\bef\big(g^{\uparrow R^{A,\bullet}}\boxtimes g^{\uparrow S^{A,\bullet}}\big)\\
{\color{greenunder}\text{composition law~(\ref{eq:pair-product-composition-law})}:}\quad & =\big(\gunderline{f^{\downarrow R^{\bullet,B}}\bef g^{\uparrow R^{A,\bullet}}}\big)\boxtimes\big(\gunderline{f^{\downarrow S^{\bullet,B}}\bef g^{\uparrow S^{A,\bullet}}}\big)\\
{\color{greenunder}\text{inductive assumption}:}\quad & =\big(g^{\uparrow R^{C,\bullet}}\bef f^{\downarrow R^{\bullet,D}}\big)\boxtimes\big(g^{\uparrow S^{C,\bullet}}\bef f^{\downarrow S^{\bullet,D}}\big)\\
{\color{greenunder}\text{composition law~(\ref{eq:pair-product-composition-law})}:}\quad & =\big(\gunderline{g^{\uparrow R^{C,\bullet}}\boxtimes g^{\uparrow S^{C,\bullet}}}\big)\bef\big(\gunderline{f^{\downarrow R^{\bullet,D}}\boxtimes f^{\downarrow S^{\bullet,D}}}\big)=g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}\quad.
\end{align*}


\paragraph{Co-product type}

In this case, we must have $P^{A,B}=R^{A,B}+S^{A,B}$ where $R^{A,B}$
and $S^{A,B}$ are some profunctors. The code for liftings follows
from Statement~\ref{subsec:functor-Statement-functor-coproduct}
and Exercise~\ref{subsec:functor-Exercise-functor-laws}:
\[
f^{\downarrow P^{\bullet,B}}\triangleq f^{\downarrow R^{\bullet,B}}\boxplus f^{\downarrow S^{\bullet,B}}=\,\begin{array}{||cc|}
f^{\downarrow R^{\bullet,B}} & \bbnum 0\\
\bbnum 0 & f^{\downarrow S^{\bullet,B}}
\end{array}\quad,\quad\quad g^{\uparrow P^{A,\bullet}}\triangleq g^{\uparrow R^{A,\bullet}}\boxplus g^{\uparrow S^{A,\bullet}}=\,\begin{array}{||cc|}
g^{\uparrow R^{A,\bullet}} & \bbnum 0\\
\bbnum 0 & g^{\uparrow S^{A,\bullet}}
\end{array}\quad.
\]
By inductive assumption, $R$ and $S$ already satisfy the commutativity
law~(\ref{eq:profunctor-commutativity-law}). So, we compute:
\begin{align*}
{\color{greenunder}\text{expect to equal }g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}:}\quad & f^{\downarrow P^{\bullet,B}}\bef g^{\uparrow P^{A,\bullet}}=\,\begin{array}{||cc|}
f^{\downarrow R^{\bullet,B}} & \bbnum 0\\
\bbnum 0 & f^{\downarrow S^{\bullet,B}}
\end{array}\,\bef\,\begin{array}{||cc|}
g^{\uparrow R^{A,\bullet}} & \bbnum 0\\
\bbnum 0 & g^{\uparrow S^{A,\bullet}}
\end{array}\\
{\color{greenunder}\text{matrix composition}:}\quad & =\,\begin{array}{||cc|}
\gunderline{f^{\downarrow R^{\bullet,B}}\bef g^{\uparrow R^{A,\bullet}}} & \bbnum 0\\
\bbnum 0 & \gunderline{f^{\downarrow S^{\bullet,B}}\bef g^{\uparrow S^{A,\bullet}}}
\end{array}\\
{\color{greenunder}\text{inductive assumption}:}\quad & =\,\,\begin{array}{||cc|}
g^{\uparrow R^{C,\bullet}}\bef f^{\downarrow R^{\bullet,D}} & \bbnum 0\\
\bbnum 0 & g^{\uparrow S^{C,\bullet}}\bef f^{\downarrow S^{\bullet,D}}
\end{array}\quad.
\end{align*}
The right-hand side of the law, $g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}$,
can be rewritten as
\begin{align*}
 & g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}=\,\begin{array}{||cc|}
g^{\uparrow R^{C,\bullet}} & \bbnum 0\\
\bbnum 0 & g^{\uparrow S^{C,\bullet}}
\end{array}\,\bef\,\begin{array}{||cc|}
f^{\downarrow R^{\bullet,D}} & \bbnum 0\\
\bbnum 0 & f^{\downarrow S^{\bullet,D}}
\end{array}\\
{\color{greenunder}\text{matrix composition}:}\quad & =\,\begin{array}{||cc|}
g^{\uparrow R^{C,\bullet}}\bef f^{\downarrow R^{\bullet,D}} & \bbnum 0\\
\bbnum 0 & g^{\uparrow S^{C,\bullet}}\bef f^{\downarrow S^{\bullet,D}}
\end{array}\quad.
\end{align*}
This now coincides with the left-hand side of the law.

\paragraph{Function type}

The type expression $P^{A,B}\triangleq R^{B,A}\rightarrow S^{A,B}$
(note the swapped type parameters in $R^{B,A}$) is contravariant
in $A$ and covariant in $B$ when $R^{A,B}$ and $S^{A,B}$ are any
profunctors. The lifting code follows from Statement~\ref{subsec:functor-Statement-functor-exponential}
and Exercise~\ref{subsec:functor-Exercise-contrafunctor-exponential}:
\[
(f^{:A\rightarrow C})^{\downarrow P^{\bullet,B}}\triangleq p^{:P^{C,B}}\rightarrow f^{\uparrow R^{B,\bullet}}\bef p\bef f^{\downarrow S^{\bullet,B}}\quad,\quad\quad(g^{:B\rightarrow D})^{\uparrow P^{A,\bullet}}\triangleq p^{:P^{A,B}}\rightarrow g^{\downarrow R^{\bullet,A}}\bef p\bef g^{\uparrow S^{A,\bullet}}\quad.
\]
By inductive assumption, $R$ and $S$ satisfy the commutativity law~(\ref{eq:profunctor-commutativity-law}).
Omitting types, we write:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & f^{\downarrow P}\bef g^{\uparrow P}=\big(p\rightarrow f^{\uparrow R}\bef p\bef f^{\downarrow S}\big)\bef\big(p\rightarrow g^{\downarrow R}\bef p\bef g^{\uparrow S}\big)\\
{\color{greenunder}\text{compute composition}:}\quad & =p\rightarrow\gunderline{g^{\downarrow R}\bef f^{\uparrow R}}\bef p\bef\gunderline{f^{\downarrow S}\bef g^{\uparrow S}}\\
{\color{greenunder}\text{inductive assumption}:}\quad & =p\rightarrow f^{\uparrow R}\bef g^{\downarrow R}\bef p\bef g^{\uparrow S}\bef f^{\downarrow S}\quad.
\end{align*}
The right-hand side of the law is then transformed to the same expression:
\begin{align*}
 & g^{\uparrow P}\bef f^{\downarrow P}=\big(p\rightarrow g^{\downarrow R}\bef p\bef g^{\uparrow S}\big)\bef\big(p\rightarrow f^{\uparrow R}\bef p\bef f^{\downarrow S}\big)\\
{\color{greenunder}\text{compute composition}:}\quad & =p\rightarrow f^{\uparrow R}\bef g^{\downarrow R}\bef p\bef g^{\uparrow S}\bef f^{\downarrow S}\quad.
\end{align*}


\paragraph{Recursive type}

A recursive profunctor $P$ is defined using a type equation of the
form
\[
P^{A,B}\triangleq S^{A,B,P^{A,B}}\quad,\quad\text{equivalently denoted as}\quad P^{A,B}\triangleq\text{Fix}^{S^{A,B,\bullet}}\quad,
\]
where $S^{A,B,R}$ must be contravariant in $A$ but covariant in
$B$ and $R$. The liftings are defined by
\[
(f^{:A\rightarrow C})^{\downarrow P^{\bullet,B}}\triangleq f^{\downarrow S^{\bullet,B,P^{C,B}}}\bef\big(f^{\overline{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{A,B,\bullet}}\quad,\quad\quad(g^{:B\rightarrow D})^{\uparrow P^{A,\bullet}}\triangleq g^{\uparrow S^{A,\bullet,P^{A,B}}}\bef\big(g^{\overline{\uparrow P^{A,\bullet}}}\big)^{\uparrow S^{A,D,\bullet}}\quad.
\]
The inductive assumption is two-fold: first, that the recursively
used lifting to $P^{A,B}$ already satisfies the commutativity law.
Second, that the profunctor $S^{A,B,R}$ satisfies the commutativity
law with respect to \emph{any} two parameters (we will prove that
law for covariant type parameters in Section~\ref{subsec:Proofs-of-commutativity-for-bifunctor}).

Both sides of the commutativity law are functions of type $P^{C,B}\rightarrow P^{A,D}$
or equivalently (if we expand the type recursion) as functions of
type $S^{C,B,P^{C,B}}\rightarrow S^{A,D,P^{A,D}}$. To avoid mistakes,
we write out the type parameters in this computation:
\begin{align*}
{\color{greenunder}\text{expect to equal }g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}:}\quad & f^{\downarrow P^{\bullet,B}}\bef g^{\uparrow P^{A,\bullet}}=f^{\downarrow S^{\bullet,B,P^{C,B}}}\bef\gunderline{\big(f^{\overline{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{A,B,\bullet}}\bef g^{\uparrow S^{A,\bullet,P^{A,B}}}}\bef\big(g^{\overline{\uparrow P^{A,\bullet}}}\big)^{\uparrow S^{A,D,\bullet}}\\
{\color{greenunder}^{B,R}\text{-commutativity of }S^{A,B,R}:}\quad & =\gunderline{f^{\downarrow S^{\bullet,B,P^{C,B}}}\bef g^{\uparrow S^{A,\bullet,P^{C,B}}}}\bef\big(f^{\overline{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{A,D,\bullet}}\bef\big(g^{\overline{\uparrow P^{A,\bullet}}}\big)^{\uparrow S^{A,D,\bullet}}\\
{\color{greenunder}^{A,B}\text{-commutativity of }S^{A,B,R}:}\quad & =g^{\uparrow S^{C,\bullet,P^{C,B}}}\bef f^{\downarrow S^{\bullet,D,P^{C,B}}}\bef\big(\gunderline{f^{\overline{\downarrow P^{\bullet,B}}}\bef g^{\overline{\uparrow P^{A,\bullet}}}}\big)^{\uparrow S^{A,D,\bullet}}\\
{\color{greenunder}\text{inductive assumption}:}\quad & =g^{\uparrow S^{C,\bullet,P^{C,B}}}\bef f^{\downarrow S^{\bullet,D,P^{C,B}}}\bef\big(g^{\overline{\uparrow P^{C,\bullet}}}\bef f^{\overline{\downarrow P^{\bullet,D}}}\gunderline{\big)^{\uparrow S^{A,D,\bullet}}}\\
{\color{greenunder}\text{composition under }^{\uparrow S^{A,D,\bullet}}:}\quad & =g^{\uparrow S^{C,\bullet,P^{C,B}}}\bef\gunderline{f^{\downarrow S^{\bullet,D,P^{C,B}}}\bef\big(g^{\overline{\uparrow P^{C,\bullet}}}\big)^{\uparrow S^{A,D,\bullet}}}\bef\big(f^{\overline{\downarrow P^{\bullet,D}}}\big)^{\uparrow S^{A,D,\bullet}}\\
{\color{greenunder}^{A,R}\text{-commutativity of }S^{A,B,R}:}\quad & =g^{\uparrow S^{C,\bullet,P^{C,B}}}\bef\big(g^{\overline{\uparrow P^{C,\bullet}}}\big)^{\uparrow S^{C,D,\bullet}}\bef f^{\downarrow S^{\bullet,D,P^{C,D}}}\bef\big(f^{\overline{\downarrow P^{\bullet,D}}}\big)^{\uparrow S^{A,D,\bullet}}\\
{\color{greenunder}\text{definitions of liftings}:}\quad & =g^{\uparrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,D}}\quad.
\end{align*}

This concludes the proof of the profunctor commutativity law.

\subsection{Commutativity laws for bifunctors and bi-contrafunctors\label{subsec:Proofs-of-commutativity-for-bifunctor}}

A bi-contrafunctor $P^{A,B}$ is a type constructor contravariant
with respect to both $A$ and $B$. The commutativity law for bi-contrafunctors
is formulated as:
\begin{align}
{\color{greenunder}\text{commutativity law of }P:}\quad & \text{cmap}_{P^{\bullet,D}}(f^{:A\rightarrow C})\bef\text{cmap}_{P^{A,\bullet}}(g^{:B\rightarrow D})=\text{cmap}_{P^{C,\bullet}}(g)\bef\text{cmap}_{P^{\bullet,B}}(f)\quad.\nonumber \\
{\color{greenunder}\text{shorter notation}:}\quad & f^{\downarrow P^{\bullet,D}}\bef g^{\downarrow P^{A,\bullet}}=g^{\downarrow P^{C,\bullet}}\bef f^{\downarrow P^{\bullet,B}}\quad.\label{eq:bi-contrafunctor-commutativity-law}
\end{align}

\begin{wrapfigure}{l}{0.43\columnwidth}%
\vspace{-1.7\baselineskip}
\[
\xymatrix{\xyScaleY{2.5pc}\xyScaleX{6.0pc}P^{C,D}\ar[r]\sp(0.55){\text{cmap}_{P^{\bullet,D}}(f^{:A\rightarrow C})~~~}\ar[d]\sb(0.45){\text{cmap}_{P^{C,\bullet}}(g^{:B\rightarrow D})} & P^{A,D}\ar[d]\sb(0.45){\text{cmap}_{P^{A,\bullet}}(g)}\\
P^{C,B}\ar[r]\sb(0.45){~~~~\text{cmap}_{P^{\bullet,B}}(f^{:A\rightarrow C})} & P^{A,B}
}
\]

\vspace{-1.7\baselineskip}
\end{wrapfigure}%

Any bifunctor or bi-contrafunctor whose type expression is built up
using the six type constructions (Table~\ref{tab:six-pure-type-constructions})
will satisfy its commutativity law. To prove that, we need to repeat
the same calculations as for profunctors in Section~\ref{subsec:Proof-of-the-profunctor-commutativity-law}
except for changing \lstinline!cmap! into \lstinline!fmap! or back
when needed. We only need to check that the proof will still work
after such changes. Looking over the proof in Section~\ref{subsec:Proof-of-the-profunctor-commutativity-law},
we find that we never used the functor or contrafunctor \emph{composition
law} for $P$. The only usage of the composition law was with respect
to the type parameter $R$ in the structure functor $S^{A,B,R}$ of
the recursive type construction. However, in all cases $S^{A,B,R}$
needs to be \emph{covariant} in $R$ because that is required by the
recursive type equation. So, we are assured that the same proof with
minor changes will work for bifunctors and bi-contrafunctors.

The same techniques and proofs apply to type constructors with more
than two type parameters.

\section{Parametricity theorem formulated via relations\label{sec:Parametricity-theorem-for-relations}\label{subsec:Relations-between-types}}

Naturality laws are formulated using arbitrary functions $f^{:A\rightarrow B}$
between arbitrary types $A$ and $B$. Typically, a naturality law
is an equation that involves the function $f$ lifted to some functors.
For instance, $f^{\uparrow G}\bef t=t\bef f^{\uparrow H}$ is the
naturality law of a natural transformation $t:\forall A.\,G^{A}\rightarrow H^{A}$,
where $G$ and $H$ are some functors.

To prove that naturality laws hold for any fully parametric transformation
$t$, we need to use induction in the structure of the code of $t$.
The proof will decompose $t$ into smaller sub-expressions for which
the naturality law should hold by the inductive assumption. Some of
those sub-expressions will have types that are no longer of the form
$\forall A.\,G^{A}\rightarrow H^{A}$. So, we need to generalize naturality
laws to arbitrary type signatures of the form $\forall A.\,P^{A,A}$,
where $P^{X,Y}$ is an arbitrary exponential-polynomial profunctor
(contravariant in $X$ and covariant in $Y$). 

It is not possible to lift an arbitrary function $f^{:A\rightarrow B}$
to a function of type $P^{A,A}\rightarrow P^{B,B}$; for that, we
would need also a function of type $B\rightarrow A$. To solve this
problem, the Reynolds-Wadler approach replaces functions $f^{:A\rightarrow B}$
by arbitrary many-to-many relations (denoted by $r^{:A\leftrightarrow B}$)
between types $A$ and $B$. It turns out that any relation $r$ of
type $A\leftrightarrow B$ can be lifted to a relation (denoted by
$r^{\updownarrow P}$) of type $P^{A,A}\leftrightarrow P^{B,B}$.
The lifting operation can be defined for any exponential-polynomial
profunctor $P^{\bullet,\bullet}$. Using this technique, we will prove
the \textsf{``}relational parametricity theorem\textsf{''}: any fully parametric code
expression (not necessarily a function) of type $\forall A.\,P^{A,A}$
satisfies a specially formulated relational naturality law. That law
will then allow us to derive the wedge law~(\ref{eq:wedge-law-for-profunctors}).

\subsection{Relations between types}

Programmers are familiar with relations as tables in relational databases.
The simplest relation is a table with two columns having values of
some fixed types, say \lstinline!INT! and \lstinline!FLOAT!. As
an example, consider a table called \lstinline!R! containing this
data:
\begin{center}
$R=$ %
\begin{tabular}{|c|c|}
\hline 
\lstinline!SAMPLE_COUNT: INT! & \lstinline!MEAN_VALUE: FLOAT!\tabularnewline
\hline 
\hline 
{\footnotesize{}150} & {\footnotesize{}0.92}\tabularnewline
\hline 
{\footnotesize{}150} & {\footnotesize{}0.95}\tabularnewline
\hline 
{\footnotesize{}180} & {\footnotesize{}0.95}\tabularnewline
\hline 
{\footnotesize{}180} & {\footnotesize{}1.02}\tabularnewline
\hline 
{\footnotesize{}200} & {\footnotesize{}0.95}\tabularnewline
\hline 
\end{tabular}
\par\end{center}

Each row of the table \lstinline!R! has some value of type \lstinline!INT!
and some value of type \lstinline!FLOAT! (we disallow \lstinline!NULL!
values here). The existence of a row $\left(180,0.95\right)$ means
that the values $180$ and $0.95$ are \textsf{``}in the relation\textsf{''} \lstinline!R!.

Two values $a^{:\text{Int}}$ and $b^{:\text{Float}}$ are in the
relation \lstinline!R! only if there exists a row $\left(a,b\right)$
in the table \lstinline!R!. All other pairs of values $\left(a,b\right)$
are \emph{not} in the relation \lstinline!R!. 

Note that the table \lstinline!R! contains several values in the
first column that correspond to the same value in the second column,
and vice versa. So, the relation \lstinline!R! is many-to-many and
cannot be represented by a function of type \lstinline!Int => Float!
or by a function of type \lstinline!Float => Int!. Instead, we must
view the relation \lstinline!R! as a subset of the set of \emph{all}
possible pairs $(a^{:\text{Int}},b^{:\text{Float}})$. The table \lstinline!R!
lists the pairs in that subset.

Instead of listing the pairs, we could describe a relation by implementing
a function telling us whether two given values are in the relation.
The Scala type signature for such a function could be:
\begin{lstlisting}
def inRelationR(a: Int, b: Float): Boolean
\end{lstlisting}
In the short code notation, this function is written as $r:\text{Int}\times\text{Float}\rightarrow\bbnum 2$.
The function call $r\left(a\times b\right)$ returns \lstinline!true!
when the two given values $a$ and $b$ are in the relation $r$.

For proving the relational parametricity theorem, we will need to
use relations between arbitrary types. Generalizing to arbitrary types
$A$ and $B$ instead of \lstinline!Int! and \lstinline!Float!,
we get the following definition:

\subsubsection{Definition \label{subsec:Definition-relation-between-A-B}\ref{subsec:Definition-relation-between-A-B}}

A \textbf{relation} between\index{parametricity theorem!relation between types}
types $A$ and $B$ is a function $r:A\times B\rightarrow\bbnum 2$.
The type of the relation $r$ is denoted by $A\leftrightarrow B$,
indicating that relations are more general than functions of types
$A\rightarrow B$ or $B\rightarrow A$. If values $x$ and $y$ are
in a relation $r$, we write $(x,y)\in r$ or with full type annotations:
$(x^{:A},y^{:B})\in r^{:A\leftrightarrow B}$. The condition $(x,y)\in r$
is equivalent to $r(x,y)=\text{true}$. $\square$

A simple example of a relation is the \textbf{identity relation}\index{identity relation},
denoted by $\text{id}^{:A\leftrightarrow A}$. The identity relation
holds only when two values of type $A$ are equal:
\[
(x^{:A},y^{:A})\in\text{id}^{:A\leftrightarrow A}\quad\text{ means }\quad x=y\quad.
\]

To use the parametricity theorem in practice, we will need to convert
functions into relations. We will denote by $\left<f\right>$ the
relation of type $A\leftrightarrow B$ obtained from a function $f^{:A\rightarrow B}$.
The relation $\left<f\right>$ is defined as:
\[
(a^{:A},b^{:B})\in\left<f\right>\text{ means }b=f(a)\quad\text{or equivalently}:\quad a\triangleright f=b\quad.
\]
Functions $f$ of type $A\rightarrow B$ are converted to many-to-one
relations $\left<f\right>$ between types $A$ and $B$: one or more
values of type $A$ may be related to a single value of type $B$.

A function $g$ of type $B\rightarrow A$ can be also converted to
a relation of type $A\leftrightarrow B$. However, $\left<g\right>$
will have type $B\leftrightarrow A$ in our notation. We note that
the relation types $A\leftrightarrow B$ and $B\leftrightarrow A$
are equivalent (isomorphic as types) because we can simply swap the
order of $x^{:A}$ and $y^{:B}$ in the condition $(x,y)\in r^{:A\leftrightarrow B}$
and obtain the \index{reverse relation}\textbf{reverse relation}
of type $B\leftrightarrow A$, denoted $\text{rev}\left(r\right)$,
such that $(y,x)\in\text{rev}\left(r\right)$. So, we will write $\text{rev}\left<g\right>$
if we need to convert a function $g^{:B\rightarrow A}$ into a relation
of type $A\leftrightarrow B$.

The operation \lstinline!rev! is an isomophism because it is its
own inverse: 
\[
\text{rev}\left(\text{rev}\left(r\right)\right)=r\quad.
\]
To prove this, substitute the definition of \lstinline!rev! twice:
\[
(x^{:A},y^{:B})\in\text{rev}\,\big(\text{rev}\,(r)\big)\quad\text{is the same as}:\quad(y,x)\in\text{rev}\,(r)\quad\text{and is the same as}:\quad(x,y)\in r\quad.
\]

Here is an example of a relation $r^{:A\leftrightarrow B}$ that is
\emph{not} equivalent to a function:
\[
(x^{:A},y^{:B})\in r\text{ means }p(x)=q(y)\text{ where }p^{:A\rightarrow C}\text{ and }q^{:B\rightarrow C}\text{ are some given functions}\quad.
\]
We will call this a \textbf{pullback relation}\index{pullback relation|textit}
and denote it by $r=\text{pull}\,(p,q)$. For certain choices of the
functions $p$ and $q$, there may be many different values $x_{1}$,
$x_{2}$, $y_{1}$, $y_{2}$, etc., such that $p(x_{1})=p(x_{2})=q(y_{1})=q(y_{2})=...$
In that case, the pullback relation will be many-to-many and will
not be equivalent to the relation $\left<f\right>$ or to $\text{rev}\left<f\right>$
for any function $f$. 

In the derivations below, we will often use relations of type $P^{A,A}\leftrightarrow P^{B,B}$,
where $P^{X,Y}$ is a profunctor (contravariant in $X$ and covariant
in $Y$). For any function $f^{:A\rightarrow B}$, there is a special
pullback relation of type $P^{A,A}\leftrightarrow P^{B,B}$: 

\subsubsection{Definition \label{subsec:Definition-wedge-relation}\ref{subsec:Definition-wedge-relation}}

Given a profunctor\index{profunctor!wedge relation}\index{wedge relation|textit}
$P^{X,Y}$ and a function $f^{:A\rightarrow B}$, two values $x^{:P^{A,A}}$
and $y^{:P^{B,B}}$ are \textbf{in a} $\left(P,f\right)$-\textbf{wedge
relation} if $x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}$.
This condition can be written more concisely as $(x,y)\in\text{pull}\,(f^{\uparrow P},f^{\downarrow P})$.

The wedge relation comes up often in applications of the relational
parametricity theorem. The $\left(P,f\right)$-wedge relation generalizes
the wedge law~(\ref{eq:wedge-law-for-profunctors}) to two arbitrary
values $x$ and $y$. The wedge law of an expression $p:\forall A.\,P^{A,A}$
is equivalent to the requirement that $p^{A}$ and $p^{B}$ are in
the $\left(P,f\right)$-wedge relation for all $f^{:A\rightarrow B}$
and for all types $A$, $B$. 

\subsection{Relational product, co-product, and pair mapper. Relational lifting}

Lifting of relations to type constructors is necessary in order to
formulate naturality laws via relations. Ordinary naturality laws
involving functions, such as Eq.~(\ref{eq:naturality-law-for-functors}),
are formulated using an arbitrary function $f^{:A\rightarrow B}$
lifted to some functors or contrafunctors. Lifting a function $f^{:A\rightarrow B}$
to a functor $G$ yields a function $f^{\uparrow G}:G^{A}\rightarrow G^{B}$.
Lifting $f$ to a contrafunctor $H$ yields a function $f^{\downarrow H}:H^{B}\rightarrow H^{A}$.
As a first step towards writing naturality laws in terms of relations,
we will define a lifting of an arbitrary relation $r^{:A\leftrightarrow B}$
to any type constructor $G^{\bullet}$. The lifted relation will have
type $G^{A}\leftrightarrow G^{B}$ and will be denoted by $r^{\updownarrow G}$.

As motivation, let us reformulate some known naturality laws in terms
of relations. Consider a natural transformation $t:\forall A.\,G^{A}\rightarrow H^{A}$
with its naturality law $f^{\uparrow G}\bef t=t\bef f^{\uparrow H}$.
To obtain a relational formulation of that law, we first apply both
sides to an arbitrary value $p^{:G^{A}}$:
\[
p\triangleright f^{\uparrow G}\triangleright t^{B}=p\triangleright t^{A}\triangleright f^{\uparrow H}\quad,\quad\quad\text{or equivalently}:\quad t^{B}(p\triangleright f^{\uparrow G})=t^{A}(p)\triangleright f^{\uparrow H}\quad,
\]
and define for brevity the value $q^{:G^{B}}\triangleq p\triangleright f^{\uparrow G}$.
This definition of $q$ is equivalent to the condition $(p,q)\in\langle f^{\uparrow G}\rangle$.
Now we rewrite the naturality law of $t$ as a statement about relations:
\begin{equation}
\text{for any }f^{:A\rightarrow B},p^{:G^{A}},q^{:G^{B}}\quad:\quad\quad\text{when}\quad(p,q)\in\langle f^{\uparrow G}\rangle\quad\quad\text{then}\quad(t^{A}(p),t^{B}(q))\in\langle f^{\uparrow H}\rangle\quad.\label{eq:naturality-law-of-t-derivation1}
\end{equation}

We will define a concise notation for conditions of this form because
they are found often when working with relations.

\subsubsection{Definition \label{subsec:Definition-pair-mapper-of-relations}\ref{subsec:Definition-pair-mapper-of-relations}}

\index{pair mapper of relations}The \textbf{pair mapper of relations}
$r^{:A\leftrightarrow C}$ and $s^{:B\leftrightarrow D}$ is a relation
$r\varogreaterthan s$ defined by:
\begin{align*}
 & \big(r^{:A\leftrightarrow C}\varogreaterthan s^{:B\leftrightarrow D}\big):\left(A\rightarrow B\right)\leftrightarrow\left(C\rightarrow D\right)\quad,\\
 & (f^{:A\rightarrow B},g^{:C\rightarrow D})\in r\varogreaterthan s\quad\text{means}\quad\quad\forall x^{:A},y^{:C}\quad:\quad\text{if }(x,y)\in r\quad\text{then}\quad(f(x),g(y))\in s\quad.
\end{align*}

\begin{wrapfigure}{l}{0.24\columnwidth}%
\vspace{-2.8\baselineskip}
\[
\xymatrix{\xyScaleY{1.8pc}\xyScaleX{2.6pc}A\ar@{<->}[d]\sp(0.5){r}\ar[rr]\sp(0.5){f} & \ar@{<->}[d(0.9)]\sp(0.55){r\varogreaterthan s} & B\ar@{<->}[d]\sp(0.5){s}\\
C\ar[rr]\sb(0.5){g} &  & D
}
\]

\vspace{-2.5\baselineskip}
\end{wrapfigure}%

\noindent The type diagram at left illustrates the pair mapper construction.
To read the diagram, we need to start with two arbitrary values, $x^{:A}$
at the top left and $y^{:C}$ at the bottom left. Given two relations
$r^{:A\leftrightarrow C}$ and $s^{:B\leftrightarrow D}$, we define
a new relation $r\ogreaterthan s$ between function types $A\rightarrow B$
and $C\rightarrow D$. Two functions $\left(f,g\right)$ are in the
relation $r\varogreaterthan s$ if the diagram commutes, which means
that $(f(x),g(y))\in s$ for any $x$ and $y$ in the relation $r$.
$\square$

Using Definition~\ref{subsec:Definition-pair-mapper-of-relations},
we can rewrite the naturality law~(\ref{eq:naturality-law-of-t-derivation1})
as:
\[
\text{for all }f^{:A\rightarrow B}\quad:\quad(t^{A},t^{B})\in\langle f^{\uparrow G}\rangle\ogreaterthan\langle f^{\uparrow H}\rangle\quad.
\]

It appears reasonable to expect that the relation $\left<f\right>$
lifted to $G$ is the same relation as $\langle f^{\uparrow G}\rangle$:
\[
\left<f\right>^{\updownarrow G}=\langle f^{\uparrow G}\rangle\quad.
\]
With this assumption (to be proved below), the relational formulation
of the naturality law becomes:
\[
\text{for all }f^{:A\rightarrow B}:\quad(t^{A},t^{B})\in\left<f\right>^{\updownarrow G}\ogreaterthan\left<f\right>^{\updownarrow H}\quad.
\]
We now replace $\left<f\right>$ by an arbitrary relation $r^{:A\leftrightarrow B}$
and obtain the\index{naturality law!in terms of relations} \textbf{relational
naturality law}:
\begin{equation}
\text{for any relation }r^{:A\leftrightarrow B}:\quad(t^{A},t^{B})\in r^{\updownarrow G}\ogreaterthan r^{\updownarrow H}\quad.\label{eq:naturality-law-of-t-derivation2}
\end{equation}

Note the similarity of the relation $r^{\updownarrow G}\ogreaterthan r^{\updownarrow H}$
and the type $G^{A}\rightarrow H^{A}$ of $t$. Denote the type signature
of $t$ by $P^{A}\triangleq G^{A}\rightarrow H^{A}$. If we \emph{define}
the lifting $r^{\updownarrow P}$ as $r^{\updownarrow G}\ogreaterthan r^{\updownarrow H}$,
the naturality law~(\ref{eq:naturality-law-of-t-derivation2}) will
be written as:
\begin{equation}
\text{for any relation }r^{:A\leftrightarrow B}:\quad(t^{A},t^{B})\in r^{\updownarrow P}\quad.\label{eq:naturality-law-of-t-derivation3}
\end{equation}
The generalization of the naturality law from functions to relations
now has a concise form: The components $t^{A}$ and $t^{B}$ of a
natural transformation $t$ belong to \emph{any} relation $r$ lifted
to $t$\textsf{'}s type signature (the type constructor $P^{\bullet}$). This
prepares us to the formulation of the parametricity theorem below
and at the same time motivates using the pair mapper operation ($\ogreaterthan$)
to define the lifting a relation to a function type constructor such
as $G^{A}\rightarrow H^{A}$.

What about lifting a relation to product ($G^{A}\times H^{A}$) or
co-product ($G^{A}+H^{A}$) type constructors? Recall that lifting
a function to $G^{A}\times H^{A}$ involves the pair product of functions\index{pair product of functions}
($\boxtimes$):
\[
\big(f^{:A\rightarrow C}\boxtimes g^{:B\rightarrow D}\big):A\times B\rightarrow C\times D\quad,\quad\quad f\boxtimes g\triangleq a\times b\rightarrow f(a)\times g(b)\quad.
\]
Namely, the lifting of a function $k^{:A\rightarrow B}$ to $G^{A}\times H^{A}$
is defined by $k^{\uparrow(G\times H)}\triangleq k^{\uparrow G}\boxtimes k^{\uparrow H}$. 

The pair co-product of functions\index{pair co-product of functions}
is defined by:
\begin{align*}
 & \big(f^{:A\rightarrow C}\boxplus g^{:B\rightarrow D}\big):A+B\rightarrow C+D\quad,\\
 & (f\boxplus g)(a^{:A}+\bbnum 0^{:B})\triangleq f(a)+\bbnum 0^{:D}\quad,\quad\quad(f\boxplus g)(\bbnum 0+b^{:B})\triangleq\bbnum 0^{:C}+g(b)\quad,\\
 & (k^{:A\rightarrow B})^{\uparrow(G+H)}\triangleq k^{\uparrow G}\boxplus k^{\uparrow H}\quad.
\end{align*}

Let us now define the analogous constructions for relations.

\subsubsection{Definition \label{subsec:Definition-pair-product-of-relations}\ref{subsec:Definition-pair-product-of-relations}}

The \textbf{pair product of relations}\index{pair product of relations}
defines a relation between the product types $A\times B$ and $C\times D$
from two relations $r^{:A\leftrightarrow C}$ and $s^{:B\leftrightarrow D}$.
The new relation $r\boxtimes s$ is defined by analogy with the pair
product of functions:
\[
\big(r^{:A\leftrightarrow C}\boxtimes s^{:B\leftrightarrow D}\big):A\times B\leftrightarrow C\times D\quad,\quad\quad(a\times b,c\times d)\in r\boxtimes s\quad\text{when}\quad(a,c)\in r\text{ and }(b,d)\in s\quad.
\]

The second construction is the pair co-product of relations\index{pair co-product of relations},
which creates a relation between co-products of types. The definition
resembles that of the \index{pair co-product of functions}pair co-product
of functions:

\subsubsection{Definition \label{subsec:Definition-pair-co-product-of-relations}\ref{subsec:Definition-pair-co-product-of-relations}}

The \textbf{pair co-product of relations} $r^{:A\leftrightarrow C}$
and $s^{:B\leftrightarrow D}$ is a relation $r\boxplus s$ defined
by:
\begin{align*}
 & \big(r^{:A\leftrightarrow C}\boxplus s^{:B\leftrightarrow D}\big):A+B\leftrightarrow C+D\quad,\\
 & \text{either}\quad\quad(a^{:A}+\bbnum 0^{:B},c^{:C}+\bbnum 0^{:D})\in r\boxplus s\quad\text{when}\quad(a,c)\in r\quad,\\
 & \text{or}\quad\quad(\bbnum 0^{:A}+b^{:B},\bbnum 0^{:C}+d^{:D})\in r\boxplus s\quad\text{when}\quad(b,d)\in s\quad.
\end{align*}
Other combinations of values (such as $a+\bbnum 0$ and $\bbnum 0+d$)
are not in the relation $r\boxplus s$. $\square$

The following example illustrates some properties of the relational
operations $\boxtimes$, $\boxplus$, and $\ogreaterthan$ when applied
to relations obtained from functions.

\subsubsection{Example \label{subsec:Example-product-relation}\ref{subsec:Example-product-relation}\index{solved examples}}

For arbitrary given functions $f^{:A\rightarrow B}$ and $g^{:C\rightarrow D}$,
prove that:

\textbf{(a)} $\left<f\right>\boxtimes\left<g\right>=\left<f\boxtimes g\right>$,
both relations having type $A\times C\leftrightarrow B\times D$.

\textbf{(b)} $\left<f\right>\boxplus\left<g\right>=\left<f\boxplus g\right>$,
both relations having type $A+C\leftrightarrow B+D$.

\textbf{(c)} $\left<f\right>\ogreaterthan\text{id}^{:C\leftrightarrow C}=\text{rev}\left<l^{:B\rightarrow C}\rightarrow f\bef l\right>$,
both relations having type $\left(A\rightarrow C\right)\leftrightarrow\left(B\rightarrow C\right)$.

\textbf{(d)} $\text{id}^{:A\leftrightarrow A}\ogreaterthan\left<g\right>=\left<k^{:A\rightarrow C}\rightarrow k\bef g\right>$,
both relations having type $\left(A\rightarrow C\right)\leftrightarrow\left(A\rightarrow D\right)$.

\textbf{(e)} $\left<f\right>\ogreaterthan\left<g\right>=\text{pull}\,(k^{:A\rightarrow C}\rightarrow k\bef g,\;l^{:B\rightarrow C}\rightarrow f\bef l)$,
both relations having type $\left(A\rightarrow C\right)\leftrightarrow\left(B\rightarrow D\right)$.

\subparagraph{Solution}

We will apply various relations to arbitrary values $a^{:A}$, $b^{:B}$,
$c^{:C}$, $d^{:D}$.%
\begin{comment}
, $a_{2}^{:A}$, $b_{2}^{:B}$, $c_{2}^{:C}$, $d_{2}^{:D}$.
\end{comment}

\textbf{(a)} The following conditions are equivalent:
\begin{align*}
 & (a\times c,b\times d)\in\left<f\right>\boxtimes\left<g\right>\quad\text{means}\quad(a,b)\in\left<f\right>\quad\text{and}\quad(c,d)\in\left<g\right>\quad\text{or}:\quad f(a)=b\text{ and }g(c)=b\quad.\\
 & (a\times c,b\times d)\in\left<f\boxtimes g\right>\quad\text{means}\quad(a\times c)\triangleright(f\boxtimes g)=b\times d\quad\text{or}:\quad f(a)\times g(c)=b\times d\quad,
\end{align*}

\textbf{(b)} The following conditions are equivalent:
\begin{align*}
 & (a+\bbnum 0,b+\bbnum 0)\in\left<f\right>\boxplus\left<g\right>\quad\text{means}\quad(a,b)\in\left<f\right>\quad\text{or equivalently}:\quad a\triangleright f=b\quad,\\
 & (a+\bbnum 0,b+\bbnum 0)\in\left<f\boxplus g\right>\quad\text{means}\quad(a+\bbnum 0)\triangleright(f\boxplus g)=b+\bbnum 0\quad\text{or equivalently}:\quad a\triangleright f=b\quad.
\end{align*}
Similarly we can show that the conditions $(\bbnum 0+c,\bbnum 0+d)\in\left<f\right>\boxplus\left<g\right>$
and $(\bbnum 0+c,\bbnum 0+d)\in\left<f\boxplus g\right>$ are equivalent.
Values from different parts of the disjunctions, such as $(a+\bbnum 0,\bbnum 0+d)$
are not in the relation $\left<f\right>\boxplus\left<g\right>$ by
definition of $\boxplus$. They are also not in the relation $\left<f\boxplus g\right>$
since the function $f\boxplus g$ preserves the left and right parts
of the disjunctions.

\textbf{(c)} Two functions $k^{:A\rightarrow C}$ and $l^{:B\rightarrow C}$
are in the relation $\left<f\right>\ogreaterthan\text{id}^{:C\leftrightarrow C}$
if:
\[
\text{for all }x^{:A},y^{:B}\quad:\quad\text{when}\quad(x,y)\in\left<f\right>\quad\text{then}\quad(k(x),l(y))\in\text{id}^{:C\leftrightarrow C}\quad.
\]
Simplifying the last condition, we get:
\[
\text{for all }x^{:A}\quad:\quad k(x)=l(f(x))\quad,\quad\text{or equivalently}:\quad k=f\bef l\quad.
\]
So, the relation between $k$ and $l$ may be expressed as $k=\psi(l)$
where:
\[
\psi:\left(B\rightarrow C\right)\rightarrow A\rightarrow C\quad,\quad\quad\psi\triangleq l^{:B\rightarrow C}\rightarrow f\bef l\quad.
\]
This relation is denoted by $\text{rev}\left<\psi\right>$. It follows
that $\left<f\right>\ogreaterthan\text{id}^{:C\leftrightarrow C}=\text{rev}\left<\psi\right>=\text{rev}\left<l^{:B\rightarrow C}\rightarrow f\bef l\right>$.

\textbf{(d)} Two functions $k^{:A\rightarrow C}$ and $l^{:A\rightarrow D}$
are in the relation $\text{id}^{:A\leftrightarrow A}\ogreaterthan\left<g\right>$
if:
\[
\text{for all }x^{:A},y^{:A}\quad:\quad\text{when}\quad(x,y)\in\text{id}^{:A\leftrightarrow A}\quad\text{then}\quad(k(x),l(y))\in\left<g\right>\quad.
\]
Simplifying the last condition, we get:
\[
\text{for all }x^{:A}\quad:\quad g(k(x))=l(x)\quad,\quad\text{or equivalently}:\quad l=k\bef g\quad.
\]
So, the relation between $k$ and $l$ may be expressed as $\phi(k)=l$
where:
\[
\phi:\left(A\rightarrow C\right)\rightarrow A\rightarrow D\quad,\quad\phi\triangleq k^{:A\rightarrow C}\rightarrow k\bef g\quad.
\]
This relation is denoted by $\left<\phi\right>$. It follows that
we may write $\text{id}^{:A\leftrightarrow A}\ogreaterthan\left<g\right>=\left<\phi\right>=\left<k^{:A\rightarrow C}\rightarrow k\bef g\right>$.

\textbf{(e)} Two functions $k^{:A\rightarrow C}$ and $l^{:B\rightarrow D}$
are in the relation $\left<f\right>\ogreaterthan\left<g\right>$ if:
\[
\text{for all }x^{:A},y^{:B}\quad:\quad\text{when}\quad(x,y)\in\left<f\right>\quad\text{then}\quad(k(x),l(y))\in\left<g\right>\quad.
\]
Simplifying the last condition, we get:
\[
\text{for all }x^{:A}\quad:\quad g(k(x))=l(f(x))\quad,\quad\text{or equivalently}:\quad f\bef l=k\bef g\quad.
\]
So, the relation between $k$ and $l$ may be expressed as $\phi(f)=\psi(g)$
where: 
\[
\phi:\left(A\rightarrow C\right)\rightarrow A\rightarrow D\quad,\quad\phi\triangleq k^{:A\rightarrow C}\rightarrow k\bef g\quad;\quad\quad\psi:\left(B\rightarrow D\right)\rightarrow A\rightarrow D\quad,\quad\psi\triangleq l^{:B\rightarrow C}\rightarrow f\bef l\quad.
\]
 This is a pullback relation that we denote by $\text{pull}\,(\phi,\psi)=\text{pull}\,\big(k^{:A\rightarrow C}\rightarrow k\bef g,\;l^{:B\rightarrow C}\rightarrow f\bef l\big)$.
$\square$

We now turn to the question of how to define the lifting $r^{\updownarrow G}$
to an arbitrary type constructor $G^{\bullet}$. It turns out that
we will actually need to define a \emph{simultaneous} lifting of several
relations to a type constructor with several type parameters. For
clarity, we postpone that definition and begin by lifting a single
relation.

\subsubsection{Definition \label{subsec:Definition-relational-lifting}\ref{subsec:Definition-relational-lifting}
(relational lifting)}

Given a relation $r^{:A\leftrightarrow B}$ and a type constructor
$G^{\bullet}$, the relational lifting of $r$ to $G$, denoted by
$r^{\updownarrow G}$, is a new relation of type $G^{A}\leftrightarrow G^{B}$.
The relation $r^{\updownarrow G}$ is defined by induction on the
structure of $G$.

Below, the type constructors $K^{A}$, $L^{A}$, $H^{X,A}$, and $S^{A,R}$
are not assumed to be covariant or contravariant. They can be any
exponential-polynomial, fully parametric type constructors. 

\textbf{(a)} If $G^{A}\triangleq Z$ with a fixed type $Z$, we define
$r^{\updownarrow G}\triangleq\text{id}^{:Z\leftrightarrow Z}$.

\textbf{(b)} If $G=\text{Id}$ (that is, $G^{A}\triangleq A$), we
define $r^{\updownarrow\text{Id}}\triangleq r$. 

\textbf{(c)} If $G^{A}\triangleq K^{A}\times L^{A}$ then we define:
\[
r^{\updownarrow G}\triangleq r^{\updownarrow K}\boxtimes r^{\updownarrow L}\quad.
\]
The inductive assumption is that the relational liftings $r^{\updownarrow K}$
and $r^{\updownarrow L}$ are already defined.

\textbf{(d)} If $G^{A}\triangleq K^{A}+L^{A}$ then we define: 
\[
r^{\updownarrow G}\triangleq r^{\updownarrow K}\boxplus r^{\updownarrow L}\quad.
\]
The inductive assumption is that the liftings $r^{\updownarrow K}$
and $r^{\updownarrow L}$ are already defined.

\textbf{(e)} If $G^{A}\triangleq K^{A}\rightarrow L^{A}$ then we
define:
\[
r^{\updownarrow G}\triangleq r^{\updownarrow K}\ogreaterthan r^{\updownarrow L}\quad.
\]
The relation $r^{\updownarrow G}$ is between functions of types $K^{A}\rightarrow L^{A}$
and $K^{B}\rightarrow L^{B}$. The inductive assumption is that the
liftings $r^{\updownarrow K}$ and $r^{\updownarrow L}$ are already
defined.

\textbf{(f)} If $G^{A}\triangleq S^{A,G^{A}}$ is defined recursively
via a recursion scheme $S^{\bullet,\bullet}$, we define $r^{\updownarrow G}$
by:
\[
r^{\updownarrow G}\triangleq\big(r,\overline{r^{\uparrow G}}\big)^{\updownarrow S}\quad.
\]
Here the notation $(r,s)^{\updownarrow S}$ means the \emph{simultaneous}
lifting of the two relations $r$, $s$ to the type constructor $S^{\bullet,\bullet}$
(see Definition~\ref{subsec:Definition-simultaneous-relational-lifting}
below for details). Also note that we use $\overline{r^{\updownarrow G}}$
recursively within the definition of $r^{\updownarrow G}$. This is
allowed since we understand $r^{\updownarrow G}$ to be a function
(of type $G^{A}\times G^{B}\rightarrow\bbnum 2$), and it is permitted
to define functions recursively. The inductive assumption is that
simultaneous liftings to $S^{\bullet,\bullet}$ are already defined.

\textbf{(g)} If $G^{A}\triangleq\forall X.\,H^{X,A}$, we define $r^{\updownarrow G}$
of type $(\forall X.\,H^{X,A})\leftrightarrow(\forall Y.\,H^{Y,B})$
by:
\[
(p^{:\forall X.\,H^{X,A}},q^{:\forall X.\,H^{X,B}})\in r^{\updownarrow\forall X.\,H^{X,\bullet}}\quad\text{means}\quad\forall(X,Y).\,\forall s^{X\leftrightarrow Y}.\,(p^{X},q^{Y})\in(s,r)^{\updownarrow H}\quad.
\]
Here $(s,r)^{\updownarrow H}$ denotes the \emph{simultaneous} lifting
of $s$ and $r$ to $H$ (see Definition~\ref{subsec:Definition-simultaneous-relational-lifting}
below). A shorter way of writing the formula above is by formulating
a relation between $p^{X}$ and $q^{Y}$ directly:
\begin{equation}
\forall(X,Y):\quad\big(r^{\updownarrow\forall Z.\,H^{Z,\bullet}}\big)^{X,Y}\triangleq\forall s^{:X\leftrightarrow Y}.\,(s,r)^{\updownarrow H}\quad.\label{eq:relational-lifting-quantified-types-short}
\end{equation}
The inductive assumption is that simultaneous liftings to $H$ are
already defined. $\square$

To show that Definition~\ref{subsec:Definition-relational-lifting}
gives a useful operation, let us look at some examples.

When the relation $r$ is obtained from a function as $r=\left<f\right>$,
one can sometimes (but not always) express the lifted relation $\left<f\right>^{\updownarrow P}$
through ordinary liftings $f^{\uparrow P}$ and $f^{\downarrow P}$:

\subsubsection{Example \label{subsec:Example-relational-lifting}\ref{subsec:Example-relational-lifting}\index{solved examples}}

A function $f^{:A\rightarrow B}$ gives rise to a relation $\left<f\right>$
of type $A\leftrightarrow B$. Use Definition~\ref{subsec:Definition-relational-lifting}
to compute the lifting $\left<f\right>^{\updownarrow P}$ for the
following type constructors $P^{\bullet}$:

\textbf{(a)} $P^{A}\triangleq A+A\times A\quad$.

\textbf{(b)} $P^{A}\triangleq(R\rightarrow A)\times(A\rightarrow S)$
where $R$ and $S$ are fixed types.

\textbf{(c)} $P^{A}\triangleq A\rightarrow A\quad$.

\textbf{(d)} $P^{A}\triangleq A\rightarrow\left(A\rightarrow R\right)\rightarrow R$
where $R$ is a fixed type.

\textbf{(e)} $P^{A}\triangleq\left(A\rightarrow A\right)\rightarrow A\quad$.

In each case \textbf{(a)}\textendash \textbf{(e)}, express $P^{A}\triangleq N^{A,A}$
where $N^{X,Y}$ is a suitable profunctor and compare the condition
$(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}$ with the $\left(N,f\right)$-wedge
relation ($p_{1}\triangleright f^{\uparrow N}=p_{2}\triangleright f^{\downarrow N}$). 

\subparagraph{Solution}

In all cases, the lifted relation $\left<f\right>^{\updownarrow P}$
has type $P^{A}\leftrightarrow P^{B}$.

\textbf{(a)} At the top level, $P^{A}$ is a disjunction: $P^{A}=\text{Id}^{A}+H^{A}$
where $H^{A}\triangleq A\times A$. So, we use Definition~\ref{subsec:Definition-relational-lifting}(d)
to get the condition for some values $(x^{:P^{A}},y^{:P^{B}})$ to
be in the relation $\left<f\right>^{\updownarrow P}$:
\[
(x,y)\in\left<f\right>^{\updownarrow P}\text{ if }x=x_{1}+\bbnum 0,\,y=y_{1}+\bbnum 0,\,(x_{1},y_{1})\in\left<f\right>^{\updownarrow\text{Id}}\text{ or }x=\bbnum 0+x_{2},\,y=\bbnum 0+y_{2},\,(x_{2},y_{2})\in\left<f\right>^{\updownarrow H}\quad.
\]
By Definition~\ref{subsec:Definition-relational-lifting}(b), we
have $\left<f\right>^{\updownarrow\text{Id}}=\left<f\right>$. The
condition $(x_{1},y_{1})\in\left<f\right>^{\updownarrow\text{Id}}$
is then simplified to $f(x_{1})=y_{1}$. Finally, we use Definition~\ref{subsec:Definition-relational-lifting}(c)
with $H^{A}=\text{Id}^{A}\times\text{Id}^{A}$ to obtain the lifting
$\left<f\right>^{\updownarrow H}$:
\[
(a_{1}^{:A}\times a_{2}^{:A},b_{1}^{:B}\times b_{2}^{:B})\in\left<f\right>^{\updownarrow H}\text{ if }(a_{1},b_{1})\in\left<f\right>^{\updownarrow\text{Id}}\text{ and }(a_{2},b_{2})\in\left<f\right>^{\updownarrow\text{Id}}\quad.
\]
The last condition is simplified to:
\[
(a_{1}^{:A}\times a_{2}^{:A},b_{1}^{:B}\times b_{2}^{:B})\in\left<f\right>^{\updownarrow H}\text{ if }f(a_{1})=b_{1}\text{ and }f(a_{2})=b_{2}\quad.
\]

Putting the pieces together, we obtain the following definition of
the relation $\left<f\right>^{\updownarrow P}$:
\[
(x^{:P^{A}},y^{:P^{B}})\in\left<f\right>^{\updownarrow P}\text{ if }x=x_{1}+\bbnum 0\text{ and }y=f(x_{1})+\bbnum 0,\text{ or }x=\bbnum 0+a_{1}\times a_{2}\text{ and }y=\bbnum 0+f(a_{1})\times f(a_{2})\quad.
\]
We note that this condition is equivalent to applying $f^{\uparrow P}$:
\[
(x^{:P^{A}},y^{:P^{B}})\in\left<f\right>^{\updownarrow P}\text{ if }x\triangleright f^{\uparrow P}=y\quad,\quad\text{ or more concisely}:\quad\left<f\right>^{\updownarrow P}=\langle f^{\uparrow P}\rangle\quad.
\]
We will see below (Statement~\ref{subsec:Statement-lifting-function-relation-covariant})
that this property holds for all functors $P$.

With $N^{X,Y}\triangleq P^{Y}$, the $\left(N,f\right)$-wedge relation
is equivalent to the equation $x\triangleright f^{\uparrow P}=y$.

\textbf{(b)} At the top level, $P^{A}$ is a product: $P^{A}=G^{A}\times H^{A}$
where $G^{A}\triangleq R\rightarrow A$ and $H^{A}\triangleq A\rightarrow S$.
By Definition~\ref{subsec:Definition-relational-lifting}(c), we
get:
\[
(g_{1}^{:G^{A}}\times h_{1}^{:H^{A}},g_{2}^{:G^{B}}\times h_{2}^{:H^{B}})\in\left<f\right>^{\updownarrow G}\boxtimes\left<f\right>^{\updownarrow H}\text{ if }(g_{1},g_{2})\in\left<f\right>^{\updownarrow G}\text{ and }(h_{1},h_{2})\in\left<f\right>^{\updownarrow H}\quad.
\]

The condition $(g_{1},g_{2})\in\left<f\right>^{\updownarrow G}$ is
rewritten using Definition~\ref{subsec:Definition-relational-lifting}(a),
(b), and (e):
\[
(g_{1}^{:R\rightarrow A},g_{2}^{:R\rightarrow B})\in\left<f\right>^{\updownarrow G}\text{ means if }(r_{1}^{:R},r_{2}^{:R})\in\text{id}^{R}\text{ then }(g_{1}(r_{1}),g_{2}(r_{2}))\in\left<f\right>^{\updownarrow\text{Id}}\quad.
\]
We have $\left<f\right>^{\updownarrow\text{Id}}=\left<f\right>$.
The identity relation ($\text{id}^{R}$) between $r_{1}$ and $r_{2}$
holds only if $r_{1}=r_{2}$. We find:
\[
(g_{1}^{:R\rightarrow A},g_{2}^{:R\rightarrow B})\in\left<f\right>^{\updownarrow G}\text{ means }\forall r^{:R}.\,f(g_{1}(r))=g_{2}(r)\quad\text{or equivalently}:\quad g_{1}\bef f=g_{2}\quad.
\]

The condition $(h_{1},h_{2})\in\left<f\right>^{\updownarrow H}$ is
rewritten using the same definitions:
\[
(h_{1}^{:A\rightarrow S},h_{2}^{:B\rightarrow S})\in\left<f\right>^{\updownarrow H}\text{ means if }(a^{:A},b^{:B})\in\left<f\right>^{\updownarrow\text{Id}}\text{ then }(h_{1}(a),h_{2}(b))\in\text{id}\quad.
\]
We simplify this to:
\[
(h_{1}^{:A\rightarrow S},h_{2}^{:B\rightarrow S})\in\left<f\right>^{\updownarrow H}\text{ means }h_{1}(a)=h_{2}(f(a))\quad\text{or equivalently}:\quad h_{1}=f\bef h_{2}\quad.
\]

Finally, the lifted relation $\left<f\right>^{\updownarrow P}$ is
expressed as:
\[
(g_{1}\times h_{1},g_{2}\times h_{2})\in\left<f\right>^{\updownarrow P}\text{ if }g_{1}\bef f=g_{2}\text{ and }h_{1}=f\bef h_{2}\quad.
\]
Defining the profunctor $P^{X,Y}\triangleq G^{Y}\times H^{X}$, we
find that $g_{1}\times h_{1}$ has type $P^{A,A}$ and $g_{2}\times h_{2}$
has type $P^{B,B}$. Then the condition $(g_{1}\times h_{1},g_{2}\times h_{2})\in\left<f\right>^{\updownarrow P}$
is equivalent to the $\left(N,f\right)$-wedge relation with $N^{X,Y}\triangleq\left(R\rightarrow Y\right)\times\left(X\rightarrow S\right)$:
\[
(g_{1}\times h_{1})\triangleright f^{\uparrow N}=(g_{2}\times h_{2})\triangleright f^{\downarrow N}\quad.
\]
In this case, we can express the lifted relation by a single equation.

\textbf{(c)} Using Definition~\ref{subsec:Definition-relational-lifting}(b)
and (e), we get:
\[
(p_{1}^{:A\rightarrow A},p_{2}^{:B\rightarrow B})\in\left<f\right>^{\updownarrow P}\text{ means if }(a^{:A},b^{:B})\in\left<f\right>\text{ then }(p_{1}(a),p_{2}(b))\in\left<f\right>\quad.
\]
The condition can be rewritten as a single equation:
\[
(p_{1}^{:A\rightarrow A},p_{2}^{:B\rightarrow B})\in\left<f\right>^{\updownarrow P}\text{ means }f(p_{1}(a))=p_{2}(f(a))\quad\text{or equivalently}:\quad p_{1}\bef f=f\bef p_{2}\quad.
\]
This is the same as the $\left(N,f\right)$-wedge relation $p_{1}\triangleright f^{\uparrow N}=p_{2}\triangleright f^{\downarrow N}$
with the profunctor $N^{X,Y}\triangleq X\rightarrow Y$.

\textbf{(d)} Since $P^{A}\triangleq A\rightarrow Q^{A}$ is a function
type (with $Q^{A}\triangleq\left(A\rightarrow R\right)\rightarrow R$),
we use Definition~\ref{subsec:Definition-relational-lifting}(e):
\[
(p_{1}^{:P^{A}},p_{2}^{:P^{B}})\in\left<f\right>^{\updownarrow\text{Id}}\ogreaterthan\left<f\right>^{\updownarrow Q}\text{ means if }(a^{:A},b^{:B})\in\left<f\right>\text{ then }(p_{1}(a),p_{2}(b))\in\left<f\right>^{\updownarrow Q}\quad.
\]
The definition of $\left<f\right>^{\updownarrow Q}$ is obtained similarly
if we define $K^{A}\triangleq A\rightarrow R$:
\[
(q_{1}^{:Q^{A}},q_{2}^{:Q^{B}})\in\left<f\right>^{\updownarrow Q}\text{ means if }(k_{1}^{:A\rightarrow R},k_{2}^{:B\rightarrow R})\in\left<f\right>^{\updownarrow K}\text{ then }(q_{1}(k_{1}),q_{2}(k_{2}))\in\text{id}^{R}\quad.
\]
Expanding the definition of $\left<f\right>^{\updownarrow K}$ as
we did in part \textbf{(b)} of this example, we get:
\[
(q_{1},q_{2})\in\left<f\right>^{\updownarrow Q}\text{ means if }b=f(a)\text{ and }k_{1}(a)=k_{2}(b)\text{ then }q_{1}(k_{1})=q_{2}(k_{2})\quad.
\]
This means $k_{1}(a)=k_{2}(f(a))$, or $k_{1}=f\bef k_{2}$. So, we
finally express $\left<f\right>^{\updownarrow Q}$ and $\left<f\right>^{\updownarrow P}$
as:
\begin{align*}
 & (q_{1}^{:Q^{A}},q_{2}^{:Q^{B}})\in\left<f\right>^{\updownarrow Q}\text{ means }(k\rightarrow f\bef k)\bef q_{1}=q_{2}\quad,\\
 & (p_{1}^{:P^{A}},p_{2}^{:P^{B}})\in\left<f\right>^{\updownarrow P}\text{ means }a\rightarrow\big((k\rightarrow f\bef k)\bef p_{1}(a)\big)=f\bef p_{2}\quad.
\end{align*}
Defining $N^{X,Y}\triangleq X\rightarrow\left(Y\rightarrow R\right)\rightarrow R$,
we find that the last formula is equivalent to the $\left(N,f\right)$-wedge
relation for $p_{1}$ and $p_{2}$:
\[
p_{1}\triangleright f^{\uparrow N}=p_{1}\bef(q\rightarrow(k\rightarrow f\bef k)\bef q)\quad,\quad\quad p_{2}\triangleright f^{\downarrow N}=f\bef p_{2}\quad.
\]

\textbf{(e)} We may write $P^{A}=Q^{A}\rightarrow\text{Id}^{A}$ with
$Q^{A}\triangleq A\rightarrow A$ and use Definition~\ref{subsec:Definition-relational-lifting}(e):
\begin{align*}
 & (p_{1}^{:P^{A}},p_{2}^{:P^{B}})\in\left<f\right>^{\updownarrow Q}\ogreaterthan\left<f\right>^{\updownarrow\text{Id}}\text{ means if }(q_{1}^{:A\rightarrow A},q_{2}^{:B\rightarrow B})\in\left<f\right>^{\updownarrow Q}\text{ then }(p_{1}(q_{1}),p_{2}(q_{2}))\in\left<f\right>\quad.\\
 & (q_{1}^{:A\rightarrow A},q_{2}^{:B\rightarrow B})\in\left<f\right>^{\updownarrow Q}\text{ means if }(a^{:A},b^{:B})\in\left<f\right>\text{ then }(q_{1}(a),q_{2}(b))\in\left<f\right>\quad.
\end{align*}
Simplifying these conditions, we get:
\begin{align}
 & (q_{1},q_{2})\in\left<f\right>^{\updownarrow Q}\text{ means }q_{1}\bef f=f\bef q_{2}\quad,\nonumber \\
 & (p_{1}^{:P^{A}},p_{2}^{:P^{B}})\in\left<f\right>^{\updownarrow P}\text{ means if }q_{1}\bef f=f\bef q_{2}\text{ then }f(p_{1}(q_{1}))=p_{2}(q_{2})\quad.\label{eq:relational-lifting-example-e-derivation1}
\end{align}
The condition for $(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}$
\emph{cannot} be written as a single equation involving $p_{1}$ and
$p_{2}$ because we cannot express $q_{1}$ through $q_{2}$ or vice
versa. The relation between $q_{1}$ and $q_{2}$ is many-to-many
and is not equivalent to a function.

Let us define $N^{X,Y}\triangleq\left(Y\rightarrow X\right)\rightarrow Y$
and write the $\left(N,f\right)$-wedge relation for $p_{1}^{:N^{A,A}}$
and $p_{2}^{:N^{B,B}}$:
\begin{equation}
k^{:B\rightarrow A}\triangleright(p_{1}\triangleright f^{\uparrow N})=k\triangleright(p_{2}\triangleright f^{\downarrow N})\quad\text{or equivalently}:\quad f(p_{1}(f\bef k))=p_{2}(k\bef f)\quad.\label{eq:N-f-wedge-relation-example-derivation1}
\end{equation}
We notice that if we set $q_{1}\triangleq f\bef k$ and $q_{2}\triangleq k\bef f$
then the precondition $q_{1}\bef f=f\bef q_{2}$ in Eq.~(\ref{eq:relational-lifting-example-e-derivation1})
will be satisfied, and it will follow that $f(p_{1}(q_{1}))=p_{2}(q_{2})$.
So, the $\left(N,f\right)$-wedge relation is a consequence of the
relation $(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}$. However,
if we have some other functions $f$, $q_{1}$, and $q_{2}$ such
that the precondition $q_{1}\bef f=f\bef q_{2}$ holds, it does not
follow that there exists a function $k$ such that $q_{1}=f\bef k$
and $q_{2}=k\bef f$. A simple counterexample is found when $f$ is
a constant function that ignores its argument and always returns a
fixed value:
\[
f^{:A\rightarrow B}\triangleq\_^{:A}\rightarrow b_{0}\quad,\quad\quad\text{where }b_{0}\text{ is a fixed value of type }B\quad.
\]
Then the precondition $q_{1}\bef f=f\bef q_{2}$ is reduced to the
condition $q_{2}(b_{0})=b_{0}$ with \emph{no} restrictions on $q_{1}$.
But if $q_{1}^{:A\rightarrow A}$ were expressed as $q_{1}=f\bef k$
with some $k^{:B\rightarrow A}$, we would have $q_{1}=\_\rightarrow k(b_{0})$.
So, functions $q_{1}$ expressed as $f\bef k$ are constant functions
that ignore their argument. This is a stronger restriction than $q_{1}\bef f=f\bef q_{2}$
because more pairs $\left(q_{1},q_{2}\right)$ satisfy $q_{1}\bef f=f\bef q_{2}$.
Accordingly, the $\left(N,f\right)$-wedge relation~(\ref{eq:N-f-wedge-relation-example-derivation1})
constrains the functions $\left(p_{1},p_{2}\right)$ weaker and is
satisfied by more pairs $\left(p_{1},p_{2}\right)$ than the relation
$\left<f\right>^{\updownarrow P}$. So, the $\left(N,f\right)$-wedge
relation is\emph{ weaker} than the relation $\left<f\right>^{\updownarrow P}$.
$\square$

\subsection{Properties of relational lifting. Simultaneous lifting}

Definition~\ref{subsec:Definition-relational-lifting} is complicated
and requires additional work to be fully usable. This section proves
the properties of relational lifting that will be required further
in this Appendix.

We begin by pointing out an ambiguity in applying Definition~\ref{subsec:Definition-relational-lifting}(g)
to a type constructor $G^{A}\triangleq\forall X.\,H^{X,A}$ where
$H^{X,A}$ does \emph{not} actually depend on the type parameter $A$
(that is, $H^{X,A}\cong K^{X}$ with some $K^{\bullet}$). In that
case, the type constructor $G$ is equivalent to a fixed type: $G^{A}\cong Z\triangleq\forall X.\,K^{X}$.
We may lift a relation $r^{:A\leftrightarrow B}$ to $G^{A}\triangleq Z$
using Definition~\ref{subsec:Definition-relational-lifting}(a) and
obtain $r^{\updownarrow G}=\text{id}^{:Z\leftrightarrow Z}$. To show
that Definition~\ref{subsec:Definition-relational-lifting} is consistent
and free of contradictions, we need to prove that Definitions~\ref{subsec:Definition-relational-lifting}(a)
and~(g) define the same lifted relation $r^{\updownarrow G}$ in
case $G^{A}=Z$. This will be shown in Statement~\ref{subsec:Statement-parametricity-gives-identity-relation}
below by using the relational naturality law~(\ref{eq:relational-naturality-law-simplified}).
That law is the main conclusion of the relational parametricity theorem
(Section~\ref{subsec:Relational-parametricity-theorem}). So, the
relational parametricity theorem is in fact a necessary requirement
for the consistency of Definition~\ref{subsec:Definition-relational-lifting}.

\subsubsection{Statement \label{subsec:Statement-parametricity-gives-identity-relation}\ref{subsec:Statement-parametricity-gives-identity-relation}}

For a quantified type $Z\triangleq\forall X.\ H^{X}$, the relation
$\text{id}^{:Z\leftrightarrow Z}$ may be expressed as:
\[
(p^{:\forall X.\ H^{X}},q^{:\forall Y.\ H^{Y}})\in\text{id}\quad\text{ means }\quad\forall(X,Y).\:\forall s^{:X\leftrightarrow Y}.\,(p^{X},q^{Y})\in s^{\updownarrow H}\quad.
\]


\subparagraph{Proof}

We need to prove the equivalence in both directions. If some $p$
and $q$ satisfy $(p^{X},q^{Y})\in s^{\updownarrow H}$ where we may
choose $s$ arbitrarily, let us choose $s$ to be the identity relation
that forces $X$ and $Y$ to be the same type:
\[
(x^{:X},y^{:Y})\in s\text{ only when }X=Y\text{ and }x=y\quad.
\]
Lifting the identity relation will produce an identity relation (Statement~\ref{subsec:Statement-lifting-relation-laws}),
so:
\[
\forall(X,Y).\,(p^{X},q^{Y})\in s^{\updownarrow H}\text{ where }s^{\updownarrow H}=\text{id}\quad\text{ means }\quad X=Y\text{ and }p^{X}=q^{X}\quad.
\]
This is the same as the identity relation between $p$ and $q$.

It remains to show that the relation $(p^{X},q^{Y})\in s^{\updownarrow H}$
is satisfied when $p=q$. In other words, $(q^{X},q^{Y})\in s^{\updownarrow H}$.
With the assumption that the value $q$ is implemented via fully parametric
code, we may apply the relational naturality law~(\ref{eq:relational-naturality-law-simplified}),
which we will prove below without using this statement. We set $Q^{X}\triangleq H^{X}$,
$r\triangleq s$, and $t\triangleq q$ in the naturality law~(\ref{eq:relational-naturality-law-simplified})
and obtain:
\[
\forall(X,Y).\,\forall s^{:X\leftrightarrow Y}.\,(q^{X},q^{Y})\in s^{\updownarrow H}\quad.
\]
$\square$

Another construction required for Definition~\ref{subsec:Definition-relational-lifting}
is the simultaneous lifting of two relations to a type constructor
with two type parameters.

\subsubsection{Definition \label{subsec:Definition-simultaneous-relational-lifting}\ref{subsec:Definition-simultaneous-relational-lifting}
(simultaneous relational lifting)}

Given any fully parametric type constructor with two type parameters
($G^{\bullet,\bullet}$) and any two relations $r^{:A\leftrightarrow B}$
and $s^{:X\leftrightarrow Y}$, the \textbf{simultaneous} \textbf{lifting}
of $r$ and $s$ to $G$, denoted by $(r,s)^{\updownarrow G}$, is
a new relation of type $G^{A,X}\leftrightarrow G^{B,Y}$. The relation
$(r,s)^{\updownarrow G}$ is defined by induction on the structure
of $G$:

\textbf{(a)} If $G^{A,X}\triangleq Z$ with a fixed type $Z$, we
define $(r,s)^{\updownarrow G}\triangleq\text{id}^{:Z\leftrightarrow Z}$.

\textbf{(b)} If $G^{A,X}=A$, we define $(r,s)^{\updownarrow G}\triangleq r$.
If $G^{A,X}=X$, we define $(r,s)^{\updownarrow G}\triangleq s$.

\textbf{(c)} If $K$ and $L$ are any fully parametric type constructors,
we define: 
\begin{align*}
\text{for}\quad G^{A,X}\triangleq K^{A,X}\times L^{A,X}\quad: & \quad(r,s)^{\updownarrow G}\triangleq(r,s)^{\updownarrow K}\boxtimes(r,s)^{\updownarrow L}\quad;\\
\text{for}\quad G^{A,X}\triangleq K^{A,X}+L^{A,X}\quad: & (r,s)^{\updownarrow G}\triangleq(r,s)^{\updownarrow K}\boxplus(r,s)^{\updownarrow L}\quad;\\
\text{for}\quad G^{A,X}\triangleq K^{A,X}\rightarrow L^{A,X}\quad: & (r,s)^{\updownarrow G}\triangleq(r,s)^{\updownarrow K}\ogreaterthan(r,s)^{\updownarrow L}\quad.
\end{align*}
The last relation is between functions of types $K^{A,X}\rightarrow L^{A,X}$
and $K^{B,Y}\rightarrow L^{B,Y}$. 

The inductive assumption is that relational liftings to $K$ and $L$
are already defined.

\textbf{(d)} If $G^{A,X}\triangleq S^{A,X,G^{A,X}}$ is defined recursively
via a recursion scheme $S^{\bullet,\bullet,\bullet}$, we define:
\[
(r,s)^{\updownarrow G}\triangleq\big(r,s,\overline{(r,s)^{\updownarrow G}}\big)^{\updownarrow S}\quad.
\]
Here we use $\overline{(r,s)^{\updownarrow G}}$ recursively within
the definition of $(r,s)^{\updownarrow G}$. This is allowed since
we understand $(r,s)^{\updownarrow G}$ to be a function, and it is
permitted to define functions recursively. The inductive assumption
is that simultaneous liftings of any \emph{three} relations to $S^{\bullet,\bullet,\bullet}$
are already defined. 

\textbf{(e)} If $G^{A,X}\triangleq\forall Z.\,H^{Z,A,X}$, we define
$(r,s)^{\updownarrow G}$ of type $(\forall Z.\,H^{Z,A,X})\leftrightarrow(\forall Z.\,H^{Z,B,Y})$
by:
\[
(p^{:\forall Z.\,H^{Z,A,X}},q^{:\forall Z.\,H^{Z,B,Y}})\in(r,s)^{\updownarrow\forall Z.\,H^{Z,\bullet,\bullet}}\quad\text{means}\quad\forall(U,V).\,\forall w^{U\leftrightarrow V}.\,(p^{U},q^{V})\in(w,r,s)^{\updownarrow H^{\bullet,\bullet,\bullet}}\quad.
\]
A shorter way of writing this formula is by formulating a relation
between $p^{U}$ and $q^{V}$ directly:
\begin{equation}
\forall(U,V):\quad\big((r,s)^{\updownarrow\forall Z.\,H^{Z,\bullet,\bullet}}\big)^{U,V}\triangleq\forall w^{:U\leftrightarrow V}.\,(w,r,s)^{\updownarrow H^{\bullet,\bullet,\bullet}}\quad.\label{eq:relational-lifting-quantified-types-short-1}
\end{equation}
The inductive assumption is that any simultaneous liftings to $H^{\bullet,\bullet,\bullet}$
are already defined. $\square$

Parts (d) and (e) of Definition~\ref{subsec:Definition-simultaneous-relational-lifting}
use a simultaneous lifting of three relations. Comparing Definitions~\ref{subsec:Definition-relational-lifting}
and~\ref{subsec:Definition-simultaneous-relational-lifting}, we
see that a similar inductive definition can be given for simultaneous
liftings of $n$ relations (with $n=1,2,...$) to a type constructor
with $n$ type parameters. We omit the details. This book will only
need simultaneous liftings of two relations.

When working with dinaturality laws (Section~\ref{sec:Naturality-laws-for-fully-parametric-functions}),
we use profunctors $P^{X,Y}$ whose type parameters are set to the
same type (e.g., $N^{A}\triangleq P^{A,A}$). In that case, we find
an ambiguity in lifting a relation $r^{:A\leftrightarrow B}$ to $N$:
First, we may compute $r^{\updownarrow N}$ according to Definition~\ref{subsec:Definition-relational-lifting}
and obtain a relation of type $N^{A}\leftrightarrow N^{B}$, which
is the same type as $P^{A,A}\leftrightarrow P^{B,B}$. Alternatively,
we may lift the pair of two relations $\left(r,r\right)$ simultaneously
to $P^{\bullet,\bullet}$ according to Definition~\ref{subsec:Definition-simultaneous-relational-lifting}
and obtain another relation $\left(r,r\right)^{\updownarrow P}$ of
type $P^{A,A}\leftrightarrow P^{B,B}$. The following statement shows
that Definitions~\ref{subsec:Definition-relational-lifting} and~\ref{subsec:Definition-simultaneous-relational-lifting}
are consistent in that case.

\subsubsection{Statement \label{subsec:Statement-relational-lifting-consistency-PAA}\ref{subsec:Statement-relational-lifting-consistency-PAA}}

Given any fully parametric type constructor $P^{X,Y}$, define $N^{A}\triangleq P^{A,A}$.
For any relation $r^{:A\leftrightarrow B}$, the liftings $r^{\updownarrow N}$
and $\left(r,r\right)^{\updownarrow P}$ will produce the same relation
of type $P^{A,A}\leftrightarrow P^{B,B}$.

\subparagraph{Proof}

{*}{*}{*}

$\square$

The proof in Statement~\ref{subsec:Statement-wedge-law-from-parametricity}
will also use the identity laws\index{identity laws!of relational lifting}
of the relational lifting operation:

\subsubsection{Statement \label{subsec:Statement-relational-lifting-identity-law}\ref{subsec:Statement-relational-lifting-identity-law}}

\textbf{(a)} Lifting an identity relation gives again an identity
relation: 
\[
\text{for any }G^{\bullet}\quad:\quad(\text{id}^{:A\leftrightarrow A})^{\updownarrow G}=\text{id}^{:G^{A}\leftrightarrow G^{A}}\quad.
\]

\textbf{(b)} An identity relation can be removed from a simultaneous
lifting. For instance, given any type constructor $H^{\bullet,\bullet,\bullet}$
and any relations $r^{:A\leftrightarrow B}$ and $s^{:X\leftrightarrow Y}$,
we have: 
\[
(r,s,\text{id}^{:Z\leftrightarrow Z})^{\updownarrow H^{\bullet,\bullet,\bullet}}=(r,s)^{\updownarrow H^{\bullet,\bullet,Z}}\quad,
\]
where the last lifting is to a type constructor $H^{\bullet,\bullet,Z}$
that has its third type parameter set to $Z$, leaving only two free
type parameters.

\subparagraph{Proof}

\textbf{(a)} Lifting

\textbf{(b)} An

{*}{*}{*}

\subsection{Relational parametricity theorem\label{subsec:Relational-parametricity-theorem}}

Our goal is to prove that any fully parametric function $t:\forall A.\,P^{A}\rightarrow Q^{A}$
obeys the relational naturality law~(\ref{eq:naturality-law-of-t-derivation2}):
\[
\forall A,B.\,\forall r^{:A\leftrightarrow B}.\,(t^{A},t^{B})\in r^{\updownarrow P}\ogreaterthan r^{\updownarrow Q}\quad.
\]
The proof will need to go by induction on the structure of the code
of the function $t$, which is built from smaller sub-expressions
using the nine code constructions of Table~\ref{tab:nine-pure-code-constructions}.
The inductive assumption is that all sub-expressions already satisfy
the relational naturality law. An immediate difficulty with this approach
is that the sub-expressions of $t$ may not have the type signature
of a function and may also contain free variables. To illustrate,
write the code of $t$ as $t=z^{:P^{A}}\rightarrow\text{expr}(z)$,
or in Scala:
\begin{lstlisting}
def t[A] = { z: P[A] => expr(z) }
\end{lstlisting}
where \textsf{``}\lstinline!expr(z)!\textsf{''} is the function\textsf{'}s body. That function
body \emph{itself} does not necessarily have a type signature of a
function (i.e., it is not necessarily of the form $K^{A}\rightarrow L^{A}$).
Also, \textsf{``}\lstinline!expr(z)!\textsf{''} may contain $z$ as a free variable
defined outside the scope of \textsf{``}\lstinline!expr(z)!\textsf{''}, and the type
of $z$ may depend on the type parameter $A$. So, we cannot directly
apply the relational naturality law~(\ref{eq:naturality-law-of-t-derivation2})
to the subexpressions, which prevents us from using the inductive
assumption. The relational naturality law needs to be reformulated
to describe function \emph{bodies}, i.e., arbitrary expressions that
may contain externally defined variables. A suitable formulation of
the relational naturality law is given in the next definition and
will be proved below in Statement~\ref{subsec:Statement-main-relational-parametricity-1}.

\subsubsection{Definition \label{subsec:Definition-relational-naturality-law}\ref{subsec:Definition-relational-naturality-law}
(relational naturality law)}

Consider any expression $t:\forall A.\,Q^{A}$ containing a single
free variable $x^{:P^{A}}$, where $P^{\bullet}$ and $Q^{\bullet}$
are any type constructors. Define the \textbf{binding function} $t^{\prime}:\forall A.\,P^{A}\rightarrow Q^{A}$
such that $t=\tilde{t}(x)$ and $\tilde{t}$ has no free variables.
(The binding function describes how the expression $t$ depends on
its free variable $x$.) Then the \textbf{relational naturality law}\index{naturality law!in terms of relations}
is written as:
\begin{equation}
\forall(A,B).\,\forall r^{:A\leftrightarrow B}.\,(\tilde{t}^{A},\tilde{t}^{B})\in r^{\updownarrow P}\ogreaterthan r^{\updownarrow Q}\quad.\label{eq:relational-naturality-law-1}
\end{equation}
If $t$ contains several free variables ($x_{1}^{:P_{1}^{A}}$, $x_{2}^{:P_{2}^{A}}$,
etc.), we define the binding function $\tilde{t}$ as a curried function
of all the free variables. For example, with \emph{two} free variables
we will have $t=\tilde{t}(x_{1})(x_{2})$, so that $\tilde{t}^{A}$
will have type $P_{1}^{A}\rightarrow P_{2}^{A}\rightarrow Q^{A}$.
Then the relational naturality law is written as:
\begin{equation}
\forall(A,B).\,\forall r^{:A\leftrightarrow B}.\,(\tilde{t}^{A},\tilde{t}^{B})\in r^{\updownarrow P_{1}}\ogreaterthan(r^{\updownarrow P_{2}}\ogreaterthan r^{\updownarrow Q})\quad.\label{eq:relational-naturality-law-two-free-vars}
\end{equation}
The generalization to any number of free variables is straightforward.
$\square$

If $t$ contains no free variables, we may still write $t$ as $\tilde{t}(x)$
where $\tilde{t}$ does not depend on $x$, setting $P^{A}\triangleq\bbnum 1$
for simplicity. In that case, Eq.~(\ref{eq:relational-naturality-law-1})
is simplified to:
\begin{align}
 & \forall(A,B).\,\forall r^{:A\leftrightarrow B}.\,\forall x_{1}^{:\bbnum 1},x_{2}^{:\bbnum 1}.\,\quad\text{if }x_{1}=x_{2}\text{ then }(\tilde{t}^{A}(x_{1}),\tilde{t}^{B}(x_{2}))\in r^{\updownarrow Q}\quad,\nonumber \\
{\color{greenunder}\text{or equivalently}:}\quad & \forall(A,B).\,\forall r^{:A\leftrightarrow B}.\,(t^{A},t^{B})\in r^{\updownarrow Q}\quad.\label{eq:relational-naturality-law-simplified-1}
\end{align}
Even in that case, to save time, we will keep using Eq.~(\ref{eq:relational-naturality-law-1})
and writing $t=\tilde{t}(x)$.

When an expression contains more than one free variable, we can gather
all the free variables into a tuple. This creates an equivalent expression
with just one free variable:

\subsubsection{Statement \label{subsec:Statement--relational-naturality-tuple-1}\ref{subsec:Statement--relational-naturality-tuple-1} }

The relational naturality law for an expression $t^{:\forall A.\,Q^{A}}$
containing two free variables $z_{1}^{:K^{A}}$ and $z_{2}^{:L^{A}}$
and the binding function $\tilde{t}(z_{1})(z_{2})=t$ is equivalent
to the relational naturality law for the expression $u$ with \emph{one}
free variable $h$ defined as:
\[
h^{:K^{A}\times L^{A}}\triangleq z_{1}\times z_{2}\quad,\quad\quad u\triangleq\tilde{u}(h)\triangleq\tilde{t}(h\triangleright\pi_{1})(h\triangleright\pi_{2})\quad.
\]


\subparagraph{Proof}

The relational naturality laws for $t$ and $u$ say that, for all
types $A$, $B$, and relations $r^{:A\leftrightarrow B}$:
\[
(\tilde{t}^{A},\tilde{t}^{B})\in r^{\updownarrow K}\ogreaterthan(r^{\updownarrow L}\ogreaterthan r^{\updownarrow Q})\quad,\quad\quad(\tilde{u}^{A},\tilde{u}^{B})\in r^{\updownarrow(K\times L)}\ogreaterthan r^{\updownarrow Q}\quad.
\]
We need to show that these two relations are equivalent given the
definition of $\tilde{u}$ via $\tilde{t}$. This equivalence is similar
to the equivalence of curried and uncurried function types ($A\rightarrow(B\rightarrow C)\cong A\times B\rightarrow C$).
We write out the definition of the pair mapper operation ($\ogreaterthan$)
and obtain: 
\begin{align*}
(\tilde{t}^{A},\tilde{t}^{B})\in r^{\updownarrow K}\ogreaterthan(r^{\updownarrow L}\ogreaterthan r^{\updownarrow Q})\quad\text{means}\quad: & \quad\forall x_{1}^{:K^{A}},x_{2}^{:L^{A}},y_{1}^{:K^{B}},y_{2}^{:L^{B}}:\,\text{ if }(x_{1},y_{1})\in r^{\updownarrow K}\text{ and }(x_{2},y_{2})\in r^{\updownarrow L}\\
 & \quad\quad\text{ then }(\tilde{t}(x_{1})(x_{2}),\tilde{t}(y_{1})(y_{2}))\in r^{\updownarrow Q}\quad;\\
(\tilde{u}^{A},\tilde{u}^{B})\in r^{\updownarrow(K\times L)}\ogreaterthan r^{\updownarrow Q}\quad\text{means}\quad: & \quad\forall h^{:K^{A}\times L^{A}},w^{:K^{B}\times L^{B}}:\,\text{ if }(h,w)\in r^{\updownarrow(K\times L)}\\
 & \quad\quad\text{ then }(\tilde{u}(h),\tilde{u}(w))\in r^{\updownarrow Q}\quad.
\end{align*}
By Definition~\ref{subsec:Definition-relational-lifting}(c) we have
$r^{\updownarrow(K\times L)}=r^{\updownarrow K}\boxtimes r^{\updownarrow L}$,
and so we may write: 
\[
(h,w)\in r^{\updownarrow(K\times L)}\quad\text{ if }\quad h=x_{1}\times x_{2}\quad,\quad w=y_{1}\times y_{2}\quad,\quad(x_{1},x_{2})\in r^{\updownarrow K}\quad,\quad\text{ and }\quad(y_{1},y_{2})\in r^{\updownarrow L}\quad.
\]
When $h=x_{1}\times x_{2}$ and $w=y_{1}\times y_{2}$, we will have
$\tilde{u}(h)=\tilde{t}(x_{1},x_{2})$ and $\tilde{u}(w)=\tilde{t}(y_{1},y_{2})$.
It follows that the relational naturality laws for $t$ and $u$ are
equivalent. $\square$

Due to Statement~\ref{subsec:Statement--relational-naturality-tuple-1},
we are allowed to assume that the expression $t$ always has a \emph{single}
free variable. This simplifies the formulation of the relational parametricity
theorem:

\subsubsection{Statement \label{subsec:Statement-main-relational-parametricity-1}\ref{subsec:Statement-main-relational-parametricity-1}
(relational parametricity theorem)}

Let $H$ and $Q$ be any fully parametric, exponential-polynomial
type constructors. Any fully parametric expression $t:\forall A.\,Q^{A}$
of the form $t=\forall A.\:\tilde{t}^{A}(h)$ containing a single
free variable $h^{:H^{A}}$ satisfies the relational naturality law~(\ref{eq:relational-naturality-law-1}).

\subparagraph{Proof }

By assumption, $t$ is built up from the nine constructions of Table~\ref{tab:nine-pure-code-constructions}.
So, one of these nine constructions is at the top level in the syntax
tree of $t$. For each of those constructions, we will prove that
$t$ satisfies Eq.~(\ref{eq:relational-naturality-law-1}) as long
as all its sub-expressions do. Throughout the proof, all relational
naturality laws will involve an arbitrary relation $r^{:A\leftrightarrow B}$
between arbitrary types $A$, $B$. For brevity, we will not write
the quantifiers $\forall A,B,r^{:A\leftrightarrow B}$ in front of
all formulas.

To shorten the proof further, we note that in every inductive case
the expression $t$ and all its sub-expressions will contain the same
free variable $h$. So, all relational naturality laws will follow
the pattern \textsf{``}for all $h_{1}^{:H^{A}}$ and $h_{2}^{:H^{B}}$ satisfying
the relation $(h_{1},h_{2})\in r^{\updownarrow H}$, some other values
are in some other relation\textsf{''}. Let us choose arbitrary but fixed values
$h_{1}^{:H^{A}}$ and $h_{2}^{:H^{B}}$ satisfying $(h_{1},h_{2})\in r^{\updownarrow H}$,
denote $t_{1}\triangleq\tilde{t}^{A}(h_{1})$ and $t_{2}\triangleq\tilde{t}^{B}(h_{2})$,
and simplify the naturality law~(\ref{eq:relational-naturality-law-1})
to:
\begin{equation}
(t_{1},t_{2})\in r^{\updownarrow Q}\quad\text{where}\quad t_{1}\triangleq\tilde{t}^{A}(h_{1})\text{ and }t_{2}\triangleq\tilde{t}^{B}(h_{2})\quad.\label{eq:relational-naturality-law-with-fixed-h}
\end{equation}
We will now prove this form of the law. Inductive assumptions will
always begin with \textsf{``}for all $h_{1}$ and $h_{2}$ ...\textsf{''}, so we are
allowed to substitute the fixed values $h_{1}$, $h_{2}$ into each
of the inductive assumptions and write them also in the form of Eq.~(\ref{eq:relational-naturality-law-with-fixed-h}).

\paragraph{Use unit value}

In this case, $t\triangleq1$ and has the unit type ($Q^{A}\triangleq\bbnum 1$).
Since $t$ contains no bound variables, the law~(\ref{eq:relational-naturality-law-with-fixed-h})
becomes:
\[
(1,1)\in r^{\updownarrow Q}\quad.
\]
This holds because by Definition~\ref{subsec:Definition-relational-lifting}(a)
the relation $r^{\updownarrow Q}=\text{id}$ regardless of $r$.

The same proof applies for $t\triangleq c$ where $c^{:C}$ is a fixed
value of a fixed type $C$.

\paragraph{Use argument}

In this case, $t\triangleq h$ where $h^{:H^{A}}$ is a free variable
(say, the argument of the function whose body is $t$). So, we must
have $Q=H$ and $\tilde{t}=\forall A.\:\text{id}^{:H^{A}\rightarrow H^{A}}$.
The law~(\ref{eq:relational-naturality-law-with-fixed-h}) then becomes:
\begin{align*}
 & (t_{1},t_{2})\in r^{\updownarrow H}\quad\text{where}\quad t_{1}\triangleq h_{1}\text{ and }t_{2}\triangleq h_{2}\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & (h_{1},h_{2})\in r^{\updownarrow H}\quad.
\end{align*}
The last condition holds trivially, since $(h_{1},h_{2})\in r^{\updownarrow H}$
is already assumed.

\paragraph{Create function}

In this case, $t\triangleq\forall A.\,p^{:P^{A}}\rightarrow\tilde{g}^{A}(h)(p)$
where the sub-expression $\tilde{g}^{A}(h)(p):G^{A}$ contains two
free variables ($h^{:H^{A}}$ and $p^{:P^{A}}$). Because $\tilde{g}(h)(p)$
is defined with curried arguments, the binding function $\tilde{t}$
satisfies:
\[
\tilde{t}(h)=t=p\rightarrow\tilde{g}(h)(p)=\tilde{g}(h)\quad,\quad\quad\text{or equivalently}:\quad\tilde{t}=\tilde{g}.
\]
The type of $\tilde{g}$ is $H^{A}\rightarrow Q^{A}$, where we set
$Q^{A}\triangleq P^{A}\rightarrow G^{A}$. Definition~\ref{subsec:Definition-relational-lifting}(e)
gives $r^{\updownarrow Q}=r^{\updownarrow P}\ogreaterthan r^{\updownarrow G}$.
So, the inductive assumption for $g$ may be written as:
\[
(g_{1},g_{2})\in r^{\updownarrow P}\ogreaterthan r^{\updownarrow G}\quad\text{where}\quad g_{1}\triangleq\tilde{g}^{A}(h_{1})=t_{1}\text{ and }g_{2}\triangleq\tilde{g}^{B}(h_{2})=t_{2}\quad.
\]
This is equivalent to:
\[
(t_{1},t_{2})\in r^{\updownarrow P}\ogreaterthan r^{\updownarrow G}=r^{\updownarrow Q}\quad,
\]
which is the law~(\ref{eq:relational-naturality-law-with-fixed-h})
we needed to prove.

\paragraph{Use function}

In this case, $t=\tilde{t}(h)\triangleq k(p)$ where $k^{:\forall A.\,P^{A}\rightarrow Q^{A}}$
and $p^{:\forall A.\,P^{A}}$ are some sub-expressions. Both $k$
and $p$ may contain $h^{:H^{A}}$ as a free variable: $k\triangleq\tilde{k}(h)$
and $p\triangleq\tilde{p}(h)$. Denote for brevity $K^{A}\triangleq P^{A}\rightarrow Q^{A}$.
The inductive assumptions say that the laws~(\ref{eq:relational-naturality-law-with-fixed-h})
already hold for both $k$ and $p$:
\begin{align*}
{\color{greenunder}\text{for }k:}\quad & (k_{1},k_{2})\in r^{\updownarrow K}=r^{\updownarrow P}\ogreaterthan r^{\updownarrow Q}\quad\quad\text{where}\quad k_{1}\triangleq\tilde{k}^{A}(h_{1})\text{ and }k_{2}\triangleq\tilde{k}^{B}(h_{2})\quad,\\
{\color{greenunder}\text{for }p:}\quad & (p_{1},p_{2})\in r^{\updownarrow P}\quad\quad\text{where}\quad p_{1}\triangleq\tilde{p}^{A}(h_{1})\text{ and }p_{2}\triangleq\tilde{p}^{B}(h_{2})\quad.
\end{align*}
We need to show that:
\[
(t_{1},t_{2})\in r^{\updownarrow Q}\quad\quad\text{where}\quad t_{1}\triangleq\tilde{t}^{A}(h_{1})=k_{1}(p_{1})\text{ and }t_{2}\triangleq\tilde{t}^{B}(h_{2})=k_{2}(p_{2})\quad.
\]
By definition of $\ogreaterthan$, the inductive assumption $(k_{1},k_{2})\in r^{\updownarrow P}\ogreaterthan r^{\updownarrow Q}$
means:
\[
(k_{1}(p_{1}),k_{2}(p_{2}))\in r^{\updownarrow Q}\quad\text{whenever}\quad(p_{1},p_{2})\in r^{\updownarrow P}\quad.
\]
Since $(p_{1},p_{2})\in r^{\updownarrow P}$ already holds by the
other inductive assumption, we obtain $(t_{1},t_{2})\in r^{\updownarrow Q}$
as required.

\paragraph{Create tuple}

In this case, $t=\tilde{t}(h)\triangleq k^{:K^{A}}\times l^{:L^{A}}$,
where the sub-expressions $k$ and $l$ contain the free variable
$h^{:H^{A}}$ via $k\triangleq\tilde{k}(h)$ and $l\triangleq\tilde{l}(h)$.
We have $Q^{A}\triangleq K^{A}\times L^{A}$, so we use Definition~\ref{subsec:Definition-relational-lifting}(c)
for the lifting $r^{\updownarrow Q}$ to get $r^{\updownarrow Q}=r^{\updownarrow K}\boxtimes r^{\updownarrow L}$.
We need to prove that:
\begin{align*}
 & (t_{1},t_{2})\in r^{\updownarrow Q}\quad\text{or equivalently}:\quad(k_{1}\times l_{1},k_{2}\times l_{2})\in r^{\updownarrow K}\boxtimes r^{\updownarrow L}\quad,\\
 & \quad\text{where}\quad t_{1}\triangleq k_{1}\times l_{1}\triangleq\tilde{k}^{A}(h_{1})\times\tilde{l}^{A}(h_{1})\quad,\quad\quad t_{2}\triangleq k_{2}\times l_{2}\triangleq\tilde{k}^{A}(h_{2})\times\tilde{l}^{A}(h_{2})\quad.
\end{align*}
The inductive assumptions are that the relational naturality law~(\ref{eq:relational-naturality-law-with-fixed-h})
holds for $k$ and $l$:
\[
(k_{1},k_{2})\in r^{\updownarrow K}\quad,\quad\quad(l_{1},l_{2})\in r^{\updownarrow L}\quad.
\]
By definition of $\boxtimes$ for relations, we obtain the required
property: $(k_{1}\times l_{1},k_{2}\times l_{2})\in r^{\updownarrow K}\boxtimes r^{\updownarrow L}$.

\paragraph{Use tuple}

It is sufficient to consider the case $t\triangleq\pi_{1}(g)$ where
$g^{:\forall A.\,Q^{A}\times L^{A}}$ is a sub-expression that contains
the free variable $h^{:H^{A}}$. The proof for $t=\pi_{2}(g)$ is
analogous. 

We need to prove that $t$ satisfies the relational naturality law
if $g$ does. Define $g=\tilde{g}(h)$ and:
\[
g_{1}\triangleq\tilde{g}^{A}(h_{1})\quad,\quad\quad g_{2}\triangleq\tilde{g}^{B}(h_{2})\quad,\quad\quad t_{1}\triangleq\pi_{1}(g_{1})\quad,\quad\quad t_{2}\triangleq\pi_{1}(g_{2})\quad.
\]

The relational naturality law of $g$, which holds by the inductive
assumption, is:
\[
(g_{1},g_{2})\in r^{\updownarrow(Q\times L)}\quad.
\]
By Definition~\ref{subsec:Definition-relational-lifting}(c) for
the lifting $r^{\updownarrow(Q\times L)}$, we have:
\[
(g_{1},g_{2})\in r^{\updownarrow(Q\times L)}\quad\text{ means }\quad(\pi_{1}(g_{1}),\pi_{1}(g_{2}))\in r^{\updownarrow Q}\text{ and }(\pi_{2}(g_{1}),\pi_{2}(g_{2}))\in r^{\updownarrow L}\quad.
\]
The condition $(\pi_{1}(g_{1}),\pi_{1}(g_{2}))\in r^{\updownarrow Q}$
is the same as the relational naturality law of $t$.

\paragraph{Create disjunction}

We consider the case $t\triangleq g+\bbnum 0$ where $Q^{A}\triangleq K^{A}+L^{A}$
and $g^{:\forall A.\,K^{A}}$ is a sub-expression that contains the
free variable $h^{:H^{A}}$ via $g\triangleq\tilde{g}(h)$. Define
for convenience: 
\[
g_{1}\triangleq\tilde{g}^{A}(h_{1})\quad,\quad\quad g_{2}\triangleq\tilde{g}^{B}(h_{2})\quad,\quad\quad t_{1}=g_{1}+\bbnum 0\quad,\quad\quad t_{2}=g_{2}+\bbnum 0\quad.
\]
The inductive assumption is that $g(h)$ satisfies its relational
naturality law, which is $(g_{1},g_{2})\in r^{\updownarrow K}$. By
Definition~\ref{subsec:Definition-relational-lifting}(d) for the
lifting $r^{\updownarrow(K+L)}$, we have:
\[
(t_{1},t_{2})=(g_{1}+\bbnum 0,g_{2}+\bbnum 0)\in r^{\updownarrow(K+L)}=r^{\updownarrow K}\boxplus r^{\updownarrow L}\text{ when }(g_{1},g_{2})\in r^{\updownarrow K}\quad.
\]
So, the relational naturality law of $t$ holds. The proof for $t\triangleq\bbnum 0+g$
is analogous.

\paragraph{Use disjunction}

In this case, $t\triangleq\forall A.\,\,\begin{array}{|c||c|}
 & G^{A}\\
\hline K^{A} & p\\
L^{A} & q
\end{array}\,\,$ is a pattern-matching function of type $\forall A.\,Q^{A}$ with
$Q^{A}\triangleq K^{A}+L^{A}\rightarrow G^{A}$. The sub-expressions
$p^{:\forall A.\,K^{A}\rightarrow G^{A}}$ and $q^{:\forall A.\,L^{A}\rightarrow G^{A}}$
contain the free variable $h^{:H^{A}}$. We define:
\[
p\triangleq\tilde{p}(h)\quad,\quad p_{1}\triangleq\tilde{p}^{A}(h_{1})\quad,\quad p_{2}\triangleq\tilde{p}^{B}(h_{2})\quad,\quad\quad q\triangleq\tilde{q}(h)\quad,\quad q_{1}\triangleq\tilde{q}^{A}(h_{1})\quad,\quad q_{2}\triangleq\tilde{q}^{B}(h_{2})\quad.
\]
Then we have:
\[
t_{1}=\,\begin{array}{|c||c|}
 & G^{A}\\
\hline K^{A} & p_{1}\\
L^{A} & q_{1}
\end{array}\quad,\quad\quad t_{2}=\,\begin{array}{|c||c|}
 & G^{B}\\
\hline K^{B} & p_{2}\\
L^{B} & q_{2}
\end{array}\quad.
\]
By the inductive assumption, the relational naturality law already
holds for $p$ and $q$:
\begin{align*}
 & \text{if }(k_{1}^{:K^{A}},k_{2}^{:K^{B}})\in r^{\updownarrow K}\text{ then }\big(p_{1}(k_{1}),p_{2}(k_{2})\big)\in r^{\updownarrow G}\quad,\\
 & \text{if }(l_{1}^{:L^{A}},l_{2}^{:L^{B}})\in r^{\updownarrow L}\text{ then }\big(q_{1}(l_{1}),q_{2}(l_{2})\big)\in r^{\updownarrow G}\quad.
\end{align*}
To derive the specific form of the law for $t$, we use Definition~\ref{subsec:Definition-relational-lifting}(e)
for the lifting $r^{\updownarrow Q}$:
\[
\text{if }(x_{1}^{:K^{A}+L^{A}},x_{2}^{:K^{B}+L^{B}})\in r^{\updownarrow(K+L)}\text{ then }\big(t_{1}(x_{1}),t_{2}(x_{2})\big)\in r^{\updownarrow G}\quad.
\]
By Definition~\ref{subsec:Definition-relational-lifting}(d), the
values $x_{1}$ and $x_{2}$ are in relation $r^{\updownarrow(K+L)}$
only if both $x_{1}$ and $x_{2}$ are in the same part of the disjunction
($K^{\bullet}+L^{\bullet}$). We consider separately the case when
they are in the left part or in the right part.

If both $x_{1}$ and $x_{2}$ in the left part, we can write $x_{1}=k_{1}^{:K^{A}}+\bbnum 0$
and $x_{2}=k_{2}^{:K^{B}}+\bbnum 0$ with some $k_{1}$ and $k_{2}$.
Then the condition $(x_{1},x_{2})\in r^{\updownarrow(K+L)}$ is equivalent
to $(k_{1},k_{2})\in r^{\updownarrow K}$, while the values $t_{1}(x_{1})$
and $t_{2}(x_{2})$ are expressed as:
\[
t_{1}(x_{1})=(k_{1}+\bbnum 0)\triangleright\,\begin{array}{|c||c|}
 & G^{A}\\
\hline K^{A} & p_{1}\\
L^{A} & q_{1}
\end{array}\,=p_{1}(k_{1})\quad,\quad\quad t_{2}(x_{2})=(k_{2}+\bbnum 0)\triangleright\,\begin{array}{|c||c|}
 & G^{B}\\
\hline K^{B} & p_{2}\\
L^{B} & q_{2}
\end{array}\,=p_{2}(k_{2})\quad.
\]
So, the conclusion of the relational naturality law of $t$ holds
due to the relational law of $p$:
\[
\big(t_{1}(x_{1}),t_{2}(x_{2})\big)=\big(p_{1}(k_{1}),p_{2}(k_{2})\big)\in r^{\updownarrow G}\quad.
\]

A similar argument proves the law for the case when both $x_{1}$
and $x_{2}$ in the right part. We write $x_{1}=\bbnum 0+l_{1}^{:L^{A}}$
and $x_{2}=\bbnum 0+l_{2}^{:L^{B}}$ and reduce the relational naturality
law of $t$ to that of $q$.

\paragraph{Recursive call}

In this case, $t\triangleq g$ where $g$ is a recursive call to a
function (possibly defined outside that expression). When proving
a law of a recursively defined function, we may assume that the law
holds for recursive calls to that function. So, the inductive assumption
says that $g$ is some expression for which the relational naturality
law already holds. It then holds for $t$ since $t=g$.

This completes the proof of the relational parametricity theorem.
$\square$ 

The relational formulation of parametricity is powerful but hard to
use for deriving laws as equations rather than as relations. To convert
the relational parametricity law such as Eq.~(\ref{eq:relational-naturality-law-simplified-1})
into an equation with functions, we first choose the relation $r$
as $r\triangleq\left<f\right>$ with a suitable function $f$. Then
we need to express the lifted relation $r^{\updownarrow Q}$ via an
equation and derive the corresponding equation for $t$. However,
Example~\ref{subsec:Example-relational-lifting}(e) shows that for
sufficiently complicated type constructors $Q$ the lifted relation
$\left<f\right>^{\updownarrow Q}$ will \emph{not} be expressible
as an equation. 

Nevertheless, this problem does not arise for most type signatures
found in practice, and the relational naturality law can be reduced
to the dinaturality law~(\ref{eq:dinaturality-law-for-profunctors})
via the wedge law~(\ref{eq:wedge-law-for-profunctors}). In a few
cases, a more powerful property called the \textsf{``}strong\textsf{''} dinaturality
law is required. In this book, strong dinaturality is only used to
prove Statements~\ref{subsec:relational-property-for-foldFn}, \ref{subsec:Statement-Church-encoding-recursive-type-covariant},
and~\ref{subsec:Statement-strong-dinaturality-property-of-fix},
while in all other places the ordinary dinaturality law is sufficient.
The following sections will prove that the wedge law and the strong
dinaturality law are consequences of the relational parametricity
theorem.

\subsection{Wedge law derived from relational parametricity}

We will show two proofs of the wedge law. One proof follows a blog
post by B.~Milewski\index{Bartosz Milewski} that uses the relational
naturality law of \lstinline!xmap!.\footnote{See \texttt{\href{https://bartoszmilewski.com/2017/04/11/profunctor-parametricity/}{https://bartoszmilewski.com/2017/04/11/profunctor-parametricity/}}}
Another, more illuminating but longer proof uses the composition of
relations and detailed properties of relational lifting.

\subsubsection{Statement \label{subsec:Statement-wedge-law-from-parametricity}\ref{subsec:Statement-wedge-law-from-parametricity}}

Any fully parametric value $t$ of type $\forall A.\,P^{A,A}$ where
$P^{X,Y}$ is a profunctor (contravariant in $X$ and covariant in
$Y$) satisfies the wedge law~(\ref{eq:wedge-law-for-profunctors}):
\[
\text{for all }A,B,f^{:A\rightarrow B}\quad:\quad t^{A}\triangleright f^{\uparrow P^{A,\bullet}}=t^{B}\triangleright f^{\downarrow P^{\bullet,B}}\quad.
\]


\subparagraph{Proof}

All profunctors $P$ have a \lstinline!xmap! method with the type
signature:
\[
\text{xmap}_{P}:\forall A,B,C,D.\,\left(B\rightarrow A\right)\rightarrow\left(C\rightarrow D\right)\rightarrow P^{A,C}\rightarrow P^{B,D}\quad.
\]
Since $t$ is assumed to be fully parametric, the type constructor
$P^{X,Y}$ must be a fully parametric profunctor. So, the \lstinline!xmap!
method is also fully parametric and satisfies the commutativity law
(Section~\ref{subsec:Proof-of-the-profunctor-commutativity-law}):
\[
\text{xmap}_{P}(f^{:B\rightarrow A})(g^{:C\rightarrow D})=f^{\downarrow P^{\bullet,C}}\bef g^{\uparrow P^{A,\bullet}}=g^{\uparrow P^{B,\bullet}}\bef f^{\downarrow P^{\bullet,D}}\quad.
\]
 Statement~\ref{subsec:Statement-main-relational-parametricity-1}
then shows that \lstinline!xmap! satisfies a relational naturality
law. Since dimap has four type parameters ($A$, $B$, $C$, $D$),
its relational naturality law involves eight arbitrary types ($A$,
$B$, $C$, $D$, $A^{\prime}$, $B^{\prime}$, $C^{\prime}$, $D^{\prime}$)
and four arbitrary relations that we denote by $a^{:A\leftrightarrow A^{\prime}}$,
$b^{:B\leftrightarrow B^{\prime}}$, $c^{:C\leftrightarrow C^{\prime}}$,
and $d^{:D\leftrightarrow D^{\prime}}$. With these notations, we
write the relational naturality law of \lstinline!xmap! as:
\[
\forall(a,b,c,d)\quad:\quad(\text{xmap}_{P}^{A,B,C,D},\text{xmap}_{P}^{A^{\prime},B^{\prime},C^{\prime},D^{\prime}})\in(b\ogreaterthan a)\ogreaterthan\big((c\ogreaterthan d)\ogreaterthan\big((a,c)^{\updownarrow P}\ogreaterthan(b,d)^{\updownarrow P}\big)\big)\quad,
\]
where $(a,c)^{\updownarrow P}$ and $(b,d)^{\updownarrow P}$ are
simultaneous relational liftings to the profunctor $P$ (see Definition~\ref{subsec:Definition-simultaneous-relational-lifting}).

We now substitute the following (carefully chosen) types and relations
into this law:
\begin{align*}
 & A=B=B^{\prime}=C\quad,\quad\quad A^{\prime}=C^{\prime}=D=D^{\prime}\quad,\\
 & a=c\quad,\quad\quad b=\text{id}\quad,\quad\quad d=\text{id}\quad.
\end{align*}
The identity laws of Statement~\ref{subsec:Statement-relational-lifting-identity-law}
give: $(b,d)^{\updownarrow P}=(\text{id},\text{id})^{\updownarrow P}=\text{id}$.
The law of \lstinline!xmap! becomes:
\[
\forall(A,D,c^{:A\leftrightarrow D})\quad:\quad(\text{xmap}_{P}^{A,A,A,D},\text{xmap}_{P}^{D,A,D,D})\in(\text{id}^{:A\leftrightarrow A}\ogreaterthan c)\ogreaterthan\big((c\ogreaterthan\text{id}^{:D\leftrightarrow D})\ogreaterthan\big((c,c)^{\updownarrow P}\ogreaterthan\text{id}^{:P^{A,D}\leftrightarrow P^{A,D}}\big)\big)\quad.
\]
Writing out the definitions of $\boxtimes$ and $\ogreaterthan$,
we obtain:
\begin{align*}
 & \forall\big(f^{:A\rightarrow A},g^{:A\rightarrow D},h^{:A\rightarrow D},k^{:D\rightarrow D},p^{:P^{A,A}},q^{:P^{D,D}}\big)\quad:\\
 & \quad\text{if}\quad\quad(f,g)\in\text{id}^{:A\leftrightarrow A}\ogreaterthan c\quad,\quad(h,k)\in c\ogreaterthan\text{id}^{:D\leftrightarrow D}\quad,\quad(p,q)\in(c,c)^{\updownarrow P}\quad,\\
 & \quad\text{then}\quad\quad\text{xmap}_{P}(f)(h)(p)=\text{xmap}_{P}(g)(k)(q)\quad.
\end{align*}
Further, we choose $f$, $g$, $k$, $h$, and the relation $c$ as
follows:
\[
f\triangleq\text{id}^{:A\rightarrow A}\quad,\quad\quad k\triangleq\text{id}^{:D\rightarrow D}\quad,\quad\quad g=h\quad,\quad\quad c\triangleq\left<g\right>\quad.
\]
With these choices, the conditions $(f,g)\in\text{id}^{:A\leftrightarrow A}\ogreaterthan c$
and $(h,k)\in c\ogreaterthan\text{id}^{:D\leftrightarrow D}$ will
always hold. Substituting these definitions into the relational naturality
law of \lstinline!xmap!, we get:
\[
\forall(g^{:A\rightarrow D},p^{:P^{A,A}},q^{:P^{D,D}})\quad:\quad\text{if}\quad(p,q)\in(\left<g\right>,\left<g\right>)^{\updownarrow P}\quad\text{then}\quad\text{xmap}_{P}(\text{id})(g)(p)=\text{xmap}_{P}(g)(\text{id})(q)\quad.
\]
Now we choose $p\triangleq t^{A}$ and $q\triangleq t^{D}$ where
$t$ is a given fully parametric value of type $\forall A.\,P^{A,A}$.
Denote temporarily $T^{A}\triangleq P^{A,A}$. The relational parametricity
theorem (Statement~\ref{subsec:Statement-main-relational-parametricity-1})
gives $(t^{A},t^{D})\in c^{\updownarrow T}$ for any relation $c^{:A\leftrightarrow D}$.
Statement~\ref{subsec:Statement-relational-lifting-consistency-PAA}
shows that $c^{\updownarrow T}=(c,c)^{\updownarrow P}$. The definition
$(p,q)=(t^{A},t^{D})$ means that the condition $(p,q)\in(c,c)^{\updownarrow P}$
holds. It follows that $t^{A}$ and $t^{D}$ satisfy the equation:
\[
\forall g^{:A\rightarrow D}\quad:\quad\text{xmap}_{P}(\text{id})(g)(t^{A})=\text{xmap}_{P}(g)(\text{id})(t^{D})\quad.
\]
 The profunctor commutativity law allows us to express \lstinline!xmap!
via separate liftings:
\[
\text{xmap}_{P}(\text{id})(g)(t^{A})=t^{A}\triangleright g^{\uparrow P^{A,\bullet}}\quad,\quad\quad\text{xmap}_{P}(g)(\text{id})(t^{B})=t^{B}\triangleright g^{\downarrow P^{\bullet,D}}\quad.
\]
We obtain $t^{A}\triangleright g^{\uparrow P^{A,\bullet}}=t^{B}\triangleright g^{\downarrow P^{\bullet,D}}$,
which is the wedge law of $t$. $\square$

\begin{comment}
commutativity law does not hold!
\end{comment}


\subsection{Composition of relations}

Like functions, relations can be composed when their types match.
The composition of relations is motivated by the \lstinline!JOIN!
operation in relational databases. Two tables are joined using a shared
column by selecting the rows where the values in the shared column
are equal. Here is an example:
\begin{center}
\hfill{}$T_{1}=$ %
\begin{tabular}{|c|c|}
\hline 
{\footnotesize{}}\lstinline!NAME! & {\footnotesize{}}\lstinline!IDNUM!\tabularnewline
\hline 
\hline 
{\footnotesize{}Chandrasekhar} & {\footnotesize{}4}\tabularnewline
\hline 
{\footnotesize{}Einstein} & {\footnotesize{}2}\tabularnewline
\hline 
{\footnotesize{}Noether} & {\footnotesize{}3}\tabularnewline
\hline 
{\footnotesize{}Poincar\'e} & {\footnotesize{}1}\tabularnewline
\hline 
\end{tabular}\hfill{}$T_{2}=$ %
\begin{tabular}{|c|c|}
\hline 
{\footnotesize{}}\lstinline!IDNUM! & {\footnotesize{}}\lstinline!BORN!\tabularnewline
\hline 
\hline 
{\footnotesize{}1} & {\footnotesize{}1854}\tabularnewline
\hline 
{\footnotesize{}2} & {\footnotesize{}1879}\tabularnewline
\hline 
{\footnotesize{}3} & {\footnotesize{}1882}\tabularnewline
\hline 
{\footnotesize{}4} & {\footnotesize{}1910}\tabularnewline
\hline 
\end{tabular}\hfill{}\hfill{}$T_{3}=$ %
\begin{tabular}{|c|c|}
\hline 
{\footnotesize{}}\lstinline!NAME! & {\footnotesize{}}\lstinline!BORN!\tabularnewline
\hline 
\hline 
{\footnotesize{}Poincar\'e} & {\footnotesize{}1854}\tabularnewline
\hline 
{\footnotesize{}Einstein} & {\footnotesize{}1879}\tabularnewline
\hline 
{\footnotesize{}Noether} & {\footnotesize{}1882}\tabularnewline
\hline 
{\footnotesize{}Chandrasekhar} & {\footnotesize{}1910}\tabularnewline
\hline 
\end{tabular}\hfill{}
\par\end{center}

The third table is the result of joining the first two tables by the
column \lstinline!IDNUM! and then dropping that column. A SQL command
for this operation would be:
\begin{lstlisting}
SELECT NAME, BORN from T1 NATURAL JOIN T2 ON IDNUM;    //IGNORETHIS
\end{lstlisting}
The resulting relation $T_{3}$ between \lstinline!NAME! and \lstinline!BORN!
can be described by the following condition: 
\[
(n^{:\text{String}},b^{:\text{Int}})\in T_{3}\quad\text{ means }\exists i^{:\text{Int}}\text{ such that }(n,i)\in T_{1}\text{ and }(i,b)\in T_{2}\quad.
\]
We require the \emph{existence} of a value $i$ in the \lstinline!IDNUM!
column, but we do not add that column to $T_{3}$.

The \lstinline!JOIN! operation motivates the following definition
of relational composition:

\subsubsection{Definition \label{subsec:Definition-composition-of-relations}\ref{subsec:Definition-composition-of-relations}}

The \textbf{composition} of relations $r^{:A\leftrightarrow B}$ and
$s^{:B\leftrightarrow C}$ is a relation $r\circ s$ defined by:
\[
(r\circ s):A\leftrightarrow C\quad;\quad\quad(a^{:A},c^{:C})\in r\circ s\quad\text{ means }\quad\exists b^{:B}\text{ such that }(a,b)\in r\text{ and }(b,c)\in s\quad.
\]

We cannot implement a Scala function for composing two arbitrary relations.
The code would have to enumerate all possible values of type $B$
in order to check whether a value $b^{:B}$ exists such that $(a,b)\in r$
and $(b,c)\in s$. This is guaranteed to terminate only if the type
$B$ has a small number of distinct values. In practical coding, we
have to use functions rather than relations.

The relational composition is compatible with the composition of functions:

\subsubsection{Statement \label{subsec:Statement-function-composition-relation}\ref{subsec:Statement-function-composition-relation}}

\textbf{(a)} An identity function is converted by $\left<\right>$
into an identity relation.

\textbf{(b)} The composition of functions is converted by $\left<\right>$
into the composition of relations.

\subparagraph{Proof}

\textbf{(a)} The relation $\left<\text{id}\right>$ corresponding
to an identity function $\text{id}^{:A\rightarrow A}$ is defined
by:
\[
(x^{:A},y^{:A})\in\left<\text{id}\right>\text{ means }x\triangleright\text{id}=y\quad\text{or equivalently}:\quad x=y\quad.
\]
This is the same as the definition of the identity relation $\text{id}^{:A\leftrightarrow A}$.
So, we can write $\left<\text{id}\right>=\text{id}^{:A\leftrightarrow A}$.

\textbf{(b)} Consider two functions $f^{:A\rightarrow B}$ and $g^{:B\rightarrow C}$.
The relation $\left<f\bef g\right>$ is defined like this:
\[
(a^{:A},c^{:C})\in\left<f\bef g\right>\text{ means }a\triangleright f\bef g=c\quad\text{or equivalently}:\quad c=g(f(a))\quad.
\]
Now write out the definitions of the relations $\left<f\right>$,
$\left<g\right>$, and $\left<f\right>\circ\left<g\right>$:
\begin{align*}
 & (a^{:A},b^{:B})\in\left<f\right>\text{ means }b=f(a)\quad,\quad\quad(b^{:B},c^{:C})\in\left<g\right>\text{ means }c=g(b)\quad,\\
 & (a^{:A},c^{:C})\in\left<f\right>\circ\left<g\right>\text{ means }\exists b^{:B}\text{ such that }(a,b)\in\left<f\right>\text{ and }(b,c)\in\left<g\right>\quad,\\
 & \quad\quad\text{or equivalently}:\quad\exists b^{:B}\text{ such that }b=f(a)\text{ and }c=g(b)\quad.
\end{align*}
Since the function $f$ is given, $b=f(a)$ always exists. Then we
get:
\[
(a^{:A},c^{:C})\in\left<f\right>\circ\left<g\right>\text{ means }c=g(f(a))\quad.
\]
This is the same as the definition of the relation $\left<f\bef g\right>$.
$\square$

For relations, there is no difference between forward and backward
composition (i.e., relations are \textsf{``}undirected\textsf{''}). Also, relation
types $A\leftrightarrow B$ and $B\leftrightarrow A$ are equivalent
via the reversal operation (\lstinline!rev!).

\subsubsection{Statement \label{subsec:Statement-reverse-relation-properties}\ref{subsec:Statement-reverse-relation-properties}}

\textbf{(a)} The relation $\text{rev}\left(r\right)$ is \emph{not}
an inverse of $r^{:A\leftrightarrow B}$ in the sense of composition:
\[
r\circ\text{rev}\,(r)\neq\text{id}^{:A\leftrightarrow A}\quad.
\]

\textbf{(b)} For any relations $r^{:B\leftrightarrow A}$ and $s^{:C\leftrightarrow B}$:
\[
\text{rev}\left(r\right)\circ\text{rev}\left(s\right)=\text{rev}\left(s\circ r\right)\quad.
\]


\subparagraph{Proof}

\textbf{(a)} Two values $x^{:A}$, $y^{:A}$ are in the relation $r\circ\text{rev}\,(r)$
if:
\[
\exists b^{:B}:(x,b)\in r\text{ and }(y,b)\in r\quad.
\]
Since the relation $r$ may be many-to-many, it does not follow that
$x=y$. So, the identity relation is \emph{not} a consequence of $r\circ\text{rev}\,(r)$.
It also does not follow that for a given $x^{:A}$ there will exist
a $b^{:B}$ such that $(x,b)\in r$. So, the pair $\left(x,x\right)$
is \emph{not} necessarily in the relation $r\circ\text{rev}\,(r)$.
In other words, the relation $r\circ\text{rev}\,(r)$ is not a consequence
of the identity relation.

\textbf{(b)} Use the definition of relational composition:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (x,y)\in\text{rev}\left(r\right)\circ\text{rev}\left(s\right)\text{ means }\exists z.\,(x,z)\in\text{rev}\left(r\right)\text{ and }(z,y)\in\text{rev}\left(s\right)\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & (x,y)\in\text{rev}\left(r\right)\circ\text{rev}\left(s\right)\text{ means }\exists z.\,(z,x)\in r\text{ and }(y,z)\in s\quad.\\
{\color{greenunder}\text{right-hand side}:}\quad & (x,y)\in\text{rev}(s\circ r)\text{ means }(y,x)\in s\circ r\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & (x,y)\in\text{rev}(s\circ r)\text{ means }\exists z.\,(y,z)\in s\text{ and }(z,x)\in r\quad.
\end{align*}
The two conditions are now the same.

\subsubsection{Statement \label{subsec:Statement-identity-composition-laws-of-relations}\ref{subsec:Statement-identity-composition-laws-of-relations}}

The composition of relations satisfies the identity\index{identity laws!of relational composition}
and \index{associativity law!of relational composition}associativity
laws:
\[
r^{:A\leftrightarrow B}\circ\text{id}^{:B\leftrightarrow B}=r\quad,\quad\quad\text{id}^{:A\leftrightarrow A}\circ r^{:A\leftrightarrow B}=r\quad,\quad(r^{:A\leftrightarrow B}\circ s^{:B\leftrightarrow C})\circ t^{:C\leftrightarrow D}=r\circ(s\circ t)\quad.
\]
The associativity law allows us to write relational composition without
parentheses, e.g., $s\circ p\circ q\circ r$.

\subparagraph{Proof}

To verify the right identity law, assume an arbitrary relation $r^{:A\leftrightarrow B}$
and write the condition for some values $x^{:A}$ and $y^{:B}$ to
be in the relation $r\circ\text{id}$:
\[
(x,y)\in r\circ\text{id}\quad\text{ means }\quad\exists b^{:B}\text{ such that }(x,b)\in r\text{ and }(b,y)\in\text{id}\quad.
\]
By definition of the identity relation ($\text{id}^{:B\leftrightarrow B}$),
the condition $(b,y)\in\text{id}$ means $b=y$. So, the condition
$(x,y)\in r\circ\text{id}$ is the same as $(x,y)\in r$. This means
$r\circ\text{id}=r$.

Similarly, we verify the left identity law:
\[
(x,y)\in\text{id}\circ r\quad\text{ means }\quad\exists a^{:A}\text{ such that }(x,a)\in\text{id}\text{ and }(a,y)\in r\quad.
\]
By definition of $\text{id}^{:A\leftrightarrow A}$, we have $x=a$,
and so the conditions $(x,y)\in\text{id}\circ r$ and $(x,y)\in r$
are equivalent. This means $\text{id}\circ r=r$.

To verify the associativity law, write the definitions of relational
composition:
\begin{align*}
{\color{greenunder}\text{right-hand side}:}\quad & (x,y)\in r\circ(s\circ t)\\
{\color{greenunder}\text{is equivalent to}:}\quad & \exists b^{:B}\text{ such that }(x,b)\in r\text{ and }(b,y)\in s\circ t\\
{\color{greenunder}\text{is equivalent to}:}\quad & \exists b^{:B}\text{ such that }(x,b)\in r\text{ and }\exists c^{:C}\text{ such that }(b,c)\in s\text{ and }(c,y)\in t\quad.
\end{align*}
The right-hand side of the associativity law is rewritten similarly:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (x,y)\in(r\circ s)\circ t\\
{\color{greenunder}\text{is equivalent to}:}\quad & \exists c^{:C}\text{ such that }(x,c)\in r\circ s\text{ and }(c,y)\in t\\
{\color{greenunder}\text{is equivalent to}:}\quad & \exists c^{:C}\text{ such that }\exists b^{:B}\text{ such that }(x,b)\in r\text{ and }(b,c)\in s\text{ and }(c,y)\in t\quad.
\end{align*}
To make the two sides equal, place the quantifiers $\exists b^{:B}\exists c^{:C}$
together in front of the expression.

{*}{*}{*}

The pair mapper does not work if we replace relations $r^{:A\leftrightarrow C}$
and $s^{:B\leftrightarrow D}$ by \emph{functions} of types $A\rightarrow C$
and $B\rightarrow D$ because the types do not match: we cannot create
a function of type $\left(A\rightarrow B\right)\rightarrow C\rightarrow D$
given $p^{:A\rightarrow C}$ and $q^{:B\rightarrow D}$. However,
we can use the relational pair mapper if we replace the functions
$p$, $q$ by relations $\left<p\right>$ and $\left<q\right>$. The
result is, in general, a many-to-many relation $\left<p\right>\ogreaterthan\left<q\right>$
that cannot be expressed as a function but has a pullback form:

\subsubsection{Example \label{subsec:Example-pair-mapper-of-functions}\ref{subsec:Example-pair-mapper-of-functions}}

Consider two functions $p^{:A\rightarrow C}$, $q^{:B\rightarrow D}$
and the corresponding relations $\left<p\right>$, $\left<q\right>$.

\textbf{(a)} Show that the relation $\left<p\right>\ogreaterthan\left<q\right>$
has a pullback form.

\textbf{(b)} Show that the relation $\big(\text{rev}\langle p\rangle\big)\ogreaterthan\left<q\right>$
is equivalent to a function.

\textbf{(c)} Show that the relation $\left<p\right>\ogreaterthan\big(\text{rev}\langle q\rangle\big)$
is equivalent to a function.

\subparagraph{Solution}

\textbf{(a)} Definition~\ref{subsec:Definition-pair-mapper-of-relations}
gives:
\[
(f^{:A\rightarrow B},g^{:C\rightarrow D})\in\left<p\right>\ogreaterthan\left<q\right>\quad\text{means}:\quad\text{if }(x^{:A},y^{:C})\in\left<p\right>\quad\text{then}\quad(f(x),g(y))\in\left<q\right>\quad.
\]
Writing out the definitions of $\left<p\right>$ and $\left<q\right>$,
we get:
\[
(f,g)\in\left<p\right>\ogreaterthan\left<q\right>\quad\text{means}:\quad\text{if }x\triangleright p=y\quad\text{then}\quad x\triangleright f\triangleright q=y\triangleright g\quad.
\]
The last condition can be simplified to $x\triangleright f\triangleright q=x\triangleright p\triangleright g$
and to a point-free formula:
\[
(f,g)\in\left<p\right>\ogreaterthan\left<q\right>\quad\text{means}:\quad f\bef q=p\bef g\quad.
\]
This is a relation in a pullback form:
\[
\left<p\right>\ogreaterthan\left<q\right>=\text{pull}\,(f^{:A\rightarrow B}\rightarrow f\bef q,\,g^{:C\rightarrow D}\rightarrow p\bef g)\quad.
\]

\textbf{(b)} Definition~\ref{subsec:Definition-pair-mapper-of-relations}
gives:
\[
(f^{:C\rightarrow B},g^{:A\rightarrow D})\in\big(\text{rev}\left<p\right>\big)\ogreaterthan\left<q\right>\quad\text{means}:\quad\text{if }(y^{:C},x^{:A})\in\text{rev}\left<p\right>\quad\text{then}\quad(f(y),g(x))\in\left<q\right>\quad.
\]
Writing out the definitions of $\text{rev}\left<p\right>$ and $\left<q\right>$,
we get:
\[
(f,g)\in\big(\text{rev}\left<p\right>\big)\ogreaterthan\left<q\right>\quad\text{means}:\quad\text{if }x\triangleright p=y\quad\text{then}\quad y\triangleright f\triangleright q=x\triangleright g\quad.
\]
Substituting $y=x\triangleright p$ into the last condition, we get
\[
x\triangleright p\triangleright f\triangleright q=x\triangleright g\quad,\quad\quad\text{or equivalently}:\quad p\bef f\bef q=g\quad.
\]
So, the relation between $f$ and $g$ has the form of a function:
\[
(f,g)\in\big(\text{rev}\left<p\right>\big)\ogreaterthan\left<q\right>\quad\text{means}\quad g=\phi(f)\quad,\quad\quad\phi\triangleq f^{:C\rightarrow B}\rightarrow p\bef f\bef q\quad.
\]
We can write the conclusion as: 
\[
\big(\text{rev}\left<p\right>\big)\ogreaterthan\left<q\right>=\langle f^{:C\rightarrow B}\rightarrow p\bef f\bef q\rangle\quad.
\]

\textbf{(c)} Using Statement~\ref{subsec:Statement-pair-product-distributive-law-rev}(c)
proved below, we have:
\[
\text{rev}\big((\text{rev}\left<p\right>)\ogreaterthan\left<q\right>\big)=p\ogreaterthan(\text{rev}\left<q\right>)\quad.
\]
So, we just need to reverse the relation found at the end of part
\textbf{(b)} above: 
\[
\left<p\right>\ogreaterthan\big(\text{rev}\left<q\right>\big)=\text{rev}\langle f^{:C\rightarrow B}\rightarrow p\bef f\bef q\rangle\quad.
\]
 $\square$

We have defined several operations on relations: the function lifting
$\left<f\right>$, the reversal ($\text{rev}$), and four binary operations
($\circ$, $\boxtimes$, $\boxplus$, $\ogreaterthan$). To illustrate
the usefulness of these operations, we will prove a number of their
properties. All these properties have the form of distributive laws
for two of these operations. We already proved one of these distributive
laws in Statement~\ref{subsec:Statement-reverse-relation-properties}(c).
We will now prove other laws that will be used below.

\subsubsection{Statement \label{subsec:Statement-pair-product-distributive-law}\ref{subsec:Statement-pair-product-distributive-law}}

For any functions $f^{:A\rightarrow B}$ and $g^{:C\rightarrow D}$,
the following distributive laws hold:

\textbf{(a)} $\left<f\boxtimes g\right>=\left<f\right>\boxtimes\left<g\right>$,
both relations having the type $A\times C\leftrightarrow B\times D$.

\textbf{(b)} $\left<f\boxplus g\right>=\left<f\right>\boxplus\left<g\right>$,
both relations having the type $A+C\leftrightarrow B+D$.

\subparagraph{Proof}

Consider arbitrary values $a^{:A}$, $b^{:B}$, $c^{:C}$, and $d^{:D}$.

\textbf{(a)} By definitions of the pair product for functions and
for relations, we have:
\begin{align*}
 & (a\times c,b\times d)\in\left<f\boxtimes g\right>\text{ means }b\times d=(f\boxtimes g)(a\times c)=f(a)\times g(c)\quad,\\
 & (a\times c,b\times d)\in\left<f\right>\boxtimes\left<g\right>\text{ means }(a,b)\in\left<f\right>\text{ and }(c,d)\in\left<g\right>\quad\text{or equivalently}:\,b=f(a)\text{ and }d=g(c)\quad.
\end{align*}
The two conditions are now the same.

\textbf{(b)} By definitions of the pair co-product for functions and
for relations, we have:
\begin{align*}
 & (a+\bbnum 0,b+\bbnum 0)\in\left<f\boxplus g\right>\text{ means }b=f(a)\quad,\quad\quad(\bbnum 0+c,\bbnum 0+d)\in\left<f\boxplus g\right>\text{ means }d=g(c)\quad,\\
 & (a+\bbnum 0,b+\bbnum 0)\in\left<f\right>\boxplus\left<g\right>\text{ means }(a,b)\in\left<f\right>\quad,\quad\quad(\bbnum 0+c,\bbnum 0+d)\in\left<f\boxplus g\right>\text{ means }(c,d)\in\left<g\right>\quad,
\end{align*}
while all other combinations of disjunctive values (say, $a+\bbnum 0$
with $\bbnum 0+d$) will not belong to these relations. We find that
the two conditions are the same.

\subsubsection{Statement \label{subsec:Statement-pair-product-distributive-law-rev}\ref{subsec:Statement-pair-product-distributive-law-rev}}

For any relations $r^{:A\leftrightarrow B}$ and $s^{:C\leftrightarrow D}$,
the following distributive laws hold:

\textbf{(a)} $\text{rev}\left(r\boxtimes s\right)=\text{rev}\left(r\right)\boxtimes\text{rev}\left(s\right)$,
both relations having the type $B\times D\leftrightarrow A\times C$.

\textbf{(b)} $\text{rev}\left(r\boxplus s\right)=\text{rev}\left(r\right)\boxplus\text{rev}\left(s\right)$,
both relations having the type $B+D\leftrightarrow A+C$.

\textbf{(c)} $\text{rev}\left(r\ogreaterthan s\right)=\text{rev}\left(r\right)\ogreaterthan\text{rev}\left(s\right)$,
both relations having the type $B\rightarrow D\leftrightarrow A\rightarrow C$.

The analogous distributive law for \lstinline!rev! and $\circ$ was
already proved in Statement~\ref{subsec:Statement-reverse-relation-properties}(c).

\subparagraph{Proof}

Consider arbitrary values $a^{:A}$, $b^{:B}$, $c^{:C}$, $d^{:D}$,
$f^{:B\rightarrow D}$, and $g^{:A\rightarrow C}$.

\textbf{(a)} Write the definitions of the relations in the two sides
of the law:
\begin{align*}
 & (b\times d,a\times c)\in\text{rev}\left(r\boxtimes s\right)\text{ means }(a\times c,b\times d)\in r\boxtimes s\quad\text{or equivalently}:\quad(a,b)\in r\text{ and }(c,d)\in s\quad,\\
 & (b\times d,a\times c)\in\text{rev}\left(r\right)\boxtimes\text{rev}\left(s\right)\text{ means }(b,a)\in\text{rev}\left(r\right)\text{ and }(d,c)\in\text{rev}\left(s\right)\quad.
\end{align*}
The two conditions are now equivalent.

\textbf{(b)} Write the definitions of the relations in the two sides
of the law:
\begin{align*}
 & (b+\bbnum 0,a+\bbnum 0)\in\text{rev}\left(r\boxplus s\right)\text{ means }(a+\bbnum 0,b+\bbnum 0)\in r\boxplus s\quad\text{or}:\quad(a,b)\in r\quad,\\
 & (b+\bbnum 0,a+\bbnum 0)\in\text{rev}\left(r\right)\boxplus\text{rev}\left(s\right)\text{ means }(b,a)\in\text{rev}\left(r\right)\quad\text{or}:\quad(a,b)\in r\quad;\\
 & (\bbnum 0+d,\bbnum 0+c)\in\text{rev}\left(r\boxplus s\right)\text{ means }(\bbnum 0+c,\bbnum 0+d)\in r\boxplus s\quad\text{or}:\quad(c,d)\in s\quad,\\
 & (\bbnum 0+d,\bbnum 0+c)\in\text{rev}\left(r\right)\boxplus\text{rev}\left(s\right)\text{ means }(d,c)\in\text{rev}\left(s\right)\quad\text{or}:\quad(c,d)\in s\quad.
\end{align*}
The two conditions are now equivalent.

\textbf{(c)} Write the definitions of the relations in the two sides
of the law:
\begin{align*}
 & (f^{:B\rightarrow D},g^{:A\rightarrow C})\in\text{rev}\left(r\ogreaterthan s\right)\text{ means if }(x^{:A},y^{:B})\in r\text{ then }(g(x),f(y))\in s\quad,\\
 & (f,g)\in\text{rev}\left(r\right)\ogreaterthan\text{rev}\left(s\right)\text{ means if }(y^{:B},x^{:A})\in\text{rev}\left(r\right)\text{ then }(f(y),g(x))\in\text{rev}\left(s\right)\quad.
\end{align*}
The two conditions are now equivalent.

\subsubsection{Statement \label{subsec:Statement-relational-composition-distributive-law}\ref{subsec:Statement-relational-composition-distributive-law}}

For any relations $r^{:A_{1}\leftrightarrow B_{1}}$, $s^{:B_{1}\leftrightarrow C_{1}}$,
$u^{:A_{2}\leftrightarrow B_{2}}$, and $v^{:B_{2}\leftrightarrow C_{2}}$,
the following distributive laws hold:

\textbf{(a)} $(r\circ s)\boxtimes(u\circ v)=(r\boxtimes u)\circ(s\boxtimes v)$,
both relations having the type $A_{1}\times A_{2}\leftrightarrow C_{1}\times C_{2}$.

\textbf{(b)} $(r\circ s)\boxplus(u\circ v)=(r\boxplus u)\circ(s\boxplus v)$,
both relations having the type $A_{1}+A_{2}\leftrightarrow C_{1}+C_{2}$.

\textbf{(c)} The relation $(r\circ s)\ogreaterthan(u\circ v)$ is
\emph{not} equivalent to $(r\ogreaterthan u)\circ(s\ogreaterthan v)$.
For any values $p^{:A_{1}\rightarrow A_{2}}$ and $q^{:C_{1}\rightarrow C_{2}}$,
if $\left(p,q\right)\in(r\ogreaterthan u)\circ(s\ogreaterthan v)$
then $\left(p,q\right)\in(r\circ s)\ogreaterthan(u\circ v)$, but
the converse does not hold.

{*}{*}{*} This is used in D.3.4.4 \ref{subsec:Statement-lifting-relation-commutativity-law}
and D.3.4.7 \ref{subsec:Statement-lifting-relation-laws}, so we need
to revise those proofs or to prove a weaker version of the distributive
law here. It\textsf{'}s weird that the distributive law does not hold but we
are able to obtain other laws that use it!

\subparagraph{Proof}

Consider arbitrary values $a_{1}^{:A_{1}}$, $a_{2}^{:A_{2}}$, $c_{1}^{:C_{1}}$,
and $c_{2}^{:C_{2}}$.

\textbf{(a)} Write the conditions for $(a_{1}\times a_{2},c_{1}\times c_{2})$
to be in the relations $(r\circ s)\boxtimes(u\circ v)$ and $(r\boxtimes u)\circ(s\boxtimes v)$:
\begin{align*}
 & (a_{1}\times a_{2},c_{1}\times c_{2})\in(r\circ s)\boxtimes(u\circ v)\text{ means }(a_{1},c_{1})\in\gunderline{r\circ s}\text{ and }(a_{2},c_{2})\in\gunderline{u\circ v}\quad,\\
{\color{greenunder}\text{rewrite}:}\quad & \quad\exists b_{1}^{:B_{1}},b_{2}^{:B_{2}}\text{ such that}:\quad(a_{1},b_{1})\in r\quad,\quad(b_{1},c_{1})\in s\quad,\quad(a_{2},b_{2})\in u\quad,\quad(b_{2},c_{2})\in v\quad;\\
 & (a_{1}\times a_{2},c_{1}\times c_{2})\in(r\boxtimes u)\circ(s\boxtimes v)\text{ means }\exists(b_{1}\times b_{2})^{:B_{1}\times B_{2}}\text{ such that}:\\
 & \quad\quad(a_{1}\times a_{2},b_{1}\times b_{2})\in\gunderline{r\boxtimes u}\text{ and }(b_{1}\times b_{2},c_{1}\times c_{2})\in\gunderline{s\boxtimes v}\quad,\\
{\color{greenunder}\text{rewrite}:}\quad & \quad(a_{1},b_{1})\in r\quad,\quad(a_{2},b_{2})\in u\quad,\quad(b_{1},c_{1})\in s\quad,\quad(b_{2},c_{2})\in v\quad.
\end{align*}
We see that the two conditions are equivalent.

\textbf{(b)} Values $x^{:A_{1}+A_{2}}$ and $z^{:C_{1}+C_{2}}$ belong
to the disjunctive relation $(r\circ s)\boxplus(u\circ v)$ if:
\begin{align*}
{\color{greenunder}\text{either}:}\quad & x=a_{1}+\bbnum 0\quad,\quad z=c_{1}+\bbnum 0\quad,\quad(a_{1},c_{1})\in r\circ s\quad;\\
{\color{greenunder}\text{or}:}\quad & x=\bbnum 0+a_{2}\quad,\quad z=\bbnum 0+c_{2}\quad,\quad(a_{2},c_{2})\in u\circ v\quad.
\end{align*}
Writing out the definitions of $r\circ s$ and $u\circ v$, we get
the equivalent conditions:
\begin{align*}
{\color{greenunder}\text{either}:}\quad & x=a_{1}+\bbnum 0\quad,\quad z=c_{1}+\bbnum 0\quad,\quad\exists b_{1}^{:B_{1}}\text{ such that }:\quad(a_{1},b_{1})\in r\quad,\quad(b_{1},c_{1})\in s\quad;\\
{\color{greenunder}\text{or}:}\quad & x=\bbnum 0+a_{2}\quad,\quad z=\bbnum 0+c_{2}\quad,\quad\exists b_{2}^{:B_{2}}\text{ such that }:\quad(a_{2},b_{2})\in u\quad,\quad(b_{2},c_{2})\in v\quad.
\end{align*}

The same values $x$ and $z$ belong to the relation $(r\boxplus u)\circ(s\boxplus v)$
if:
\[
\exists y^{:B_{1}+B_{2}}\text{ such that }:\quad(x,y)\in r\boxplus u\quad,\quad(y,z)\in s\boxplus v\quad.
\]
We obtain two cases where all of $x$, $y$, $z$ are either on the
left or on the right of their disjunctive types:
\begin{align*}
{\color{greenunder}\text{either}:}\quad & x=a_{1}+\bbnum 0\quad,\quad y=b_{1}^{:B_{1}}+\bbnum 0\quad,\quad z=c_{1}+\bbnum 0\quad,\quad(a_{1},b_{1})\in r\quad,\quad(b_{1},c_{1})\in s\quad;\\
{\color{greenunder}\text{or}:}\quad & x=\bbnum 0+a_{2}\quad,\quad y=\bbnum 0+b_{2}^{:B_{2}}\quad,\quad z=\bbnum 0+c_{2}\quad,\quad(a_{2},b_{2})\in u\quad,\quad(b_{2},c_{2})\in v\quad.
\end{align*}
These conditions are the same as the conditions for $(x,z)\in(r\circ s)\boxplus(u\circ v)$. 

\textbf{(c)} First consider the relation $(r\ogreaterthan u)\circ(s\ogreaterthan v)$
and look at the condition $(p,q)\in(r\ogreaterthan u)\circ(s\ogreaterthan v)$:
\begin{align*}
 & (p,q)\in(r\ogreaterthan u)\circ(s\ogreaterthan v)\quad\text{ means }\quad\exists f^{:B_{1}\rightarrow B_{2}}\text{ with }(p,f)\in\gunderline{r\ogreaterthan u}\text{ and }(f,q)\in\gunderline{s\ogreaterthan v}\quad;\\
{\color{greenunder}\text{rewrite}:}\quad & \quad\exists f^{:B_{1}\rightarrow B_{2}}\quad\text{ such that }\quad\forall\big(a^{:A_{1}},b^{:B_{1}}\big)\quad:\quad\text{ if }(a,b)\in r\text{ then }(p(a),f(b))\in u\\
 & \quad\quad\text{and }\quad\forall\big(b^{:B_{1}},c^{:C_{1}}\big)\quad:\quad\text{ if }(b,c)\in s\text{ then }(f(b),q(c))\in v\quad.
\end{align*}
In the last formula, the conditions for arbitrary values $a$, $b$,
$c$ may be brought together:
\begin{align}
 & (p,q)\in(r\ogreaterthan u)\circ(s\ogreaterthan v)\quad\text{ means }\quad\exists f^{:B_{1}\rightarrow B_{2}}\quad\text{ such that }\forall\big(a^{:A_{1}},b^{:B_{1}},c^{:C_{1}}\big)\quad:\quad\nonumber \\
 & \quad\text{ if }(a,b)\in r\text{ and }(b,c)\in s\text{ then }(p(a),f(b))\in u\text{ and }(f(b),q(c))\in v\quad.\label{eq:composition-of-pair-mappers-derivation1}
\end{align}

Turning now to the relation $(r\circ s)\ogreaterthan(u\circ v)$,
we write the condition $\left(p,q\right)\in(r\circ s)\ogreaterthan(u\circ v)$
as:
\begin{align*}
 & \left(p,q\right)\in(r\circ s)\ogreaterthan(u\circ v)\quad\text{ means }\quad\forall\big(a^{:A_{1}},c^{:C_{1}}\big):\,\text{if }(a,c)\in r\circ s\text{ then }(p(a),q(c))\in u\circ v\quad;\\
{\color{greenunder}\text{rewrite}:}\quad & \quad\text{if }\exists b_{1}^{:B_{1}}\text{ with }(a,b_{1})\in r\text{ and }(b_{1},c)\in s\text{ then }\exists b_{2}^{:B_{2}}\text{ with }(p(a),b_{2})\in u\text{ and }(b_{2},q(c))\in v\quad.
\end{align*}
The last condition requires the existence of a value $b_{2}^{:B_{2}}$
given the existence of a value $b_{1}^{:B_{1}}$. This differs from
Eq.~(\ref{eq:composition-of-pair-mappers-derivation1}) where the
last condition requires the existence of a \emph{function} of type
$B_{1}\rightarrow B_{2}$. How can we make these two conditions more
similar? The condition for $\left(p,q\right)\in(r\circ s)\ogreaterthan(u\circ v)$
says that we need to find a value $b_{2}$ for any given value of
$b_{1}$ satisfying $(a,b_{1})\in r$ and $(b_{1},c)\in s$. This
is equivalent to having a function $f:B_{1}\rightarrow B_{2}$ such
that $b_{2}=f(b_{1})$. (We only need to have $b_{2}=f(b_{1})$ for
values $b_{1}$ satisfying the conditions $(a,b_{1})\in r$ and $(b_{1},c)\in s$.
The value $f(b_{1})$ may be arbitrary for other $b_{1}$.) So, renaming
$b_{1}$ to $b$, we may rewrite the last formula as:
\begin{align}
 & (p,q)\in(r\circ s)\ogreaterthan(u\circ v)\quad\text{ means }\quad\forall\big(a^{:A_{1}},c^{:C_{1}}\big)\quad\exists f^{:B_{1}\rightarrow B_{2}}\quad\text{ such that }\quad\forall b^{:B_{1}}\quad:\nonumber \\
 & \quad\text{if }(a,b)\in r\text{ and }(b,c)\in s\text{ then }(x(a),f(b))\in u\text{ and }(f(b),z(c))\in v\quad.\label{eq:composition-of-pair-mappers-derivation2}
\end{align}

The difference between Eqs.~(\ref{eq:composition-of-pair-mappers-derivation1})
and~(\ref{eq:composition-of-pair-mappers-derivation2}) is now only
in the order of quantifiers: $\exists f\forall(a,b,c)$ instead of
$\forall(a,c)\exists f\forall b$. The conditions after the quantifier
are the same in both formulas.

Let us show that Eq.~~(\ref{eq:composition-of-pair-mappers-derivation2})
follows from Eq.~(\ref{eq:composition-of-pair-mappers-derivation1}).
Suppose we have $\exists f.\,\forall(a,b,c).\,S(f,a,b,c)$ where $S(f,a,b,c)$
is some statement about $f$, $a$, $b$, $c$. This gives is a certain
function $f$. We will make the property $\forall(a,c).\,\exists f.\,\forall b.\,S(f,a,b,c)$
also hold if we use the same $f$ (regardless of $a$ and $c$).

The converse is not always true: Eq.~(\ref{eq:composition-of-pair-mappers-derivation1})
does not follow from Eq.~~(\ref{eq:composition-of-pair-mappers-derivation2}).
Accordingly, the relations $(r\circ s)\ogreaterthan(u\circ v)$ and
$(r\ogreaterthan u)\circ(s\ogreaterthan v)$ are in general \emph{not}
equivalent. 

As an explicit example, consider a constant-valued function $h^{:A\rightarrow B}\triangleq\_^{:A}\rightarrow b_{0}$
where $b_{0}$ is a fixed value of type $B$, and set $r=v=\left<h\right>$.
For simplicity, choose $s$ and $u$ to be identity relations ($s\triangleq\text{id}^{:B\leftrightarrow B}$
and $u\triangleq\text{id}^{:A\leftrightarrow A}$). Then compare the
following two relations:
\begin{align*}
 & t_{1}:A\rightarrow A\leftrightarrow B\rightarrow B\quad,\quad\quad t_{1}\triangleq(r\circ s)\ogreaterthan(u\circ v)=\big(\left<h\right>\circ\text{id}^{:B\leftrightarrow B}\big)\ogreaterthan\big(\text{id}^{:A\leftrightarrow A}\circ\left<h\right>\big)\quad,\\
 & t_{2}:A\rightarrow A\leftrightarrow B\rightarrow B\quad,\quad\quad t_{2}\triangleq(r\ogreaterthan u)\circ(s\ogreaterthan v)=\big(\left<h\right>\ogreaterthan\text{id}^{:A\leftrightarrow A}\big)\circ\big(\text{id}^{:B\leftrightarrow A}\ogreaterthan\left<h\right>\big)\quad.
\end{align*}
We can simplify $t_{1}$ using the identity law from Statement~\ref{subsec:Statement-identity-composition-laws-of-relations}:
\[
t_{1}=\left<h\right>\ogreaterthan\left<h\right>\quad.
\]
 We need to write out the conditions for any $p^{:A\rightarrow A}$
and $q^{:B\rightarrow B}$ to belong to $t_{1}$ or to $t_{2}$. Begin
with $t_{1}$:
\begin{align*}
 & (p,q)\in t_{1}\quad\text{ means }\quad\forall a^{:A},b^{:B}.\,\text{ if }(a,b)\in\left<h\right>\text{ then }(p(a),q(b))\in\left<h\right>\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & \quad\text{ if }h(a)=b\text{ then }h(p(a))=q(b)\quad.
\end{align*}
Since $h(x)=b_{0}$ for any $x$, we can simplify the last condition
to $q(b_{0})=b_{0}$. There are no restrictions on $p$. So, any function
$p^{:A\rightarrow A}$ and any $q^{:B\rightarrow B}$ belong to $t_{1}$
as long as $q(b_{0})=b_{0}$.

Turning now to the condition for $\left(p,q\right)\in t_{2}$, we
write:
\[
(p,q)\in t_{2}\quad\text{ means }\quad\exists f^{:B\rightarrow A}\text{ such that }(p,f)\in\left<h\right>\ogreaterthan\text{id}^{:A\leftrightarrow A}\text{ and }(f,q)\in\text{id}^{:B\leftrightarrow B}\ogreaterthan\left<h\right>\quad.
\]
Write the definitions of $\left<h\right>\ogreaterthan\text{id}$ and
$\text{id}\ogreaterthan\left<h\right>$:
\begin{align*}
 & (p,f)\in\left<h\right>\ogreaterthan\text{id}^{:A\leftrightarrow A}\quad\text{ means }\quad\forall a^{:A},b^{:B}.\,\text{ if }(a,b)\in\left<h\right>\text{ then }p(a)=f(b)\quad,\\
 & (f,q)\in\text{id}^{:B\leftrightarrow B}\ogreaterthan\left<h\right>\quad\text{ means }\quad\forall b_{1}^{:B},b_{2}^{:B}.\,\text{ if }b_{1}=b_{2}\text{ then }(f(b_{1}),q(b_{2}))\in\left<h\right>\quad.
\end{align*}
Simplifying these conditions, we get:
\begin{align*}
 & (p,f)\in\left<h\right>\ogreaterthan\text{id}^{:A\leftrightarrow A}\quad\text{ means }\quad\forall a^{:A}.\,p(a)=f(h(a))=f(b_{0})\quad,\quad\quad\text{or}:\quad p=\_^{:A}\rightarrow f(b_{0})\quad,\\
 & (f,q)\in\text{id}^{:B\leftrightarrow B}\ogreaterthan\left<h\right>\quad\text{ means }\quad\forall b^{:B}.\,h(f(b))=q(b)\quad,\quad\quad\text{or}:\quad q=\_^{:B}\rightarrow b_{0}\quad.
\end{align*}
The only use of $f$ is for computing $f(b_{0})$, which is a value
of type $A$ that we can denote by $a_{0}$. So, the condition for
$\left(p,q\right)\in t_{2}$ becomes:
\[
(p,q)\in t_{2}\quad\text{ means }\quad\exists a_{0}^{:A}\text{ such that }p=\_^{:A}\rightarrow a_{0}\text{ and }q=\_^{:B}\rightarrow b_{0}\quad.
\]
Only constant-valued functions $p$, $q$ belong to $t_{2}$, and
only if $q=h$. We find that $t_{2}$ is stronger than $t_{1}$.

\subsubsection{Statement \label{subsec:Statement-relational-identity-distributive-laws}\ref{subsec:Statement-relational-identity-distributive-laws}}

The identity relation ($\text{id}^{:Z\leftrightarrow Z}$) satisfies:
\[
\text{id}\circ\text{id}=\text{id}\quad,\quad\text{id}\boxtimes\text{id}=\text{id}^{:Z\times Z\leftrightarrow Z\times Z}\quad,\quad\quad\text{id}\boxplus\text{id}=\text{id}^{:Z+Z\leftrightarrow Z+Z}\quad,\quad\quad\text{id}\ogreaterthan\text{id}=\text{id}^{:Z\rightarrow Z\leftrightarrow Z\rightarrow Z}\quad.
\]


\subparagraph{Proof}

The property $\text{id}\circ\text{id}=\text{id}$ follows from the
identity law in Statement~\ref{subsec:Statement-identity-composition-laws-of-relations}.

For other properties, assume arbitrary values $a^{:Z}$, $b^{:Z}$,
$c^{:Z}$, and $d^{:Z}$. To verify $\text{id}\boxtimes\text{id}=\text{id}$,
write:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (a\times b,c\times d)\in\text{id}\boxtimes\text{id}\quad\text{ means }\quad(a,c)\in\text{id}\text{ and }(b,d)\in\text{id}\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & (a\times b,c\times d)\in\text{id}\boxtimes\text{id}\quad\text{ means }\quad a=c\text{ and }b=d\quad;\\
{\color{greenunder}\text{right-hand side}:}\quad & (a\times b,c\times d)\in\text{id}^{:Z\times Z\leftrightarrow Z\times Z}\quad\text{ means }\quad a\times b=c\times d\quad.
\end{align*}
The conditions ($a\times b=c\times d$) and ($a=c$ and $b=d$) are
the same.

To verify $\text{id}\boxplus\text{id}=\text{id}$, write the two cases
where values are in the relation $\text{id}\boxplus\text{id}$:
\begin{align*}
 & (a+\bbnum 0,b+\bbnum 0)\in\text{id}\boxplus\text{id}\quad\text{ when }\quad(a,b)\in\text{id}\quad\text{or equivalently}:\quad a=b\quad;\\
 & (a+\bbnum 0,b+\bbnum 0)\in\text{id}^{:Z+Z\leftrightarrow Z+Z}\quad\text{ when }\quad a+\bbnum 0=b+\bbnum 0\quad\text{ or equivalently }:\quad a=b\quad;\\
 & (\bbnum 0+c,\bbnum 0+d)\in\text{id}\boxplus\text{id}\quad\text{ when }\quad(c,d)\in\text{id}\quad\text{or equivalently}:\quad c=d\quad;\\
 & (\bbnum 0+c,\bbnum 0+d)\in\text{id}^{:Z+Z\leftrightarrow Z+Z}\quad\text{ when }\bbnum 0+c=\bbnum 0+d\quad\text{ or equivalently }:\quad c=d\quad.
\end{align*}
In both cases, the conditions are the same.

To verify $\text{id}\ogreaterthan\text{id}=\text{id}$, assume arbitrary
functions $f^{:Z\rightarrow Z}$ and $g^{:Z\rightarrow Z}$:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (f,g)\in\text{id}\ogreaterthan\text{id}\quad\text{ means }\quad\text{whenever }(a,b)\in\text{id}\text{ then }(f(a),g(b))\in\text{id}\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & \quad(f,g)\in\text{id}\ogreaterthan\text{id}\quad\text{ means }\quad\text{whenever }a=b\text{ then }f(a)=g(b)\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & \quad(f,g)\in\text{id}\ogreaterthan\text{id}\quad\text{ means }\quad\text{for all }a^{:Z}\quad,\quad f(a)=g(a)\quad.\\
{\color{greenunder}\text{right-hand side}:}\quad & (f,g)\in\text{id}^{:Z\rightarrow Z\leftrightarrow Z\rightarrow Z}\quad\text{ means }f=g\quad.
\end{align*}
Two functions are equal when their results on any equal input arguments
are equal. So, the condition $f=g$ is the same condition as $f(a)=g(a)$
for all $a^{:Z}$. $\square$

\subsection{Strong dinaturality: definition and general properties\label{subsec:Strong-dinaturality.-General-properties}}

\subsubsection{Definition \label{subsec:Definition-strong-dinaturality}\ref{subsec:Definition-strong-dinaturality}}

A function $t^{A}:P^{A,A}\rightarrow Q^{A,A}$ is \textbf{strongly
dinatural} \index{strong dinaturality law}if for any $f^{:A\rightarrow B}$,
$x^{:P^{A,A}}$, and $y^{:P^{B,B}}$ the following property holds:\vspace{-0.3\baselineskip}
\begin{equation}
\text{when}\quad x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then}\quad x\triangleright t^{A}\triangleright f^{\uparrow Q^{A,\bullet}}=y\triangleright t^{B}\triangleright f^{\downarrow Q^{\bullet,B}}\quad.\label{eq:strong-dinaturality-law}
\end{equation}

\begin{wrapfigure}{l}{0.32\columnwidth}%
\vspace{-2.5\baselineskip}
\[
\xymatrix{\xyScaleY{1.2pc}\xyScaleX{1.0pc}P^{A,A}\ar[rd]\sb(0.35){f^{\uparrow P^{A,\bullet}}\negthickspace\negthickspace}\ar[rr]\sp(0.5){t^{A}} &  & Q^{A,A}\ar[rd]\sp(0.5){f^{\uparrow Q^{A,\bullet}}}\\
 & P^{A,B} & \negthickspace(\Rightarrow) & Q^{A,B}\\
P^{B,B}\ar[ru]\sp(0.5){f^{\downarrow P^{\bullet,B}}\negthickspace\negthickspace\negthickspace}\ar[rr]\sp(0.5){t^{B}} &  & Q^{B,B}\ar[ru]\sb(0.65){\negthickspace\negthickspace f^{\downarrow Q^{\bullet,B}}}
}
\]
\vspace{-2\baselineskip}
\end{wrapfigure}%

\noindent The strong dinaturality law has an unusual form: it is an
equation (the \textsf{``}conclusion\textsf{''}) that is required to hold only when
some values satisfy another equation (the law\textsf{'}s \textsf{``}precondition\textsf{''}).
The type diagram for that law, shown at left, is a truncated form
of the type diagram for Eq.~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors}).
The implication symbol ($\Rightarrow$) means that the left part of
the diagram is an assumption used by the right part.

In terms of the wedge relation (see Definition~\ref{subsec:Definition-wedge-relation}),
strong dinaturality\textsf{'}s \textsf{``}precondition\textsf{''} is the $\left(P,f\right)$-wedge
relation for $(x,y)$, and the \textsf{``}conclusion\textsf{''} is the $\left(Q,f\right)$-wedge
relation for $(t(x),t(y))$.

The name \textsf{``}strong dinaturality\textsf{''} suggests that this property is
\textsf{``}stronger\textsf{''} than the ordinary dinaturality. Indeed, strongly dinatural
transformations are always dinatural:

\subsubsection{Statement \label{subsec:Statement-strong-dinaturality-entails-dinaturality}\ref{subsec:Statement-strong-dinaturality-entails-dinaturality}}

\textbf{(a)} Any function $t^{A}:P^{A,A}\rightarrow Q^{A,A}$ satisfying
Eq.~(\ref{eq:strong-dinaturality-law}) will also satisfy Eq.~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors}).

\textbf{(b)} A natural transformation $t^{A}:F^{A}\rightarrow G^{A}$
(where $F$ and $G$ are both functors or both contrafunctors) is
strongly dinatural.

\textbf{(c)} A natural transformation $t^{X,Y}:P^{X,Y}\rightarrow Q^{X,Y}$
between profunctors $P$ and $Q$ gives a strongly dinatural transformation
$t^{A,A}:P^{A,A}\rightarrow Q^{A,A}$ when we set $X\triangleq A$
and $Y\triangleq A$ in $t^{X,Y}$.

\subparagraph{Proof}

\textbf{(a)} The law~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors})
is an equality of functions of type $P^{B,A}\rightarrow Q^{A,B}$.
We will now show that those functions will give equal results when
applied to an arbitrary value $p:P^{B,A}$. Choose $x\triangleq p\triangleright f^{\downarrow P^{\bullet,A}}$
and $y\triangleq p\triangleright f^{\uparrow P^{B,\bullet}}$. The
precondition in Eq.~(\ref{eq:strong-dinaturality-law}) is satisfied
with these $x$ and $y$:
\begin{align*}
{\color{greenunder}\text{expect to equal }(y\triangleright f^{\downarrow P^{\bullet,B}}):}\quad & x\triangleright f^{\uparrow P^{A,\bullet}}=p\triangleright f^{\downarrow P^{\bullet,A}}\triangleright f^{\uparrow P^{A,\bullet}}\\
{\color{greenunder}\text{profunctor commutativity law of }P:}\quad & =p\triangleright f^{\uparrow P^{B,\bullet}}\triangleright f^{\downarrow P^{\bullet,B}}=y\triangleright f^{\downarrow P^{\bullet,B}}\quad.
\end{align*}
So, we can use the law~(\ref{eq:strong-dinaturality-law})\textsf{'}s conclusion
and obtain Eq.~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors})
applied to $p$, completing the proof:
\begin{align*}
 & \gunderline x\triangleright t^{A}\triangleright f^{\uparrow Q^{A,\bullet}}\overset{!}{=}\gunderline y\triangleright t^{B}\triangleright f^{\downarrow Q^{\bullet,B}}\quad,\\
{\color{greenunder}\text{definitions of }x,y:}\quad & p\triangleright f^{\downarrow P^{\bullet,A}}\bef t^{A}\bef f^{\uparrow Q^{A,\bullet}}\overset{!}{=}p\triangleright f^{\uparrow P^{B,\bullet}}\bef t^{B}\bef f^{\downarrow Q^{\bullet,B}}\quad.
\end{align*}

\textbf{(b)} Consider the case where $F$ and $G$ are both functors.
We may view $t:F^{A}\rightarrow G^{A}$ as a function with the type
signature $t:P^{A,A}\rightarrow Q^{A,A}$ if we define the profunctors
$P^{X,Y}\triangleq F^{Y}$ and $Q^{X,Y}\triangleq G^{Y}$. Since $P^{X,Y}$
and $Q^{X,Y}$ ignore the type parameter $X$, we have the liftings
$f^{\uparrow P}=f^{\uparrow F}$, $f^{\downarrow P}=\text{id}$, $f^{\uparrow Q}=f^{\uparrow G}$,
and $f^{\downarrow Q}=\text{id}$. The strong dinaturality law of
$t$ is then written as
\[
\text{when}\quad x^{:F^{A}}\triangleright f^{\uparrow F}=y^{:F^{B}}\quad\text{ then}\quad x\triangleright t\triangleright f^{\uparrow G}\overset{?}{=}y\triangleright t\quad.
\]
This is equivalent to $x\triangleright t\bef f^{\uparrow G}\overset{?}{=}x\triangleright f^{\uparrow F}\bef t$,
which holds by the assumed naturality of $t$.

A similar proof works for the case where $F$ and $G$ are both contrafunctors.

\textbf{(c)} A natural transformation $t^{X,Y}:P^{X,Y}\rightarrow Q^{X,Y}$
needs to satisfy naturality laws separately with respect to the type
parameters $X$ and $Y$: for all $f^{:A\rightarrow B}$,
\[
f^{\uparrow P^{X,\bullet}}\bef t^{X,B}=t^{X,A}\bef f^{\uparrow Q^{X,\bullet}}\quad,\quad\quad f^{\downarrow P^{\bullet,Y}}\bef t^{A,Y}=t^{B,Y}\bef f^{\downarrow Q^{\bullet,Y}}\quad.
\]
The strong dinaturality law of $t^{A,A}$ says that
\[
\text{when}\quad x^{:P^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=y^{:P^{B,B}}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then}\quad x\triangleright t^{A,A}\bef f^{\uparrow Q^{A,\bullet}}\overset{?}{=}y\triangleright t^{B,B}\bef f^{\downarrow Q^{\bullet,B}}\quad.
\]
Using the precondition of that law and the naturality laws of $t^{X,Y}$,
rewrite the conclusion of that law:
\begin{align*}
 & x\triangleright\gunderline{t^{A,A}\bef f^{\uparrow Q^{A,\bullet}}}=\gunderline{x\triangleright f^{\uparrow P^{A,\bullet}}}\bef t^{A,B}=y\triangleright f^{\downarrow P^{\bullet,B}}\bef t^{A,B}\\
 & \overset{?}{=}y\triangleright\gunderline{t^{B,B}\bef f^{\downarrow Q^{\bullet,B}}}=y\triangleright f^{\downarrow P^{\bullet,B}}\bef t^{A,B}\quad.
\end{align*}
The two sides of the strong dinaturality law are now equal. $\square$

Strongly dinatural transformations compose, just as natural transformations
do:

\subsubsection{Statement \label{subsec:Statement-natural-transformations-composition}\ref{subsec:Statement-natural-transformations-composition}}

For any functors $P$, $Q$, $R$ and natural transformations $u:P^{A}\rightarrow Q^{A}$
and $v:Q^{A}\rightarrow R^{A}$, the composition $t\triangleq u\bef v$
is a natural transformation (of type $P^{A}\rightarrow R^{A}$).

\subparagraph{Proof}

The naturality laws for $u$ and $v$ are written with an arbitrary
$f^{:A\rightarrow B}$ as
\[
f^{\uparrow P}\bef u=u\bef f^{\uparrow Q}\quad,\quad\quad f^{\uparrow Q}\bef v=v\bef f^{\uparrow R}\quad.
\]
The required naturality law for $t$ is derived by a direct calculation:

\begin{wrapfigure}{l}{0.3\columnwidth}%
\vspace{-1.6\baselineskip}
\[
\xymatrix{\xyScaleY{1.8pc}\xyScaleX{2.0pc}P^{A}\ar[d]\sp(0.45){f^{\uparrow P}}\ar[r]\sp(0.5){u} & Q^{A}\ar[d]\sp(0.45){f^{\uparrow Q}}\ar[r]\sp(0.5){v} & R^{A}\ar[d]\sp(0.45){f^{\uparrow R}}\\
P^{B}\ar[r]\sp(0.5){u} & Q^{B}\ar[r]\sp(0.5){v} & R^{B}
}
\]
\vspace{-2.8\baselineskip}
\end{wrapfigure}%

~\vspace{-1\baselineskip}
\begin{align*}
{\color{greenunder}\text{expect to equal }t\bef f^{\uparrow R}:}\quad & f^{\uparrow P}\bef\gunderline t=\gunderline{f^{\uparrow P}\bef u}\bef v\\
{\color{greenunder}\text{naturality of }u:}\quad & =u\bef\gunderline{f^{\uparrow Q}\bef v}\\
{\color{greenunder}\text{naturality of }v:}\quad & =\gunderline{u\bef v}\bef f^{\uparrow R}=t\bef f^{\uparrow R}\quad.
\end{align*}
This calculation shows that the above type diagram commutes. $\square$

While a direct proof fails for the composition of dinatural transformations
(as we have seen in Section~\ref{subsec:Composition-properties-of-dinatural}),
the additional requirement of \emph{strong} dinaturality makes the
composition property straightforward to prove:

\subsubsection{Statement \label{subsec:Statement-strongly-dinatural-transformation-composition}\ref{subsec:Statement-strongly-dinatural-transformation-composition}}

For any profunctors $P$, $Q$, $R$ and any two strongly dinatural
transformations $u:P^{A,A}\rightarrow Q^{A,A}$ and $v:Q^{A,A}\rightarrow R^{A,A}$,
the transformation $t\triangleq u\bef v$ is also strongly dinatural.

\subparagraph{Proof}

We assume that the strong dinaturality law holds for $u$ and $v$:
given $f^{:A\rightarrow B}$,
\begin{align*}
 & \text{for }x^{:P^{A,A}}\text{ and }y^{:P^{B,B}}:\quad\text{ when}\quad x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then}\quad x\triangleright u^{A}\triangleright f^{\uparrow Q^{A,\bullet}}=y\triangleright u^{B}\triangleright f^{\downarrow Q^{\bullet,B}}\quad,\\
 & \text{for }m^{:Q^{A,A}}\text{ and }n^{:Q^{B,B}}:\quad\text{ when}\quad m\triangleright f^{\uparrow Q^{A,\bullet}}=n\triangleright f^{\downarrow Q^{\bullet,B}}\quad\text{ then}\quad m\triangleright v^{A}\triangleright f^{\uparrow R^{A,\bullet}}=n\triangleright v^{B}\triangleright f^{\downarrow R^{\bullet,B}}\quad.
\end{align*}
We need to show that the strong naturality law holds for $t\triangleq u\bef v$:
\[
\text{when}\quad x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then}\quad x\triangleright u^{A}\triangleright v^{A}\triangleright f^{\uparrow R^{A,\bullet}}\overset{?}{=}y\triangleright u^{B}\triangleright v^{B}\triangleright f^{\downarrow R^{\bullet,B}}\quad.
\]
For any $x$ and $y$ satisfying the above assumption, let $m\triangleq x\triangleright u^{A}$
and $n\triangleq y\triangleright u^{B}$. Then the precondition of
the strong dinaturality law of $v$ is satisfied due to the strong
dinaturality law of $u$:
\[
m\triangleright f^{\uparrow Q^{A,\bullet}}=x\triangleright u^{A}\triangleright f^{\uparrow Q^{A,\bullet}}=y\triangleright u^{B}\triangleright f^{\downarrow Q^{\bullet,B}}=n\triangleright f^{\downarrow Q^{\bullet,B}}\quad.
\]
So, we can use the conclusion of the strong dinaturality law of $v$
with these $m$ and $n$:
\[
x\triangleright u^{A}\triangleright v^{A}\triangleright f^{\uparrow R^{A,\bullet}}=m\triangleright v^{A}\triangleright f^{\uparrow R^{A,\bullet}}\overset{!}{=}n\triangleright v^{B}\triangleright f^{\downarrow R^{\bullet,B}}=y\triangleright u^{B}\triangleright v^{B}\triangleright f^{\downarrow R^{\bullet,B}}\quad.
\]
This is the same as the conclusion of the strong dinaturality law
of $t$. $\square$

It turns out that the type parameters of a natural transformation
can be substituted by some profunctors to obtain a strongly dinatural
transformation. Two versions of this result are proved next.

\subsubsection{Statement \label{subsec:Statement-strong-dinatural-substitute-functor-in-natural}\ref{subsec:Statement-strong-dinatural-substitute-functor-in-natural}}

Suppose $u^{A}:G^{A}\rightarrow H^{A}$ is a natural transformation,
and suppose $P^{X,Y}$ is some fully parametric profunctor. Define
the function $t$ by
\[
t^{A}:G^{P^{A,A}}\rightarrow H^{P^{A,A}}\quad,\quad\quad t^{A}\triangleq u^{P^{A,A}}\quad,
\]
i.e., as the unmodified code of the function $u^{A}$ with a substituted
type parameter ($P^{A,A}$ instead of $A$). Then $t$ is a strongly
dinatural transformation between profunctors $G^{P^{A,A}}$ and $H^{P^{A,A}}$.

\subparagraph{Proof}

By assumption, the naturality law of $u$ holds for any $k^{:R\rightarrow S}$
and says that
\[
k^{\uparrow G}\bef u^{S}=u^{R}\bef k^{\uparrow H}\quad.
\]
The strong dinaturality law of $t$ says that for arbitrary $x^{:G^{P^{A,A}}}$,
$y^{:G^{P^{B,B}}}$, and $f^{:A\rightarrow B}$,
\[
\text{when}\quad x\triangleright f^{\uparrow P^{A,\bullet}\uparrow G}=y\triangleright f^{\downarrow P^{\bullet,B}\uparrow G}\quad\text{ then}\quad x\triangleright t^{A}\triangleright f^{\uparrow P^{A,\bullet}\uparrow H}\overset{?}{=}y\triangleright t^{B}\triangleright f^{\downarrow P^{\bullet,B}\uparrow H}\quad.
\]
Substitute $t^{A}\triangleq u^{P^{A,A}}$ into the above equation
and use the naturality law of $u$:
\begin{align*}
{\color{greenunder}\text{expect to equal }(y\triangleright t^{B}\triangleright f^{\downarrow P^{\bullet,B}\uparrow H}):}\quad & x\triangleright\gunderline{t^{A}}\triangleright f^{\uparrow P^{A,\bullet}\uparrow H}=x\triangleright\gunderline{u^{P^{A,A}}\bef f^{\uparrow P^{A,\bullet}\uparrow H}}\\
{\color{greenunder}\text{naturality law of }u\text{ with }R\triangleq P^{A,A}\text{ and }S\triangleq P^{A,B}:}\quad & =\gunderline{x\triangleright f^{\uparrow P^{A,\bullet}\uparrow G}}\bef u^{P^{A,B}}\\
{\color{greenunder}\text{assumption }(x\triangleright f^{\uparrow P\uparrow G}=y\triangleright f^{\downarrow P\uparrow G}):}\quad & =y\triangleright\gunderline{f^{\downarrow P^{\bullet,B}\uparrow G}\bef u^{P^{A,B}}}\\
{\color{greenunder}\text{naturality law of }u\text{ with }R\triangleq P^{B,B}\text{ and }S\triangleq P^{A,B}:}\quad & =y\triangleright u^{P^{B,B}}\bef f^{\downarrow P^{\bullet,B}\uparrow H}=y\triangleright t^{B}\triangleright f^{\downarrow P^{\bullet,B}\uparrow H}\quad.
\end{align*}


\subsubsection{Statement \label{subsec:Statement-strong-dinatural-substitute-profunctor-in-natural}\ref{subsec:Statement-strong-dinatural-substitute-profunctor-in-natural}}

Assume that $G^{X,Y}$ and $H^{X,Y}$ are bifunctors, $P^{X,Y}$ and
$Q^{X,Y}$ are profunctors, and $u^{A,B}:G^{A,B}\rightarrow H^{A,B}$
is a natural transformation (separately with respect to the type parameters
$A$ and $B$). Assume that $G$ and $H$ satisfy their commutativity
laws. Define the function $t$ by
\[
t^{A}:G^{P^{A,A},Q^{A,A}}\rightarrow H^{P^{A,A},Q^{A,A}}\quad,\quad\quad t\triangleq u^{P^{A,A},Q^{A,A}}\quad,
\]
i.e., as the unmodified code of $u^{A,B}$ with substituted type parameters
($P^{A,A}$ and $Q^{A,A}$ instead of $A$ and $B$). Then $t$ is
a strongly dinatural transformation between profunctors $G^{P^{A,A},Q^{A,A}}$
and $H^{P^{A,A},Q^{A,A}}$.

\subparagraph{Proof}

The assumed naturality law of $u$ holds separately with respect to
each type parameter:
\[
f^{\uparrow G^{\bullet,B}}\bef u^{B,B}=u^{A,B}\bef f^{\uparrow H^{\bullet,B}}\quad,\quad\quad f^{\uparrow G^{A,\bullet}}\bef u^{A,B}=u^{A,A}\bef f^{\uparrow H^{A,\bullet}}\quad.
\]
To formulate the strong dinaturality law of $t$, we need to start
with arbitrary $x^{:G^{P^{A,A}}}$, $y^{:G^{P^{B,B}}}$, and $f^{:A\rightarrow B}$,
and lift $f$ to functors such as $G^{P^{A,\bullet},Q^{A,\bullet}}$.
Since, by assumption, $G$ obeys the commutativity law, we may lift
$f$ to $G^{P^{A,\bullet},Q^{A,\bullet}}$ via a composition of liftings
with respect to $P$ and $Q$ in any order, for example $x\triangleright f^{\uparrow P^{A,\bullet}\uparrow G^{\bullet,Q}}\triangleright f^{\uparrow Q^{A,\bullet}\uparrow G^{P,\bullet}}$.
We handle the lifting to the contrafunctor $G^{P^{\bullet,A},Q^{\bullet,A}}$
similarly. So, let us write the strong dinaturality law of $t$: 
\begin{align*}
 & \text{when}\quad x\triangleright f^{\uparrow P^{A,\bullet}\uparrow G^{\bullet,Q}}\triangleright f^{\uparrow Q^{A,\bullet}\uparrow G^{P,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}\uparrow G^{\bullet,Q}}\triangleright f^{\downarrow Q^{\bullet,B}\uparrow G^{P,\bullet}}\\
 & \quad\text{then}\quad x\triangleright t^{A}\triangleright f^{\uparrow P^{A,\bullet}\uparrow H^{\bullet,Q}}\triangleright f^{\uparrow Q^{A,\bullet}\uparrow H^{P,\bullet}}\overset{?}{=}y\triangleright t^{B}\triangleright f^{\downarrow P^{\bullet,B}\uparrow H^{\bullet,Q}}\triangleright f^{\downarrow Q^{\bullet,B}\uparrow H^{P,\bullet}}\quad.
\end{align*}
Substitute $t\triangleq u^{P^{A,A},Q^{A,A}}$ into the above equation
and use the naturality laws of $u$:
\begin{align*}
 & x\triangleright t^{A}\triangleright f^{\uparrow P^{A,\bullet}\uparrow H^{\bullet,Q}}\triangleright f^{\uparrow Q^{A,\bullet}\uparrow H^{P,\bullet}}=x\triangleright\gunderline{u\bef f^{\uparrow P\uparrow H}\bef}f^{\uparrow Q\uparrow H}=x\triangleright f^{\uparrow P\uparrow G}\bef\gunderline{u\bef f^{\uparrow Q\uparrow H}}=\gunderline{x\triangleright f^{\uparrow P\uparrow G}\triangleright f^{\uparrow Q\uparrow G}}\bef u\\
 & =y\triangleright f^{\downarrow P\uparrow G}\bef\gunderline{f^{\downarrow Q\uparrow G}\bef u}=y\triangleright\gunderline{f^{\downarrow P\uparrow G}\bef u}\bef f^{\downarrow Q\uparrow H}=y\triangleright u\bef f^{\downarrow P\uparrow H}\bef f^{\downarrow Q\uparrow H}\\
 & =y\triangleright t^{B}\triangleright f^{\downarrow P^{\bullet,B}\uparrow H^{\bullet,Q}}\triangleright f^{\downarrow Q^{\bullet,B}\uparrow H^{P,\bullet}}\quad.
\end{align*}
We omitted the full type annotations for brevity, since we already
checked that types match. $\square$

\subsubsection{Statement \label{subsec:Statement-pair-product-coproduct-strongly-dinatural}\ref{subsec:Statement-pair-product-coproduct-strongly-dinatural}}

For any two strongly dinatural transformations $u^{:P^{A,A}\rightarrow R^{A,A}}$
and $v^{:Q^{A,A}\rightarrow S^{A,A}}$:

\textbf{(a)} The pair product $t\triangleq u\boxtimes v$ (of type
$P^{A,A}\times Q^{A,A}\rightarrow R^{A,A}\times S^{A,A}$) is strongly
dinatural.

\textbf{(b)} The pair co-product\index{pair co-product of functions}
$t\triangleq u\boxplus v$ (of type $P^{A,A}+Q^{A,A}\rightarrow R^{A,A}+S^{A,A}$)
is strongly dinatural.

\subparagraph{Proof}

\textbf{(a)} The strong dinaturality law of $t\triangleq u\boxtimes v$
says that, for any $f^{:A\rightarrow B}$:
\begin{align*}
 & \text{when}\quad(x_{1}^{:P^{A,A}}\times y_{1}^{:Q^{A,A}})\triangleright f^{\uparrow(P^{A,\bullet}\times Q^{A,\bullet})}=(x_{2}^{:P^{B,B}}\times y_{2}^{:Q^{B,B}})\triangleright f^{\downarrow(P^{\bullet,B}\times Q^{\bullet,B})}\\
 & \quad\text{then}\quad(x_{1}\times y_{1})\triangleright t\triangleright f^{\uparrow(R^{A,\bullet}\times S^{A,\bullet})}=(x_{2}\times y_{2})\triangleright t\triangleright f^{\downarrow(R^{\bullet,B}\times S^{\bullet,B})}\quad.
\end{align*}
Substituting the definitions of liftings to $P\times Q$ and $R\times S$,
as well as the definition $t\triangleq u\boxtimes v$, we get:
\begin{align*}
 & \text{when}\quad(x_{1}\triangleright f^{\uparrow P^{A,\bullet}})\times(y_{1}\triangleright f^{\uparrow Q^{A,\bullet}})=(x_{2}\triangleright f^{\downarrow P^{\bullet,B}})\times(y_{2}\triangleright f^{\downarrow Q^{\bullet,B}})\\
 & \quad\text{then}\quad(x_{1}\triangleright u\bef f^{\uparrow R^{A,\bullet}})\times(y_{1}\triangleright v\bef f^{\uparrow S^{A,\bullet}})=(x_{2}\triangleright u\bef f^{\downarrow R^{\bullet,B}})\times(y_{2}\triangleright v\bef f^{\downarrow S^{\bullet,B}})\quad.
\end{align*}
An equality between pairs is the same as two equalities for the elements
of the pairs. So, the strong naturality law of $t$ is the same as
the following statement:
\begin{align*}
 & \text{when}\quad x_{1}^{:P^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=x_{1}^{:P^{B,B}}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ and }\quad y_{1}^{:Q^{A,A}}\triangleright f^{\uparrow Q^{A,\bullet}}=y_{2}^{:Q^{B,B}}\triangleright f^{\downarrow Q^{\bullet,B}}\\
 & \quad\text{then}\quad x_{1}\triangleright u\bef f^{\uparrow R^{A,\bullet}}=x_{2}\triangleright u\bef f^{\downarrow R^{\bullet,B}}\quad\text{ and }\quad y_{1}\triangleright v\bef f^{\uparrow S^{A,\bullet}}=y_{2}\triangleright v\bef f^{\downarrow S^{\bullet,B}}\quad.
\end{align*}
These are exactly the preconditions and the conclusions of the strong
naturality laws of $u$ and $v$.

\textbf{(b)} The strong dinaturality law of $t\triangleq u\boxplus v$
says that, for any $f^{:A\rightarrow B}$:
\begin{align*}
 & \text{when}\quad x^{:P^{A,A}+Q^{A,A}}\triangleright f^{\uparrow(P^{A,\bullet}+Q^{A,\bullet})}=y^{:P^{B,B}+Q^{B,B}}\triangleright f^{\downarrow(P^{\bullet,B}+Q^{\bullet,B})}\\
 & \quad\text{then}\quad x\triangleright t\triangleright f^{\uparrow(R^{A,\bullet}+S^{A,\bullet})}=y\triangleright t\triangleright f^{\downarrow(R^{\bullet,B}+S^{\bullet,B})}\quad.
\end{align*}
Substituting the definitions of liftings to $P+Q$ and $R+S$, as
well as the definition $t\triangleq u\boxplus v$, we get:
\begin{align*}
 & \text{when}\quad x^{:P^{A,A}+Q^{A,A}}\triangleright\,\begin{array}{|c||cc|}
 & P^{A,B} & Q^{A,B}\\
\hline P^{A,A} & f^{\uparrow P^{A,\bullet}} & \bbnum 0\\
Q^{A,A} & \bbnum 0 & f^{\uparrow Q^{A,\bullet}}
\end{array}\,=y^{:P^{B,B}+Q^{B,B}}\triangleright\,\begin{array}{|c||cc|}
 & P^{A,B} & Q^{A,B}\\
\hline P^{B,B} & f^{\downarrow P^{\bullet,B}} & \bbnum 0\\
Q^{B,B} & \bbnum 0 & f^{\downarrow Q^{\bullet,B}}
\end{array}\,\quad\text{ then}\\
 & x\triangleright\,\begin{array}{|c||cc|}
 & R^{A,A} & S^{A,A}\\
\hline P^{A,A} & u & \bbnum 0\\
Q^{A,A} & \bbnum 0 & v
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & R^{A,B} & S^{A,B}\\
\hline R^{A,A} & f^{\uparrow R^{A,\bullet}} & \bbnum 0\\
S^{A,A} & \bbnum 0 & f^{\uparrow S^{A,\bullet}}
\end{array}\,=y\triangleright\,\begin{array}{|c||cc|}
 & R^{B,B} & S^{B,B}\\
\hline P^{B,B} & u & \bbnum 0\\
Q^{B,B} & \bbnum 0 & v
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & R^{A,B} & S^{A,B}\\
\hline R^{B,B} & f^{\downarrow R^{\bullet,B}} & \bbnum 0\\
S^{B,B} & \bbnum 0 & f^{\downarrow S^{\bullet,B}}
\end{array}
\end{align*}
or, simplifying the matrix composition:
\[
x\triangleright\,\begin{array}{|c||cc|}
 & R^{A,B} & S^{A,B}\\
\hline P^{A,A} & u\bef f^{\uparrow R^{A,\bullet}} & \bbnum 0\\
Q^{A,A} & \bbnum 0 & v\bef f^{\uparrow S^{A,\bullet}}
\end{array}\,=y\triangleright\,\begin{array}{|c||cc|}
 & R^{A,B} & S^{A,B}\\
\hline P^{B,B} & u\bef f^{\downarrow R^{\bullet,B}} & \bbnum 0\\
Q^{B,B} & \bbnum 0 & v\bef f^{\downarrow S^{\bullet,B}}
\end{array}\quad.
\]
The precondition of the strong dinaturality law of $t$ is an equality
between values of disjunctive type $P^{A,B}+Q^{A,B}$ computed by
applying some diagonal code matrices to $x$ and $y$. So, the precondition
can hold only if both $x$ and $y$ are in the same part of the disjunction.
We must have either $x=p_{1}^{:P^{A,A}}+\bbnum 0$ and $y=p_{2}^{:P^{B,B}}+\bbnum 0$,
or $x=\bbnum 0+q_{1}^{:Q^{A,A}}$ and $y=\bbnum 0+q_{2}^{:Q^{B,B}}$.
In the first case, we can simplify the strong dinaturality law of
$t$ to
\[
\text{when }\quad p_{1}^{:P^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=p_{2}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then }\quad p_{1}\triangleright u\bef f^{\uparrow R^{A,\bullet}}=p_{2}\triangleright u\bef f^{\downarrow R^{\bullet,B}}\quad,
\]
and in the second case to 
\[
\text{when }\quad q_{1}^{:P^{A,A}}\triangleright f^{\uparrow Q^{A,\bullet}}=q_{2}\triangleright f^{\downarrow Q^{\bullet,B}}\quad\text{ then }\quad q_{1}\triangleright v\bef f^{\uparrow S^{A,\bullet}}=q_{2}\triangleright v\bef f^{\downarrow S^{\bullet,B}}\quad.
\]
These are exactly the statements of the strong dinaturality laws of
$u$ and $v$. $\square$

\subsubsection{Statement \label{subsec:Statement-strongly-dinatural-product-of-functions}\ref{subsec:Statement-strongly-dinatural-product-of-functions}}

\textbf{(a)} If both $u:P^{A,A}\rightarrow Q^{A,A}$ and $v:P^{A,A}\rightarrow R^{A,A}$
are strongly dinatural then 
\[
t:P^{A,A}\rightarrow Q^{A,A}\times R^{A,A}\quad,\quad\quad t\triangleq p^{:P^{A,A}}\rightarrow u(p)\times v(p)
\]
is also strongly dinatural. 

\textbf{(b)} If both $u:P^{A,A}\rightarrow R^{A,A}$ and $v:Q^{A,A}\rightarrow R^{A,A}$
are strongly dinatural then
\[
t:P^{A,A}+Q^{A,A}\rightarrow R^{A,A}\quad,\quad\quad t\triangleq\,\begin{array}{|c||c|}
 & R^{A,A}\\
\hline P^{A,A} & u\\
Q^{A,A} & v
\end{array}
\]
is also strongly dinatural.

\subparagraph{Proof}

\textbf{(a)} The function $t$ can be expressed as
\[
t=(p^{:P^{A,A}}\rightarrow p\times p)\bef\big(p^{:P^{A,A}}\times p^{:P^{A,A}}\rightarrow u(p)\times v(p)\big)=\Delta^{P^{A,A}}\bef(u\boxtimes v)\quad.
\]
Note that $\Delta^{A}:A\rightarrow A\times A$ is a natural transformation
(see Section~\ref{subsec:Deriving-laws-for-functions-}). By Statement~\ref{subsec:Statement-strong-dinatural-substitute-functor-in-natural},
setting the type parameter of $\Delta$ to $P^{A,A}$ yields a strongly
dinatural transformation ($\Delta^{P^{A,A}}$). The pair product $u\boxtimes v$
is strongly dinatural by Statement~\ref{subsec:Statement-pair-product-coproduct-dinatural}(a).
Finally, the composition $\Delta\bef(u\boxtimes v)$ is strongly dinatural
by Statement~\ref{subsec:Statement-strongly-dinatural-transformation-composition}.

\textbf{(b)} The function $t$ can be expressed via the natural transformation
\lstinline!merge! (see Section~\ref{subsec:Working-with-disjunctive-functions}):
\[
t=\,\begin{array}{|c||cc|}
 & R^{A,A} & R^{A,A}\\
\hline P^{A,A} & u & \bbnum 0\\
Q^{A,A} & \bbnum 0 & v
\end{array}\,\bef\,\begin{array}{|c||c|}
 & R^{A,A}\\
\hline R^{A,A} & \text{id}\\
R^{A,A} & \text{id}
\end{array}\,=(u\boxplus v)\bef\text{merge}^{R^{A,A}}\quad.
\]
The pair co-product ($u\boxplus v$) is strongly dinatural by Statement~\ref{subsec:Statement-pair-product-coproduct-dinatural}(b).
The function \lstinline!merge! with a substituted type parameter
($R^{A,A}$) is strongly dinatural by Statement~\ref{subsec:Statement-strong-dinatural-substitute-functor-in-natural}.
Finally, the composition $(u\boxplus v)\bef\text{merge}$ is strongly
dinatural by Statement~\ref{subsec:Statement-strongly-dinatural-transformation-composition}.
$\square$

\subsubsection{Statement \label{subsec:Statement-strong-dinaturality-of-function-application}\ref{subsec:Statement-strong-dinaturality-of-function-application}}

Consider the (fully parametric) function $w\triangleq q\times z\rightarrow q(z)$.

\textbf{(a)} The most general type signature for $w$ is $w^{Q,Z}:(Z\rightarrow Q)\times Z\rightarrow Q$
with two type parameters ($Q$, $Z$). With this type signature, $w$
is strongly dinatural with respect to both $Q$ and $Z$.

\textbf{(b)} If we substitute the types $Q^{A,A}$ and $Z^{A,A}$
instead of $Q$ and $Z$, where $Q^{X,Y}$ and $Z^{X,Y}$ are some
profunctors obeying their commutativity laws, we obtain the function
\[
w:\forall A.\,(Z^{A,A}\rightarrow Q^{A,A})\times Z^{A,A}\rightarrow Q^{A,A}\quad,\quad\quad w^{A}\triangleq q^{:Z^{A,A}\rightarrow Q^{A,A}}\times z^{:Z^{A,A}}\rightarrow q(z)\quad.
\]
Then the function $w$ is a dinatural transformation.

\textbf{(c)} Under the same assumptions as in \textbf{(b)}, the function
$w$ is strongly dinatural if the profunctor$Z^{X,Y}$ has the \textsf{``}weak
pullback property\textsf{''} (see Definition~\ref{subsec:Definition-weak-pullback-property}
and Statement~\ref{subsec:Statement-weak-pullback-property}). %
\begin{comment}
Otherwise, $w$ is \emph{not} necessarily strongly dinatural. \textemdash{}
not clear if we can show a counterexample
\end{comment}


\subparagraph{Proof}

\textbf{(a)} The type signature $\forall(Q,Z).\,(Z\rightarrow Q)\times Z\rightarrow Q$
has the form of a natural transformation with respect to the parameter
$Q$ (but not with respect to $Z$). By Statement~\ref{subsec:Statement-strong-dinaturality-entails-dinaturality}(b),
strong dinaturality holds with respect to $Q$. 

The strong dinaturality law of $w$ with respect to $Z$ needs to
be checked directly. For convenience, define the profunctors $P^{X,Y}\triangleq(X\rightarrow Q)\times Y$
and $R^{X,Y}\triangleq Q$, so that $w$ has the type signature $P^{Z,Z}\rightarrow R^{Z,Z}$.
We need to prove that, for any $f^{:A\rightarrow B}$,
\[
\text{when}\quad x^{:P^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=y^{:P^{B,B}}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{ then}\quad x\triangleright w\triangleright f^{\uparrow R}\overset{?}{=}y\triangleright w\triangleright f^{\downarrow R}\quad.
\]
Since $R^{X,Y}$ is a constant profunctor, we have $f^{\uparrow R}=\text{id}$
and $f^{\downarrow R}=\text{id}$. Defining $g^{:A\rightarrow Q}\times a^{:A}\triangleq x^{:P^{A,A}}$
and $h^{:B\rightarrow Q}\times b^{:B}\triangleq y^{:P^{B,B}}$, and
using the liftings 
\[
f^{\uparrow P^{A,\bullet}}=\text{id}^{A\rightarrow Q}\boxtimes f\quad,\quad\quad f^{\downarrow P^{\bullet,B}}=(g^{:B\rightarrow Q}\rightarrow f\bef g)\boxtimes\text{id}^{B}\quad,
\]
we can rewrite the law as
\[
\text{when}\quad g^{:A\rightarrow Q}\times f(a^{:A})=(f\bef h^{:B\rightarrow Q})\times b^{:B}\quad\text{ then}\quad g(a)\overset{?}{=}h(b)\quad.
\]
Since we must have $b=f(a)$ and $g=f\bef h$, the law holds:
\[
g(a)=h(f(a))=h(b).
\]

\textbf{(b)} Define the profunctor $P^{X,Y}\triangleq(Z^{Y,X}\rightarrow Q^{X,Y})\times Z^{X,Y}$
(note the reverse order of type parameters in $Z^{Y,X}$). Then $w^{A}$
has type signature $P^{A,A}\rightarrow Q^{A,A}$, and the naturality
law~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors})
is
\[
f^{\downarrow P}\bef w\bef f^{\uparrow Q}\overset{?}{=}f^{\uparrow P}\bef w\bef f^{\downarrow Q}\quad.
\]
It helps to write out the code for the liftings to $P$. For arbitrary
$f^{:A\rightarrow B}$,
\begin{align*}
 & (g^{:Z^{Y,B}\rightarrow Q^{B,Y}}\times a^{:Z^{B,Y}})\triangleright f^{\downarrow P^{\bullet,Y}}=\big(f^{\uparrow Z^{Y,\bullet}}\bef g\bef f^{\downarrow Q^{\bullet,Y}}\big)\times(a\triangleright f^{\downarrow Z^{\bullet,Y}})\quad,\\
 & (g^{:Z^{A,X}\rightarrow Q^{X,A}}\times a^{:Z^{X,A}})\triangleright f^{\uparrow P^{X,\bullet}}=\big(f^{\downarrow Z^{\bullet,X}}\bef g\bef f^{\uparrow Q^{X,\bullet}}\big)\times(a\triangleright f^{\uparrow Z^{X,\bullet}})\quad.
\end{align*}

The naturality law of $w$ is written as an equation between functions
of type $P^{B,A}\rightarrow Q^{A,B}$:
\[
f^{\downarrow P^{\bullet,A}}\bef w^{A}\bef f^{\uparrow Q^{A,\bullet}}\overset{?}{=}f^{\uparrow P^{B,\bullet}}\bef w^{B}\bef f^{\downarrow Q^{\bullet,B}}\quad.
\]
Apply both sides of the law to an arbitrary value $g\times a:P^{B,A}$:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (g^{:Z^{A,B}\rightarrow Q^{B,A}}\times a^{:Z^{B,A}})\,\gunderline{\triangleright f^{\downarrow P}}\bef w\bef f^{\uparrow Q}=\big((f^{\uparrow Z}\bef g\bef f^{\downarrow Q})\times(a\triangleright f^{\downarrow Z})\big)\triangleright w\triangleright f^{\uparrow Q}\\
{\color{greenunder}\text{definition of }w:}\quad & \quad=(a\triangleright f^{\downarrow Z})\triangleright(f^{\uparrow Z}\bef g\bef f^{\downarrow Q})\triangleright f^{\uparrow Q}=a\triangleright f^{\downarrow Z}\bef f^{\uparrow Z}\bef g\bef f^{\downarrow Q}\bef f^{\uparrow Q}\\
{\color{greenunder}\text{right-hand side}:}\quad & \overset{?}{=}(g\times a)\,\gunderline{\triangleright f^{\uparrow P}}\bef w\bef f^{\downarrow Q}=\big((f^{\downarrow Z}\bef g\bef f^{\uparrow Q})\times(a\triangleright f^{\uparrow Z})\big)\triangleright w\triangleright f^{\downarrow Q}\\
{\color{greenunder}\text{definition of }w:}\quad & \quad=(a\triangleright f^{\uparrow Z})\triangleright(f^{\downarrow Z}\bef g\bef f^{\uparrow Q})\triangleright f^{\downarrow Q}=a\triangleright f^{\uparrow Z}\bef f^{\downarrow Z}\bef g\bef f^{\uparrow Q}\bef f^{\downarrow Q}\quad.
\end{align*}
The remaining difference between two sides disappears due to the commutativity
laws of $Q$ and $Z$ (namely $f^{\uparrow Q}\bef f^{\downarrow Q}=f^{\downarrow Q}\bef f^{\uparrow Q}$
and $f^{\uparrow Z}\bef f^{\downarrow Z}=f^{\downarrow Z}\bef f^{\uparrow Z}$).

\textbf{(c)} The strong dinaturality of $w$ says that, for arbitrary
$f^{:A\rightarrow B}$,
\[
\text{when}\quad x^{:G^{A,A}\times Z^{A,A}}\triangleright f^{\uparrow(G\times Z)}=y^{:G^{B,B}\times Z^{B,B}}\triangleright f^{\downarrow(G\times Z)}\quad\text{ then}\quad x\triangleright w\triangleright f^{\uparrow Q}\overset{?}{=}y\triangleright w\triangleright f^{\downarrow Q}\quad.
\]
Since $f^{\uparrow(G\times Z)}=f^{\uparrow G}\boxtimes f^{\uparrow Z}$,
the precondition of the law splits into two equations:
\begin{align*}
{\color{greenunder}\text{defining }g_{1},z_{1},g_{2},z_{2}:}\quad & g_{1}^{:G^{A,A}}\times z_{1}^{:Z^{A,A}}\triangleq x\quad,\quad\quad g_{2}^{:G^{B,B}}\times z_{2}^{:Z^{B,B}}\triangleq y\quad,\\
{\color{greenunder}\text{the precondition is equivalent to}:}\quad & g_{1}\triangleright f^{\uparrow G}=g_{2}\triangleright f^{\downarrow G}\quad\text{ and }\quad z_{1}\triangleright f^{\uparrow Z}=z_{2}\triangleright f^{\downarrow Z}\quad.
\end{align*}
The precondition $g_{1}\triangleright f^{\uparrow G}=g_{2}\triangleright f^{\downarrow G}$
is equivalent to
\begin{equation}
f^{\downarrow Z}\bef g_{1}\bef f^{\uparrow Q}=f^{\uparrow Z}\bef g_{2}\bef f^{\downarrow Q}\quad.\label{eq:strong-dinaturality-precondition-derivation1}
\end{equation}
Now we express the conclusion of the law via the new variables:
\begin{equation}
x\triangleright w\triangleright f^{\uparrow Q}=z_{1}\triangleright g_{1}\bef f^{\uparrow Q}\overset{?}{=}y\triangleright w\triangleright f^{\downarrow Q}=z_{2}\triangleright g_{2}\bef f^{\downarrow Q}\quad.\label{eq:strong-dinaturality-conclusion-derivation1}
\end{equation}
From here, the proof cannot proceed without more information. We need
to show that $z_{1}\triangleright g_{1}\bef f^{\uparrow Q}\overset{?}{=}z_{2}\triangleright g_{2}\bef f^{\downarrow Q}$,
but we do not have any equations relating $g_{1}\bef f^{\uparrow Q}$
to $g_{2}\bef f^{\downarrow Q}$. We cannot use Eq.~(\ref{eq:strong-dinaturality-precondition-derivation1})
because it contains additional lifted functions $f^{\downarrow Z}$
and $f^{\uparrow Z}$ that we cannot insert into the law~(\ref{eq:strong-dinaturality-conclusion-derivation1}),
given that $z_{1}$ and $z_{2}$ are unknown. 

If the profunctor $Z$ has the weak pullback property, the $\left(Z,f\right)$-wedge
relation $z_{1}\triangleright f^{\uparrow Z}=z_{2}\triangleright f^{\downarrow Z}$
gives us a value $u^{:Z^{B,A}}$ such that $z_{1}=u\triangleright f^{\downarrow Z}$
and $z_{2}=u\triangleright f^{\uparrow Z}$. Then the law~(\ref{eq:strong-dinaturality-conclusion-derivation1})
becomes
\[
z_{1}\triangleright g_{1}\bef f^{\uparrow Q}=u\triangleright f^{\downarrow Z}\bef g_{1}\bef f^{\uparrow Q}\overset{?}{=}z_{2}\triangleright g_{2}\bef f^{\downarrow Q}=u\triangleright f^{\uparrow Z}\bef g_{2}\bef f^{\downarrow Q}\quad.
\]
Both sides are now equal due to Eq.~(\ref{eq:strong-dinaturality-precondition-derivation1}).
$\square$

We will now formulate the weak pullback property used in the previous
statement. 

\subsubsection{Definition \label{subsec:Definition-weak-pullback-property}\ref{subsec:Definition-weak-pullback-property}
(weak pullback property)\index{weak pullback property of profunctors}}

A profunctor $P^{X,Y}$ (contravariant in $X$ and covariant%
\begin{comment}
empty arrow
\end{comment}

\begin{wrapfigure}{l}{0.41\columnwidth}%
\vspace{-2\baselineskip}
\[
\xymatrix{\xyScaleY{2.3pc}\xyScaleX{1.5pc} & P^{A,A}\ar[d]\sb(0.5){f^{\uparrow P^{A,\bullet}}} & \ar@{}[d]\sb(0.5){\displaystyle \Rightarrow} & P^{B,A}\ar[r]\sp(0.55){f^{\downarrow P^{\bullet,A}}}\ar[d]\sb(0.5){f^{\uparrow P^{B,\bullet}}} & P^{A,A}\ar[d]\sb(0.5){f^{\uparrow P^{A,\bullet}}}\\
P^{B,B}\ar[r]\sb(0.5){f^{\downarrow P^{\bullet,B}}} & P^{A,B} &  & P^{B,B}\ar[r]\sb(0.5){f^{\downarrow P^{\bullet,B}}} & P^{A,B}
}
\]

\vspace{-2\baselineskip}
\end{wrapfigure}%

\noindent in $Y$) has the \textbf{weak pullback property} if for
any $f^{:A\rightarrow B}$ and any values $x^{:P^{A,A}}$ and $y^{:P^{B,B}}$
in the $\left(P,f\right)$-wedge relation ($x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}$),
we can compute a value $p^{:P^{B,A}}$ such that $x=p\triangleright f^{\downarrow P^{\bullet,A}}$
and $y=p\triangleright f^{\uparrow P^{B,\bullet}}$. The type diagram
means that a wedge starting with $x$ and $y$ can be always completed
to a commuting square.

Note that the function $f$ in this definition does not need to be
invertible.

\subsubsection{Example \label{subsec:Example-weak-pullback-property-1}\ref{subsec:Example-weak-pullback-property-1}}

\textbf{(a)} Suppose a profunctor $P^{X,Y}$ does not depend on the
type parameter $X$ (so, $P^{X,Y}\triangleq G^{Y}$ with some functor
$G$). Then $P$ has the weak pullback property. 

\textbf{(b)} The weak pullback property holds for profunctors $P^{X,Y}\triangleq H^{X}$,
where $H$ is a contrafunctor.

\subparagraph{Proof}

\textbf{(a)} With the choice $P^{X,Y}\triangleq G^{Y}$, we have the
liftings $f^{\uparrow P}=f^{\uparrow G}$ and $f^{\downarrow P}=f$.
The weak pullback property of $P^{X,Y}$ says that, for any $f^{:A\rightarrow B}$,
$x^{:G^{A}}$, $y^{:G^{B}}$:
\[
\text{when}\quad x\triangleright f^{\uparrow G}=y\quad\text{ then }\quad\exists p^{:G^{A}}\text{ such that }x=p\text{ and }y=p\triangleright f^{\uparrow G}\quad.
\]
This property is satisfied by choosing $p\triangleq x$.

\textbf{(b)} With the choice $P^{X,Y}\triangleq H^{X}$, we have the
liftings $f^{\uparrow P}=f$ and $f^{\downarrow P}=f^{\downarrow H}$.
The weak pullback property of $P^{X,Y}$ says that, for any $f^{:A\rightarrow B}$,
$x^{:H^{A}}$, $y^{:H^{B}}$:
\[
\text{when}\quad x=y\triangleright f^{\downarrow H}\quad\text{ then }\quad\exists p^{:H^{A}}\text{ such that }x=p\triangleright f^{\downarrow H}\text{ and }y=p\quad.
\]
This property is satisfied by choosing $p\triangleq y$.

\subsubsection{Example \label{subsec:Example-weak-pullback-property}\ref{subsec:Example-weak-pullback-property}}

The profunctor $P^{X,Y}\triangleq X\rightarrow Y$ does \emph{not}
have the weak pullback property.

\subparagraph{Proof}

With $P^{X,Y}\triangleq X\rightarrow Y$, we rewrite the $\left(P,f\right)$-wedge
relation ($x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}$)
for arbitrary $f^{:A\rightarrow B}$, $x^{:A\rightarrow A}$, and
$y^{:B\rightarrow B}$ as:
\[
x\bef f=f\bef y\quad.
\]
A commuting square diagram requires to find a value $p^{:B\rightarrow A}$
such that $x=f\bef p$ and $y=p\bef f$. Choose $f$ as a constant
function ($f\triangleq\_\rightarrow b_{0}$ with a fixed value $b_{0}^{:B}$).
The $\left(P,f\right)$-wedge relation applied to an arbitrary value
$a^{:A}$ gives
\[
a\triangleright x\bef f=b_{0}\overset{!}{=}a\triangleright f\bef y=y(b_{0})\quad.
\]
So, the wedge relation will hold for any $x^{:A\rightarrow A}$ and
for any $y^{:B\rightarrow B}$ such that $y(b_{0})=b_{0}$. In particular,
$y$ is not necessarily a constant function. But the condition $y=p\bef f=\_\rightarrow b_{0}$
can be satisfied only if $y$ is a constant function. So, we are unable
to find $p^{:P^{B,A}}$ that completes the square diagram. $\square$

We can perform structural analysis to discover profunctors that have
the weak pullback property:

\subsubsection{Statement \label{subsec:Statement-weak-pullback-property}\ref{subsec:Statement-weak-pullback-property}}

A profunctor $P^{X,Y}$ has the weak pullback property if:

\textbf{(a)} $P^{X,Y}\triangleq F^{Y}$ with some (covariant) functor
$F$,

\textbf{(b)} $P^{X,Y}\triangleq G^{X}$ with some contrafunctor $G$,

\textbf{(c)} $P^{X,Y}\triangleq Z\rightarrow Q^{X,Y}$ with a fixed
type $Z$ and a profunctor $Q$ that has the weak pullback property,

\textbf{(d)} $P^{X,Y}\triangleq Q^{X,Y}\times R^{X,Y}$ with profunctors
$Q$ and $R$ that both have the weak pullback property,

\textbf{(e)} $P^{X,Y}\triangleq Q^{X,Y}+R^{X,Y}$ with profunctors
$Q$ and $R$ that both have the weak pullback property.

\subparagraph{Proof}

Consider some values $f^{:A\rightarrow B}$, $x^{:P^{A,A}}$, and
$y^{:P^{B,B}}$, for which the $\left(P,f\right)$-wedge relation
holds:
\[
x\triangleright f^{\uparrow P^{A,\bullet}}=y\triangleright f^{\downarrow P^{\bullet,B}}\quad.
\]

\textbf{(a)} If $P^{X,Y}=F^{Y}$ with a functor $F$, the wedge relation
is simplified to $x\triangleright f^{\uparrow F}=y$ with $x^{:F^{A}}$
and $y^{:F^{B}}$, and we need to find $p^{:F^{A}}$ such that $x=p$
and $y=x\triangleright f^{\uparrow F}$. So, we define $p\triangleq x$
and complete the commuting square.

\textbf{(b)} If $P^{X,Y}=G^{X}$ with a contrafunctor $G$, the wedge
relation is simplified to $x=y\triangleright f^{\downarrow G}$ with
$x^{:G^{A}}$ and $y^{:G^{B}}$, and we need to find $p^{:G^{B}}$
such that $x=p\triangleright f^{\downarrow G}$ and $y=p$. So, we
define $p\triangleq y$ and complete the commuting square.

\textbf{(c)} Write the wedge relation for $x^{:Z\rightarrow Q^{A,A}}$
and $y^{:Z\rightarrow Q^{B,B}}$, applied to an arbitrary $z^{:Z}$:
\[
z\triangleright x\triangleright f^{\uparrow Q^{A,\bullet}}=z\triangleright y\triangleright f^{\downarrow Q^{\bullet,B}}\quad.
\]
Note that this is the same as the $\left(Q,f\right)$-wedge relation
for the values $z\triangleright x$ and $z\triangleright y$. By assumption,
$Q^{X,Y}$ has the weak pullback property; so, we may apply that property
to $z\triangleright x$ and $z\triangleright y$ and obtain a value
$q^{:Q^{B,A}}$ that satisfies the following two equations:
\begin{equation}
q\triangleright f^{\uparrow Q^{B,\bullet}}=z\triangleright y\quad,\quad\quad q\triangleright f^{\downarrow Q^{\bullet,A}}=z\triangleright x\quad.\label{eq:weak-pullback-c-derivation1}
\end{equation}
We need to find $p^{:Z\rightarrow Q^{B,A}}$ such that $x=p\bef f^{\downarrow Q^{\bullet,A}}$
and $y=p\bef f^{\uparrow Q^{B,\bullet}}$. These equations are the
same as Eq.~(\ref{eq:weak-pullback-c-derivation1}) if we define
$p(z)\triangleq q$. So, we have found the value $p$ that completes
the square diagram.

\textbf{(d)} Write the wedge relation for $x^{:Q^{A,A}\times R^{A,A}}=x_{1}^{:Q^{A,A}}\times x_{2}^{:R^{A,A}}$
and $y^{:Q^{B,B}\times R^{B,B}}=y_{1}^{:Q^{B,B}}\times y_{2}^{:R^{B,B}}$
as:
\[
(x_{1}\times x_{2})\triangleright f^{\uparrow(Q\times R)}=(x_{1}\triangleright f^{\uparrow Q})\times(x_{2}\triangleright f^{\uparrow R})\overset{!}{=}(y_{1}\times y_{2})\triangleright f^{\downarrow(Q\times R)}=(y_{1}\triangleright f^{\downarrow Q})\times(y_{2}\triangleright f^{\downarrow R})\quad.
\]
This equation is equivalent to the two equations, 
\[
x_{1}\triangleright f^{\uparrow Q}\overset{!}{=}y_{1}\triangleright f^{\downarrow Q}\quad\text{ and }\quad x_{2}\triangleright f^{\uparrow R}\overset{!}{=}y_{2}\triangleright f^{\downarrow R}\quad,
\]
which are the $\left(Q,f\right)$- and $\left(R,f\right)$-wedge relations.
By assumption, $Q$ and $R$ have the weak pullback property. So,
we may compute values $p_{1}^{:Q^{B,A}}$ and $p_{2}^{:R^{B,A}}$
such that
\[
p_{1}\triangleright f^{\downarrow Q}=x_{1}\quad,\quad\quad p_{1}\triangleright f^{\uparrow Q}=y_{1}\quad,\quad\quad p_{2}\triangleright f^{\downarrow R}=x_{2}\quad,\quad\quad p_{2}\triangleright f^{\uparrow R}=y_{2}\quad.
\]
If we define $p\triangleq p_{1}\times p_{2}$, we complete the square
diagram for $\left(Q\times R,f\right)$-wedge relation because:
\[
p\triangleright f^{\downarrow(Q\times R)}=(p_{1}\triangleright f^{\downarrow Q})\times(p_{2}\triangleright f^{\downarrow R})=x_{1}\times x_{2}\quad,\quad\quad p\triangleright f^{\uparrow(Q\times R)}=(p_{1}\triangleright f^{\uparrow Q})\times(p_{2}\triangleright f^{\uparrow R})=y_{1}\times y_{2}\quad.
\]

\textbf{(e)} Write the wedge relation for $x^{:Q^{A,A}+R^{A,A}}$
and $y^{:Q^{B,B}+R^{B,B}}$ as:
\[
x\triangleright\,\begin{array}{|c||cc|}
 & Q^{A,B} & R^{A,B}\\
\hline Q^{A,A} & f^{\uparrow Q^{A,\bullet}} & \bbnum 0\\
R^{A,A} & \bbnum 0 & f^{\uparrow R^{A,\bullet}}
\end{array}\,=y\triangleright\,\begin{array}{|c||cc|}
 & Q^{A,B} & R^{A,B}\\
\hline Q^{B,B} & f^{\downarrow Q^{\bullet,B}} & \bbnum 0\\
R^{B,B} & \bbnum 0 & f^{\downarrow R^{\bullet,B}}
\end{array}\quad.
\]
This equation can be satisfied only if both sides are either of type
$Q^{A,B}+\bbnum 0$ or of type $\bbnum 0+R^{A,B}$. Since the lifting
matrices are diagonal, this can happen only if $x$ has type $Q^{A,A}+\bbnum 0$
and $y$ has type $Q^{B,B}+\bbnum 0$, of if $x$ has type $\bbnum 0+R^{A,A}$
and $y$ has type $\bbnum 0+R^{B,B}$. In the first case, the property
becomes equivalent to that of the profunctor $Q$; in the second case,
to that of the profunctor $R$.

To see this more explicitly, consider two cases: $x\triangleq x_{1}^{:Q^{A,A}}+\bbnum 0$
and $x\triangleq\bbnum 0+x_{2}^{:R^{A,A}}$. In the first case, $x\triangleright f^{\uparrow P}=(x_{1}\triangleright f^{\uparrow Q})+\bbnum 0$.
This can be equal to $y\triangleright f^{\downarrow P}$ only if $y\triangleright f^{\downarrow P}$
is in the left part of the disjunction type $Q^{A,B}+R^{A,B}$. This
happens only when $y=y_{1}^{:Q^{B,B}}+\bbnum 0$ with some $y_{1}$.
So, the $\left(P,f\right)$-wedge relation implies $x_{1}\triangleright f^{\uparrow Q}=y_{1}\triangleright f^{\downarrow Q}$
\textemdash{} the $\left(Q,f\right)$-wedge relation. By assumption,
$Q$ has the weak pullback property. So, we can compute some $p_{1}^{:Q^{B,A}}$
such that $x_{1}=p_{1}\triangleright f^{\downarrow Q}$ and $y_{1}=p_{1}\triangleright f^{\uparrow Q}$.
If we now define $p\triangleq p_{1}+\bbnum 0$, we will have $p\triangleright f^{\downarrow P}=x$
and $p\triangleright f^{\uparrow P}=y$, and the square diagram is
completed.

The case $x\triangleq\bbnum 0+x_{2}^{:R^{A,A}}$ is proved similarly,
by using the weak pullback property of $R$. %
\begin{comment}
\textbf{(f)} Rewrite the wedge relation for $x^{:S^{A,A,P^{A,A}}}$
and $y^{:S^{B,B,P^{B,B}}}$ by using the explicit liftings to $S$:
\[
x\triangleright f^{\uparrow S^{A,\bullet,P^{A,A}}}\bef\big(\overline{f^{\uparrow P^{A,\bullet}}}\big)^{\uparrow S^{A,B,\bullet}}=y\triangleright f^{\downarrow S^{\bullet,B,P^{B,B}}}\bef\big(\overline{f^{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{A,B,\bullet}}\quad.
\]
Due to the commutativity law of $S$, we may exchange the order of
compositions here:
\[
f^{\uparrow S^{A,\bullet,P^{A,B}}}\big(x\triangleright(f^{\uparrow P^{A,\bullet}})^{\uparrow S^{A,A,\bullet}}\big)=y\triangleright\big(\overline{f^{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{B,B,\bullet}}\triangleright f^{\downarrow S^{\bullet,B,P^{A,B}}}\quad.
\]
Now we can use the weak pullback property of $S$ to obtain some $z:S^{B,A,P^{A,B}}$
such that
\[
x\triangleright\big(\overline{f^{\uparrow P^{A,\bullet}}}\big)^{\uparrow S^{A,A,\bullet}}=z\triangleright f^{\downarrow S^{\bullet,A,P^{A,B}}}\text{ and }y\triangleright\big(\overline{f^{\downarrow P^{\bullet,B}}}\big)^{\uparrow S^{B,B,\bullet}}=z\triangleright f^{\uparrow S^{B,\bullet,P^{A,B}}}\quad.
\]
We need to produce a value $t:S^{B,A,P^{B,A}}$ such that 
\[
x=t\triangleright f^{\downarrow S^{\bullet,A,P^{B,A}}}\bef\big(\overline{f^{\downarrow P^{\bullet,A}}}\big)^{\uparrow S^{A,A,\bullet}}\text{ and }y=t\triangleright f^{\uparrow S^{B,\bullet,P^{B,A}}}\bef\big(\overline{f^{\uparrow P^{B,\bullet}}}\big)^{\uparrow S^{B,B,\bullet}}\quad.
\]
\end{comment}
$\square$

The list of constructions in Statement~\ref{subsec:Statement-weak-pullback-property}
does not include the recursive type construction and limits the function
arguments to types not involving $A$. It remains an open question
whether, say, a suitably limited form of the recursive type construction
would produce new profunctors having the weak pullback property.

\subsection{Strong dinaturality derived from relational parametricity\label{subsec:Strong-dinaturality-derived-from-parametricity}}

To find out which fully parametric functions are strongly dinatural,
we will perform structural analysis on the functions\textsf{'} type signatures.
The plan\footnote{The derivations in this section are based on talk slides here: \texttt{\href{https://www.ioc.ee/~tarmo/tday-voore/vene-slides.pdf}{https://www.ioc.ee/$\sim$tarmo/tday-voore/vene-slides.pdf}}}
is to show that a fully parametric function $g:P^{A,A}\rightarrow Q^{A,A}$
will satisfy the strong dinaturality law,
\[
p_{1}^{:P^{A,A}}\triangleright f^{\uparrow P}=p_{2}^{:P^{B,B}}\triangleright f^{\downarrow P}\quad\Rightarrow\quad g(p_{1})\triangleright f^{\uparrow Q}=g(p_{2})\triangleright f^{\downarrow Q}\quad,
\]
if the profunctors $P^{A,A}$ and $Q^{A,A}$ are built up via a certain
subset of type constructions. 

The starting point is to note that $g$ satisfies the relational naturality
law with the relation $\left<f\right>$:
\[
(p_{1}^{:P^{A,A}},p_{2}^{:P^{B,B}})\in\left<f\right>^{\updownarrow P}\quad\Rightarrow\quad(g(p_{1}),g(p_{2}))\in\left<f\right>^{\updownarrow Q}\quad.
\]
We will prove the strong dinaturality law of $g$ if we show that
the relation $(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}$ follows
from the $\left(P,f\right)$-wedge relation for $(p_{1},p_{2})$,
and that the $\left(P,f\right)$-wedge relation for $(g(p_{1}),g(p_{2}))$
follows from the relation $(g(p_{1}),g(p_{2}))\in\left<f\right>^{\updownarrow Q}$.
We will then have proved the following chain of implications:
\[
\big(p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}\big)\Rightarrow\big((p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\big)\Rightarrow\big((g(p_{1}),g(p_{2}))\in\left<f\right>^{\updownarrow Q}\big)\Rightarrow\big(g(p_{1})\triangleright f^{\uparrow Q}=g(p_{2})\triangleright f^{\downarrow Q}\big)\quad.
\]
Let us call the first of these implications the \textsf{``}post-wedge property\textsf{''}
of the profunctor $P$ and the last the \textsf{``}pre-wedge\textsf{''} property of
the profunctor $Q$:

\subsubsection{Definition \label{subsec:Definition-pre-post-wedge-property}\ref{subsec:Definition-pre-post-wedge-property}}

A profunctor $P$ has the \textbf{pre-wedge property} if, for any
$f^{:A\rightarrow B}$, $p_{1}^{:P^{A,A}}$, and $p_{2}^{:P^{B,B}}$:
\[
\text{if }(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\text{ then }p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}\quad.
\]
A profunctor $P$ has the \textbf{post-wedge property} if, for any
$f^{:A\rightarrow B}$, $p_{1}^{:P^{A,A}}$, and $p_{2}^{:P^{B,B}}$:
\[
\text{if }p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}\text{ then }(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\quad.
\]

In other words, the pre-wedge property means that the $\left(P,f\right)$-wedge
relation\index{wedge relation} follows from the lifted relation $\left<f\right>^{\updownarrow P}$.
The post-wedge property means that the lifted relation $\left<f\right>^{\updownarrow P}$
follows from the $\left(P,f\right)$-wedge relation.

It turns out that there is a connection between the pre-wedge property
and the weak pullback property (see Definition~\ref{subsec:Definition-weak-pullback-property}):

\subsubsection{Statement \label{subsec:Statement-weak-pullback-entails-pre-wedge}\ref{subsec:Statement-weak-pullback-entails-pre-wedge}}

If a profunctor $P$ is constructed as in Statement~\ref{subsec:Statement-weak-pullback-property}
(which assures the weak pullback property), $P$ will also have the
pre-wedge property.

\subparagraph{Proof}

We use induction on the top-level type construction within $P$.

\paragraph{Constant type}

Here $P^{X,Y}$ does not depend on one of its type parameters ($P^{X,Y}\triangleq Q^{Y}$
or $P^{X,Y}\triangleq R^{X}$). We know from Statement~\ref{subsec:Statement-weak-pullback-property}(a,
b) that $P$ will then have the weak pullback property.

Suppose $P^{X,Y}\triangleq Q^{Y}$ where $Q$ is a functor. By Statement~\ref{subsec:Statement-lifting-function-relation-covariant}(a),
we then have:
\[
(x^{:Q^{A}},y^{:Q^{B}})\in\left<f\right>^{\updownarrow P}\text{ means }x\triangleright f^{\uparrow Q}=y\quad.
\]
This is equivalent to the $\left(P,f\right)$-wedge relation for $(x,y)$
because $f^{\uparrow P}=f^{\uparrow Q}$ and $f^{\downarrow P}=\text{id}$.

Similarly, if $P^{X,Y}\triangleq R^{X}$ where $R$ is a contrafunctor,
we have by Statement~\ref{subsec:Statement-lifting-function-relation-covariant}(b):
\[
(x^{:R^{A}},y^{:R^{B}})\in\left<f\right>^{\updownarrow P}\text{ means }x=y\triangleright f^{\downarrow R}\quad.
\]
This is equivalent to the $\left(P,f\right)$-wedge relation for $(x,y)$
because $f^{\uparrow P}=\text{id}$ and $f^{\downarrow P}=f^{\downarrow R}$.

So, we have shown that the $\left(P,f\right)$-wedge relation for
$(x,y)$ is \emph{equivalent} to the relation $(x,y)\in\left<f\right>^{\updownarrow P}$.
This equivalence means that $P$ has both the pre-wedge and the post-wedge
properties.

\paragraph{Products}

Here $P^{X,Y}\triangleq K^{X,Y}\times L^{X,Y}$ with some profunctors
$K$ and $L$ that already have the pre-wedge property. We need to
show that $P$ has the pre-wedge property:
\[
\text{if }(k_{1}^{:K^{A,A}}\times l_{1}^{:L^{A,A}},k_{2}^{:K^{B,B}}\times l_{2}^{:L^{B,B}})\in\left<f\right>^{\updownarrow P}\text{ then }(k_{1}\times l_{1})\triangleright f^{\uparrow P}=(k_{2}\times l_{2})\triangleright f^{\downarrow P}\quad.
\]
Using the definitions of the liftings $^{\updownarrow P}$, $^{\uparrow P}$,
and $^{\downarrow P}$, we rewrite the above condition as:
\[
\text{if }(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\text{ and }(l_{1},l_{2})\in\left<f\right>^{\updownarrow L}\text{ then }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\text{ and }l_{1}\triangleright f^{\uparrow L}=l_{2}\triangleright f^{\downarrow L}\quad.
\]
This is the same as the conjunction of the pre-wedge properties of
$K$ and $L$.

\paragraph{Co-products}

Here $P^{X,Y}\triangleq K^{X,Y}+L^{X,Y}$ with some profunctors $K$
and $L$ that already have the pre-wedge property. We need to show
that $P$ has the pre-wedge property:
\[
\text{if }(p_{1}^{:K^{A,A}+L^{A,A}},p_{2}^{:K^{B,B}+L^{B,B}})\in\left<f\right>^{\updownarrow P}\text{ then }p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}\quad.
\]
By Definition~\ref{subsec:Definition-relational-lifting}(d), two
values $p_{1}^{:K^{A,A}+L^{A,A}}$ and $p_{2}^{:K^{B,B}+L^{B,B}}$
can be in the relation $\left<f\right>^{\updownarrow P}$ only if
both $p_{1}$ and $p_{2}$ are in the same part of the disjunction
$K+L$. If both $p_{1}$ and $p_{2}$ are in the left part of the
disjunction, we have $p_{1}\triangleq k_{1}^{:K^{A,A}}+\bbnum 0$
and $p_{2}\triangleq k_{2}^{:K^{B,B}}+\bbnum 0$. Then the pre-wedge
property of $P$ is:
\[
\text{if }(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\text{ then }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\quad.
\]
This holds because it is just the pre-wedge property of $K$. Similarly,
we prove that if both $p_{1}$ and $p_{2}$ are in the right part
of the disjunction then the pre-wedge property of $P$ is reduced
to the pre-wedge property of $L$.

\paragraph{Function types}

Here we have $P^{X,Y}=Z\rightarrow L^{X,Y}$ where $L$ has the weak
pullback property and $Z$ is a fixed type. The liftings to $P$ are
expressed as:
\begin{align*}
 & p_{1}^{:P^{A,A}}\triangleright f^{\uparrow P}=p_{1}\bef f^{\uparrow L}\quad,\quad\quad p_{2}^{:P^{B,B}}\triangleright f^{\downarrow P}=p_{2}\bef f^{\downarrow L}\quad,\\
 & (p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\text{ when }\forall z^{:Z}.\,(p_{1}(z),p_{2}(z))\in\left<f\right>^{\updownarrow L}\quad.
\end{align*}
We need to prove the pre-wedge property of $P$:
\[
\text{if }(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\text{ then }p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}\quad\text{or equivalently}:\quad p_{1}\bef f^{\uparrow L}=p_{2}\bef f^{\downarrow L}\quad.
\]
Choose arbitrary $z^{:Z}$, $p_{1}^{:P^{A,A}}$, and $p_{2}^{:P^{A,A}}$
satisfying $(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}$. It
follows that $(p_{1}(z),p_{2}(z))\in\left<f\right>^{\updownarrow L}$.
By the inductive assumption, the profunctor $L$ already has the pre-wedge
property:
\[
\text{if }(l_{1},l_{2})\in\left<f\right>^{\updownarrow L}\text{ then }l_{1}\triangleright f^{\uparrow L}=l_{2}\triangleright f^{\downarrow L}\quad.
\]
Substituting $l_{1}\triangleq p_{1}(z)$ and $l_{2}\triangleq p_{2}(z)$,
we obtain: 
\[
p_{1}(z)\triangleright f^{\uparrow L}=p_{2}(z)\triangleright f^{\downarrow L}\quad\text{or equivalently}:\quad z\triangleright p_{1}\bef f^{\uparrow L}=z\triangleright p_{2}\bef f^{\downarrow L}\quad.
\]
Since this holds for an arbitrary $z^{:Z}$, we have proved that $p_{1}\bef f^{\uparrow L}=p_{2}\bef f^{\downarrow L}$.

\subsubsection{Statement \label{subsec:Statement-post-wedge}\ref{subsec:Statement-post-wedge}}

A profunctor $P$ will have the post-wedge property if:

\textbf{(a)} The type expression $P^{X,Y}$ does not depend either
on $X$ or on $Y$.

\textbf{(b)} We have $P^{X,Y}\triangleq K^{X,Y}\times L^{X,Y}$, where
the profunctors $K$ and $L$ have the post-wedge property.

\textbf{(c)} We have $P^{X,Y}\triangleq K^{X,Y}+L^{X,Y}$, where the
profunctors $K$ and $L$ have the post-wedge property.

\textbf{(d)} We have $P^{X,Y}\triangleq K^{Y,X}\rightarrow L^{X,Y}$,
where the profunctor $K$ has the weak pullback property and the profunctor
$L$ has the post-wedge property.%
\begin{comment}
\textbf{(e)} We have a recursive type $P^{X,Y}\triangleq S^{X,Y,P^{X,Y}}$,
where $S^{X,Y,R}$ is contravariant in $X$ and covariant in $Y$
and $R$, and has the post-wedge property when viewed as a profunctor
with respect to $X$ and $Y$. Does the post-wedge really hold for
$P$?
\end{comment}


\subparagraph{Proof}

In each case, assuming the $\left(P,f\right)$-wedge relation for
$(x,y)$, we will show that $(x,y)\in\left<f\right>^{\updownarrow P}$. 

\textbf{(a)} The proof of Statement~\ref{subsec:Statement-weak-pullback-entails-pre-wedge}
also derived the post-wedge property of $P$.

\textbf{(b)} We need to show that $P$ has the post-wedge property:
\[
\text{if }(k_{1}^{:K^{A,A}}\times l_{1}^{:L^{A,A}})\triangleright f^{\uparrow P}=(k_{2}^{:K^{B,B}}\times l_{2}^{:L^{B,B}})\triangleright f^{\downarrow P}\text{ then }(k_{1}\times l_{1},k_{2}\times l_{2})\in\left<f\right>^{\updownarrow P}\quad.
\]
Using the definitions of the liftings $^{\updownarrow P}$, $^{\uparrow P}$,
and $^{\downarrow P}$, we rewrite the above condition as:
\[
\text{if }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\text{ and }l_{1}\triangleright f^{\uparrow L}=l_{2}\triangleright f^{\downarrow L}\text{ then }(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\text{ and }(l_{1},l_{2})\in\left<f\right>^{\updownarrow L}\quad.
\]
This is the same as the conjunction of the post-wedge properties of
$K$ and $L$. 

\textbf{(c)} We need to show that $P$ has the post-wedge property:
\[
\text{if }p_{1}^{:K^{A,A}+L^{A,A}}\triangleright f^{\uparrow P}=p_{2}^{:K^{B,B}+L^{B,B}}\triangleright f^{\downarrow P}\text{ then }(p_{1},p_{2})\in\left<f\right>^{\updownarrow P}\quad.
\]
The liftings $f^{\uparrow P}$ and $f^{\downarrow P}$ are defined
via the standard pattern-matching code for disjunctive types:
\[
f^{\uparrow P^{A,\bullet}}\triangleq\,\begin{array}{|c||cc|}
 & K^{A,B} & L^{A,B}\\
\hline K^{A,A} & f^{\uparrow K^{A,\bullet}} & \bbnum 0\\
L^{A,A} & \bbnum 0 & f^{\uparrow L^{A,\bullet}}
\end{array}\quad,\quad\quad f^{\downarrow P^{\bullet,B}}\triangleq\,\begin{array}{|c||cc|}
 & K^{A,B} & L^{A,B}\\
\hline K^{B,B} & f^{\downarrow K^{\bullet,B}} & \bbnum 0\\
L^{B,B} & \bbnum 0 & f^{\downarrow L^{\bullet,B}}
\end{array}\quad.
\]
Since both code matrices are diagonal, the two parts of the disjunctive
type $K+L$ do not mix. It follows that $p_{1}\triangleright f^{\uparrow P}=p_{2}\triangleright f^{\downarrow P}$
only if both $p_{1}$ and $p_{2}$ are in the same part of the disjunction
$K+L$. If both $p_{1}$ and $p_{2}$ are in the left part of the
disjunction, we have $p_{1}\triangleq k_{1}^{:K^{A,A}}+\bbnum 0$
and $p_{2}\triangleq k_{2}^{:K^{B,B}}+\bbnum 0$, and the $\left(P,f\right)$-wedge
relation for $p_{1}$ and $p_{2}$ reduces to the $\left(K,f\right)$-wedge
relation for $k_{1}$ and $k_{2}$:
\[
k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\quad.
\]
By Definition~\ref{subsec:Definition-relational-lifting}(d), the
values $p_{1}$ and $p_{2}$ will be in the relation $\left<f\right>^{\updownarrow P}$
only if $(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}$. Then the
post-wedge property of $P$ becomes:
\[
\text{if }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\text{ then }(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\quad.
\]
This holds because it is just the post-wedge property of $K$. Similarly,
we prove that if both $p_{1}$ and $p_{2}$ are in the right part
of the disjunction then the post-wedge property of $P$ is reduced
to the post-wedge property of $L$.

\textbf{(d)} Express the $\left(P,f\right)$-wedge relation and the
relation $\left<f\right>^{\updownarrow P}$ for $(x,y)$ through liftings
to $K$ and $L$:
\begin{align*}
 & x^{:K^{A,A}\rightarrow L^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=y^{:K^{B,B}\rightarrow L^{B,B}}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{or equivalently}:\quad f^{\downarrow K^{\bullet,A}}\bef x\bef f^{\uparrow L^{A,\bullet}}=f^{\uparrow K^{B,\bullet}}\bef y\bef f^{\downarrow L^{\bullet,B}}\quad,\\
 & (x^{:K^{A,A}\rightarrow L^{A,A}},y^{:K^{B,B}\rightarrow L^{B,B}})\in\left<f\right>^{\updownarrow P}\text{ means if }(k_{1}^{:K^{A,A}},k_{2}^{:K^{B,B}})\in\left<f\right>^{\updownarrow K}\text{ then }(x(k_{1}),y(k_{2}))\in\left<f\right>^{\updownarrow L}\quad.
\end{align*}
Fix any $k_{1}^{:K^{A,A}}$ and $k_{2}^{:K^{B,B}}$ such that $(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}$;
we need to prove that $(x(k_{1}),y(k_{2}))\in\left<f\right>^{\updownarrow L}$.

By Statement~\ref{subsec:Statement-weak-pullback-entails-pre-wedge},
$K$ has the pre-wedge property in addition to the weak pullback property:
\begin{align*}
 & \text{if }(k_{1}^{:K^{A,A}},k_{2}^{:K^{B,B}})\in\left<f\right>^{\updownarrow K}\text{ then }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\quad;\\
 & \text{if }k_{1}^{:K^{A,A}}\triangleright f^{\uparrow K}=k_{2}^{:K^{B,B}}\triangleright f^{\downarrow K}\text{ then }\exists k_{0}^{:K^{B,A}}\text{ such that }k_{1}=k_{0}\triangleright f^{\downarrow K^{\bullet,A}}\text{ and }k_{2}=k_{0}\triangleright f^{\uparrow K^{B,\bullet}}\quad.
\end{align*}
So, we may assume that a suitable value $k_{0}^{:K^{B,A}}$ is known.
Apply the $\left(P,f\right)$-wedge relation to $k_{0}$:
\[
k_{0}\triangleright f^{\downarrow K^{\bullet,A}}\bef x\bef f^{\uparrow L^{A,\bullet}}=k_{0}\triangleright f^{\uparrow K^{B,\bullet}}\bef y\bef f^{\downarrow L^{\bullet,B}}\quad\text{or equivalently}:\quad k_{1}\triangleright x\triangleright f^{\uparrow L}=k_{2}\triangleright y\triangleright f^{\downarrow L}\quad.
\]
 It is also given that $L$ has the post-wedge property:
\[
\text{if }l_{1}^{:L^{A,A}}\triangleright f^{\uparrow L}=l_{2}^{:L^{B,B}}\triangleright f^{\downarrow L}\text{ then }(l_{1},l_{2})\in\left<f\right>^{\updownarrow L}\quad.
\]
We now set $l_{1}\triangleq k_{1}\triangleright x$ and $l_{2}\triangleq k_{2}\triangleright y$
to obtain $(x(k_{1}),y(k_{2}))\in\left<f\right>^{\updownarrow L}$
as required.

\subsubsection{Statement \label{subsec:Statement-pre-wedge}\ref{subsec:Statement-pre-wedge}}

A profunctor $P$ will have the pre-wedge property if:

\textbf{(a)} The type expression $P^{X,Y}$ does not depend either
on $X$ or on $Y$.

\textbf{(b)} We have $P^{X,Y}\triangleq K^{X,Y}\times L^{X,Y}$, where
the profunctors $K$ and $L$ have the pre-wedge property.

\textbf{(c)} We have $P^{X,Y}\triangleq K^{X,Y}+L^{X,Y}$, where the
profunctors $K$ and $L$ have the pre-wedge property.

\textbf{(d)} We have $P^{X,Y}\triangleq K^{Y,X}\rightarrow L^{X,Y}$,
where the profunctor $K$ has the \emph{post}-wedge property and the
profunctor $L$ has the pre-wedge property.

\begin{comment}
\textbf{(e)} We have a recursive type $P^{X,Y}\triangleq S^{X,Y,P^{X,Y}}$,
where $S^{X,Y,R}$ is contravariant in $X$ and covariant in $Y$
and $R$, and has the pre-wedge property when viewed as a profunctor
with respect to $X$ and $Y$. \textendash{} Not sure if this is correct.
\end{comment}


\subparagraph{Proof}

In each case, we will verify the $\left(P,f\right)$-wedge relation
for $(x,y)$ assuming that $(x,y)\in\left<f\right>^{\updownarrow P}$. 

\textbf{(a)} The proof of Statement~\ref{subsec:Statement-weak-pullback-entails-pre-wedge}
derived also the pre-wedge property of $P$.

\textbf{(b)} and \textbf{(c)} were already proved as part of the proof
of Statement~\ref{subsec:Statement-weak-pullback-entails-pre-wedge}
for product and co-product types.

\textbf{(d)} The $\left(P,f\right)$-wedge relation for $(x,y)$ is
an equation for functions of type $K^{B,A}\rightarrow L^{A,B}$:
\[
x^{:K^{A,A}\rightarrow L^{A,A}}\triangleright f^{\uparrow P^{A,\bullet}}=y^{:K^{B,B}\rightarrow L^{B,B}}\triangleright f^{\downarrow P^{\bullet,B}}\quad\text{or equivalently}:\quad f^{\downarrow K^{\bullet,A}}\bef x\bef f^{\uparrow L^{A,\bullet}}=f^{\uparrow K^{B,\bullet}}\bef y\bef f^{\downarrow L^{\bullet,B}}\quad.
\]
We will apply both sides of this equation to an arbitrary $k_{0}^{:P^{B,A}}$
and show that the results are equal:
\[
k_{0}^{:P^{B,A}}\triangleright f^{\downarrow K^{\bullet,A}}\bef x\bef f^{\uparrow L^{A,\bullet}}\overset{?}{=}k_{0}\triangleright f^{\uparrow K^{B,\bullet}}\bef y\bef f^{\downarrow L^{\bullet,B}}\quad.
\]
We may use the relation $\left<f\right>^{\updownarrow P}$ for $(x,y)$:
\[
(x^{:K^{A,A}\rightarrow L^{A,A}},y^{:K^{B,B}\rightarrow L^{B,B}})\in\left<f\right>^{\updownarrow P}\text{ means if }(k_{1}^{:K^{A,A}},k_{2}^{:K^{B,B}})\in\left<f\right>^{\updownarrow K}\text{ then }(x(k_{1}),y(k_{2}))\in\left<f\right>^{\updownarrow L}\quad.
\]
Define $k_{1}\triangleq k_{0}\triangleright f^{\downarrow K^{\bullet,A}}$,
$k_{2}\triangleq k_{0}\triangleright f^{\uparrow K^{B,\bullet}}$
and note that the $\left(K,f\right)$-wedge relation for $(k_{1},k_{2})$
holds due to the commutativity law of the profunctor $K$:
\[
k_{1}\triangleright f^{\uparrow K^{A,\bullet}}=k_{2}\triangleright f^{\downarrow K^{\bullet,B}}\quad\text{or equivalently}:\quad k_{0}\triangleright f^{\downarrow K^{\bullet,A}}\bef f^{\uparrow K^{A,\bullet}}=k_{0}\triangleright f^{\uparrow K^{B,\bullet}}\bef f^{\downarrow K^{\bullet,B}}\quad.
\]
Since it is given that $K$ has the post-wedge property, we have:
\[
\text{if }k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\text{ then }(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\quad.
\]
So, the values $k_{1}$ and $k_{2}$ are in the relation $\left<f\right>^{\updownarrow K}$
and satisfy the precondition of the relation $\left<f\right>^{\updownarrow P}$.
So, we get $(x(k_{1}),y(k_{2}))\in\left<f\right>^{\updownarrow L}$.
It is given that $L$ has the pre-wedge property:
\[
\text{if }(l_{1}^{:L^{A,A}},l_{2}^{:L^{B,B}})\in\left<f\right>^{\updownarrow L}\text{ then }l_{1}\triangleright f^{\uparrow L}=l_{2}\triangleright f^{\downarrow L}\quad.
\]
It follows that $x(k_{1})$ and $y(k_{2})$ are in the $\left(L,f\right)$-wedge
relation. Using the definitions of $k_{1}$ and $k_{2}$, we may now
rewrite that relation as
\[
x(k_{0}\triangleright f^{\downarrow K})\triangleright f^{\uparrow L}=y(k_{0}\triangleright f^{\uparrow K})\triangleright f^{\downarrow L}\quad\text{or equivalently}:\quad k_{0}\triangleright f^{\downarrow K}\bef x\bef f^{\uparrow L}=k_{0}\triangleright f^{\uparrow K}\bef y\bef f^{\downarrow L}\quad.
\]
This is the property we needed to prove.

\subsubsection{Statement \label{subsec:Statement-post-pre-wedge}\ref{subsec:Statement-post-pre-wedge}}

If $P^{X,Y}\triangleq K^{Y,X}\rightarrow L^{X,Y}$, where the profunctor
$K$ has the post-wedge property and the profunctor $L$ has the pre-wedge
property, then any fully parametric function $g:\forall A.\,P^{A,A}$
is strongly dinatural.

\subparagraph{Proof}

The strong dinaturality law of $g$ is written as:
\[
\text{when }\quad k_{1}^{:K^{A,A}}\triangleright f^{\uparrow K}=k_{2}^{:K^{B,B}}\triangleright f^{\downarrow K}\quad\text{ then }\quad g(k_{1})\triangleright f^{\uparrow L}=g(k_{2})\triangleright f^{\downarrow L}\quad.
\]
Since $K$ has the post-wedge property, we have:
\[
\big(k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\big)\quad\Rightarrow\quad(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\quad.
\]
The full parametricity of $g$ means that the relational naturality
law holds:
\[
(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\quad\Rightarrow\quad(g(k_{1}),g(k_{2}))\in\left<f\right>^{\updownarrow L}\quad.
\]
The profunctor $L$ has the pre-wedge property, so
\[
(g(k_{1}),g(k_{2}))\in\left<f\right>^{\updownarrow L}\quad\Rightarrow\quad\big(g(k_{1})\triangleright f^{\uparrow L}=g(k_{2})\triangleright f^{\downarrow L}\big)\quad.
\]
This gives us a chain of implications that proves the strong dinaturality
law of $g$:
\[
\big(k_{1}\triangleright f^{\uparrow K}=k_{2}\triangleright f^{\downarrow K}\big)\Rightarrow\big((k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\big)\Rightarrow\big((g(k_{1}),g(k_{2}))\in\left<f\right>^{\updownarrow L}\big)\Rightarrow\big(g(k_{1})\triangleright f^{\uparrow L}=g(k_{2})\triangleright f^{\downarrow L}\big)\quad.
\]
$\square$

The results of Statement~\ref{subsec:Statement-post-pre-wedge} hold
only for type signatures $\forall A.\,K^{A,A}\rightarrow L^{A,A}$
with profunctors $K$ and $L$ having certain specific properties.
Nevertheless, a broad range of practically encountered functions have
type signatures of this form. Here are some examples.

\subsubsection{Example \label{subsec:Example-strong-dinaturality-for-some-type-signatures}\ref{subsec:Example-strong-dinaturality-for-some-type-signatures}\index{solved examples}}

Use Statement~\ref{subsec:Statement-post-pre-wedge} with an appropriately
chosen profunctor $P^{X,Y}$ to show that strong dinaturality holds
for all fully parametric functions with the following type signatures:

\textbf{(a)} $\forall A.\,G^{A}\rightarrow H^{A}\quad.$

\textbf{(b)} $\forall A.\,F^{A}\rightarrow G^{A}\rightarrow H^{A}\quad.$

\textbf{(c)} $\forall A.\,(F^{A}\rightarrow G^{A})\rightarrow H^{A}\quad.$

Here each of $F^{\bullet}$, $G^{\bullet}$, $H^{\bullet}$ is independently
either a functor or a contrafunctor. 

\subparagraph{Solution}

\textbf{(a)} To represent the type $\forall A.\,G^{A}\rightarrow H^{A}$
as $\forall A.\,P^{A,A}$ via a profunctor $P^{X,Y}$, we need to
consider the four cases where $G$ and $H$ are either covariant or
contravariant. If $G$ and $H$ are both covariant, we define $P^{X,Y}\triangleq G^{X}\rightarrow H^{Y}$.
If $G$ is covariant but $H$ is contravariant, $P^{X,Y}\triangleq G^{X}\rightarrow H^{X}$.
If $G$ is contravariant but $H$ is covariant, $P^{X,Y}\triangleq G^{Y}\rightarrow H^{Y}$.
If $G$ and $H$ are both contravariant, $P^{X,Y}\triangleq G^{Y}\rightarrow H^{X}$.
In each case, $P^{X,Y}=K^{Y,X}\rightarrow L^{X,Y}$ where $K$ and
$L$ depend on only one of their type parameters. By Statement~\ref{subsec:Statement-post-wedge}(a)
and Statement~\ref{subsec:Statement-pre-wedge}(a), both $K$ and
$L$ with have the pre-wedge and the post-wedge properties in all
cases. So, the profunctors $K$ and $L$ satisfy the conditions of
Statement~\ref{subsec:Statement-post-pre-wedge}.

\textbf{(b)} Since the function arrows group to the right, the type
$\forall A.\,F^{A}\rightarrow G^{A}\rightarrow H^{A}$ is of the form
$\forall A.\,K^{A,A}\rightarrow L^{A,A}$ where $K^{A,A}\cong F^{A}$
and $L^{A,A}\cong G^{A}\rightarrow H^{A}$. Depending on whether $F$,
$G$, $H$ are covariant or contravariant, we need to define the profunctors
$K$, $L$, and $P$ differently. The profunctor $L$ will be of the
form $L^{X,Y}\triangleq M^{Y,X}\rightarrow N^{X,Y}$ with some profunctors
$M$ and $N$. For instance, if $F$, $G$, $H$ are all covariant,
we need to define $P^{X,Y}\triangleq F^{X}\rightarrow G^{X}\rightarrow H^{Y}$,
$M^{X,Y}\triangleq G^{Y}$, $N^{X,Y}\triangleq H^{Y}$. Reasoning
as in the proof of part \textbf{(a)}, we conclude that in all cases
the profunctors $K$, $M$, and $N$ will depend only on one of their
type parameters. So, the profunctors $K$, $M$, and $N$ will have
both the pre-wedge and the post-wedge properties. By Statement~\ref{subsec:Statement-pre-wedge}(d),
the profunctor $L$ will have the pre-wedge property. So, the profunctors
$K$ and $L$ satisfy the conditions of Statement~\ref{subsec:Statement-post-pre-wedge}.

\textbf{(c)} The type $\forall A.\,(F^{A}\rightarrow G^{A})\rightarrow H^{A}$
is of the form $\forall A.\,K^{A,A}\rightarrow L^{A,A}$ with the
profunctors $K^{A,A}\cong F^{A}\rightarrow G^{A}$ and $L^{A,A}\cong H^{A}$.
We need to define $K^{X,Y}\triangleq M^{Y,X}\rightarrow N^{X,Y}$
with some profunctors $M$ and $N$, depending on the covariance of
$F$, $G$, and $H$. Reasoning as before, we find that the profunctors
$L$, $M$, $N$ must have both the pre-wedge and the post-wedge properties.
In addition, Example~\ref{subsec:Example-weak-pullback-property-1}(a)
shows that $M$ has the weak pullback property. The post-wedge property
of $K$ is then established via Statement~\ref{subsec:Statement-post-wedge}(d).
So, the profunctors $K$ and $L$ satisfy the conditions of Statement~\ref{subsec:Statement-post-pre-wedge}.
$\square$

Example~\ref{subsec:Example-strong-dinaturality-for-some-type-signatures}(c)
is used in Statement~\ref{subsec:Statement-Church-encoding-recursive-type-covariant}
for the Church encoding of recursive types.

In some cases, strong dinaturality holds for type signatures not covered
by Statement~\ref{subsec:Statement-post-pre-wedge}:

\subsubsection{Statement \label{subsec:Statement-functor-post-pre-wedge}\ref{subsec:Statement-functor-post-pre-wedge}}

If $L^{\bullet}$ is a (covariant) functor and $K^{\bullet,\bullet}$
is a profunctor having both the pre-wedge and the post-wedge properties
then the profunctor $P^{X,Y}\triangleq L^{K^{X,Y}}$ has the post-wedge
property.

\subparagraph{Proof}

Assume any two values $x^{:L^{K^{A,A}}}$ and $y^{:L^{K^{B,B}}}$
in the $\left(P,f\right)$-wedge relation:
\[
x\triangleright f^{\uparrow K^{A,\bullet}\uparrow L}=y\triangleright f^{\downarrow K^{\bullet,B}\uparrow L}\quad.
\]
We need to show that $\left(x,y\right)\in\left<f\right>^{\updownarrow P}$.
Lifting a relation to $P$ means first lifting to $K$ and then to
$L$, so $\left<f\right>^{\updownarrow P}=\left<f\right>^{\updownarrow K\updownarrow L}$.
Since $K$ has both the pre-wedge and the post-wedge property, the
$\left(K,f\right)$-wedge relation is equivalent to the relation $\left<f\right>^{\updownarrow K}$.
The $\left(K,f\right)$-wedge relation has the form of a pullback
relation, so we can write $\left<f\right>^{\updownarrow K}$ as:
\[
(k_{1},k_{2})\in\left<f\right>^{\updownarrow K}\text{ means }k_{1}\triangleright f^{\uparrow K^{A,\bullet}}=k_{2}\triangleright f^{\downarrow K^{B,\bullet}}\quad.
\]
By Statement~\ref{subsec:Statement-pullback-as-composition}(b),
lifting a pullback relation to $L$ gives again a pullback relation:
\[
\left(x,y\right)\in\left<f\right>^{\updownarrow K\updownarrow L}\text{ means }x\triangleright f^{\uparrow K^{A,\bullet}\uparrow L}=y\triangleright f^{\downarrow K^{B,\bullet}\uparrow L}\quad.
\]
So, $\left(x,y\right)$ are in the relation $\left<f\right>^{\updownarrow K\updownarrow L}=\left<f\right>^{\updownarrow P}$.
$\square$

We can now show that the function \lstinline!foldFn! used in Section~\ref{subsec:Equivalence-of-foldLeft,foldMap,reduce,and-toList}
is strongly dinatural.

\subsubsection{Example \label{subsec:Example-strong-dinaturality-proof-of-foldFn-law}\ref{subsec:Example-strong-dinaturality-proof-of-foldFn-law}\index{solved examples}}

Show that any fully parametric function with type $\forall A.\,L^{A\rightarrow A}\rightarrow A\rightarrow A$
is strongly dinatural when $L$ is a (covariant) functor.

\subparagraph{Solution}

The type can be shortened to $L^{E^{A}}\rightarrow E^{A}$ if we denote
$E^{A}\triangleq A\rightarrow A$. Define the profunctor $K^{X,Y}\triangleq X\rightarrow Y$
so that $E^{A}=K^{A,A}$. The profunctor $K$ has the post-wedge property
by Statement~\ref{subsec:Statement-post-wedge}(a) and (d). It also
has the pre-wedge property by Statement~\ref{subsec:Statement-pre-wedge}(a)
and (d). Then Statement~\ref{subsec:Statement-functor-post-pre-wedge}
means that the profunctor $L^{K^{X,Y}}$ has the post-wedge property.
We can now use Statement~\ref{subsec:Statement-functor-post-pre-wedge}
to find that $\phi$ satisfies the strong dinaturality law. 

\section{Summary}

We have proved several results that apply to all fully parametric
code:
\begin{itemize}
\item The lifting methods of any fully parametric bifunctor, profunctor,
or bi-contrafunctor obey the commutativity law such as Eq.~(\ref{eq:f-fmap-fmap-bifunctor-commutativity}). 
\item Any function of type $P^{A,A}\rightarrow Q^{A,A}$ (where $P$, $Q$
are profunctors) obeys the dinaturality law~(\ref{eq:dinaturality-law-for-dinatural-transformations-of-profunctors}),
which allows us to derive a specific naturality law for any fully
parametric function. The form of the law depends only on the function\textsf{'}s
type signature and applies to all fully parametric implementations
of that type signature. 
\item There is only one fully parametric and lawful implementation of the
\lstinline!Functor! or \lstinline!Contrafunctor! typeclass instances
for a given type constructor. (For many other typeclasses, such as
\lstinline!Filterable! or \lstinline!Monad!, many type constructors
have several inequivalent and lawful typeclass instances.) The unique
lawful implementations are defined by the procedures shown in Chapter~\ref{chap:Functors,-contrafunctors,-and}.
\item Any fully parametric expression $t:\forall A.\,Q^{A}$ satisfies the
relational naturality law~(\ref{eq:relational-naturality-law}).
In general, the relational naturality law expresses a property of
\emph{relations} rather than functions and is not equivalent to any
equation satisfied by the function $t$. 
\item If the type signature of $t$ satisfies the conditions of Statement~\ref{subsec:Statement-post-pre-wedge},
the function $t$ satisfies the \textsf{``}strong dinaturality\textsf{''} law, which
is simpler to use than the relational naturality law.
\end{itemize}
Section~\ref{subsec:Statement-parametricity-theorem-via-dinaturality}
derives the dinaturality law of $t$ by structural analysis of the
code of $t$, but the proof depends on Statement~\ref{subsec:Statement-dinatural-transformation-composition},
whose proof in this book is incomplete. We also did not derive the
dinaturality law of $t$ directly from the relational parametricity
law. It appears that the dinaturality law cannot be derived purely
by structural analysis of code but requires more complicated methods
that are beyond the scope of this book.\footnote{Janis Voigtl\"ander\index{Janis@Janis Voigtl\"ander} gave a general
derivation of the dinaturality law from the relational parametricity
theorem in the paper \texttt{\href{https://arxiv.org/pdf/1908.07776.pdf}{https://arxiv.org/pdf/1908.07776.pdf}}.
Another derivation of the dinaturality law, based on obtaining the
syntactic normal form of dinatural transformations, was given by Joachim
de Lataillade\index{Joachim de Lataillade} in the paper \textsf{``}Dinatural
terms in System $F$\textsf{''}: see \texttt{\href{https://www.irif.fr/~delatail/dinat.pdf}{https://www.irif.fr/$\sim$delatail/dinat.pdf}}} 

\chapter{Solutions of some exercises}

\subsection*{Chapter \ref{chap:1-Values,-types,-expressions,}}

\subsubsection*{Exercise \ref{subsec:ch1-aggr-Exercise-1}}

~
\begin{lstlisting}
def at(n: Double, maxN: Int) = (0 to maxN)
  .map { k => 1.0*(1 - k % 2 * 2) / (2 * k + 1) / math.pow(n, 2 * k + 1) }
  .sum
def p(n: Int) = 16 * at(5, n) - 4 * at(239, n)

scala> p(12)
res0: Double = 3.141592653589794
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch1-aggr-Exercise-2}}

~
\begin{lstlisting}
def isPrime(n: Int) = (2 to n - 1).takeWhile(k => k * k <= n).forall(k => n % k != 0)
def ep(n: Int): Double = (2 to n)
  .filter(isPrime)
  .map  { k => 1.0 / (1.0 - 1.0 / k / k / k /k) }
  .product
val pi = 3.1415926535897932

scala> ep(100); pi*pi*pi*pi/90
res0_0: Double = 1.0823231553280295
res0_1: Double = 1.082323233711138 
\end{lstlisting}


\subsection*{Chapter \ref{chap:2-Mathematical-induction}}

\subsubsection*{Exercise \ref{tuples-Exercise-10}}

~
\begin{lstlisting}
numsLists.map(_.sortBy(- _).take(3))
\end{lstlisting}


\subsubsection*{Exercise \ref{tuples-Exercise-11}}

~
\begin{lstlisting}
a.flatMap(x => b.map(y => (x, y)))
\end{lstlisting}


\subsubsection*{Exercise \ref{tuples-Exercise-12}}

~
\begin{lstlisting}
def payments[Person, Amount](data: Seq[Map[Person, Amount]]): Map[Person, Seq[Amount]] =
  data.flatMap(_.toSeq).groupBy(_._1).mapValues(_.map(_._2))
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-2.2-foldleft-5}}

~
\begin{lstlisting}
def batching[A](xs: Seq[A], size: Int): Seq[Seq[A]] = {  
  type Acc = (Seq[Seq[A]], Seq[A], Int)
  val init: Acc = ((Seq(), Seq(), 0))
  val (result, rem, _) = xs.foldLeft(init){ case ((seq, rem, len), x) =>
    val newLen = len + 1
    if (newLen > size) (seq ++ Seq(rem), Seq(x), 1)
    else (seq, rem ++ Seq(x), newLen)
  }
  result ++ Seq(rem)
}
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-2.2-foldleft-5-1}}

~
\begin{lstlisting}
def weightBatching[A](xs: Seq[A], maxW: Double)(w: A => Double): Seq[Seq[A]] = {  
  type Acc = (Seq[Seq[A]], Seq[A], Double)
  val init: Acc = ((Seq(), Seq(), 0.0))
  val (result, rem, _) = xs.foldLeft(init) { case ((seq, rem, weight), x) =>
    val wx = w(x)
    if (wx > maxW) (seq ++ Seq(rem, Seq(x)), Seq(), 0.0)
    else {
      val newWeight = weight + wx
      if (newWeight > maxW) (seq ++ Seq(rem), Seq(x), wx)
      else (seq, rem ++ Seq(x), newWeight)
    }
  }
  result ++ Seq(rem)
}
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-2.2-foldleft-6}}

~
\begin{lstlisting}
def groupBy[A, K](xs: Seq[A])(by: A => K): Map[K, Seq[A]] = {  
  val init: Map[K, Seq[A]] = Map()
  xs.foldLeft(init) { (res, x) =>
    val key = by(x)
    val seq = res.getOrElse(key, Seq()) ++ Seq(x)
    res.updated(key, seq)
  }
}
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-3}}

~
\begin{lstlisting}
def digitsOf(n: Int): Seq[Int] = Stream.iterate(n)(_ / 10).takeWhile(_ != 0).map(_ % 10).toList
def cubeDigits(n: Int): Int = digitsOf(n).map(x => x*x*x).sum
def cubes(n: Int): Stream[Int] = Stream.iterate(n)(cubeDigits)

def stopRepeats[T](str: Stream[T]): Stream[T] = {
  val halfSpeed = str.flatMap(x => Seq(x, x))
  val result = halfSpeed.zip(str).drop(1).takeWhile{ case (h, s) => h != s }.map(_._2)
  if (result.isEmpty) str.take(1) else str
}
def cubesReach1(n: Int): Boolean = stopRepeats(cubes(n)).contains(1)
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-4}}

~
\begin{lstlisting}
def prod3(a: Set[Int], b: Set[Int], c: Set[Int]): Set[Set[Int]] =
  a.flatMap(x => b.flatMap(y => c.map(z => Set(x, y, z))))
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-5}}

~
\begin{lstlisting}
def prodSet(sets: Set[Set[Int]]): Set[Set[Int]] =
    sets.foldLeft(Set[Set[Int]](Set())) {
      // Combine each of results so far with each element in current set
      case (accumSet: Set[Set[Int]], currSet: Set[Int]) =>
        for {
          s <- accumSet
          i <- currSet
        } yield s + i
      }
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-4-1}}

~
\begin{lstlisting}
@tailrec def pairs(goal: Int, xs: Array[Int])(
  res: Set[(Int, Int)] = Set(), left: Int = 0, right: Int = xs.length): Set[(Int, Int)] =
    if (left == right) res else {
      val sum = xs(left) + xs(right - 1)
      val (newLeft, newRight, newRes) = if (sum == goal) 
                  (left + 1, right, res + ((xs(left), xs(right - 1)))) 
                else if (sum < goal)
                  (left + 1, right, res)
                else (left, right - 1, res)
      pairs(goal, xs)(newRes, newLeft, newRight)
}
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-6}}

~
\begin{lstlisting}
def revSentence(sentence: String): String = sentence.split(" ").reverse.mkString(" ")
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2revdigits-Exercise-seq-7}}

~
\begin{lstlisting}
def digitsOf(n: Int): Seq[Int] = Stream.iterate(n)(_ / 10).takeWhile(_ != 0).map(_ % 10).toList
def revDigits(n: Int): Int = digitsOf(n).foldLeft(0){case (acc, d) => acc * 10 + d }
def isPalindrome(n: Int): Boolean = revDigits(n) == n
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-8}}

~
\begin{lstlisting}
def findPalindrome(n: Int): Int = 
  Stream.iterate(n) { x => x + revDigits(x) } .filter(isPalindrome).head
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:ch2Exercise-seq-9-1}}

~
\begin{lstlisting}
def unfold2[A,B](init: A)(next: A => Option[(A,B)]): Stream[B] = next(init) match {
   case None           => Stream()
   case Some((a, b))   => Stream.cons(b, unfold2(a)(next))
}
\end{lstlisting}


\subsection*{Chapter \ref{chap:Disjunctive-types}}

\subsubsection*{Exercise \ref{subsec:Disjunctive-Exercise-non-empty-list-1}}

~
\begin{lstlisting}
def toList[A](nel: NEL[A]): List[A] = nel match {
  case Last(x)         => List(x)
  case More(x, tail)   => x :: toList(tail)
} // Not tail-recursive.
def toList[A](nel: NEL[A]): List[A] = foldLeft(nel)(Nil:List[A]) {
  (prev, x) =>  x :: prev
}.reverse // Tail-recursive, but performs two traversals.
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-disjunctive-EvenList}}

~
\begin{lstlisting}
sealed trait EvenList[A]
final case class Lempty[A]() extends EvenList[A]
final case class Lpair[A](x: A, y: A, tail: EvenList[A]) extends EvenList

def fmap[A, B](f: A => B): EvenList[A] => EvenList[B] = {
  case Lempty()            => Lempty[B]()
  case Lpair(x, y, tail)   => Lpair[B](f(x), f(y), fmap(f)(tail))
}
\end{lstlisting}


\subsection*{Chapter \ref{chap:Higher-order-functions}}

\subsubsection*{Exercise \ref{subsec:Exercise-hof-simple-8}}

~
\begin{lstlisting}
@tailrec def convergeN[X](p: X => Boolean)(x:X)(m:Int)(f: X => X): Option[X] =  {
              if (m <= 0) None
              else if (p(x)) Some(x) else converge(p)(f(x))(m-1)(f)             }
// Defining it as def convergeN[X]: (X => Boolean) => X => Int => (X => X) => Option[X] = ???
// will break tail recursion!
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-hof-simple-7-1}}

~
\begin{lstlisting}
def recover[E, A]: Option[Either[E, A]] => (E => A) => Option[A] = {
  case None             => _ => None
  case Some(Right(a))   => _ => Some(a)
  case Some(Left(e))    => f => Some(f(e))
}
\end{lstlisting}


\subsubsection*{Exercise \ref{subsec:Exercise-hof-composition-1}}

\textbf{(a)} Choose $f^{:A\rightarrow A}\triangleq(\_\rightarrow z)$
and compute $(f\bef h)(x)=h(z)\overset{!}{=}h(x)$, for any $x$.
So, $h(x)$ equals a fixed value $h(z)$. It follows that $h(x)$
does not depend on $x$, i.e., $h$ is a constant function.

\textbf{(b)} Choose $f^{:A\rightarrow B}\triangleq\_\rightarrow b$
and compute $(f\bef h)(x)=h(b)\overset{!}{=}(g\bef f)(x)=f(g(x))=b$.
It follows that $h(b)=b$ for any $b^{:B}$, so $h$ is an identity
function. Substitute that into the law and get $f=g\bef f$ for any
function $f$. Substitute $f\triangleq\text{id}$ into that and derive
$\text{id}=g\bef\text{id}=g$. So, $g$ is the identity function.

\subsection*{Chapter \ref{chap:5-Curry-Howard}}

\subsubsection*{Exercise \ref{subsec:Exercise-type-notation-2}}

The type expression is: $A\times\text{Int}+A\times\text{Char}+A\times\text{Float}$.

\subsection*{Chapter \ref{chap:Reasoning-about-code}}

\subsubsection*{Exercise \ref{subsec:Exercise-reasoning-1-4-1}}

It is assumed that $\phi:\forall A.\,F^{A}\rightarrow G^{A}$ satisfies
its naturality law: for any $f^{:A\rightarrow B}$,
\[
f^{\uparrow F}\bef\phi=\phi\bef f^{\uparrow G}\quad.
\]

\textbf{(a)} To verify the naturality law of $\phi^{\uparrow K}$:
for any $f^{:A\rightarrow B}$,
\begin{align*}
 & f^{\uparrow F\uparrow K}\bef\phi^{\uparrow K}\overset{?}{=}\phi^{\uparrow K}\bef f^{\uparrow G\uparrow K}\quad,\\
{\color{greenunder}\text{composition under }^{\uparrow K}:}\quad & (f^{\uparrow F}\bef\phi)^{\uparrow K}\overset{?}{=}(\phi\bef f^{\uparrow G})^{\uparrow K}\quad.
\end{align*}
The last equation holds due to the naturality law of $\phi$.

\textbf{(b)} The naturality law of the pair product ($\phi\boxtimes\psi$)
says that for any $p^{:A\rightarrow B}$,
\[
p^{\uparrow(F\times K)}\bef(\phi\boxtimes\psi)\overset{?}{=}(\phi\boxtimes\psi)\bef p^{\uparrow(G\times L)}\quad.
\]
Begin with the left-hand side of this equation:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p^{\uparrow(F\times K)}\bef(\phi\boxtimes\psi)\\
{\color{greenunder}\text{definition of }^{\uparrow(F\times K)}:}\quad & =\big(f^{:F^{A}}\times k^{:K^{A}}\rightarrow(f\triangleright p^{\uparrow F})\times(k\triangleright p^{\uparrow K})\big)\bef(\phi\boxtimes\psi)\\
{\color{greenunder}\text{definition of }\phi\boxtimes\psi:}\quad & =f\times k\rightarrow\phi(f\triangleright p^{\uparrow F})\times\psi(k\triangleright p^{\uparrow K})=f\times k\rightarrow(f\triangleright p^{\uparrow F}\bef\phi)\times(k\triangleright p^{\uparrow K}\bef\psi)\quad.
\end{align*}
To rewrite the right-hand side, introduce the function argument into
$\phi\boxtimes\psi$:
\begin{align*}
{\color{greenunder}\text{right-hand side}:}\quad & (\phi\boxtimes\psi)\bef p^{\uparrow(G\times L)}=\big(f^{:F^{A}}\times k^{:K^{A}}\rightarrow(f\triangleright\phi)\times(k\triangleright\psi)\big)\bef p^{\uparrow(G\times L)}\\
{\color{greenunder}\text{definition of }^{\uparrow(G\times L)}:}\quad & =f\times k\rightarrow(f\triangleright\gunderline{\phi\triangleright p^{\uparrow G}})\times(k\triangleright\gunderline{\psi\triangleright p^{\uparrow L}})=f\times k\rightarrow(f\triangleright\phi\bef p^{\uparrow G})\times(k\triangleright\psi\bef p^{\uparrow L})\quad.
\end{align*}
The remaining differences between the two sides disappear due to the
naturality laws of $\phi$ and $\psi$:
\[
p^{\uparrow F}\bef\phi=\phi\bef p^{\uparrow G}\quad,\quad\quad p^{\uparrow K}\bef\psi=\psi\bef p^{\uparrow L}\quad.
\]

The naturality law of the pair co-product ($\phi\boxplus\psi$) says
that for any $p^{:A\rightarrow B}$,
\[
p^{\uparrow(F+K)}\bef(\phi\boxplus\psi)\overset{?}{=}(\phi\boxplus\psi)\bef p^{\uparrow(G+L)}\quad.
\]
Begin with the left-hand side of this equation:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p^{\uparrow(F+K)}\bef(\phi\boxplus\psi)\\
{\color{greenunder}\text{definitions of }^{\uparrow(F+K)}\text{ and of }\phi\boxplus\psi:}\quad & =\,\begin{array}{|c||cc|}
 & F^{B} & K^{B}\\
\hline F^{A} & p^{\uparrow F} & \bbnum 0\\
K^{A} & \bbnum 0 & p^{\uparrow K}
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & G^{B} & L^{B}\\
\hline F^{B} & \phi & \bbnum 0\\
K^{B} & \bbnum 0 & \psi
\end{array}\\
{\color{greenunder}\text{matrix composition}:}\quad & =\,\,\begin{array}{|c||cc|}
 & G^{B} & L^{B}\\
\hline F^{A} & p^{\uparrow F}\bef\phi & \bbnum 0\\
K^{A} & \bbnum 0 & p^{\uparrow K}\bef\psi
\end{array}\quad.
\end{align*}
The right-hand side is rewritten in a similar way:
\begin{align*}
{\color{greenunder}\text{right-hand side}:}\quad & (\phi\boxplus\psi)\bef p^{\uparrow(G+L)}\\
 & =\,\begin{array}{|c||cc|}
 & G^{A} & L^{A}\\
\hline F^{A} & \phi & \bbnum 0\\
K^{A} & \bbnum 0 & \psi
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & G^{B} & L^{B}\\
\hline G^{A} & p^{\uparrow G} & \bbnum 0\\
L^{A} & \bbnum 0 & p^{\uparrow L}
\end{array}\,=\,\begin{array}{|c||cc|}
 & G^{B} & L^{B}\\
\hline F^{A} & \phi\bef p^{\uparrow G} & \bbnum 0\\
K^{A} & \bbnum 0 & \psi\bef p^{\uparrow L}
\end{array}\quad.
\end{align*}
The remaining differences between the two sides disappear due to the
naturality laws of $\phi$ and $\psi$. 

\subsection*{Chapter \ref{chap:Typeclasses-and-functions}}

\subsubsection*{Exercise \ref{subsec:tc-Exercise-3}}

\textbf{(a)}
\begin{lstlisting}
def monoidFunc[A: Monoid, R] = Monoid[R => A](
  { (x, y) => r => x(r) |+| y(r) }, _ => implicitly[Monoid[A]].empty
)
\end{lstlisting}

In the code notation:
\[
f^{:R\rightarrow A}\oplus g^{:R\rightarrow A}\triangleq a\rightarrow f(a)\oplus_{A}g(a)\quad,\quad\quad e\triangleq(\_\rightarrow e_{A})\quad.
\]
Proof of monoid laws:
\begin{align*}
 & a\triangleright\left(\left(f\oplus g\right)\oplus h\right)=\left(a\triangleright(f\oplus g)\right)\oplus_{A}h(a)=f(a)\oplus_{A}g(a)\oplus_{A}h(a)\quad.\\
 & a\triangleright\left(f\oplus\left(g\oplus h\right)\right)=f(a)\oplus_{A}\left(a\triangleright(g\oplus h)\right)=f(a)\oplus_{A}g(a)\oplus_{A}h(a)\quad.\\
 & a\triangleright\left(e\oplus f\right)=e(a)\oplus_{A}f(a)=e_{A}\oplus_{A}f(a)=f(a)=a\triangleright f\quad.\\
 & a\triangleright(f\oplus e)=f(a)\oplus_{A}e(a)=f(a)\oplus_{A}e_{A}=f(a)=a\triangleright f\quad.
\end{align*}


\subsubsection*{Exercise \ref{subsec:tc-Exercise-9-1} }

\textbf{(a)} A counterexample is the functor $F^{A}\triangleq R\rightarrow A$,
where $R$ is a fixed type.

\textbf{(b)} The function $C^{A}\times C^{B}\rightarrow C^{A+B}$
cannot be implemented for $C^{A}\triangleq\left(A\rightarrow P\right)+\left(A\rightarrow Q\right)$.
This more complicated contrafunctor $C$ is necessary because the
simpler contrafunctor $C^{A}\triangleq A\rightarrow P$ does not provide
a counterexample.

\subsubsection*{Exercise \ref{subsec:tc-Exercise-9-1-1-1}}

Define the method $\text{ex}_{F}$ as:
\[
\text{ex}_{F}\triangleq x^{:F^{A}}\rightarrow x\triangleright(a^{:A}\rightarrow a\times1)^{\uparrow F}\triangleright q^{A,\bbnum 1}\triangleright\pi_{1}\quad\text{ or equivalently: }\quad\text{ex}_{F}\triangleq(a^{:A}\rightarrow a\times1)^{\uparrow F}\bef q^{A,\bbnum 1}\bef\pi_{1}\quad.
\]
To show that the naturality law ($f^{\uparrow F}\bef\text{ex}_{F}=\text{ex}_{F}\bef f$)
holds, use the identity $(f\boxtimes g)\bef\pi_{1}=\pi_{1}\bef f$:
\begin{align*}
 & f^{\uparrow F}\bef\text{ex}_{F}=\gunderline{f^{\uparrow F}\bef(a^{:A}\rightarrow a\times1)^{\uparrow F}}\bef q^{A,\bbnum 1}\bef\pi_{1}=(a^{:A}\rightarrow a\times1)^{\uparrow F}\bef\gunderline{(f\boxtimes\text{id})^{\uparrow F}\bef q^{A,\bbnum 1}}\bef\pi_{1}\\
 & =(a^{:A}\rightarrow a\times1)^{\uparrow F}\bef q^{A,\bbnum 1}\bef\gunderline{(f\boxtimes\text{id}^{\uparrow F})\bef\pi_{1}}=\gunderline{(a^{:A}\rightarrow a\times1)^{\uparrow F}\bef q^{A,\bbnum 1}\bef\pi_{1}}\bef f=\text{ex}_{F}\bef f\quad.
\end{align*}

Given a method $\text{ex}_{F}$, define $q$ as:
\[
q^{A,B}\triangleq f^{:F^{A\times B}}\rightarrow(f\triangleright\pi_{1}^{\uparrow F}\triangleright\text{ex}_{F}^{A})\times(f\triangleright\pi_{2}^{\uparrow F})\quad\text{ or equivalently: }\quad q^{A,B}\triangleq\Delta\bef(\pi_{1}^{\uparrow F}\boxtimes\pi_{2}^{\uparrow F})\bef(\text{ex}_{F}^{A}\boxtimes\text{id})\quad.
\]
Show that the required laws hold for $q$. Identity law: 
\[
f\triangleright q^{\bbnum 1,B}=(f\triangleright\pi_{1}^{\uparrow F}\triangleright\text{ex}_{F}^{\bbnum 1})\times(f\triangleright\pi_{2}^{\uparrow F})=1\times(f\triangleright\pi_{2}^{\uparrow F})
\]
because $\text{ex}_{F}^{\bbnum 1}$ produces a value of type $\bbnum 1$,
which can only be $1$.

To prove the naturality law:
\begin{align*}
(f^{:A\rightarrow C}\boxtimes g^{:B\rightarrow D})^{\uparrow F}\bef\Delta\bef(\pi_{1}^{\uparrow F}\boxtimes\pi_{2}^{\uparrow F})\bef(\text{ex}_{F}^{A}\boxtimes\text{id}) & =\Delta\bef(\pi_{1}^{\uparrow F}\boxtimes\pi_{2}^{\uparrow F})\bef(\text{ex}_{F}^{A}\boxtimes\text{id})\bef f\boxtimes(g^{\uparrow F})\\
\Delta\bef((f\boxtimes g)^{\uparrow F}\boxtimes(f\boxtimes g)^{\uparrow F})\bef(\pi_{1}^{\uparrow F}\boxtimes\pi_{2}^{\uparrow F})\bef(\text{ex}_{F}^{A}\boxtimes\text{id}) & =\Delta\bef(\pi_{1}^{\uparrow F}\bef\text{ex}_{F}\bef f)\boxtimes(\pi_{2}^{\uparrow F}\bef g^{\uparrow F})\\
\Delta\bef(\pi_{1}^{\uparrow F}\bef\,\gunderline{f^{\uparrow F}\bef\text{ex}_{F}})\boxtimes(\pi_{2}^{\uparrow F}\bef g^{\uparrow F}) & =\Delta\bef(\pi_{1}^{\uparrow F}\bef\,\gunderline{\text{ex}_{F}\bef f})\boxtimes(\pi_{2}^{\uparrow F}\bef g^{\uparrow F})
\end{align*}
To prove the associativity law, write the left-hand side as:
\begin{align*}
 & f^{:F^{A\times B\times C}}\triangleright q^{A,B\times C}\bef(\text{id}^{A}\boxtimes q^{B,C})=(f\triangleright\pi_{1}^{\uparrow F}\triangleright\text{ex}_{F}^{A})\times(f\triangleright(a\times b\times c\rightarrow b\times c)^{\uparrow F}\triangleright q^{B,C})\\
 & =(f\triangleright\pi_{1}^{\uparrow F}\triangleright\text{ex}_{F}^{A})\times(f\triangleright(a\times b\times c\rightarrow b\times c)^{\uparrow F}\triangleright\pi_{1}^{\uparrow F}\triangleright\text{ex}_{F}^{B})\times(f\triangleright(a\times b\times c\rightarrow b\times c)^{\uparrow F}\triangleright\pi_{2}^{\uparrow F})\\
 & =\left(f\triangleright\text{ex}_{F}\triangleright(a\times b\times c\rightarrow a)\right)\times\left(f\triangleright\text{ex}_{F}\triangleright(a\times b\times c\rightarrow b)\right)\times(f\triangleright(a\times b\times c\rightarrow c)^{\uparrow F})\quad.
\end{align*}
The right-hand side is then simplified to the same expression:
\begin{align*}
 & f^{:F^{A\times B\times C}}\triangleright q^{A\times B,C}=(f\triangleright\text{ex}_{F}\triangleright(a\times b\times c\rightarrow a\times b))\times(f\triangleright(a\times b\times c\rightarrow c)^{\uparrow F})\\
 & =\left(f\triangleright\text{ex}_{F}\triangleright(a\times b\times c\rightarrow a)\right)\times\left(f\triangleright\text{ex}_{F}\triangleright(a\times b\times c\rightarrow b)\right)\times(f\triangleright(a\times b\times c\rightarrow c)^{\uparrow F})\quad.
\end{align*}


\subsection*{Chapter \ref{chap:Filterable-functors}}

\subsubsection*{Exercise \ref{subsec:filt-exercise-derive-liftOpt-equivalence-1}}

Starting from \lstinline!liftOpt!\textsf{'}s law, derive the naturality law:
\[
\text{liftOpt}\left(g\right)\bef\text{liftOpt}\,(f\bef\text{pu}_{\text{Opt}})=\text{liftOpt}(f\bef\text{pu}_{\text{Opt}}\diamond_{_{\text{Opt}}}g)=\text{liftOpt}(f\bef g)\quad.
\]
Now use the naturality-identity law and get $\text{liftOpt}\,(f\bef\text{pu}_{\text{Opt}})=f^{\downarrow C}$.
The result is the naturality law $\text{liftOpt}\left(f\bef g\right)=\text{liftOpt}\left(g\right)\bef f^{\downarrow C}$.

\subsubsection*{Exercise \ref{subsec:Exercise-filterable-laws-4}}

A counterexample is the functor $F^{A}\triangleq\bbnum 1+A\times\left(Z\rightarrow A\right)$.
For this functor, one can implement \lstinline!deflate!\textsf{'}s type signature,
but the code cannot obey the identity law because it must always return
$1+\bbnum 0$.

\subsection*{Chapter \ref{chap:Semimonads-and-monads}}

\subsubsection*{Exercise \ref{subsec:Exercise-flatten-concat-distributive-law}}

The values $p$ and $q$ must have type $\text{List}^{\text{List}^{A}}$.
There are two possibilities: $p$ is an empty list ($p=1+\bbnum 0$),
and $p=\bbnum 0+h\times t$. If $p$ is empty, so is $p\triangleright\text{ftn}$
and the law holds. In the other case, we have (due to the code of
$\pplus$) that:
\[
\left(\bbnum 0+h\times t\right)\pplus q=\bbnum 0+h\times\left(t\pplus q\right)\quad,
\]
and so:
\begin{align*}
 & \left(p\pplus q\right)\triangleright\text{ftn}=\left(\bbnum 0+h\times\left(t\pplus q\right)\right)\triangleright\text{ftn}=h\pplus\gunderline{\left(t\pplus q\right)\triangleright\overline{\text{ftn}}}\\
{\color{greenunder}\text{inductive assumption}:}\quad & =\gunderline{h\pplus(t\triangleright\overline{\text{ftn}})}\pplus(q\triangleright\overline{\text{ftn}})\\
{\color{greenunder}\text{code of }\text{ftn}:}\quad & =\left(\bbnum 0+h\times t\right)\triangleright\overline{\text{ftn}}\pplus(q\triangleright\overline{\text{ftn}})=\left(p\triangleright\text{ftn}\right)\pplus\left(q\triangleright\text{ftn}\right)\quad.
\end{align*}


\subsubsection*{Exercise \ref{subsec:Exercise-monad-of-monoid-is-monoid}}

Define the empty element ($e_{M}$) of the monoid $M^{W}$ as:
\[
e_{M}\triangleq\text{pu}_{M}(e_{W})\quad.
\]
The binary operation $\oplus_{M}$ of the monoid $M^{W}$ may be implemented
through $\oplus_{W}$ as:
\[
p\oplus_{M}q\triangleq p\triangleright\text{flm}_{M}\big(u^{:W}\rightarrow q\triangleright(v^{:W}\rightarrow u\oplus_{W}v)^{\uparrow M}\big)\quad.
\]
To check the left identity law of $M^{W}$:
\begin{align*}
{\color{greenunder}\text{expect to equal }p:}\quad & e_{M}\oplus_{M}p=e_{W}\triangleright\gunderline{\text{pu}_{M}\triangleright\text{flm}_{M}}\big(u^{:W}\rightarrow p\triangleright(v^{:W}\rightarrow u\oplus_{W}v)^{\uparrow M}\big)\\
 & =e_{W}\triangleright\big(u^{:W}\rightarrow p\triangleright(v^{:W}\rightarrow u\oplus_{W}v)^{\uparrow M}\big)=p\triangleright(v\rightarrow\gunderline{e_{W}\oplus_{W}v})^{\uparrow M}\\
{\color{greenunder}\text{left identity law of }W:}\quad & =p\triangleright(v\rightarrow v)^{\uparrow M}=p\triangleright\text{id}^{\uparrow M}=p\quad.
\end{align*}
To check the right identity law of $M^{W}$:
\begin{align*}
{\color{greenunder}\text{expect to equal }p:}\quad & p\oplus_{M}e_{M}=p\triangleright\text{flm}_{M}\big(u^{:W}\rightarrow e_{W}\triangleright\gunderline{\text{pu}_{M}\triangleright(v^{:W}\rightarrow u\oplus_{W}v)^{\uparrow M}}\big)\\
{\color{greenunder}\text{naturality of }\text{pu}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(u^{:W}\rightarrow\gunderline{e_{W}\triangleright(v^{:W}}\rightarrow u\oplus_{W}v)\triangleright\text{pu}_{M}\big)\\
{\color{greenunder}\text{apply function}:}\quad & =p\triangleright\text{flm}_{M}\big(u^{:W}\rightarrow(\gunderline{u\oplus_{W}e_{W}})\triangleright\text{pu}_{M}\big)\\
{\color{greenunder}\text{right identity law of }W:}\quad & =p\triangleright\text{flm}_{M}\big(u^{:W}\rightarrow u\triangleright\text{pu}_{M}\big)=p\triangleright\text{id}=p\quad.
\end{align*}
To check the associativity law of $M^{W}$, we use the associativity
law of $\text{flm}_{M}$:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (p\oplus_{M}q)\oplus_{M}r\\
 & =p\triangleright\gunderline{\text{flm}_{M}}\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{W}v)^{\uparrow M}\big)\,\gunderline{\triangleright\text{flm}_{M}}\big(t\rightarrow r\triangleright(w\rightarrow t\oplus_{W}w)^{\uparrow M}\big)\\
{\color{greenunder}\text{associativity of }\text{flm}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(\gunderline (u\rightarrow q\triangleright(v\rightarrow u\oplus_{W}v)^{\uparrow M}\gunderline{)\bef}\,\text{flm}_{M}(t\rightarrow r\triangleright(w\rightarrow t\oplus_{W}w)^{\uparrow M})\big)\\
{\color{greenunder}\text{compute composition}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{W}v\gunderline{)^{\uparrow M}\triangleright\text{flm}_{M}}(t\rightarrow r\triangleright(w\rightarrow t\oplus_{W}w)^{\uparrow M})\big)\\
{\color{greenunder}\text{naturality of }\text{flm}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright\text{flm}_{M}(v\rightarrow\gunderline{u\oplus_{W}v})\,\gunderline{\bef(t}\rightarrow r\triangleright(w\rightarrow\gunderline t\oplus_{W}w)^{\uparrow M}))\big)\\
{\color{greenunder}\text{compute composition}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright\text{flm}_{M}(v\rightarrow r\triangleright(w\rightarrow(u\oplus_{W}v)\oplus_{W}w)^{\uparrow M})\big)\quad.
\end{align*}
Now write the right-hand side of the law:
\begin{align*}
{\color{greenunder}\text{right-hand side}:}\quad & p\oplus_{M}(q\oplus_{M}r)=p\triangleright\text{flm}_{M}\big(u\rightarrow(q\oplus r)\triangleright(t\rightarrow u\oplus_{W}t)^{\uparrow M}\big)\\
{\color{greenunder}\text{substitute }q\oplus r:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright\text{flm}_{M}(v\rightarrow r\triangleright(w\rightarrow v\oplus_{W}w)^{\uparrow M})\,\gunderline{\triangleright\,(}t\rightarrow u\oplus_{W}t\gunderline{)^{\uparrow M}}\big)\\
{\color{greenunder}\text{naturality of }\text{flm}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright\text{flm}_{M}(v\rightarrow r\triangleright(w\rightarrow u\oplus_{W}w\gunderline{)^{\uparrow M}\bef}\,(t\rightarrow u\oplus_{W}t\gunderline{)^{\uparrow M}})\big)\\
{\color{greenunder}\text{composition under }^{\uparrow M}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright\text{flm}_{M}(v\rightarrow r\triangleright(w\rightarrow u\oplus_{W}(v\oplus_{W}w))^{\uparrow M})\big)\quad.
\end{align*}
 The difference between the two sides now vanishes due to the assumed
associativity law of $W$:
\[
(u\oplus_{W}v)\oplus_{W}w\overset{!}{=}u\oplus_{W}(v\oplus_{W}w)\quad.
\]


\subsubsection*{Exercise \ref{subsec:Exercise-1-monads-9-1}}

\textbf{(b)} The code is converted into monad methods like this:
\begin{align*}
 & r_{1}=p\triangleright\text{flm}_{M}(x\rightarrow q\triangleright(y\rightarrow f(x,y))^{\uparrow M})\quad,\\
 & r_{2}=q\triangleright\text{flm}_{M}(y\rightarrow p\triangleright(x\rightarrow f(x,y))^{\uparrow M})\quad.
\end{align*}
For a commutative monad $M$, we have $r_{1}=r_{2}$. The monoid operation
$\oplus_{M}$ is defined by:
\[
p^{:M^{\bbnum 1}}\oplus_{M}q^{:M^{\bbnum 1}}\triangleq p\triangleright\text{flm}_{M}(1\rightarrow q)\quad.
\]
Commutativity of $\oplus_{M}$ means that:
\[
p\triangleright\text{flm}_{M}(1\rightarrow q)=p\oplus_{M}q\overset{?}{=}q\oplus_{M}p=q\triangleright\text{flm}_{M}(1\rightarrow p)\quad.
\]
Use $f(x,y)\triangleq1$ in the definitions of $r_{1}$ and $r_{2}$;
then the above equation is equivalent to $r_{1}=r_{2}$.

\subsubsection*{Exercise \ref{subsec:Exercise-1-monads-12}}

The second definition, $\text{pu}_{L}\triangleq a\rightarrow\bbnum 0+\text{pu}_{F}(a)$,
fails the right identity law:
\begin{align*}
 & \text{pu}_{L}^{\uparrow L}\bef\text{ftn}_{L}=\,\begin{array}{|c||ccc|}
 & A & F^{A} & F^{L^{A}}\\
\hline A & \bbnum 0 & \text{pu}_{F} & \bbnum 0\\
F^{A} & \bbnum 0 & \bbnum 0 & \text{pu}_{L}^{\uparrow F}
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & A & F^{A}\\
\hline A & \text{id} & \bbnum 0\\
F^{A} & \bbnum 0 & \text{id}\\
F^{L^{A}} & \bbnum 0 & \gamma^{\uparrow F}\bef\text{ftn}_{F}
\end{array}\,=\,\begin{array}{|c||cc|}
 & A & F^{A}\\
\hline A & \bbnum 0 & \text{pu}_{F}\\
F^{A} & \bbnum 0 & ...
\end{array}\,\neq\text{id}\quad.
\end{align*}
This matrix cannot be equal to the identity function because it has
a missing diagonal element.

\subsubsection*{Exercise \ref{subsec:Exercise-monad-composition-mm}}

If $M$ is a semimonad, we have the Kleisli composition $\diamond_{_{M}}$
that satisfies the associativity law. Define $\diamond_{_{L}}$ by:
\[
f^{:A\rightarrow M^{M^{B}}}\diamond_{_{L}}g^{:B\rightarrow M^{M^{C}}}\triangleq f\bef\text{ftn}_{M}\diamond_{_{M}}g\quad.
\]
Here, parentheses are unnecessary as shown in Statement~\ref{subsec:Statement-equivalence-kleisli-composition-and-flatMap}.

It is inconvenient to mix the Kleisli composition and the \lstinline!flatten!
method, so we express \lstinline!flatten! as:
\[
\text{ftn}_{M}^{:M^{M^{A}}\rightarrow M^{A}}=\text{flm}_{M}(\text{id}^{:M^{A}\rightarrow M^{A}})=\text{id}^{:M^{M^{A}}\rightarrow M^{M^{A}}}\diamond_{_{M}}\text{id}^{:M^{A}\rightarrow M^{A}}\quad.
\]
For brevity, we will omit type annotations from now on. So, we can
express the Kleisli composition $\diamond_{_{L}}$ through $\diamond_{_{M}}$
by:
\[
f\diamond_{_{L}}g\triangleq f\bef\gunderline{\left(\text{id}\diamond_{_{M}}\text{id}\right)\diamond_{_{M}}g}=f\bef\left(\text{id}\diamond_{_{M}}\left(\text{id}\diamond_{_{M}}g\right)\right)=\left(f\bef\text{id}\right)\diamond_{_{M}}\left(\text{id}\diamond_{_{M}}g\right)=f\diamond_{_{M}}\text{id}\diamond_{_{M}}g\quad.
\]

Associativity of $\diamond_{_{L}}$ then follows from associativity
of $\diamond_{_{M}}$ as:
\begin{align*}
\left(f\diamond_{_{L}}g\right)\diamond_{_{L}}h & =\left(f\diamond_{_{M}}\text{id}\diamond_{_{M}}g\right)\diamond_{_{M}}\text{id}\diamond_{_{M}}h=f\diamond_{_{M}}\text{id}\diamond_{_{M}}g\diamond_{_{M}}\text{id}\diamond_{_{M}}h\quad,\\
f\diamond_{_{L}}\left(g\diamond_{_{L}}h\right) & =f\diamond_{_{M}}\text{id}\diamond_{_{M}}\left(g\diamond_{_{M}}\text{id}\diamond_{_{M}}h\right)=f\diamond_{_{M}}\text{id}\diamond_{_{M}}g\diamond_{_{M}}\text{id}\diamond_{_{M}}h\quad.
\end{align*}

This definition of $\diamond_{_{L}}$ corresponds to a definition
of $\text{ftn}_{L}$ that flattens the \emph{first} three layers of
$M$ in $M\circ M\circ M\circ M$:
\[
\text{ftn}_{L}\triangleq\text{ftn}_{M}\bef\text{ftn}_{M}\quad.
\]

An alternative definition will flatten the \emph{last} three layers:
\[
\text{ftn}_{L}\triangleq(\text{ftn}_{M}\bef\text{ftn}_{M})^{\uparrow M}\quad.
\]
Both definitions satisfy the associativity law and so define a semimonad
$L$,
\[
\text{ftn}_{L}^{\uparrow L}\bef\text{ftn}_{L}=\text{ftn}_{L}\bef\text{ftn}_{L}\quad.
\]
Let us verify that directly. For the first definition:
\begin{align*}
 & \gunderline{(\text{ftn}_{M}\bef\text{ftn}_{M})^{\uparrow M\uparrow M}\bef\text{ftn}_{M}}\bef\text{ftn}_{M}=\text{ftn}_{M}\bef\big(\gunderline{\text{ftn}_{M}\bef\text{ftn}_{M}}\big)^{\uparrow M}\bef\text{ftn}_{M}\\
{\color{greenunder}\text{associativity of }M:}\quad & =\gunderline{\text{ftn}_{M}\bef\text{ftn}_{M}^{\uparrow M\uparrow M}}\bef\text{ftn}_{M}^{\uparrow M}\bef\text{ftn}_{M}=\gunderline{\text{ftn}_{M}^{\uparrow M}\bef\text{ftn}_{M}}\bef\gunderline{\text{ftn}_{M}^{\uparrow M}\bef\text{ftn}_{M}}\\
{\color{greenunder}\text{associativity of }M:}\quad & =\text{ftn}_{M}\bef\text{ftn}_{M}\bef\text{ftn}_{M}\bef\text{ftn}_{M}\quad.
\end{align*}
For the second definition, we just apply $^{\uparrow M}$ to the preceding
derivation.

A definition of \lstinline!flatten! that flattens separately the
first two and the last two layers of $M$ ($\text{ftn}_{L}\triangleq\text{ftn}_{M}\bef\text{ftn}_{M}^{\uparrow M}$)
will fail the associativity law when $M$ is chosen in a suitable
way.

If $M$ is a full monad, we have $\text{pu}_{M}$ that satisfies the
identity laws. Define $\text{pu}_{L}$ by:
\[
\text{pu}_{L}\triangleq\text{pu}_{M}\bef\text{pu}_{M}\quad.
\]
There is no other way of defining $\text{pu}_{L}$ since we cannot
obtain a value of type $M^{A}$ other than via $\text{pu}_{M}$.

With any of the two possible definitions of $\text{ftn}_{L}$, at
least one of the identity laws for $\diamond_{_{L}}$ fails.

For the definition of $\text{ftn}_{L}$ that flattens the first three
layers, the right identity law will fail:
\[
\text{pu}_{L}^{\uparrow L}\bef\text{ftn}_{L}=\gunderline{(\text{pu}_{M}\bef\text{pu}_{M})^{\uparrow M\uparrow M}\bef\text{ftn}_{M}}\bef\text{ftn}_{M}=\text{ftn}_{M}\bef\text{pu}_{M}^{\uparrow M}\bef\gunderline{\text{pu}_{M}^{\uparrow M}\bef\text{ftn}_{M}}=\text{ftn}_{M}\bef\text{pu}_{M}^{\uparrow M}\quad.
\]
In general, this function is not equal to an identity function, because
\lstinline!flatten! merges the two monadic layers of $M$ and in
that way loses information about a value of type $M^{M^{A}}$.

For the definition of $\text{ftn}_{L}$ that flattens the last three
layers, the left identity law will fail:
\[
\text{pu}_{L}\bef\text{ftn}_{L}=\text{pu}_{M}\bef\gunderline{\text{pu}_{M}\bef(\text{ftn}_{M}}\bef\text{ftn}_{M})^{\uparrow M}=\gunderline{\text{pu}_{M}\bef\text{ftn}_{M}}\bef\text{pu}_{M}\bef\text{ftn}_{M}^{\uparrow M}=\text{pu}_{M}\bef\text{ftn}_{M}^{\uparrow M}=\text{ftn}_{M}\bef\text{pu}_{M}\quad.
\]
In general, this function is not equal to an identity function.

\subsection*{Chapter \ref{chap:8-Applicative-functors,-contrafunctors}}

\subsubsection*{Exercise \ref{subsec:Exercise-simplify-law-omit-lifted-function}}

We may choose $f=\text{id}$ and derive $u=v$ from the given law.
Conversely, if $u=v$ then $u\bef f^{\uparrow F}=v\bef f^{\uparrow F}$
for any function $f$.

\subsubsection*{Exercise \ref{subsec:Exercise-applicative-II-4-1}}

To verify the law, write:
\begin{align*}
 & \text{ap}\,(r)(\text{pu}_{L}(a))=\text{zip}\big(r\times\text{pu}_{L}(a)\big)\triangleright\text{eval}^{\uparrow L}\\
{\color{greenunder}\text{right identity law of }\text{zip}:}\quad & =r\triangleright\left(f\rightarrow f\times a\right)^{\uparrow L}\bef\text{eval}^{\uparrow L}\\
{\color{greenunder}\text{composition under }^{\uparrow L}:}\quad & =r\triangleright(f\rightarrow f(a))\quad.
\end{align*}


\subsubsection*{Exercise \ref{subsec:Exercise-function-type-construction-not-applicative}}

\textbf{(a)} We cannot implement \lstinline!zip! as a fully parametric
function with this type signature:
\[
\text{zip}_{F}(p^{:(A\rightarrow P)\rightarrow Q}\times q^{:(B\rightarrow P)\rightarrow Q})\triangleq h^{:A\times B\rightarrow P}\rightarrow\text{???}^{:Q}\quad.
\]
\begin{lstlisting}
import io.chymyst.ch._
type S[A] = (A => P) => Q

scala> def zip[A, B](p: S[A], q: S[B]): S[(A, B)] = implement
type ((A => P) => Q) => ((B => P) => Q) => (Tuple2[A,B] => P) => Q cannot be implemented
\end{lstlisting}

The reason it cannot be implemented is that the only way of getting
a value of type $Q$ is to call the given functions $p$ or $q$.
But we cannot call $p$ or $q$ since we cannot supply their arguments:
we have a function of type $A\times B\rightarrow P$, and we cannot
produce a function of type $A\rightarrow P$ or $B\rightarrow P$
out of it.

The only solution is for \lstinline!zip! to ignore its arguments
and always return the empty value $e_{Q}$. However, this implementation
loses information and will fail the identity laws.

\textbf{(b)} We \emph{can} implement the type signature of \lstinline!zip!,
but only in a trivial way:
\[
\text{zip}_{F}(p^{:(A\rightarrow P)\rightarrow\bbnum 1+A}\times q^{:(B\rightarrow P)\rightarrow\bbnum 1+B})\triangleq h^{:A\times B\rightarrow P}\rightarrow\text{???}^{:\bbnum 1+A\times B}\quad.
\]
The functions $p$ and $q$ cannot be called since we cannot supply
their arguments, just as in part \textbf{(a)}. The only solution is
that \lstinline!zip! should ignore its arguments and always return
$1+\bbnum 0^{:A\times B}$. However, this implementation loses information
and will fail the identity laws.

\subsubsection*{Exercise \ref{subsec:Exercise-additional-law-of-ap}}

We define $\text{ex}_{F}(h\times g)\triangleq\text{ex}_{H}(h)$ and
obtain:
\begin{align*}
 & \text{ex}_{F}(\text{zip}_{F}(h_{1}\times g_{1}\times h_{2}\times g_{2}))=\text{ex}_{F}(\text{zip}_{H}(h_{1}\times h_{2})\times\text{zip}_{G}(g_{1}\times g_{2}))=\text{ex}_{H}(\text{zip}_{H}(h_{1}\times h_{2}))\quad,\\
 & \text{ex}_{F}(h_{1}\times g_{1})\times\text{ex}_{F}(h_{2}\times g_{2})=\text{ex}_{H}(h_{1})\times\text{ex}_{H}(h_{2})\quad.
\end{align*}
The two sides are equal due to the compatibility law of $\text{zip}_{H}$
and $\text{ex}_{H}$.

\subsubsection*{Exercise \ref{subsec:Exercise-profunctor-example}}

The first and the third occurrences of $A$ in $Q^{A}$ are contravariant
while the others are covariant. So, we define a profunctor $P^{X,Y}$
and get:
\[
P^{X,Y}\triangleq\left(X\rightarrow\text{Int}\right)\times Y\times\left(X\rightarrow Y\right)\quad,\quad\quad Q^{A}=P^{A,A}\quad.
\]


\subsubsection*{Exercise \ref{subsec:Exercise-applicative-II-11}(b)}

Hint: Use $P^{A}\triangleq A\rightarrow Z$ as a counterexample.

\subsection*{Chapter \ref{chap:9-Traversable-functors-and}}

\subsubsection*{Exercise \ref{subsec:Exercise-traversables-1}}

First, we express \lstinline!consume! via \lstinline!sequence! and
back:
\begin{align}
 & \text{consume}_{L}(f^{:L^{A}\rightarrow B})(p^{:L^{F^{A}}})=p\triangleright\text{seq}_{L}\triangleright f^{\uparrow F}\quad,\quad\text{or equivalently}:\quad\text{consume}_{L}(f)=\text{seq}_{L}\bef f^{\uparrow F}\quad,\label{eq:consume-via-seq}\\
 & \text{seq}_{L}=\text{consume}_{L}(\text{id}^{:L^{A}\rightarrow L^{A}})\quad.\label{eq:seq-via-consume}
\end{align}
The pattern is similar to that in Statement~\ref{subsec:Statement-tr-equivalent-to-ftr}
and to the equivalence of \lstinline!sequence! and \lstinline!traverse!.

We need to derive the equivalence between \lstinline!consume! and
\lstinline!sequence! in both directions. To figure out the necessary
naturality law, we begin with the direction that restores \lstinline!consume!
from \lstinline!sequence! because \lstinline!consume! is the more
complicated function (having two type parameters).

\textbf{(a)} Given a function \lstinline!consume!, we define \lstinline!sequence!
via Eq.~(\ref{eq:seq-via-consume}) and then define a new function
\lstinline!consume!$^{\prime}$ via Eq.~(\ref{eq:consume-via-seq}).
Then we need to show that \lstinline!consume!$^{\prime}$ equals
\lstinline!consume!. For an arbitrary $f^{:L^{A}\rightarrow B}$,
we write:
\begin{align*}
{\color{greenunder}\text{expect to equal }\text{consume}_{L}(f):}\quad & \text{consume}_{L}^{\prime}(f)=\text{seq}_{L}\bef f^{\uparrow F}=\text{consume}_{L}(\text{id})\bef f^{\uparrow F}=???
\end{align*}
If we know nothing about \lstinline!consume!, we cannot conclude
that $\text{consume}_{L}(f)\overset{?}{=}\text{consume}_{L}(\text{id})\bef f^{\uparrow F}$
because the function $f$ is arbitrary and the value $\text{consume}_{L}(f)$
does not need to be related in any way to $\text{consume}_{L}(\text{id})$.
We must use a naturality law involving a lifted function applied after
\lstinline!consume!:
\begin{equation}
\text{consume}_{L}(f^{:L^{A}\rightarrow B})\bef(g^{:B\rightarrow C})^{\uparrow F}=\text{consume}_{L}(f\bef g)\quad.\label{eq:consume-naturality-law}
\end{equation}
If this law holds, we can derive the required equation:
\[
\text{consume}_{L}^{\prime}(f)=\text{consume}_{L}(\text{id})\bef f^{\uparrow F}=\text{consume}_{L}(\text{id}\bef f)=\text{consume}_{L}(f)\quad.
\]

\textbf{(b)} Given a function \lstinline!sequence!, we first define
\lstinline!consume! via Eq.~(\ref{eq:consume-via-seq}) and then
define a new function \lstinline!sequence!$^{\prime}$ via Eq.~(\ref{eq:seq-via-consume}).
We then show that \lstinline!sequence!$^{\prime}$ equals \lstinline!sequence!:
\[
\text{seq}_{L}^{\prime}=\text{consume}_{L}(\text{id})=\text{seq}_{L}\bef\gunderline{\text{id}^{\uparrow F}}=\text{seq}\quad.
\]

If a function \lstinline!consume! is defined via \lstinline!sequence!,
the naturality law~(\ref{eq:consume-naturality-law}) will hold automatically:
\[
\text{consume}_{L}(f)\bef g^{\uparrow F}=\text{seq}_{L}\bef\gunderline{f^{\uparrow F}\bef g^{\uparrow F}}=\text{seq}_{L}\bef(f\bef g)^{\uparrow F}=\text{consume}_{L}\left(f\bef g\right)\quad.
\]


\subsubsection*{Exercise \ref{subsec:Exercise-traversables-laws-1}}

The naturality law of \lstinline!ftr! with respect to the type parameter
$B$ is:
\begin{equation}
\text{ftr}\,(f^{:A\rightarrow G^{B}})\bef(g^{:B\rightarrow C})^{\uparrow H}=\text{ftr}\,(f\bef g^{\uparrow G})\quad.\label{eq:ftr-right-naturality-law}
\end{equation}
The type signature of \lstinline!tr! has only one type parameter,
so \lstinline!tr! has only one naturality law:
\begin{equation}
(f^{:A\rightarrow B})^{\uparrow G\uparrow F}\bef\text{tr}=\text{tr}\bef f^{\uparrow H}\quad.\label{eq:tr-naturality-law}
\end{equation}

1) We show that the naturality law~(\ref{eq:tr-naturality-law})
of \lstinline!tr! follows from the two naturality laws of \lstinline!ftr!.
Express \lstinline!tr! as $\text{ftr}\left(\text{id}\right)$ and
substitute into the two sides of \lstinline!tr!\textsf{'}s naturality law:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & f^{\uparrow G\uparrow F}\bef\text{tr}=f^{\uparrow G\uparrow F}\bef\text{ftr}\left(\text{id}\right)\\
{\color{greenunder}\text{naturality law~(\ref{eq:ftr-left-naturality-law})}:}\quad & \quad=\text{ftr}\,(f^{\uparrow G}\bef\text{id})=\text{ftr}\,(f^{\uparrow G})\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & \text{tr}\bef f^{\uparrow H}=\text{ftr}\left(\text{id}\right)\bef f^{\uparrow H}\\
{\color{greenunder}\text{naturality law~(\ref{eq:ftr-right-naturality-law})}:}\quad & \quad=\text{ftr}\,(\text{id}\bef f^{\uparrow G})=\text{ftr}\,(f^{\uparrow G})\quad.
\end{align*}
The two sides of the law are now equal.

2) We show that the second naturality law of \lstinline!ftr! follows
from the naturality law of \lstinline!tr!. Express \lstinline!ftr!
through \lstinline!tr! as $\text{ftr}\,(f)=f^{\uparrow F}\bef\text{tr}$,
and substitute into the two sides of \lstinline!ftr!\textsf{'}s naturality
law~(\ref{eq:ftr-right-naturality-law}): 
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & \text{ftr}\,(f\big)\bef g^{\uparrow H}=f^{\uparrow F}\bef\gunderline{\text{tr}\bef g^{\uparrow H}}\\
{\color{greenunder}\text{naturality law~(\ref{eq:tr-naturality-law}) of }\text{tr}:}\quad & =f^{\uparrow F}\bef g^{\uparrow G\uparrow F}\bef\text{tr}=(f\bef g^{\uparrow G})^{\uparrow F}\bef\text{tr}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & \text{ftr}\,(f\bef g^{\uparrow G})=(f\bef g^{\uparrow G})^{\uparrow F}\bef\text{tr}\quad.
\end{align*}
The two sides of the law are now equal.

\subsubsection*{Exercise \ref{subsec:Exercise-traversables-5}}

To verify the identity law~(\ref{eq:identity-law-of-sequence}),
set $F=\text{Id}$ in the definition~(\ref{eq:def-sequence-for-functor-composition})
of $\text{seq}_{L}$:
\begin{align*}
{\color{greenunder}\text{expect to equal }\text{id}:}\quad & \text{seq}_{L}^{\text{Id},A}=(\text{seq}_{N}^{\text{Id},A})^{\uparrow M}\bef\text{seq}_{M}^{\text{Id},N^{A}}\\
{\color{greenunder}\text{identity laws of }\text{seq}_{M}\text{ and }\text{seq}_{N}:}\quad & =\text{id}^{\uparrow M}\bef\text{id}=\text{id}\quad.
\end{align*}

To verify the composition law~(\ref{eq:composition-law-of-sequence}),
begin with the left-hand side:
\begin{align*}
{\color{greenunder}\text{expect to equal }\text{seq}_{L}^{F\circ G,A}:}\quad & \text{seq}_{L}^{F,G^{A}}\bef(\text{seq}_{L}^{G,A})^{\uparrow F}\\
 & \quad=(\text{seq}_{N}^{F,G^{A}})^{\uparrow M}\bef\gunderline{\text{seq}_{M}^{F,N^{G^{A}}}\bef\big((\text{seq}_{N}^{G,A})^{\uparrow M}}\bef\text{seq}_{M}^{G,N^{A}}\big)^{\uparrow F}\\
{\color{greenunder}\text{naturality law of }\text{seq}_{M}:}\quad & \quad=(\text{seq}_{N}^{F,G^{A}})^{\uparrow M}\bef(\text{seq}_{N}^{G,A})^{\uparrow F\uparrow M}\bef\text{seq}_{M}^{F,G^{N^{A}}}\bef(\text{seq}_{M}^{G,N^{A}})^{\uparrow F}\\
{\color{greenunder}\text{composition laws of }\text{seq}_{M}\text{ and }\text{seq}_{N}:}\quad & \quad=(\text{seq}_{N}^{F\circ G,A})^{\uparrow M}\bef\text{seq}_{M}^{F\circ G,N^{A}}=\text{seq}_{L}^{F\circ G,A}\quad.
\end{align*}


\subsubsection*{Exercise \ref{subsec:Exercise-traversables-10}}

The empty elements of the monoids $M^{R}$ and $M^{S}$ are, by definition,
$\text{pu}_{M}(e_{R})$ and $\text{pu}_{M}(e_{S})$ respectively.
The binary operations of $M^{R}$ and $M^{S}$ are:
\begin{align*}
 & p\underset{M^{R}}{\oplus}q\triangleq p\triangleright\text{flm}_{M}\big(u^{:R}\rightarrow q\triangleright(v^{:R}\rightarrow u\oplus_{R}v)^{\uparrow M}\big)\quad,\\
 & p\underset{M^{S}}{\oplus}q\triangleq p\triangleright\text{flm}_{M}\big(u^{:S}\rightarrow q\triangleright(v^{:S}\rightarrow u\oplus_{S}v)^{\uparrow M}\big)\quad.
\end{align*}
To verify the properties of the monoid morphism, begin with the identity
law:
\begin{align*}
{\color{greenunder}\text{expect to equal }e_{M^{S}}:}\quad & \gunderline{e_{M^{R}}}\triangleright\phi^{\uparrow M}=e_{R}\triangleright\gunderline{\text{pu}_{M}\triangleright\phi^{\uparrow M}}=\gunderline{e_{R}\triangleright\phi}\triangleright\text{pu}_{M}=e_{S}\triangleright\text{pu}_{M}=e_{M^{S}}\quad.
\end{align*}
Now check the composition law. For any $p^{:M^{R}}$ and $q^{:M^{R}}$:
\begin{align*}
{\color{greenunder}\text{expect to equal }(p\underset{M^{R}}{\oplus}q)\triangleright\phi^{\uparrow M}:}\quad & (p\triangleright\phi^{\uparrow M})\underset{M^{S}}{\oplus}(q\triangleright\phi^{\uparrow M})\\
 & =(p\triangleright\gunderline{\phi^{\uparrow M})\triangleright\text{flm}_{M}}\big(u\rightarrow q\triangleright\phi^{\uparrow M}\triangleright(v\rightarrow u\oplus_{S}v)^{\uparrow M}\big)\\
{\color{greenunder}\text{naturality of }\text{flm}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(\gunderline{\phi\bef(}v\rightarrow\phi(u)\oplus_{S}v))^{\uparrow M}\big)\\
{\color{greenunder}\text{compute composition}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow\gunderline{\phi(u)\oplus_{S}\phi(v)}))^{\uparrow M}\big)\\
{\color{greenunder}\text{composition law of }\phi:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow\phi(u\oplus_{R}v))^{\uparrow M}\big)\\
{\color{greenunder}\text{move }\phi^{\uparrow M}\text{ outside}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\bef\phi^{\uparrow M}\big)\\
{\color{greenunder}\text{naturality of }\text{flm}_{M}:}\quad & =\gunderline{p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\big)}\bef\phi^{\uparrow M}\\
 & =(p\underset{M^{R}}{\oplus}q)\triangleright\phi^{\uparrow M}\quad.
\end{align*}


\subsubsection*{Exercise \ref{subsec:Exercise-traversables-10-1}}

The empty elements of the monoids $M^{R}$ and $N^{R}$ are, by definition,
are $\text{pu}_{M}(e_{R})$ and $\text{pu}_{N}(e_{R})$ respectively.
The binary operations of $M^{R}$ and $N^{R}$ are:
\begin{align*}
 & p\underset{M^{R}}{\oplus}q\triangleq p\triangleright\text{flm}_{M}\big(u^{:R}\rightarrow q\triangleright(v^{:R}\rightarrow u\oplus_{R}v)^{\uparrow M}\big)\quad,\\
 & p\underset{N^{R}}{\oplus}q\triangleq p\triangleright\text{flm}_{N}\big(u^{:R}\rightarrow q\triangleright(v^{:R}\rightarrow u\oplus_{R}v)^{\uparrow N}\big)\quad.
\end{align*}
To verify the properties of the monoid morphism, begin with the identity
law:
\begin{align*}
{\color{greenunder}\text{expect to equal }e_{N^{R}}:}\quad & \gunderline{e_{M^{R}}}\triangleright\phi=\gunderline{e_{R}\triangleright\text{pu}_{M}}\triangleright\phi=e_{R}\triangleright\text{pu}_{N}=e_{N^{R}}\quad.
\end{align*}
Now check the composition law. For any $p^{:M^{R}}$ and $q^{:M^{R}}$:
\begin{align*}
{\color{greenunder}\text{expect to equal }(p\underset{M^{R}}{\oplus}q)\triangleright\phi:}\quad & (p\triangleright\phi)\underset{N^{R}}{\oplus}(q\triangleright\phi)=(p\triangleright\phi)\triangleright\text{flm}_{N}\big(u\rightarrow q\triangleright\gunderline{\phi\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow N}}\big)\\
{\color{greenunder}\text{naturality of }\phi:}\quad & =p\triangleright\gunderline{\phi\bef\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\bef\phi\big)^{\uparrow N}}\bef\text{ftn}_{N}\\
{\color{greenunder}\text{naturality of }\phi:}\quad & =p\triangleright\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\big)^{\uparrow M}\bef\gunderline{\phi\bef\phi^{\uparrow N}\bef\text{ftn}_{N}}\\
{\color{greenunder}\text{composition law of }\phi:}\quad & =p\triangleright\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\gunderline{\big)^{\uparrow M}\bef\text{ftn}_{M}}\bef\phi\\
{\color{greenunder}\text{definition of }\text{ftn}_{M}:}\quad & =p\triangleright\text{flm}_{M}\big(u\rightarrow q\triangleright(v\rightarrow u\oplus_{R}v)^{\uparrow M}\big)\bef\phi=(p\underset{M^{R}}{\oplus}q)\triangleright\phi\quad.
\end{align*}


\subsection*{Chapter \ref{chap:monad-transformers}}

\subsubsection*{Exercise \ref{subsec:Exercise-monad-transformer-extra-layer-5}}

The identity law of monad morphisms says that $\text{pu}_{\bbnum 1}\bef\phi=\text{pu}_{M}$.
However, $\text{pu}_{\bbnum 1}=\_^{:A}\rightarrow1$ and ignores its
argument. So, $\text{pu}_{M}$ is also a function that ignores its
argument. By the monad $M$\textsf{'}s left identity law, $\text{pu}_{M}\bef\text{ftn}_{M}=\text{id}^{:M^{A}\rightarrow M^{A}}$.
So, the identity function $\text{id}:M^{A}\rightarrow M^{A}$ ignores
its argument. It always returns the same value of type $M^{A}$, say
$m_{0}$. It follows that the type $M^{A}$ has only one distinct
value, namely $m_{0}$. So $M^{A}\cong\bbnum 1$. This argument holds
for each type $A$. 

\subsubsection*{Exercise \ref{subsec:Exercise-monad-transformer-extra-layer-3}}

\textbf{(a)} For any monad morphism $\phi:M\leadsto N$, the monadic
naturality law must hold:
\[
\text{dbl}\bef\phi\overset{?}{=}\phi\bef\text{dbl}\quad.
\]
The left-hand side applied to an arbitrary value $m^{:M^{A}}$ is:
\begin{align*}
 & m\triangleright\text{dbl}\bef\phi=m\triangleright(\_\rightarrow m)^{\uparrow M}\bef\gunderline{\text{ftn}_{M}\bef\phi}\\
{\color{greenunder}\text{monad morphism law of }\phi:}\quad & =m\triangleright\gunderline{(\_\rightarrow m)^{\uparrow M}\bef\phi^{\uparrow M}}\bef\phi\bef\text{ftn}_{M}\\
{\color{greenunder}\text{compute composition}:}\quad & =m\triangleright\gunderline{(\_\rightarrow m\triangleright\phi)^{\uparrow M}\bef\phi}\bef\text{ftn}_{M}\\
{\color{greenunder}\text{naturality of }\phi:}\quad & =m\triangleright\phi\bef(\_\rightarrow m\triangleright\phi)^{\uparrow M}\bef\text{ftn}_{M}\quad.
\end{align*}
The right-hand side applied to $m$ is:
\begin{align*}
 & m\triangleright\phi\bef\text{dbl}=m\triangleright\phi\triangleright\text{dbl}=m\triangleright\phi\triangleright(\_\rightarrow m\triangleright\phi)^{\uparrow M}\bef\text{ftn}_{M}\\
 & =m\triangleright\phi\bef(\_\rightarrow m\triangleright\phi)^{\uparrow M}\bef\text{ftn}_{M}\quad.
\end{align*}
Both sides are now equal.

\textbf{(b)} The identity law holds:
\[
a\triangleright\text{pu}_{M}\bef\text{dbl}=a\triangleright\text{pu}_{M}\triangleright\text{flm}_{M}(\_\rightarrow a\triangleright\text{pu}_{M})=a\triangleright(\_\rightarrow a\triangleright\text{pu}_{M})=a\triangleright\text{pu}_{M}\quad.
\]
The composition law:
\[
\text{dbl}^{\uparrow M}\bef\text{dbl}\bef\text{ftn}_{M}\overset{?}{=}\text{ftn}_{M}\bef\text{dbl}\quad.
\]
Simplify the left-hand side, applying to an arbitrary value $p:M^{M^{A}}$.
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & p\triangleright\text{dbl}^{\uparrow M}\bef\text{dbl}\bef\text{ftn}_{M}=p\triangleright\text{dbl}^{\uparrow M}\triangleright(\_\rightarrow p\triangleright\text{dbl}^{\uparrow M})^{\uparrow M}\bef\gunderline{\text{ftn}_{M}\bef\text{ftn}_{M}}\\
{\color{greenunder}\text{associativity law of }M:}\quad & =p\triangleright\text{dbl}^{\uparrow M}\bef(\_\rightarrow p\triangleright\text{dbl}^{\uparrow M})^{\uparrow M}\bef\text{ftn}_{M}^{\uparrow M}\bef\text{ftn}_{M}\\
 & =p\triangleright\big(\text{dbl}\bef(\_\rightarrow p\triangleright\text{dbl}^{\uparrow M})\bef\text{ftn})\big)^{\uparrow M}\bef\text{ftn}_{M}=p\triangleright\big(\_\rightarrow p\triangleright\text{dbl}^{\uparrow M}\bef\text{ftn}_{M})\big)^{\uparrow M}\bef\text{ftn}_{M}\\
 & =p\triangleright\text{flm}_{M}\big(\_\rightarrow p\triangleright\text{flm}_{M}(\text{dbl})\big)\quad.
\end{align*}
The right-hand side:
\begin{align*}
{\color{greenunder}\text{right-hand side}:}\quad & p\triangleright\text{ftn}_{M}\bef\text{dbl}=p\triangleright\gunderline{\text{ftn}_{M}\bef(\_\rightarrow p\triangleright\text{ftn}_{M})^{\uparrow M}}\bef\text{ftn}_{M}\\
{\color{greenunder}\text{naturality of }\text{ftn}_{M}:}\quad & =p\triangleright(\_\rightarrow p\triangleright\text{ftn}_{M})^{\uparrow M\uparrow M}\bef\gunderline{\text{ftn}_{M}\bef\text{ftn}_{M}}\\
{\color{greenunder}\text{associativity law of }M:}\quad & =p\triangleright(\_\rightarrow p\triangleright\text{ftn}_{M})^{\uparrow M\uparrow M}\bef\text{ftn}_{M}^{\uparrow M}\bef\text{ftn}_{M}\\
 & =p\triangleright\big((\_\rightarrow p\triangleright\text{ftn}_{M})^{\uparrow M}\bef\text{ftn}_{M}\big)^{\uparrow M}\bef\text{ftn}_{M}\\
 & =p\triangleright\text{flm}_{M}\big(\text{flm}_{M}(\_\rightarrow p\triangleright\text{ftn}_{M})\big)\quad.
\end{align*}
The simplification is stuck \textemdash{} we cannot make the two sides
coincide; no law can be applied to simplify further.

Choosing specific values $p=\text{pu}_{M}(m)$ or $p=\text{pu}_{M}^{\uparrow M}(m)$
does not help to obtain a counter-example: the laws hold for those
specific choices.

To obtain a counter-example, consider a specific non-commutative monad,
such as \lstinline!List!. For the \lstinline!List! monad, the \lstinline!double!
function repeats the entire list as many times as elements in the
list:
\begin{lstlisting}
scala> double(List(1, 2))
res0: List[Int] = List(1, 2, 1, 2)

scala> val ll = List(List(1, 2), List(3))
ll: List[List[Int]] = List(List(1, 2), List(3))

scala> double(ll.flatten)
res1: List[Int] = List(1, 2, 3, 1, 2, 3, 1, 2, 3)

scala> double(ll.map(double))
res2: List[List[Int]] = List(List(1, 2, 1, 2), List(3), List(1, 2, 1, 2), List(3)) 

scala> double(ll.map(double)).flatten
res3: List[Int] = List(1, 2, 1, 2, 3, 1, 2, 1, 2, 3)
\end{lstlisting}
The composition law does not hold.

\subsubsection*{Exercise \ref{subsec:Statement-search-and-selector-monads}}

The identity law does not hold:
\begin{align*}
{\color{greenunder}\text{expect to equal }a\triangleright\text{pu}_{\text{Search}}:}\quad & a\triangleright\text{pu}_{\text{Sel}}\bef\text{finder}=(\_^{:A\rightarrow\bbnum 1+P}\rightarrow a)\triangleright\text{finder}\\
 & =p\rightarrow p(a)\triangleright\,\begin{array}{|c||cc|}
 & \bbnum 1 & A\\
\hline \bbnum 1 & \text{id} & \bbnum 0\\
P & \bbnum 0 & a
\end{array}\,=p\rightarrow a\triangleright p\bef(\_\rightarrow a)^{\uparrow\text{Opt}}\neq p\rightarrow\bbnum 0+a\quad.
\end{align*}
The result should have been $\text{pu}_{\text{Search}}(a)=\_\rightarrow\text{pu}_{\text{Opt}}(a)=\_\rightarrow\bbnum 0+a$.

\subsubsection*{Exercise \ref{subsec:Exercise-selector-and-continuation-monads}}

To verify the identity law:
\begin{align*}
{\color{greenunder}\text{expect to equal }a\triangleright\text{pu}_{\text{Cont}}:}\quad & a\triangleright\text{pu}_{\text{Sel}}\bef\text{scc}=f\rightarrow f(f\triangleright(\gunderline{a\triangleright\text{pu}_{\text{Sel}}}))\\
{\color{greenunder}\text{use definition of }\text{pu}_{\text{Sel}}:}\quad & =f\rightarrow f(f\triangleright(\_\rightarrow a))=f\rightarrow f(a)=a\triangleright\text{pu}_{\text{Cont}}\quad.
\end{align*}
To verify the composition law, we begin with the flipped Kleisli formulation
of the two monads. The flipped Kleisli functions have types $\left(B\rightarrow P\right)\rightarrow A\rightarrow P$
(for \lstinline!Cont!) and $\left(B\rightarrow P\right)\rightarrow A\rightarrow B$
(for \lstinline!Sel!).
\begin{align*}
 & f^{:\left(B\rightarrow P\right)\rightarrow A\rightarrow P}\tilde{\diamond}_{_{\text{Cont}}}g^{:\left(C\rightarrow P\right)\rightarrow B\rightarrow P}\triangleq h^{:C\rightarrow P}\rightarrow h\triangleright g\triangleright f=g\bef f\quad,\\
 & f^{:\left(B\rightarrow P\right)\rightarrow A\rightarrow B}\tilde{\diamond}_{_{\text{Sel}}}g^{:\left(C\rightarrow P\right)\rightarrow B\rightarrow C}\triangleq h^{:C\rightarrow P}\rightarrow f(g(h)\bef h)\bef g(h)\quad.
\end{align*}
We need to check that \lstinline!Sel!\textsf{'}s Kleisli composition ($\tilde{\diamond}_{_{\text{Sel}}}$)
is mapped to \lstinline!Cont!\textsf{'}s Kleisli composition ($\tilde{\diamond}_{_{\text{Cont}}}$).
First, we need to modify \lstinline!scc! so that it works on the
flipped Kleisli functions:
\[
\tilde{\text{scc}}^{A,B}:\left(\left(B\rightarrow P\right)\rightarrow A\rightarrow B\right)\rightarrow\left(B\rightarrow P\right)\rightarrow A\rightarrow P\quad,\quad\tilde{\text{scc}}=c^{:\left(\left(B\rightarrow P\right)\rightarrow A\rightarrow B\right)}\rightarrow k^{:B\rightarrow P}\rightarrow c(k)\bef k\quad.
\]

Note that the implementation of \lstinline!scc! is uniquely determined
by its type signature, and so is the implementation of $\tilde{\text{scc}}$.
So, we can spare ourselves the effort of translating $\text{scc}$
into $\tilde{\text{scc}}$.

Now it remains to show that for any $f^{:\left(B\rightarrow P\right)\rightarrow A\rightarrow B}$
and $g^{:\left(C\rightarrow P\right)\rightarrow B\rightarrow C}$
the law holds:
\[
(f\triangleright\tilde{\text{scc}}^{A,B})\tilde{\diamond}_{_{\text{Cont}}}(g\triangleright\tilde{\text{scc}}^{B,C})=(f\tilde{\diamond}_{_{\text{Sel}}}g)\triangleright\tilde{\text{scc}}^{A,C}\quad.
\]
Rewrite the two sides of the law separately:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & (f\triangleright\tilde{\text{scc}})\tilde{\diamond}_{_{\text{Cont}}}(g\triangleright\tilde{\text{scc}})=(g\triangleright\tilde{\text{scc}})\bef(f\triangleright\tilde{\text{scc}})=(k\rightarrow g(k)\bef k)\bef(h\rightarrow f(h)\bef h)\\
 & \quad=k\rightarrow f(g(k)\bef k)\bef g(k)\bef k\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & (f\tilde{\diamond}_{_{\text{Sel}}}g)\triangleright\tilde{\text{scc}}=\big(h\rightarrow f(g(h)\bef h)\bef g(h)\big)\triangleright\tilde{\text{scc}}=k\rightarrow\big(h\rightarrow f(g(h)\bef h)\bef g(h)\big)(k)\bef k\\
 & \quad=k\rightarrow f(g(k)\bef k)\bef g(k)\bef k\quad.
\end{align*}
Both sides are now equal.

\subsubsection*{Exercise \ref{par:Exercise-mt-3-1}}

The operations $\text{pu}_{N}$ and $\text{ftn}_{N}$ are defined
by:
\[
\text{pu}_{N}\triangleq\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0
\end{array}\quad,\quad\quad\text{ftn}_{N}\triangleq\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
M^{A} & \bbnum 0 & \text{id}\\
M^{N^{A}} & \bbnum 0 & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\quad.
\]

To verify the identity law:
\begin{align*}
{\color{greenunder}\text{expect to equal }\text{pu}_{M}:}\quad & \text{pu}_{N}\bef\gamma_{M}=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0
\end{array}\,\bef\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
M^{A} & \text{id}
\end{array}\,=\text{pu}_{M}\quad.
\end{align*}
To verify the composition law, transform separately the two sides
of the law:
\begin{align*}
{\color{greenunder}\text{left-hand side}:}\quad & \gamma_{M}\bef\gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}=\,\begin{array}{|c||c|}
 & M^{N^{A}}\\
\hline N^{A} & \text{pu}_{M}\\
M^{N^{A}} & \text{id}
\end{array}\,\bef\gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline N^{A} & \gunderline{\text{pu}_{M}\bef\gamma_{M}^{\uparrow M}}\bef\text{ftn}_{M}\\
M^{N^{A}} & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\\
 & \quad=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline N^{A} & \gamma_{M}\bef\gunderline{\text{pu}_{M}\bef\text{ftn}_{M}}\\
M^{N^{A}} & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
M^{A} & \text{id}\\
M^{N^{A}} & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\quad,\\
{\color{greenunder}\text{right-hand side}:}\quad & \text{ftn}_{N}\bef\gamma_{M}=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
M^{A} & \bbnum 0 & \text{id}\\
M^{N^{A}} & \bbnum 0 & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\,\bef\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
M^{A} & \text{id}
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
M^{A} & \text{id}\\
M^{N^{A}} & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\quad.
\end{align*}
The two sides are now equal.

\subsubsection*{Exercise \ref{par:Exercise-mt-3}}

The operations $\text{ftn}_{L}$ and $\text{ftn}_{N}$ are defined
as usual for a free pointed monad:
\begin{align*}
 & \text{ftn}_{L}\triangleq\,\begin{array}{|c||c|}
 & L^{A}\\
\hline L^{A} & \text{id}\\
K^{L^{A}} & k\rightarrow\bbnum 0+k\triangleright\gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}
\end{array}\,=\,\begin{array}{|c||cc|}
 & A & K^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \text{id}\\
K^{L^{A}} & \bbnum 0 & \gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}
\end{array}\quad,\\
 & \text{ftn}_{N}\triangleq\,\begin{array}{|c||c|}
 & N^{A}\\
\hline N^{A} & \text{id}\\
M^{N^{A}} & m\rightarrow\bbnum 0+m\triangleright\gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\,=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
M^{A} & \bbnum 0 & \text{id}\\
M^{N^{A}} & \bbnum 0 & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\quad.
\end{align*}

We define $\psi$ by:
\[
\psi=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi
\end{array}\quad.
\]

To verify the identity law for $\psi$:
\[
\text{pu}_{L}\bef\psi=\,\begin{array}{|c||cc|}
 & A & K^{A}\\
\hline A & \text{id} & \bbnum 0
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi
\end{array}\,=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0
\end{array}\,=\text{pu}_{N}\quad.
\]

To verify the composition law for $\psi$, write both sides of the
law separately:
\begin{align*}
\psi^{\uparrow L}\bef\psi\bef\text{ftn}_{N} & =\,\begin{array}{|c||ccc|}
 & A & M^{A} & K^{N^{A}}\\
\hline A & \text{id} & \bbnum 0 & \bbnum 0\\
K^{A} & \bbnum 0 & \phi & \bbnum 0\\
K^{L^{A}} & \bbnum 0 & \bbnum 0 & \psi^{\uparrow K}
\end{array}\,\bef\,\begin{array}{|c||ccc|}
 & A & M^{A} & M^{N^{A}}\\
\hline A & \text{id} & \bbnum 0 & \bbnum 0\\
M^{A} & \bbnum 0 & \text{id} & \bbnum 0\\
K^{N^{A}} & \bbnum 0 & \bbnum 0 & \phi
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
M^{A} & \bbnum 0 & \text{id}\\
M^{N^{A}} & \bbnum 0 & \gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\\
 & =\,\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi\\
K^{L^{A}} & \bbnum 0 & \psi^{\uparrow K}\bef\phi\bef\gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}
\end{array}\quad,\\
\text{ftn}_{L}\bef\psi & =\,\begin{array}{|c||cc|}
 & A & K^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \text{id}\\
K^{L^{A}} & \bbnum 0 & \gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}
\end{array}\,\bef\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi
\end{array}\,=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi\\
K^{L^{A}} & \bbnum 0 & \gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\phi
\end{array}\quad.
\end{align*}
It remains to show that:
\[
\psi^{\uparrow K}\bef\phi\bef\gamma_{M}^{\uparrow M}\bef\text{ftn}_{M}\overset{?}{=}\gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\phi\quad.
\]
The monad morphism law for $\phi$ gives:
\[
\phi\bef\phi^{\uparrow M}\bef\text{ftn}_{M}=\text{ftn}_{K}\bef\phi\quad.
\]
It remains to show that:
\begin{align*}
 & \gunderline{\psi^{\uparrow K}\bef\phi}\bef\gamma_{M}^{\uparrow M}\overset{?}{=}\gunderline{\gamma_{K}^{\uparrow K}\bef\phi}\bef\phi^{\uparrow M}\quad.\\
 & \phi\bef\psi^{\uparrow M}\bef\gamma_{M}^{\uparrow M}=\phi\bef(\psi\bef\gamma_{M})^{\uparrow M}\overset{?}{=}\phi\bef\gamma_{K}^{\uparrow M}\bef\phi^{\uparrow M}=\phi\bef(\gamma_{K}\bef\phi)^{\uparrow M}\quad.
\end{align*}
It remains to show that $\psi\bef\gamma_{M}=\gamma_{K}\bef\phi$.
For that, we use the identity law of $\phi$:
\begin{align*}
 & \psi\bef\gamma_{M}=\,\begin{array}{|c||cc|}
 & A & M^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \phi
\end{array}\,\bef\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
M^{A} & \text{id}
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
K^{A} & \phi
\end{array}\quad,\\
 & \gamma_{K}\bef\phi=\,\begin{array}{|c||c|}
 & K^{A}\\
\hline A & \text{pu}_{K}\\
K^{A} & \text{id}
\end{array}\,\bef\phi=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \gunderline{\text{pu}_{K}\bef\phi}\\
K^{A} & \phi
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
K^{A} & \phi
\end{array}\quad.
\end{align*}


\subsubsection*{Exercise \ref{par:Exercise-mt-3-3}}

\textbf{(a)} Verify the naturality law of $\psi$, assuming the naturality
laws of $\phi$ and $\chi$:
\begin{align*}
 & (f^{:A\rightarrow B})^{\uparrow(K+L)}\bef\psi=\,\begin{array}{|c||cc|}
 & K^{B} & L^{B}\\
\hline K^{A} & f^{\uparrow K} & \bbnum 0\\
L^{A} & \bbnum 0 & f^{\uparrow L}
\end{array}\,\bef\,\begin{array}{|c||c|}
 & M^{B}\\
\hline K^{B} & \phi\\
L^{B} & \chi
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{B}\\
\hline K^{A} & \gunderline{f^{\uparrow K}\bef\phi}\\
L^{A} & \gunderline{f^{\uparrow L}\bef\chi}
\end{array}\,=\,\begin{array}{|c||c|}
 & M^{B}\\
\hline K^{A} & \phi\bef f^{\uparrow M}\\
L^{A} & \chi\bef f^{\uparrow M}
\end{array}\quad,\\
 & \psi\bef f^{\uparrow M}=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline K^{A} & \phi\\
L^{A} & \chi
\end{array}\,\bef f^{\uparrow M}=\,\begin{array}{|c||c|}
 & M^{B}\\
\hline K^{A} & \phi\bef f^{\uparrow M}\\
L^{A} & \chi\bef f^{\uparrow M}
\end{array}\quad.
\end{align*}

\textbf{(b)} Given the definitions of $\phi$ and $\chi$, we can
write $\psi$ as:
\[
\psi=\,\begin{array}{|c||c|}
 & M^{A}\\
\hline K^{A} & \phi\\
L^{A} & \chi
\end{array}\quad.
\]
Assuming the naturality law of $\psi$, verify the naturality law
of $\phi$ by applying to an arbitrary value $k^{:K^{A}}$ and expressing
$f^{\uparrow K}$ through $f^{\uparrow(K+L)}$: 
\begin{align*}
{\color{greenunder}\text{expect to equal }k\triangleright\phi\bef f^{\uparrow M}:}\quad & k\triangleright f^{\uparrow K}\bef\phi=k\triangleright f^{\uparrow K}\triangleright(k\rightarrow k+\bbnum 0)\triangleright\psi=\big((k\triangleright f^{\uparrow K})+\bbnum 0\big)\triangleright\psi\\
 & =(k+\bbnum 0)\triangleright\,\begin{array}{|c||cc|}
 & K^{B} & L^{B}\\
\hline K^{A} & f^{\uparrow K} & \bbnum 0\\
L^{A} & \bbnum 0 & f^{\uparrow L}
\end{array}\,\bef\psi=(k+\bbnum 0)\triangleright\gunderline{f^{\uparrow(K+L)}\bef\psi}\\
{\color{greenunder}\text{naturality law of }\psi:}\quad & =(\gunderline{k+\bbnum 0})\triangleright\psi\bef f^{\uparrow M}=k\triangleright\gunderline{(x\rightarrow x+\bbnum 0)\bef\psi}\bef f^{\uparrow M}=k\triangleright\phi\bef f^{\uparrow M}\quad.
\end{align*}

The proof of the naturality law of $\chi$ is analogous.

\subsubsection*{Exercise \ref{par:Exercise-mt-3-2-1}}

\textbf{(a)} Choose the monads $K^{A}\triangleq R\rightarrow A$ and
$M^{A}\triangleq S\rightarrow A$, where $R$, $S$ are fixed (but
arbitrary) types. Define $P^{A}\triangleq A+T_{K}^{M,A}$ and show
that there exist no monad morphisms $\phi:M^{A}\rightarrow P^{A}$.
If such $\phi$ exists, it must satisfy the identity law, $\text{pu}_{M}\bef\phi=\text{pu}_{P}$.
The type signature of $\phi$ is:
\[
\phi:(R\rightarrow A)\rightarrow A+(S\rightarrow R\rightarrow A)\quad.
\]
The only fully parametric implementation is:
\[
\phi:f^{:R\rightarrow A}\rightarrow\bbnum 0+(\_^{:S}\rightarrow f)\quad,
\]
because we cannot produce values of type $A+\bbnum 0$ since we cannot
create values of the unknown type $R$. However, this implementation
of $\phi$ does not satisfy the identity law since $\text{pu}_{P}$
must return values of type $A+\bbnum 0$.

\textbf{(b)} Given $\theta_{K}:K^{A}\rightarrow M^{A}$, we define:
\[
\theta_{L}:L^{A}\rightarrow M^{A}\quad,\quad\quad\theta_{L}\triangleq\,\begin{array}{|c||c|}
 & M^{A}\\
\hline A & \text{pu}_{M}\\
K^{A} & \theta_{K}
\end{array}\quad.
\]
The proof becomes shorter if we express $\theta_{L}$ as:
\[
\theta_{L}=\gamma_{K}\bef\theta_{K}\quad,\quad\quad\gamma_{K}\triangleq\,\begin{array}{|c||c|}
 & K^{A}\\
\hline A & \text{pu}_{K}\\
K^{A} & \text{id}
\end{array}\quad.
\]
We already know that $\gamma_{K}:L\leadsto K$ is a monad morphism
(Exercise~\ref{par:Exercise-mt-3-1}). So, $\theta_{L}$ is a composition
of monad morphisms.

\textbf{(c)} Choose $K^{A}\triangleq\bbnum 1$ (the unit monad) and
note that $L^{A}=A+\bbnum 1$ is the standard \lstinline!Option!
monad (there are no other lawful monad implementations for the type
constructor $L^{A}\triangleq A+\bbnum 1$). Choose $M\triangleq L$,
so that a monad morphism $\theta_{L}\triangleq\text{id}^{:L\leadsto M}$
exists. If it were possible to define a monad morphism $K\leadsto M$,
we would have a monad morphism $\bbnum 1\rightarrow\bbnum 1+A$, but
this is impossible: the only natural transformation of type $\bbnum 1\rightarrow\bbnum 1+A$
is $1\rightarrow1+\bbnum 0$, which does not satisfy the identity
law of monad morphisms, $\text{pu}_{K}\bef\theta_{K}=\text{pu}_{M}$,
since it never returns any values of type $\bbnum 0+A$. (Generally,
the existence of a monad morphism $\bbnum 1\leadsto M$ means that
$M=\bbnum 1$.)

\textbf{(d)} By Exercise~\ref{par:Exercise-mt-3-3}, we have a natural
transformation $\theta_{K}:K^{A}\rightarrow A$ defined by:
\[
\theta_{K}\triangleq(k\rightarrow\bbnum 0+k)\bef\theta_{L}\quad.
\]
The given runner $\theta_{L}$ is then expressed through $\theta_{K}$
as:
\[
\theta_{L}=\,\begin{array}{|c||c|}
 & A\\
\hline A & \text{id}\\
K^{A} & \theta_{K}
\end{array}\quad,
\]
because the natural transformation in the upper row of the matrix
has the type signature $A\rightarrow A$ and so must be an identity
function.

It remains to verify the monad morphism laws of $\theta_{K}:K\leadsto\text{Id}$.
The identity law is:
\[
\text{pu}_{K}\bef\theta_{K}\overset{?}{=}\text{pu}_{\text{Id}}=\text{id}\quad.
\]
 Since the function $p\triangleq\text{pu}_{K}\bef\theta_{K}$ has
type signature $A\rightarrow A$ and is a natural transformation (as
a composition of two natural transformations), that function must
satisfy $p\bef f=f\bef p$ for any function $f^{:A\rightarrow B}$,
and so $p$ must be equal to the identity function.

The composition law of $\theta_{K}$ is an equation for functions
of type $K^{K^{A}}\rightarrow A$:
\[
\text{ftn}_{K}\bef\theta_{K}\overset{?}{=}\theta_{K}^{\uparrow K}\bef\theta_{K}\bef\text{ftn}_{\text{Id}}=\theta_{K}\bef\theta_{K}\quad.
\]
Applied to an arbitrary value $k:K^{K^{A}}$, this law becomes:
\[
k\triangleright\text{ftn}_{K}\bef\theta_{K}\overset{?}{=}k\triangleright\theta_{K}^{\uparrow K}\bef\theta_{K}\quad.
\]
Since $\theta_{K}$ is defined via $\theta_{L}$, we need to use the
composition law of $\theta_{L}$:
\[
\text{ftn}_{L}\bef\theta_{L}\overset{!}{=}\theta_{L}^{\uparrow L}\bef\theta_{L}\bef\text{ftn}_{\text{Id}}=\theta_{L}^{\uparrow K}\bef\theta_{L}\quad.
\]
Rewrite this law as:
\begin{align*}
 & \text{ftn}_{L}\bef\theta_{L}=\,\begin{array}{|c||cc|}
 & A & K^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \text{id}\\
K^{L^{A}} & \bbnum 0 & \gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}
\end{array}\,\bef\,\begin{array}{|c||c|}
 & A\\
\hline A & \text{id}\\
K^{A} & \theta_{K}
\end{array}\,=\,\begin{array}{|c||c|}
 & A\\
\hline A & \text{id}\\
K^{A} & \theta_{K}\\
K^{L^{A}} & \gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\theta_{K}
\end{array}\quad,\\
 & \theta_{L}^{\uparrow L}\bef\theta_{L}=\,\begin{array}{|c||cc|}
 & A & K^{A}\\
\hline A & \text{id} & \bbnum 0\\
K^{A} & \bbnum 0 & \text{id}\\
K^{L^{A}} & \bbnum 0 & \theta_{L}^{\uparrow K}
\end{array}\,\bef\,\begin{array}{|c||c|}
 & A\\
\hline A & \text{id}\\
K^{A} & \theta_{K}
\end{array}\,=\,\begin{array}{|c||c|}
 & A\\
\hline A & \text{id}\\
K^{A} & \theta_{K}\\
K^{L^{A}} & \theta_{L}^{\uparrow K}\bef\theta_{K}
\end{array}\quad.
\end{align*}
The third rows of the matrices give the equation for functions of
type $K^{L^{A}}\rightarrow A$:
\[
\gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\theta_{K}\overset{!}{=}\theta_{L}^{\uparrow K}\bef\theta_{K}\quad.
\]
Apply both sides of this equation to an arbitrary value $p$ of type
$K^{\bbnum 0+K^{A}}$, defined via an arbitrary value $k^{:K^{K^{A}}}$
as $p\triangleq k\triangleright(x\rightarrow\bbnum 0+x)^{\uparrow K}$:
\begin{align*}
 & p\triangleright\gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\theta_{K}\overset{!}{=}p\triangleright\theta_{L}^{\uparrow K}\bef\theta_{K}\quad,\\
{\color{greenunder}\text{or equivalently}:}\quad & k\triangleright(x\rightarrow\bbnum 0+x)^{\uparrow K}\bef\gamma_{K}^{\uparrow K}\bef\text{ftn}_{K}\bef\theta_{K}\overset{!}{=}k\triangleright(x\rightarrow\bbnum 0+x)^{\uparrow K}\bef\theta_{L}^{\uparrow K}\bef\theta_{K}\quad.
\end{align*}
We compute some sub-expressions separately:
\begin{align*}
 & (x\rightarrow\bbnum 0+x)\bef\gamma_{K}=(x\rightarrow\bbnum 0+x)\bef\,\begin{array}{|c||c|}
 & K^{A}\\
\hline A & \text{pu}_{K}\\
K^{A} & \text{id}
\end{array}\,=x\rightarrow x=\text{id}\quad,\\
 & (x\rightarrow\bbnum 0+x)\bef\theta_{L}=\theta_{K}\quad.
\end{align*}
After these simplifications, the composition law of $\theta_{L}$
gives:
\[
k\triangleright\text{ftn}_{K}\bef\theta_{K}\overset{!}{=}k\triangleright\theta_{K}^{\uparrow K}\bef\theta_{K}\quad.
\]
This is the composition law of $\theta_{K}$.

\subsubsection*{Exercise \ref{par:Exercise-mt-3-2}}

Consider the monad morphism $\varepsilon:K\leadsto\bbnum 1$ defined
by $\varepsilon\triangleq\_\rightarrow1$. (This is a monad morphism
because all its laws reduce to the equation $1=1$.) By Exercise~\ref{par:Exercise-mt-3},
we can implement a corresponding monad morphism between free pointed
monads $\text{Id}+K\leadsto\text{Id}+\bbnum 1$. Define $\delta$
as that monad morphism. The monad $\text{Id}+\bbnum 1$ (the free
pointed monad on $\bbnum 1$) is the standard \lstinline!Option!
monad. So, we have implemented a monad morphism $\delta:A+K^{A}\rightarrow\text{Opt}^{A}$.

\subsubsection*{Exercise \ref{subsec:Exercise-effectful-list-not-monad}}

\textbf{(a)} Consider the requirement of mapping $L^{L^{A}}\rightarrow L^{A}$
via a \lstinline!flatten! function:
\[
\text{ftn}_{L}:\bbnum 1+L^{A}\times M^{L^{L^{A}}}\rightarrow\bbnum 1+A\times M^{L^{A}}\quad,\quad\quad\text{ftn}_{L}=\text{???}
\]
The result value of this function cannot always be $1$, or else it
will fail the monad laws. This function must sometimes return a pair
of type $A\times M^{L^{A}}$. Let us rewrite the type signature of
\lstinline!flatten! as
\[
\text{ftn}_{L}:\bbnum 1+(\bbnum 1+A\times M^{L^{A}})\times M^{L^{L^{A}}}\rightarrow\bbnum 1+A\times M^{L^{A}}\quad.
\]
Consider input values of the form:
\[
x\triangleq\bbnum 0^{:\bbnum 1}+(1+\bbnum 0^{:A\times M^{L^{A}}})\times m^{:M^{L^{L^{A}}}}\quad.
\]
the result of evaluating $\text{ftn}_{L}(x)$ must be $1+\bbnum 0$:
a fully parametric function cannot extract values of type $A$ from
a value of type $M^{A}$ with an unknown monad $M$. Note that the
value $(1+\bbnum 0^{:A\times M^{L^{A}}})$ represents an empty list;
let us denote that value by $\text{Nil}^{:L^{A}}$. So, we must have:
\[
\text{ftn}_{L}(\bbnum 0+\text{Nil}\times(...))=\text{Nil}\quad.
\]
Since the only way to extract list values is to run the $M$-effects,
the implementation of $\text{ftn}_{L}$ must proceed recursively,
extracting list elements left to right. Now consider $M=\text{Id}$
(so we can simply omit the functor layers of $M$) and the nested
list value:
\[
z\triangleq\left[p,q,\text{Nil},r,s,...\right]\quad,
\]
 where $p$, $q$, $r$, $s$ are some sub-lists of type $L^{A}$
with some type $A$. Then the value $z$ has the form:
\[
z=\bbnum 0+p\times(\bbnum 0+q\times(\bbnum 0+\text{Nil}\times(\bbnum 0+r\times(\bbnum 0+s\times(1+\bbnum 0)))))\quad.
\]
As we must have $\text{ftn}_{L}(\bbnum 0+\text{Nil}\times t)=\text{Nil}$
for any $t^{:M^{L^{L^{A}}}}$, the recursive evaluation of $\text{ftn}_{L}(z)$
will give:
\[
\text{ftn}_{L}(z)=...\text{ftn}_{L}(\bbnum 0+\text{Nil}\times(\bbnum 0+r\times(\bbnum 0+s\times(1+\bbnum 0))))=...\text{Nil}\quad.
\]
So, the result may be some function of $p$ and $q$ (which could
be $p\pplus q$) but cannot depend on $r$ and $s$. We have shown
that $\text{ftn}_{L}$ must ignore all sub-lists that follow an empty
sub-list. Regardless of how we implement $\text{ftn}_{L}$ (and whether
it satisfies the monad laws), the result of evaluating $\text{ftn}_{L}(z)$
cannot be the full concatenation $p\pplus q\pplus r\pplus s$ because
the computation must ignore the sub-lists $r$ and $s$.

\textbf{(b)} We will now implement the non-standard monad instance
for the \lstinline!List! constructor:
\begin{lstlisting}[mathescape=true]
def pure[A](x: A): List[A] = List(x) // $\color{dkgreen}\textrm{pu}_L$
def flatten[A](p: List[List[A]]): List[A] = if (p.exists(_.isEmpty)) Nil else p.flatten  // $\color{dkgreen}\textrm{ftn}_L$
\end{lstlisting}
The new \lstinline!flatten! function gives  the same results as the
standard \lstinline!flatten! method of the \lstinline!List! type
constructor, except if one of the nested sub-lists is empty. In that
case, the \lstinline!flatten! function returns an empty list (unlike
\lstinline!List!\textsf{'}s standard \lstinline!flatten! method).

To verify the monad laws, we use the known fact that the standard
\lstinline!List! monad obeys the laws. So, we only need to check
the laws in the cases when the new \lstinline!flatten! function is
applied to a value of type \lstinline!List[List[A]]! having an empty
nested sub-list. That case cannot arise in the identity laws since
neither $\text{pu}_{L}(x^{:A})$ nor $\text{pu}_{L}^{\uparrow L}(x^{:\text{List}^{A}})$
ever returns a value with a nested empty sub-list:
\begin{lstlisting}
pure(List(a, b, c)) == List(List(a, b, c))
List(a, b, c).map(pure) == List(List(a), List(b), List(c))
\end{lstlisting}
Applying \lstinline!flatten! to these values gives the initial list
\lstinline!List(a, b, c)!. So, both identity laws hold.

It remains to check the associativity law, which is an equality between
functions $\text{ftn}_{L}\bef\text{ftn}_{L}$ and $\text{ftn}_{L}^{\uparrow L}\bef\text{ftn}_{L}$
of type \lstinline!List[List[List[A]]] => List[A]!. A value $p$
of type \lstinline!List[List[List[A]]]! could contain a nested empty
list at the first and/or the second nesting depth, for instance:
\begin{lstlisting}
val p1: List[List[List[Int]]] = List(List[List[Int]](), List(List(123), List(456)))
val p2: List[List[List[Int]]] = List(List(List[Int](), List(123)), List(List(456)))
\end{lstlisting}
Whenever $p$ contains an empty sub-list at the first nesting depth,
we will have $p\triangleright\text{ftn}_{L}=\text{Nil}$ because $\text{ftn}_{L}$
explicitly checks for the existence of an empty sub-list. So, $p\triangleright\text{ftn}_{L}\bef\text{ftn}_{L}=\text{Nil}$.
On the other hand, $p\triangleright\text{ftn}_{L}^{\uparrow L}$ will
be again a list containing an empty sub-list, for example:
\begin{lstlisting}
scala> p1.map(flatten)
res0: List[List[Int]] = List(List(), List(123, 456))
\end{lstlisting}
We will then have $p\triangleright\text{ftn}_{L}^{\uparrow L}\triangleright\text{ftn}_{L}=\text{Nil}$,
and the law holds. If $p$ contains an empty sub-list at the \emph{second}
nesting depth, $p\triangleright\text{ftn}_{L}$ will contain an empty
sub-list at the first nesting depth, for example:
\begin{lstlisting}
scala> flatten(p2)
res1: List[List[Int]] = List(List(), List(123), List(456)) 
\end{lstlisting}
So $p\triangleright\text{ftn}_{L}\bef\text{ftn}_{L}=\text{Nil}$.
On the other hand, $p\triangleright\text{ftn}_{L}^{\uparrow L}$ will
also be a list with an empty sub-list at the first nesting depth,
for example:
\begin{lstlisting}
scala> p2.map(flatten)
res2: List[List[Int]] = List(List(), List(456))
\end{lstlisting}
Applying $\text{ftn}_{L}$ to the last result, we will get an empty
list. Thus, $p\triangleright\text{ftn}_{L}\triangleright\text{ftn}_{L}=\text{Nil}$,
and the law again holds. So, we have shown that the associativity
law holds for the non-standard \lstinline!List! monad.

\textbf{(c)} Try implementing the method $\text{flift}:M^{A}\rightarrow L^{A}$.
The function \lstinline!flift! must produce a value of type $L^{A}\cong\bbnum 1+A\times M^{L^{A}}$.
Since $M$ is an arbitrary monad, we cannot extract a value of type
$A$ out of $M^{A}$ while keeping the code fully parametric. So,
we can implement \lstinline!flift! only by defining $\text{flift}\triangleq\_^{:M^{A}}\rightarrow\text{Nil}^{:L^{A}}$.
However, that definition will not satisfy the identity law:
\[
\text{pu}_{M}\bef\text{flift}=(\_\rightarrow\text{Nil})\neq\text{pu}_{L}\quad.
\]


\subsubsection*{Exercise \ref{subsec:Exercise-combined-codensity-monad}}

\textbf{(a)} Denote for brevity $\text{Cod}_{F}^{M,A}\triangleq C^{A}$.
The naturality law for functions $c^{:C^{A}}$ says that for any $k^{:A\rightarrow F^{X}}$
and $q^{:X\rightarrow Y}$, we have:
\[
(k^{:A\rightarrow F^{X}}\bef q^{\uparrow F})\triangleright c^{Y}=k\triangleright c^{X}\bef q^{\uparrow M\uparrow F}\quad.
\]

The flipped Kleisli method $\tilde{\text{pu}}_{C}$ is defined by:
\[
\tilde{\text{pu}}_{C}:\forall X.\,(A\rightarrow F^{X})\rightarrow A\rightarrow F^{M^{X}}\quad,\quad\quad\tilde{\text{pu}}_{C}\triangleq\forall X.\,k^{:A\rightarrow F^{X}}\rightarrow k\bef\text{pu}_{M}^{\uparrow F}\quad.
\]

To verify the left identity law:
\begin{align*}
{\color{greenunder}\text{expect to equal }g:}\quad & \tilde{\text{pu}}_{C}\tilde{\diamond}\,g^{:\forall Y.\,(B\rightarrow F^{Y})\rightarrow A\rightarrow F^{M^{Y}}}=\forall Z.\,k^{:B\rightarrow F^{Z}}\rightarrow\big(\gunderline{k\triangleright g^{Z}\triangleright\tilde{\text{pu}}_{C}^{M^{Z}}}\big)\bef\text{ftn}_{M}^{\uparrow F}\\
 & =\forall Z.\,k^{:C\rightarrow F^{Z}}\rightarrow\big(g^{Z}(k)\bef\gunderline{\text{pu}_{M}^{\uparrow F}\big)\bef\text{ftn}_{M}^{\uparrow F}}\\
{\color{greenunder}\text{left identity law of }M:}\quad & =\forall Z.\,k^{:C\rightarrow F^{Z}}\rightarrow g^{Z}(k)=g\quad.
\end{align*}

To verify the right identity law:
\begin{align*}
{\color{greenunder}\text{expect to equal }f:}\quad & f^{:\forall X.\,(B\rightarrow F^{X})\rightarrow A\rightarrow F^{M^{X}}}\tilde{\diamond}\,\tilde{\text{pu}}_{C}=\forall Z.\,k^{:B\rightarrow F^{Z}}\rightarrow\big(\gunderline{k\triangleright\tilde{\text{pu}}_{C}^{Z}}\triangleright f^{M^{Z}}\big)\bef\text{ftn}_{M}^{\uparrow F}\\
 & =\forall Z.\,k^{:B\rightarrow F^{Z}}\rightarrow\big(\gunderline{(k\bef\text{pu}_{M}^{\uparrow F})\triangleright f^{M^{Z}}}\big)\bef\text{ftn}_{M}^{\uparrow F}\\
{\color{greenunder}\text{naturality law of }f:}\quad & =\forall Z.\,k^{:B\rightarrow F^{Z}}\rightarrow\big(k\triangleright f^{Z}\bef\gunderline{\text{pu}_{M}^{\uparrow M\uparrow F}\big)\bef\text{ftn}_{M}^{\uparrow F}}\\
{\color{greenunder}\text{right identity law of }M:}\quad & =\forall Z.\,k^{:B\rightarrow F^{Z}}\rightarrow k\triangleright f^{Z}=f\quad.
\end{align*}

To verify the associativity law, write its two sides separately; omit
all types for brevity:
\begin{align*}
 & (f\,\tilde{\diamond}\,g)\,\tilde{\diamond}\,h=l\rightarrow\big(l\triangleright h\triangleright(\gunderline{f\,\tilde{\diamond}\,g})\big)\bef\text{ftn}_{M}^{\uparrow F}=l\rightarrow\big(\gunderline{l\triangleright h\triangleright(k}\rightarrow(k\triangleright g\triangleright f)\bef\text{ftn}_{M}^{\uparrow F})\big)\bef\text{ftn}_{M}^{\uparrow F}\\
 & \quad=l\rightarrow\big((l\triangleright h\triangleright g\triangleright f)\bef\text{ftn}_{M}^{\uparrow F}\big)\bef\text{ftn}_{M}^{\uparrow F}=l\rightarrow l\triangleright h\bef g\bef f\bef\text{ftn}_{M}^{\uparrow F}\bef\text{ftn}_{M}^{\uparrow F}\quad,\\
 & f\,\tilde{\diamond}\,(g\,\tilde{\diamond}\,h)=l\rightarrow\big(l\triangleright(g\,\tilde{\diamond}\,h)\triangleright f\big)\bef\text{ftn}_{M}^{\uparrow F}=l\rightarrow\big(l\triangleright(k\rightarrow(k\triangleright h\triangleright g)\bef\text{ftn}_{M}^{\uparrow F})\triangleright f\big)\bef\text{ftn}_{M}^{\uparrow F}\\
 & \quad=l\rightarrow\big(\big((l\triangleright h\triangleright g)\bef\gunderline{\text{ftn}_{M}^{\uparrow F}\big)\triangleright f}\big)\bef\text{ftn}_{M}^{\uparrow F}=l\rightarrow l\triangleright h\triangleright g\triangleright f\bef\text{ftn}_{M}^{\uparrow M\uparrow F}\bef\text{ftn}_{M}^{\uparrow F}\quad.
\end{align*}
In the last line, we have used the naturality law of $f$. The remaining
difference between the two sides is:
\[
\text{ftn}_{M}^{\uparrow F}\bef\text{ftn}_{M}^{\uparrow F}\overset{?}{=}\text{ftn}_{M}^{\uparrow M\uparrow F}\bef\text{ftn}_{M}^{\uparrow F}\quad,
\]
which follows from the associativity law of $\text{ftn}_{M}$.

\textbf{(b)} The \lstinline!flatMap! method must have the type signature:
\[
\text{flm}_{L}:\big((A\rightarrow X)\rightarrow M^{X}\big)\rightarrow(A\rightarrow(B\rightarrow X)\rightarrow M^{X})\rightarrow(B\rightarrow X)\rightarrow M^{X}\quad.
\]
Choose $M^{A}\triangleq\bbnum 1+A$; now we need to implement the
type signature:
\begin{align*}
 & \text{flm}_{L}:\big((A\rightarrow X)\rightarrow\bbnum 1+X\big)\rightarrow\left(A\rightarrow(B\rightarrow X)\rightarrow\bbnum 1+X\right)\rightarrow(B\rightarrow X)\rightarrow\bbnum 1+X\quad,\\
 & \text{flm}_{L}\triangleq p^{:(A\rightarrow X)\rightarrow\bbnum 1+X}\rightarrow q^{:\left(A\rightarrow(B\rightarrow X)\rightarrow\bbnum 1+X\right)}\rightarrow r^{:B\rightarrow X}\rightarrow\text{???}^{:\bbnum 1+X}\quad.
\end{align*}
Can this function ever return a value of type $\bbnum 0+X$? When
we try filling out the typed hole $\text{???}^{:\bbnum 1+X}$, we
cannot apply the function $r$ since we have no available values of
type $B$. We could substitute $r$ into the second curried argument
of $q$, obtaining a function of type $A\rightarrow\bbnum 1+X$. But
we have no available values of type $A$. We also cannot apply the
function $p$ since its argument is of type $A\rightarrow X$, but
we only have $A\rightarrow\bbnum 1+X$, which is not guaranteed to
return nonempty values. So, the only way of implementing \lstinline!flatMap!
via fully parametric code is to return the constant value $1+\bbnum 0^{:X}$.
This would lose information and violate an identity law of monads.

\chapter{A humorous disclaimer}

\index{jokes}\emph{The following text is quoted in part from an anonymous
online source (\textsf{``}Project Guten Tag\textsf{''}) dating back at least to 1997.
The original text is no longer available on the Internet.}

\medskip{}

\noun{Warranto Limitensis; Disclamatantus Damagensis}

Solus exceptus \textsf{``}Rectum Replacator Refundiens\textsf{''} describitus ecci,
\begin{enumerate}
\item Projectus (etque nunquam partum quis hic etext remitibus cum \noun{Project
Guten Tag}$^{\text{TM}}$ identificator) disclamabat omni liabilitus
tuus damagensis, pecuniensisque, includibantus pecunia legalitus,
et 
\item \noun{Remedia Negligentitia Non Habet Tuus, Warrantus Destructi\-bus
Contractus Nullibus Ni Liabilitus Sumus, Inclutatibus Non Limitatus
Destructio Directibus, Consequentius, Punitio, O Incidentus, Non Sunt
Si Nos Notificat Vobis}. 
\end{enumerate}
Sit discubriatus defectus en etextum sic entram diariam noventam recibidio,
pecuniam tuum refundatorium receptorus posset, sic scribatis vendor.
Sit veniabat medium physicalis, vobis idem reternat et replacator
possit copius. Sit venitabat electronicabilis, sic viri datus chansus
segundibus. 

\noun{Hic Etext Venid \textsf{``}Como-asi\textsf{''}. Nihil Warranti Nunquam Classum,
Expressito Ni Implicato, Le Macchen Como Si Etexto Bene Sit O Il Medio
Bene Sit, Inclutat Et Non Limitat Warranti Mercatensis, Appropriatensis
Purposem. }

Statuen varias non permitatent disclamabaris ni warranti implicatoren
ni exclusioni limitatio damagaren consequentialis, ecco lo qua disclamatori
exclusato\-rique non vobis applicant, et potat optia alia legali.

\twocolumn

\chapter{GNU Free Documentation License\label{sec:GFDL} }

{\footnotesize{}Version 1.2, November 2002}{\footnotesize\par}

{\tiny{}Copyright (c) 2000,2001,2002 Free Software Foundation, Inc.
59 Temple Place, Suite 330, Boston, MA 02111-1307, USA}{\tiny\par}

{\tiny{}Everyone is permitted to copy and distribute verbatim copies
of this license document, but changing it is not allowed.}{\tiny\par}

{\tiny{}\setcounter{subsection}{-1}}{\tiny\par}

\subsection*{{\tiny{}Preamble}}

{\tiny{}The purpose of this License is to make a manual, textbook,
or other functional and useful document free in the sense of freedom:
to assure everyone the effective freedom to copy and redistribute
it, with or without modifying it, either commercially or noncommercially.
Secondarily, this License preserves for the author and publisher a
way to get credit for their work, while not being considered responsible
for modifications made by others.}{\tiny\par}

{\tiny{}This License is a kind of \textquotedblleft copyleft\textsf{''}, which
means that derivative works of the document must themselves be free
in the same sense. It complements the GNU General Public License,
which is a copyleft license designed for free software.}{\tiny\par}

{\tiny{}We have designed this License in order to use it for manuals
for free software, because free software needs free documentation:
a free program should come with manuals providing the same freedoms
that the software does. But this License is not limited to software
manuals; it can be used for any textual work, regardless of subject
matter or whether it is published as a printed book. We recommend
this License principally for works whose purpose is instruction or
reference.}{\tiny\par}

\subsection{Applicability and definitions\label{subsec:1Applicability-and-definitions}}

{\tiny{}This License applies to any manual or other work, in any medium,
that contains a notice placed by the copyright holder saying it can
be distributed under the terms of this License. Such a notice grants
a world-wide, royalty-free license, unlimited in duration, to use
that work under the conditions stated herein. The \textquotedblleft Document\textsf{''},
below, refers to any such manual or work. Any member of the public
is a licensee, and is addressed as \textquotedblleft you\textsf{''}. You accept
the license if you copy, modify or distribute the work in a way requiring
permission under copyright law.}{\tiny\par}

{\tiny{}A \textquotedblleft Modified Version\textsf{''} of the Document means
any work containing the Document or a portion of it, either copied
verbatim, or with modifications and/or translated into another language.}{\tiny\par}

{\tiny{}A \textquotedblleft Secondary Section\textsf{''} is a named appendix
or a front-matter section of the Document that deals exclusively with
the relationship of the publishers or authors of the Document to the
Document\textsf{'}s overall subject (or to related matters) and contains nothing
that could fall directly within that overall subject. (Thus, if the
Document is in part a textbook of mathematics, a Secondary Section
may not explain any mathematics.) The relationship could be a matter
of historical connection with the subject or with related matters,
or of legal, commercial, philosophical, ethical or political position
regarding them.}{\tiny\par}

{\tiny{}The \textquotedblleft Invariant Sections\textsf{''} are certain Secondary
Sections whose titles are designated, as being those of Invariant
Sections, in the notice that says that the Document is released under
this License. If a section does not fit the above definition of Secondary
then it is not allowed to be designated as Invariant. The Document
may contain zero Invariant Sections. If the Document does not identify
any Invariant Sections then there are none.}{\tiny\par}

{\tiny{}The \textquotedblleft Cover Texts\textsf{''} are certain short passages
of text that are listed, as Front-Cover Texts or Back-Cover Texts,
in the notice that says that the Document is released under this License.
A Front-Cover Text may be at most 5 words, and a Back-Cover Text may
be at most 25 words.}{\tiny\par}

{\tiny{}A \textquotedblleft Transparent\textsf{''} copy of the Document means
a machine-readable copy, represented in a format whose specification
is available to the general public, that is suitable for revising
the document straightforwardly with generic text editors or (for images
composed of pixels) generic paint programs or (for drawings) some
widely available drawing editor, and that is suitable for input to
text formatters or for automatic translation to a variety of formats
suitable for input to text formatters. A copy made in an otherwise
Transparent file format whose markup, or absence of markup, has been
arranged to thwart or discourage subsequent modification by readers
is not Transparent. An image format is not Transparent if used for
any substantial amount of text. A copy that is not \textquotedblleft Transparent\textsf{''}
is called \textquotedblleft Opaque\textsf{''}.}{\tiny\par}

{\tiny{}Examples of suitable formats for Transparent copies include
plain ASCII without markup, Texinfo input format, \LaTeX{} input format,
SGML or XML using a publicly available DTD, and standard-conforming
simple HTML, PostScript or PDF designed for human modification. Examples
of transparent image formats include PNG, XCF and JPG. Opaque formats
include proprietary formats that can be read and edited only by proprietary
word processors, SGML or XML for which the DTD and/or processing tools
are not generally available, and the machine-generated HTML, PostScript
or PDF produced by some word processors for output purposes only.}{\tiny\par}

{\tiny{}The \textsf{``}Title Page\textsf{''} means, for a printed book, the title
page itself, plus such following pages as are needed to hold, legibly,
the material this License requires to appear in the title page. For
works in formats which do not have any title page as such, \textquotedblleft Title
Page\textquotedblright{} means the text near the most prominent appearance
of the work\textsf{'}s title, preceding the beginning of the body of the text.}{\tiny\par}

{\tiny{}A section \textsf{``}Entitled XYZ\textsf{''} means a named subunit of the
Document whose title either is precisely XYZ or contains XYZ in parentheses
following text that translates XYZ in another language. (Here XYZ
stands for a specific section name mentioned below, such as \textquotedblleft Acknowledgements\textquotedblright ,
\textquotedblleft Dedications\textquotedblright , \textquotedblleft Endorsements\textquotedblright ,
or \textquotedblleft History\textquotedblright .) To \textquotedblleft Preserve
the Title\textquotedblright{} of such a section when you modify the
Document means that it remains a section \textquotedblleft Entitled
XYZ\textquotedblright{} according to this definition.}{\tiny\par}

{\tiny{}The Document may include Warranty Disclaimers next to the
notice which states that this License applies to the Document. These
Warranty Disclaimers are considered to be included by reference in
this License, but only as regards disclaiming warranties: any other
implication that these Warranty Disclaimers may have is void and has
no effect on the meaning of this License.}{\tiny\par}

\subsection{Verbatim copying\label{subsec:2Verbatim-copying}}

{\tiny{}You may copy and distribute the Document in any medium, either
commercially or noncommercially, provided that this License, the copyright
notices, and the license notice saying this License applies to the
Document are reproduced in all copies, and that you add no other conditions
whatsoever to those of this License. You may not use technical measures
to obstruct or control the reading or further copying of the copies
you make or distribute. However, you may accept compensation in exchange
for copies. If you distribute a large enough number of copies you
must also follow the conditions in section~\ref{subsec:3Copying-in-quantity}.}{\tiny\par}

{\tiny{}You may also lend copies, under the same conditions stated
above, and you may publicly display copies.}{\tiny\par}

\subsection{Copying in quantity\label{subsec:3Copying-in-quantity}}

{\tiny{}If you publish printed copies (or copies in media that commonly
have printed covers) of the Document, numbering more than 100, and
the Document\textsf{'}s license notice requires Cover Texts, you must enclose
the copies in covers that carry, clearly and legibly, all these Cover
Texts: Front-Cover Texts on the front cover, and Back-Cover Texts
on the back cover. Both covers must also clearly and legibly identify
you as the publisher of these copies. The front cover must present
the full title with all words of the title equally prominent and visible.
You may add other material on the covers in addition. Copying with
changes limited to the covers, as long as they preserve the title
of the Document and satisfy these conditions, can be treated as verbatim
copying in other respects.}{\tiny\par}

{\tiny{}If the required texts for either cover are too voluminous
to fit legibly, you should put the first ones listed (as many as fit
reasonably) on the actual cover, and continue the rest onto adjacent
pages.}{\tiny\par}

{\tiny{}If you publish or distribute Opaque copies of the Document
numbering more than 100, you must either include a machine-readable
Transparent copy along with each Opaque copy, or state in or with
each Opaque copy a computer-network location from which the general
network-using public has access to download using public-standard
network protocols a complete Transparent copy of the Document, free
of added material. If you use the latter option, you must take reasonably
prudent steps, when you begin distribution of Opaque copies in quantity,
to ensure that this Transparent copy will remain thus accessible at
the stated location until at least one year after the last time you
distribute an Opaque copy (directly or through your agents or retailers)
of that edition to the public.}{\tiny\par}

{\tiny{}It is requested, but not required, that you contact the authors
of the Document well before redistributing any large number of copies,
to give them a chance to provide you with an updated version of the
Document.}{\tiny\par}

\subsection{Modifications\label{subsec:4Modifications}}

{\tiny{}You may copy and distribute a Modified Version of the Document
under the conditions of sections~\ref{subsec:2Verbatim-copying}
and \ref{subsec:3Copying-in-quantity} above, provided that you release
the Modified Version under precisely this License, with the Modified
Version filling the role of the Document, thus licensing distribution
and modification of the Modified Version to whoever possesses a copy
of it. In addition, you must do these things in the Modified Version:}{\tiny\par}

{\tiny{}A. Use in the Title Page (and on the covers, if any) a title
distinct from that of the Document, and from those of previous versions
(which should, if there were any, be listed in the History section
of the Document). You may use the same title as a previous version
if the original publisher of that version gives permission.}{\tiny\par}

{\tiny{}B. List on the Title Page, as authors, one or more persons
or entities responsible for authorship of the modifications in the
Modified Version, together with at least five of the principal authors
of the Document (all of its principal authors, if it has fewer than
five), unless they release you from this requirement.}{\tiny\par}

{\tiny{}C. State on the Title page the name of the publisher of the
Modified Version, as the publisher.}{\tiny\par}

{\tiny{}D. Preserve all the copyright notices of the Document.}{\tiny\par}

{\tiny{}E. Add an appropriate copyright notice for your modifications
adjacent to the other copyright notices.}{\tiny\par}

{\tiny{}F. Include, immediately after the copyright notices, a license
notice giving the public permission to use the Modified Version under
the terms of this License, in the form shown in the Addendum below.}{\tiny\par}

{\tiny{}G. Preserve in that license notice the full lists of Invariant
Sections and required Cover Texts given in the Document\textsf{'}s license
notice.}{\tiny\par}

{\tiny{}H. Include an unaltered copy of this License.}{\tiny\par}

{\tiny{}I. Preserve the section Entitled \textsf{``}History\textsf{''}, Preserve its
Title, and add to it an item stating at least the title, year, new
authors, and publisher of the Modified Version as given on the Title
Page. If there is no section Entitled \textquotedblleft History\textquotedblright{}
in the Document, create one stating the title, year, authors, and
publisher of the Document as given on its Title Page, then add an
item describing the Modified Version as stated in the previous sentence.}{\tiny\par}

{\tiny{}J. Preserve the network location, if any, given in the Document
for public access to a Transparent copy of the Document, and likewise
the network locations given in the Document for previous versions
it was based on. These may be placed in the \textsf{``}History\textsf{''} section.
You may omit a network location for a work that was published at least
four years before the Document itself, or if the original publisher
of the version it refers to gives permission.}{\tiny\par}

{\tiny{}K. For any section Entitled \textsf{``}Acknowledgements\textsf{''} or \textsf{``}Dedications\textsf{''},
Preserve the Title of the section, and preserve in the section all
the substance and tone of each of the contributor acknowledgements
and/or dedications given therein.}{\tiny\par}

{\tiny{}L. Preserve all the Invariant Sections of the Document, unaltered
in their text and in their titles. Section numbers or the equivalent
are not considered part of the section titles.}{\tiny\par}

{\tiny{}M. Delete any section Entitled \textsf{``}Endorsements\textsf{''}. Such a
section may not be included in the Modified Version.}{\tiny\par}

{\tiny{}N. Do not retitle any existing section to be Entitled \textsf{``}Endorsements\textsf{''}
or to conflict in title with any Invariant Section.}{\tiny\par}

{\tiny{}O. Preserve any Warranty Disclaimers.}{\tiny\par}

{\tiny{}If the Modified Version includes new front-matter sections
or appendices that qualify as Secondary Sections and contain no material
copied from the Document, you may at your option designate some or
all of these sections as invariant. To do this, add their titles to
the list of Invariant Sections in the Modified Version\textsf{'}s license notice.
These titles must be distinct from any other section titles.}{\tiny\par}

{\tiny{}You may add a section Entitled \textsf{``}Endorsements\textsf{''}, provided
it contains nothing but endorsements of your Modified Version by various
parties \textemdash{} for example, statements of peer review or that
the text has been approved by an organization as the authoritative
definition of a standard.}{\tiny\par}

{\tiny{}You may add a passage of up to five words as a Front-Cover
Text, and a passage of up to 25 words as a Back-Cover Text, to the
end of the list of Cover Texts in the Modified Version. Only one passage
of Front-Cover Text and one of Back-Cover Text may be added by (or
through arrangements made by) any one entity. If the Document already
includes a cover text for the same cover, previously added by you
or by arrangement made by the same entity you are acting on behalf
of, you may not add another; but you may replace the old one, on explicit
permission from the previous publisher that added the old one.}{\tiny\par}

{\tiny{}The author(s) and publisher(s) of the Document do not by this
License give permission to use their names for publicity for or to
assert or imply endorsement of any Modified Version.}{\tiny\par}

\subsection*{{\tiny{}Combining documents}}

{\tiny{}You may combine the Document with other documents released
under this License, under the terms defined in section 4 above for
modified versions, provided that you include in the combination all
of the Invariant Sections of all of the original documents, unmodified,
and list them all as Invariant Sections of your combined work in its
license notice, and that you preserve all their Warranty Disclaimers.}{\tiny\par}

{\tiny{}The combined work need only contain one copy of this License,
and multiple identical Invariant Sections may be replaced with a single
copy. If there are multiple Invariant Sections with the same name
but different contents, make the title of each such section unique
by adding at the end of it, in parentheses, the name of the original
author or publisher of that section if known, or else a unique number.
Make the same adjustment to the section titles in the list of Invariant
Sections in the license notice of the combined work.}{\tiny\par}

{\tiny{}In the combination, you must combine any sections Entitled
\textquotedblleft History\textquotedblright{} in the various original
documents, forming one section Entitled \textquotedblleft History\textquotedblright ;
likewise combine any sections Entitled \textquotedblleft Acknowledgements\textquotedblright ,
and any sections Entitled \textquotedblleft Dedications\textquotedblright .
You must delete all sections Entitled \textquotedblleft Endorsements.\textquotedblright{}}{\tiny\par}

\subsection*{{\tiny{}Collections of documents}}

{\tiny{}You may make a collection consisting of the Document and other
documents released under this License, and replace the individual
copies of this License in the various documents with a single copy
that is included in the collection, provided that you follow the rules
of this License for verbatim copying of each of the documents in all
other respects.}{\tiny\par}

{\tiny{}You may extract a single document from such a collection,
and distribute it individually under this License, provided you insert
a copy of this License into the extracted document, and follow this
License in all other respects regarding verbatim copying of that document.}{\tiny\par}

\subsection*{{\tiny{}Aggregation with independent works}}

{\tiny{}A compilation of the Document or its derivatives with other
separate and independent documents or works, in or on a volume of
a storage or distribution medium, is called an \textquotedblleft aggregate\textquotedblright{}
if the copyright resulting from the compilation is not used to limit
the legal rights of the compilation\textsf{'}s users beyond what the individual
works permit. When the Document is included an aggregate, this License
does not apply to the other works in the aggregate which are not themselves
derivative works of the Document.}{\tiny\par}

{\tiny{}If the Cover Text requirement of section~\ref{subsec:3Copying-in-quantity}
is applicable to these copies of the Document, then if the Document
is less than one half of the entire aggregate, the Document\textsf{'}s Cover
Texts may be placed on covers that bracket the Document within the
aggregate, or the electronic equivalent of covers if the Document
is in electronic form. Otherwise they must appear on printed covers
that bracket the whole aggregate.}{\tiny\par}

\subsection*{{\tiny{}Translation}}

{\tiny{}Translation is considered a kind of modification, so you may
distribute translations of the Document under the terms of section~\ref{subsec:4Modifications}.
Replacing Invariant Sections with translations requires special permission
from their copyright holders, but you may include translations of
some or all Invariant Sections in addition to the original versions
of these Invariant Sections. You may include a translation of this
License, and all the license notices in the Document, and any Warranty
Disclaimers, provided that you also include the original English version
of this License and the original versions of those notices and disclaimers.
In case of a disagreement between the translation and the original
version of this License or a notice or disclaimer, the original version
will prevail.}{\tiny\par}

{\tiny{}If a section in the Document is Entitled \textquotedblleft Acknowledgements\textquotedblright ,
\textquotedblleft Dedications\textquotedblright , or \textquotedblleft History\textquotedblright ,
the requirement (section~\ref{subsec:4Modifications}) to Preserve
its Title (section~\ref{subsec:1Applicability-and-definitions})
will typically require changing the actual title.}{\tiny\par}

\subsection*{{\tiny{}Termination}}

{\tiny{}You may not copy, modify, sublicense, or distribute the Document
except as expressly provided for under this License. Any other attempt
to copy, modify, sublicense or distribute the Document is void, and
will automatically terminate your rights under this License. However,
parties who have received copies, or rights, from you under this License
will not have their licenses terminated so long as such parties remain
in full compliance.}{\tiny\par}

\subsection*{{\tiny{}Future revisions of this license}}

{\tiny{}The Free Software Foundation may publish new, revised versions
of the GNU Free Documentation License from time to time. Such new
versions will be similar in spirit to the present version, but may
differ in detail to address new problems or concerns. See \url{http://www.gnu.org/copyleft/}.}{\tiny\par}

{\tiny{}Each version of the License is given a distinguishing version
number. If the Document specifies that a particular numbered version
of this License \textquotedblleft or any later version\textquotedblright{}
applies to it, you have the option of following the terms and conditions
either of that specified version or of any later version that has
been published (not as a draft) by the Free Software Foundation. If
the Document does not specify a version number of this License, you
may choose any version ever published (not as a draft) by the Free
Software Foundation.}{\tiny\par}

\subsection*{\noun{\tiny{}Addendum}{\tiny{}: How to use this License for your
documents}}

{\tiny{}To use this License in a document you have written, include
a copy of the License in the document and put the following copyright
and license notices just after the title page:}{\tiny\par}

{\tiny{}Copyright (c) <year> <your name>. Permission is granted to
copy, distribute and/or modify this document under the terms of the
GNU Free Documentation License, Version 1.2 or any later version published
by the Free Software Foundation; with no Invariant Sections, no Front-Cover
Texts, and no Back-Cover Texts. A copy of the license is included
in the section entitled \textquotedblleft GNU Free Documentation License\textquotedblright .}{\tiny\par}

{\tiny{}If you have Invariant Sections, Front-Cover Texts and Back-Cover
Texts, replace the \textquotedblleft with...Texts.\textquotedblright{}
line with this:}{\tiny\par}

{\tiny{}with the Invariant Sections being <list their titles>, with
the Front-Cover Texts being <list>, and with the Back-Cover Texts
being <list>.}{\tiny\par}

{\tiny{}If you have Invariant Sections without Cover Texts, or some
other combination of the three, merge those two alternatives to suit
the situation.}{\tiny\par}

{\tiny{}If your document contains nontrivial examples of program code,
we recommend releasing these examples in parallel under your choice
of free software license, such as the GNU General Public License,
to permit their use in free software.}{\tiny\par}

\subsection*{{\tiny{}Copyright }}

{\tiny{}Copyright (c) 2000, 2001, 2002 Free Software Foundation, Inc.
59 Temple Place, Suite 330, Boston, MA 02111-1307, USA}{\tiny\par}

{\tiny{}Everyone is permitted to copy and distribute verbatim copies
of this license document, but changing it is not allowed.}{\tiny\par}
